### 2007-09-20 pkg/paper/article.Rnw
Added article files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@5 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{article}
\usepackage{graphicx}
\begin{document}

This article aims to set out the technical information that is
necessary for producing precise specifications
of colour within a software environment.

This article is aimed at the developers of software for
statistical graphics, so it is not an exhaustive discussion of
colour.  However, sufficient background material is included 
to allow an understanding of the terms and concepts used in this
article.

It is hoped that the information will also be generally useful
for people working with colours as data values, including 
the generation of sets of colours for statistical plots.

\section{What's wrong with RGB?}

The most common way to specify colours in a software system
is as an RGB triplet.  In this system, a colour is specified 
as an amount of red, an amount of green, and an amount of blue,
where the amounts range from 0 (none) to 1 (as much as possible).
In some systems, the amounts are specified in the range 0 to 255
(because it is common to use 8 bits to store each colour component
and $2^8 = 256$), and it is not uncommon to write values within this 
range in hexadecimal notation so that values range from \#00 to
\#FF (F has the decimal value 15 and the second digit 
specifies a multiple of
16, so $15*16 + 15 = 255$).

For example, in HTML, the colour red is specified as \#FF0000.

\subsection{RGB is unintuitive}

This system for specifying colour has some appeal.  It appears 
straightforward to specify simple colours:  red is all red and no green
or blue, green and blue are similar, white is all three colours, and black is
none.

However, intuition rapidly runs out of steam in this system.  What is the
RGB specification for the colour brown?  What about pink or orange?

It is possible to create colour palettes for the RGB system and 
to select colours by eye, but this is not appropriate for situations
where it is necessary to select sets of colours.  For example, 
it might be useful to represent an ordinal variable as different
shades of brown.  Given the RGB specification for brown, how 
can we specify lighter shades of brown?

\subsection{RGB is not perceptually uniform}

An intuitive answer to the previous question would be to linearly
interpolate between the RGB values for brown and the RGB values
for white, but this only serves to highlight another problem with the
RGB system, which is that it is not a \emph{perceptually uniform} system.
This means that a unit step within RGB space does not produce a 
constant perceptual change in colour.
Figure \ref{figure:shadesofbrown} shows the linear interpolation in
RGB space compared to a similar interpolation in a perceptually uniform space.

<<echo=FALSE>>=
library(grid)
library(colorspace)

<<rgbnotuniform, echo=FALSE, fig=TRUE, height=3, include=FALSE>>=
# grid.newpage()

# RGB
N <- 100
grid.rect(x=1:N/N, width=1/N, y=.75, height=0.4,
          just=c("right", "centre"),
          gp=gpar(col=NA,
            fill=rgb(seq(165, 255, length=N),
              seq(42, 255, length=N),
              seq(42, 255, length=N), max=255)))

# HCL

# col2rgb("brown")
#       [,1]
# red    165
# green   42
# blue    42
# N <- 20

# > as(RGB(165/255, 42/255, 42/255), "polarLAB")
#             L        C        H
# [1,] 58.72249 41.92013 24.42661

cols <- coords(as(polarLAB(seq(58.7224, 99, length=N),
                           seq(41.92013, 0, length=N),
                           24.42661), "RGB"))

grid.rect(x=1:N/N, width=1/N, y=0.25, height=0.4,
          just=c("right", "centre"),
          gp=gpar(col=NA, fill=rgb(cols[,1], cols[,2], cols[,3])))

@ 
\begin{figure}
\includegraphics[width=\textwidth]{article-rgbnotuniform}
\caption{\label{figure:shadesofbrown}A linear interpolation
from brown to white in RGB space (top) and a perceptually uniform
 space (bottom).  Equal-sized horizontal steps result in non-equal-sized
perceptual changes in colour in the RGB interpolation (the perceived 
colour changes rapidly at the left end and more slowly at the 
right end).}
\end{figure}

\subsection{RGB is device-dependent}

A further problem with RGB colour specifications is that it is unclear
what ``red'', ``green'', and ``blue'' actually mean.

The colour of light is determined by its wave-length. A typical human can 
perceive light in the range 400nm to 700nm.  A coherent
light source (e.g., a laser) at the 400nm range appears blue and a coherent
light source at the 700nm range appears red.  Most light sources
(e.g., light bulbs, or the Sun)
are not coherent and emit light at many different wave lengths.
A light source can be characterised by its Spectral Power Distribution (SPD),
which is how much of each wavelength of light the light source emits.

On an LCD computer screen, 
each pixel is made up from three sub-pixels, which are
coloured red, green, and blue.  A full red colour is produced by emitting 
light at full power from just the red sub-pixel.  However, this red is not
a single wavelength.
Figure \ref{figure:SPD} shows the SPDs might look like 
for the red, green, and blue
sub-pixels of a typical computer monitor.

% source: http://white.stanford.edu/~brian/scielab/scielab1-1-1/

<<spd, echo=FALSE, fig=TRUE, include=FALSE>>=
SPD <- read.table("displaySPD.txt", skip=5,
    col.names=c("r", "g", "b"))
wavelengths <- scan("wavelength.txt", what=numeric(), skip=5)

plot(wavelengths, SPD$r, type="l", col="red", lwd=3,
     ylab="", xlab="wavelength", axes=FALSE)
lines(wavelengths, SPD$g, type="l", col="green", lwd=3)
lines(wavelengths, SPD$b, type="l", col="blue", lwd=3)
box()
axis(1)

@ 
\begin{figure}
\includegraphics[width=\textwidth]{article-spd}
\caption{\label{figure:SPD}SPD for red, green, and blue on
a typical computer monitor.}
\end{figure}

In order to know what light is emitted in response to an RGB colour 
specification like \#FF0000, it is necessary to know the SPDs for
the device that will be producing the light.
This can vary considerably between different devices such as computer
monitors and data projectors (beamers).

For less-than-full red, the SPD stays the same shape and just
less light is emitted (the result is a darker colour).
However, there is a further complication because the decrease is not linear.
Figure \ref{figure:gamma} shows the curves relating the amount of
light emitted as a function of the sizes of the RGB components
for a typical computer monitor.
These are commonly referred to as ``display gamma curves'' because
the curve is, for convenience, approximated by a power function
with a single exponent called gamma.  Although each curve is slightly 
different, it is common to use a single gamma value for a
particular device.  Figure \ref{figure:gamma} shows the power curve
for the standard gamma value of 2.2 as a dotted line.

% source: http://white.stanford.edu/~brian/scielab/scielab1-1-1/

<<gamma, echo=FALSE, fig=TRUE, include=FALSE>>=
gamma <- read.table("gamma.txt", skip=5)

plot(seq(0, 1, length=dim(gamma)[1]), gamma[,1],
     type="l", col="red", lwd=3,
     xlab="amount of primary", ylab="intensity")
lines(seq(0, 1, length=dim(gamma)[1]), gamma[,2],
     type="l", col="green", lwd=3)
lines(seq(0, 1, length=dim(gamma)[1]), gamma[,3],
     type="l", col="blue", lwd=3)
lines(seq(0, 1, length=dim(gamma)[1]),
     seq(0, 1, length=dim(gamma)[1])^2.2,
     type="l", lty="dotted", lwd=3)

@ 
\begin{figure}
\includegraphics[width=\textwidth]{article-gamma}
\caption{\label{figure:gamma}Gamma curves for red, green, and blue on
a typical computer monitor.}
\end{figure}

In order to know how much light is emitted in response
to an RGB colour specification, for example, \#990000, it is necessary
to know the gamma for the device that will be producing the light. 
This can also vary considerably between different devices such as computer
monitors and data projectors (beamers).

In summary, we have what we will call the \emph{RGB uncertainty principle}:
in RGB space, you don't know where you are, 
you don't know how fast you are moving, and 
you don't know which way to go.

It is simply not adequate for a software system to only allow 
the user to specify colours in an RGB system.

\section{Device-independence:  CIE XYZ}

Human perception of colours.  Standard observer.

Chromaticities.

\section{Perceptual uniformity:  CIE Lab}

\section{Intuitive colour choice:  HCL}

\section{sRGB}

RGB is still important because that is how you express colours
in some systems, ergo it is necessary to allow people to work
in a nice colourspace, but then convert to RGB.

\section{Other colour spaces}

CMYK for print.  Simple-minded RGB-CMY is easy, but wrong.
Sophisticated RGB-CMY is damn nigh impossible.

\end{document}

### 2007-09-26 pkg/paper/article.Rnw
chipping away at the conversions between colorspaces
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@6 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 necessary for producing precise specifications
-of colour within a software environment.
+of color within a software environment.
 
@@
 statistical graphics, so it is not an exhaustive discussion of
-colour.  However, sufficient background material is included 
+color.  However, sufficient background material is included 
 to allow an understanding of the terms and concepts used in this
@@
 It is hoped that the information will also be generally useful
-for people working with colours as data values, including 
-the generation of sets of colours for statistical plots.
+for people working with colors as data values, including 
+the generation of sets of colors for statistical plots.
 
@@
 
-The most common way to specify colours in a software system
-is as an RGB triplet.  In this system, a colour is specified 
+The most common way to specify colors in a software system
+is as an RGB triplet.  In this system, a color is specified 
 as an amount of red, an amount of green, and an amount of blue,
@@
 In some systems, the amounts are specified in the range 0 to 255
-(because it is common to use 8 bits to store each colour component
+(because it is common to use 8 bits to store each color component
 and $2^8 = 256$), and it is not uncommon to write values within this 
@@
 
-For example, in HTML, the colour red is specified as \#FF0000.
+For example, in HTML, the color red is specified as \#FF0000.
 
@@
 
-This system for specifying colour has some appeal.  It appears 
-straightforward to specify simple colours:  red is all red and no green
-or blue, green and blue are similar, white is all three colours, and black is
+This system for specifying color has some appeal.  It appears 
+straightforward to specify simple colors:  red is all red and no green
+or blue, green and blue are similar, white is all three colors, and black is
 none.
@@
 However, intuition rapidly runs out of steam in this system.  What is the
-RGB specification for the colour brown?  What about pink or orange?
+RGB specification for the color brown?  What about pink or orange?
 
-It is possible to create colour palettes for the RGB system and 
-to select colours by eye, but this is not appropriate for situations
-where it is necessary to select sets of colours.  For example, 
+It is possible to create color palettes for the RGB system and 
+to select colors by eye, but this is not appropriate for situations
+where it is necessary to select sets of colors.  For example, 
 it might be useful to represent an ordinal variable as different
@@
 This means that a unit step within RGB space does not produce a 
-constant perceptual change in colour.
+constant perceptual change in color.
 Figure \ref{figure:shadesofbrown} shows the linear interpolation in
@@
  space (bottom).  Equal-sized horizontal steps result in non-equal-sized
-perceptual changes in colour in the RGB interpolation (the perceived 
-colour changes rapidly at the left end and more slowly at the 
+perceptual changes in color in the RGB interpolation (the perceived 
+color changes rapidly at the left end and more slowly at the 
 right end).}
@@
 
-A further problem with RGB colour specifications is that it is unclear
+A further problem with RGB color specifications is that it is unclear
 what ``red'', ``green'', and ``blue'' actually mean.
 
-The colour of light is determined by its wave-length. A typical human can 
+The color of light is determined by its wave-length. A typical human can 
 perceive light in the range 400nm to 700nm.  A coherent
@@
 each pixel is made up from three sub-pixels, which are
-coloured red, green, and blue.  A full red colour is produced by emitting 
+colored red, green, and blue.  A full red color is produced by emitting 
 light at full power from just the red sub-pixel.  However, this red is not
@@
 
-In order to know what light is emitted in response to an RGB colour 
+In order to know what light is emitted in response to an RGB color 
 specification like \#FF0000, it is necessary to know the SPDs for
@@
 For less-than-full red, the SPD stays the same shape and just
-less light is emitted (the result is a darker colour).
+less light is emitted (the result is a darker color).
 However, there is a further complication because the decrease is not linear.
@@
 for a typical computer monitor.
-These are commonly referred to as ``display gamma curves'' because
+These are commonly referred to as \emph{display gamma curves} because
 the curve is, for convenience, approximated by a power function
@@
 In order to know how much light is emitted in response
-to an RGB colour specification, for example, \#990000, it is necessary
+to an RGB color specification, for example, \#990000, it is necessary
 to know the gamma for the device that will be producing the light. 
@@
 
+\textit{Add mention of gamut as well;  RGB principle adds ... and 
+there are places that you just can't get to!}
+
 It is simply not adequate for a software system to only allow 
-the user to specify colours in an RGB system.
+the user to specify colors in an RGB system.
 
-\section{Device-independence:  CIE XYZ}
+\subsection{Why RGB?}
 
-Human perception of colours.  Standard observer.
+If RGB has all of these problems, do we need to deal with the
+RGB colorspace at all?
 
-Chromaticities.
+Yes. The RGB colorspace is still important because it is the way that
+colors need to be specified for producing graphics output on 
+computer monitors.
+The point is that RGB space may be where you need to end up, but it is
+not a good place to start.
 
-\section{Perceptual uniformity:  CIE Lab}
+\section{CIE colorspaces}
 
-\section{Intuitive colour choice:  HCL}
+One thing that has not been mentioned so far is the problem that,
+even if we could give a description of a color that everyone
+would agree upon (e.g., a spectral power distribution), 
+we cannot guarantee that
+every person would experience the same color, because perception
+of color is a subjective task.  This is the starting point for
+a discussion of alternative color spaces to RGB.
+
+Given that no two people can expect to have exactly the same
+subjective response to the same physical stimulus, we must do the
+next best thing and target a \emph{standard observer}.
+This is the approach taken by the
+International Commission on Illumination---referred to 
+as the CIE, after its French title,
+Commission Internationale de l'Eclairage.
+
+\subsection{Device-independence:  CIE XYZ}
+
+Based on experiments with human subjects, the CIE developed the
+CIE XYZ colorspace.  In this colorspace, 
+any SPD, that is any color, can be mapped to a unique triplet
+of XYZ values, which represent the standard observer's perception of
+the color.
+
+This is a device-independent colorspace.  The specification of a
+color is absolute in the sense that it corresponds to a unique
+physical stimulus and there is no ambiguity in terms of 
+what the standard observer's response to the color should be.
+
+This is a good first step, but CIE XYZ does not give us 
+everything we need.  For example, if we want to specify a
+simple color, such as ``red'', which three values should we use?
+
+In this colorspace, Y corresponds roughly to how bright a color
+appears, but the X and Z components have no simple intuitive 
+meaning.  Furthermore, unit steps within this colorspace are
+not perceptually equal, so generating a palette of colors cannot
+proceed in a rational manner.  This leads us on to consideration 
+to other colorspaces that are derived from CIE XYZ.
+
+\subsection{Perceptual uniformity:  CIE LUV}
+
+The CIE LUV colorspace, technically referred to as CIE L*u*v*,
+ is a transformation of CIE XYZ in a 
+reasonably successful attempt to achieve perceptual uniformity.
+Again, any color can be specified as a triplet of values,
+with the advantage that a unit step in any direction corresponds to a 
+fixed perceptual change in color.  However, the problem of 
+intuitively specifying colors remains.
+
+\subsection{Intuitive color choice:  CIE LCh}
+
+The CIE LCh colorspace is a re-expression of CIE LUV using
+polar coordinates for the U-V plane.  In this colorspace,
+any color may be described by a triplet of values, a unit
+step is perceptually uniform throughout the colorspace, 
+and the three dimensions of the colorspace have intuitive 
+interpretations:
+
+\begin{description}
+\item[L] corresponds to the luminance or
+brightness of the color.  This is normalised to
+the range 0 to 100.
+\item[C] corresponds to the chroma (colorfulness, or purity) of the color.
+Lowering the chroma produces a color that is closer to grey.
+The range of this parameter depends on the luminance and hue.  For example,
+the range of possible chromas is much smaller for 
+very dark colors (low luminance);  in the extreme case of luminance 0,
+which corresponds to black, chroma can only take the value 0.
+\item[h] corresponds to the hue of the color.
+This value can range from 0 to 360, with 0 corresponding to red, 
+120 to green, 240 to blue, and so on.
+\end{description}
+
+This then is a colorspace in which it makes sense to specify colors.
+The specification does not depend on a particular device, selection
+of colors is intuitive, and color palettes can be obtained in a
+rational fashion.
+
+Only one problem remains:  we can conveniently specify a color
+using CIE LCh, but in order to render that color we need a
+specification that a rendering device understands.
+For example, in order to display a color on screen,
+we need to be able to convert a CIE LCh color into an RGB
+color specification.
+
+\section{Colorspace conversions}
+
+The conversion from CIE LCh to RGB consists of a series
+of steps.  We start with conversions between the various
+CIE colorspaces.
+
+\subsection{LCh to LUV}
+
+This is simply a conversion from the polar expression to
+cartesian coordinates in the U-V plane.
+
+$$u* = C \times cos(h)$$
+
+$$v* = C \times sin(h)$$
+
+The L component 
+is unchanged.
+
+\subsection{LUV to XYZ}
+
+This conversion is non-linear, reflecting the non-linear
+nature of human color perception.
+
+The conversion also depends on the concept of a \emph{white point}.
+For computer monitors, the white point is what you get when you
+combine red, green, and blue channels at full power.  This is not as
+unambiguous as it sounds because it depends both on the SPDs 
+of the red, green, and blue channels \emph{and} on the relative 
+powerfulness of the maximum for each channel.  In effect, it is 
+possible for two monitors to have very different white points, 
+with one, for example, much ``bluer'' than another if the maximum
+of the blue channel is set higher.\footnote{It may be easier
+to think about a white point in terms of a printed image.  In that case,
+the white point is an area of the image that perfectly reflects the
+ambient lighting conditions.  In other words, the white point
+corresponds to whatever illumination is being used to view the image.}
+
+Typically, the white point for a computer monitor is set by the
+manufacturer and the common setting used these days is something called 
+\emph{D65}, which is a CIE specification of a colour that
+corresponds roughly to the SPD of daylight.
+
+\subsection{XYZ to RGB}
+
+The first point to recall is that a CIE color specification
+is device independent;  it corresponds to an unambiguous 
+perceptual response (for the standard observer).  On the
+other hand, an RGB color specification is quite ambiguous,
+depending on the physical characteristics of the screen
+pixels.
+
+This means that the conversion from CIE colorspace to RGB
+should, ideally, occur for a particular device, \emph{and}
+it can only occur then if the characteristics of that
+device, such as the SPDs of the screen pixels, are known.
+
+It is possible to achieve this level of detail by measuring
+the physical characteristics of a device, but for a general-purpose
+system it is sensible to use a default based on industry standards.
+
+\subsection{D65}
+
+The standard white point.
+
+\subsection{Rec. 709}
+
+The standard RGB chromaticities.
 
@@
 
-RGB is still important because that is how you express colours
-in some systems, ergo it is necessary to allow people to work
-in a nice colourspace, but then convert to RGB.
+The problem also remains:  given a CIE specification, how do we 
+make a specific device emit that color.  Ergo transformations
+between colorspaces.
 
-\section{Other colour spaces}
+\section{Other color spaces}
 

### 2007-09-26 pkg/paper/article.Rnw
full set of one-possible-scenario conversions complete
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@7 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass{article}
+\setlength{\parskip}{.1in}
+\setlength{\parindent}{0in}
 \usepackage{graphicx}
@@
 
-In this colorspace, Y corresponds roughly to how bright a color
-appears, but the X and Z components have no simple intuitive 
+In this colorspace, $Y$ corresponds roughly to how bright a color
+appears, but the $X$ and $Z$ components have no simple intuitive 
 meaning.  Furthermore, unit steps within this colorspace are
@@
 
-The CIE LUV colorspace, technically referred to as CIE L*u*v*,
+The CIE LUV colorspace
  is a transformation of CIE XYZ in a 
@@
 
-\subsection{Intuitive color choice:  CIE LCh}
+\subsection{Intuitive color choice:  CIE LCH}
 
-The CIE LCh colorspace is a re-expression of CIE LUV using
+The CIE LCH colorspace is a re-expression of CIE LUV using
 polar coordinates for the U-V plane.  In this colorspace,
@@
 \begin{description}
-\item[L] corresponds to the luminance or
+\item[$L$] corresponds to the luminance or
 brightness of the color.  This is normalised to
 the range 0 to 100.
-\item[C] corresponds to the chroma (colorfulness, or purity) of the color.
+\item[$C$] corresponds to the chroma (colorfulness, or purity) of the color.
 Lowering the chroma produces a color that is closer to grey.
@@
 which corresponds to black, chroma can only take the value 0.
-\item[h] corresponds to the hue of the color.
+\item[$H$] corresponds to the hue of the color.
 This value can range from 0 to 360, with 0 corresponding to red, 
@@
 Only one problem remains:  we can conveniently specify a color
-using CIE LCh, but in order to render that color we need a
+using CIE LCH, but in order to render that color we need a
 specification that a rendering device understands.
 For example, in order to display a color on screen,
-we need to be able to convert a CIE LCh color into an RGB
+we need to be able to convert a CIE LCH color into an RGB
 color specification.
@@
 
-The conversion from CIE LCh to RGB consists of a series
+The conversion from CIE LCH to RGB consists of a series
 of steps.  We start with conversions between the various
@@
 
-\subsection{LCh to LUV}
+\subsection{LCH to LUV}
 
@@
 
-$$u* = C \times cos(h)$$
+$$U = C \cos(H)$$
 
-$$v* = C \times sin(h)$$
+$$V = C \sin(H)$$
 
@@
 This conversion is non-linear, reflecting the non-linear
-nature of human color perception.
+nature of human color perception.  It is also performed in 
+several stages, reflecting the historical development of the
+CIE LUV colorspace, which builds upon a couple of other
+earlier transformations of the CIE XYZ colorspace.
 
 The conversion also depends on the concept of a \emph{white point}.
-For computer monitors, the white point is what you get when you
-combine red, green, and blue channels at full power.  This is not as
-unambiguous as it sounds because it depends both on the SPDs 
-of the red, green, and blue channels \emph{and} on the relative 
-powerfulness of the maximum for each channel.  In effect, it is 
-possible for two monitors to have very different white points, 
-with one, for example, much ``bluer'' than another if the maximum
-of the blue channel is set higher.\footnote{It may be easier
+For now, we can just think of the white point as representing the
+viewing conditions under which the colour is to be shown.
+\footnote{It is relatively straightforward 
 to think about a white point in terms of a printed image.  In that case,
@@
 corresponds to whatever illumination is being used to view the image.}
+The white point is important because the perception of colors
+is dependent on the viewing conditions.  For example, an image
+will look different when viewed under a pink light instead
+of a white light.  More subtly, 
+an image will look different when viewed outside in normal sunlight compared
+to inside with fluorescent lighting.
 
-Typically, the white point for a computer monitor is set by the
-manufacturer and the common setting used these days is something called 
-\emph{D65}, which is a CIE specification of a colour that
-corresponds roughly to the SPD of daylight.
+The white point is expressed as CIE XYZ tristimulus values.
+There are a number of established white points used in practice,
+with the most popular being a representation of normal daylight,
+called $D65$:
+
+% Source:  Ross's code, confirmed by published article
+
+$$\begin{array}{c}X_r = 95.047\\
+                  Y_r = 100.000\\
+                  Z_r = 108.883
+  \end{array}$$
+
+\textit{Put more of these in an appendix?}
+
+% Source:  http://brucelindbloom.com/
+
+The value of $Y$ is relatively straightforward to calculate from the
+value of $L$:
+
+$$Y = \left\{\begin{array}{l l}((L + 16)/116)^3 & \quad L > 7.999592 \\
+                               L/903.3 & \quad L \leq 7.999592 \\
+             \end{array} 
+      \right.$$
+
+Before we can calculate $X$ and $Z$, we need to calculate several 
+intermediate values:
+
+$$\begin{array}{l}u_r = \frac{4X_r}{X_r + 15Y_r + 3Z_r}\\
+                  v_r = \frac{9Y_r}{X_r + 15Y_r + 3Z_r}\\
+  \end{array}$$
+
+$$\begin{array}{l}a = \frac{1}{3}\left(\frac{52L}{U + 13Lu_r} - 1\right)\\
+                  b = -5Y\\
+                  c = -\frac{1}{3}\\
+                  d = Y\left(\frac{39L}{V + 13Lv_r} - 5\right)\\
+  \end{array}$$
+
+\textit{Check me: should those be straight 
+U and V, or ``normalised'' versions?}
+
+The values of $X$ and $Y$ are now calculated from these values:
+
+$$X = \frac{d - b}{a - c}$$
+
+$$Z = Xa + b$$
 
@@
 
-\subsection{D65}
+\textit{Fixme:  need to deinfe ``primaries'' earlier.}
 
-The standard white point.
+One such standard is the ITU-R BT.709-2 Recommendation (for
+high-definition television), which specifies CIE XYZ values
+for the $R$, $G$, and $B$ primaries of a display device.
 
-\subsection{Rec. 709}
+\textit{Fixme:  need to deinfe ``channels'' earlier.}
 
-The standard RGB chromaticities.
+The second point is that we again need a white point in
+order to perform the conversion.
+For computer monitors, the white point is what you get when you
+combine red, green, and blue channels at full power.  This is not as
+unambiguous as it sounds because it depends both on the SPDs 
+of the red, green, and blue channels \emph{and} on the relative 
+powerfulness of the maximum for each channel.  In effect, it is 
+possible for two monitors to have very different white points, 
+with one, for example, much ``bluer'' than another if the maximum
+of the blue channel is set higher.
+
+Typically, the white point for a computer monitor is set by the
+manufacturer and the common setting used these days is 
+D65 (as described in the previous section).
+
+Using the 709-2 Recommendation primaries and a D65 white point,
+the conversions are:
+
+% source:  Poynton's Color FAQ
+
+\textit{Check me again:  are we dealing with tristimulus values
+or chromaticities?  Ross's code says chromaticities, but Bruce
+Lindbloom's site seems to allow working in tristimulus values.}
+
+$$R = 3.240479X - 1.537150Y - 0.498535Z$$
+
+$$G = -0.969256X + 1.875992Y + 0.041556Z$$
+
+$$B = 0.055648X - 0.204043Y + 1.057311Z$$
+
+\textit{Discuss calculation of transforms from RGB primaries that
+are not Rec. 709?}
+
+\textit{Need to talk about clipping to the RGB gamut}
+
+\section{Gamma correction}
 
@@
 
-The problem also remains:  given a CIE specification, how do we 
-make a specific device emit that color.  Ergo transformations
-between colorspaces.
+The sRGB colorspace consists of colour specifications as 
+RGB tristimulus values, \emph{plus} fixed CIE specifications of
+the $R$, $G$, and $B$ primaries, \emph{plus} a fixed white point, 
+\emph{plus} a fixed gamma.
+This is in effect a device-independent colorspace because it
+dictates the characteristics of a display device (and the viewing
+conditions).
 

### 2007-10-08 pkg/paper/article.Rnw
some enhancements/comments in gradient plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@8 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<echo=FALSE>>=
-library(grid)
-library(colorspace)
+library("grid")
+library("colorspace")
+@
 
@@
 N <- 100
+
+col <- "brown"
+col.rgb <- as.vector(col2rgb(col))
+col.lch <- as.vector(coords(as(RGB(col.rgb[1]/255, col.rgb[2]/255, col.rgb[3]/255), "polarLAB")))
+
+fill.rgb <- rgb(seq(col.rgb[1], 255, length=N),
+                seq(col.rgb[2], 255, length=N),
+                seq(col.rgb[3], 255, length=N), max=255)
+fill.lch <- coords(as(polarLAB(seq(col.lch[1], 99, length=N),
+                               seq(col.lch[2], 0, length=N),
+                                   col.lch[3]), "RGB"))
+fill.lch <- rgb(fill.lch[,1], fill.lch[,2], fill.lch[,3]) ## we do not even get this behaviour with
+                                                          ## hex(..., gamma = 1)
+							  ## only for the inverse there is
+							  ## hex2RGB(..., gamma = NA)
+
 grid.rect(x=1:N/N, width=1/N, y=.75, height=0.4,
@@
           gp=gpar(col=NA,
-            fill=rgb(seq(165, 255, length=N),
-              seq(42, 255, length=N),
-              seq(42, 255, length=N), max=255)))
-
-# HCL
-
-# col2rgb("brown")
-#       [,1]
-# red    165
-# green   42
-# blue    42
-# N <- 20
-
-# > as(RGB(165/255, 42/255, 42/255), "polarLAB")
-#             L        C        H
-# [1,] 58.72249 41.92013 24.42661
-
-cols <- coords(as(polarLAB(seq(58.7224, 99, length=N),
-                           seq(41.92013, 0, length=N),
-                           24.42661), "RGB"))
+          fill=fill.rgb))
 
@@
           just=c("right", "centre"),
-          gp=gpar(col=NA, fill=rgb(cols[,1], cols[,2], cols[,3])))
+          gp=gpar(col=NA,
+          fill=fill.lch))
 
@@
 are not coherent and emit light at many different wave lengths.
-A light source can be characterised by its Spectral Power Distribution (SPD),
+A light source can be characterised by its spectral power distribution (SPD),
 which is how much of each wavelength of light the light source emits.
@@
 
-\subsection{Device-independence:  CIE XYZ}
+\subsection{Device-independence:  CIEXYZ}
 
 Based on experiments with human subjects, the CIE developed the
-CIE XYZ colorspace.  In this colorspace, 
+CIEXYZ colorspace.  In this colorspace, 
 any SPD, that is any color, can be mapped to a unique triplet
@@
 
-This is a good first step, but CIE XYZ does not give us 
+This is a good first step, but CIEXYZ does not give us 
 everything we need.  For example, if we want to specify a
@@
 proceed in a rational manner.  This leads us on to consideration 
-to other colorspaces that are derived from CIE XYZ.
+to other colorspaces that are derived from CIEXYZ.
 
-\subsection{Perceptual uniformity:  CIE LUV}
+\subsection{Perceptual uniformity:  CIELUV}
 
-The CIE LUV colorspace
- is a transformation of CIE XYZ in a 
+The CIELUV colorspace
+ is a transformation of CIEXYZ in a 
 reasonably successful attempt to achieve perceptual uniformity.
@@
 
-\subsection{Intuitive color choice:  CIE LCH}
+\subsection{Intuitive color choice:  CIELCH}
 
-The CIE LCH colorspace is a re-expression of CIE LUV using
+The CIELCH colorspace is a re-expression of CIELUV using
 polar coordinates for the U-V plane.  In this colorspace,
@@
 Only one problem remains:  we can conveniently specify a color
-using CIE LCH, but in order to render that color we need a
+using CIELCH, but in order to render that color we need a
 specification that a rendering device understands.
 For example, in order to display a color on screen,
-we need to be able to convert a CIE LCH color into an RGB
+we need to be able to convert a CIELCH color into an RGB
 color specification.
@@
 
-The conversion from CIE LCH to RGB consists of a series
+The conversion from CIELCH to RGB consists of a series
 of steps.  We start with conversions between the various
@@
 several stages, reflecting the historical development of the
-CIE LUV colorspace, which builds upon a couple of other
-earlier transformations of the CIE XYZ colorspace.
+CIELUV colorspace, which builds upon a couple of other
+earlier transformations of the CIEXYZ colorspace.
 
@@
 For now, we can just think of the white point as representing the
-viewing conditions under which the colour is to be shown.
-\footnote{It is relatively straightforward 
+viewing conditions under which the colour is to be
+shown.\footnote{It is relatively straightforward 
 to think about a white point in terms of a printed image.  In that case,
@@
 
-The white point is expressed as CIE XYZ tristimulus values.
+The white point is expressed as CIEXYZ tristimulus values.
 There are a number of established white points used in practice,
@@
 One such standard is the ITU-R BT.709-2 Recommendation (for
-high-definition television), which specifies CIE XYZ values
+high-definition television), which specifies CIEXYZ values
 for the $R$, $G$, and $B$ primaries of a display device.

### 2007-10-16 pkg/paper/article.Rnw
added a little more on gamma correction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@10 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\section{Gamma correction}
+\subsection{Gamma correction}
+
+The RGB tristimulus values that we have calculated to this point 
+have taken into account the SPDs of the red, green, and blue 
+pixels on a screen (the primaries), 
+and the relative powerfulness of each channel 
+(the white point), but we still have to worry about the non-linear
+relationship between the input values and the amount of light
+emitted from each pixel (the display gamma).
+
+We know that RGB tristimulus values will undergo an exponential 
+transformation when displayed by the monitor, so we invert that
+transformation in the tristimulus values that we store.
+
+The exponential transformation may vary between different screens
+or projectors, but again there is a standard transformation 
+that is implemented in most modern equipment.
+
+$$C' = \left\{\begin{array}{l l}12.92C & \quad C \leq 0.0031308 \\
+                                (1 + \alpha)C^{1/2.4} - 0.055 & 
+                                \quad C > 0.0031308 \\
+             \end{array} 
+      \right.$$
 
@@
 
+HLS (and others) because you may have been given a colour in
+such a (crappy) colorspace and want to convert it to something
+better in order to work with it.
+
 CMYK for print.  Simple-minded RGB-CMY is easy, but wrong.

### 2007-10-16 pkg/R/colorspace.R
converted Achims HLS R code to C code and added R front-ends
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@11 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -9,6 +9,7 @@
 ##     HSV       Hue-Saturation-Value
 ##     LUV       CIE-L*u*v*
 ##     polarLUV  CIE-L*u*v* in polar coordinates
+##     HLS       Hue-Lightness-Saturation
 ##
 ##  The ``canonical'' space here is really CIE-XYZ, but in this
 ##  implementation all spaces are treated equally, because
@@ -28,6 +29,7 @@ setClass("XYZ", contains="color")
 setClass("LAB", contains="color")
 setClass("polarLAB", contains="color")
 setClass("HSV", contains="color")
+setClass("HLS", contains="color")
 setClass("LUV", contains="color")
 setClass("polarLUV", contains="color")
 
@@ -126,6 +128,21 @@ HSV =
     new("HSV", coords = coords)
   }
 
+HLS =
+  function(H, L, S, names)
+  {
+    if (missing(H)) return(new("HLS"))
+    if (missing(names)) names = dimnames(H)[[1]]
+    coords = cbind(H, if (missing(L)) NULL else L,
+                      if (missing(S)) NULL else S)
+    CheckMatrix(coords)
+    ## CheckBounds(coords[,1], 0, 360)
+    ## CheckBounds(coords[,2], 0, 1)
+    ## CheckBounds(coords[,3], 0, 1)
+    dimnames(coords) = list(names, c("H", "L", "S"))
+    new("HLS", coords = coords)
+  }
+
 LUV =
   function(L, U, V, names)
   {
@@ -169,6 +186,10 @@ setAs("color", "polarLAB", function(from)
 
 setAs("color", "HSV", function(from)
       HSV(.Call("as_HSV", from@coords, class(from), .WhitePoint),
+          names = dimnames(from@coords)[[1]]))
+
+setAs("color", "HLS", function(from)
+      HLS(.Call("as_HLS", from@coords, class(from), .WhitePoint),
           names = dimnames(from@coords)[[1]]))
 
 setAs("color", "LUV", function(from)

### 2007-10-16 pkg/src/colorspace.c

@@ -288,6 +288,90 @@ static int HSV_to_RGB(double h, double s, double v,
     return 1;
 }
 
+/* 
+ * rgb all in [0,1] 
+ * h in [0, 360], ls in [0,1]
+ */
+static int RGB_to_HLS(double r, double g, double b,
+		      double *h, double *l, double *s)
+{
+    double min, max;
+
+    min = min3(r, g, b);
+    max = max3(r, g, b);
+
+    *l = (max + min)/2;
+
+    if (max != min) {
+        if (*l <  0.5)  
+            *s = (max - min)/(max + min);
+        if (*l >= 0.5)  
+            *s = (max - min)/(2.0 - max - min);
+
+        if (r == max) 
+            *h = (g - b)/(max - min);
+        if (g == max) 
+            *h = 2.0 + (b - r)/(max - min);
+        if (b == max) 
+            *h = 4.0 + (r - g)/(max - min);
+
+        *h = *h * 60;
+        if (*h < 0) 
+            *h = *h + 360;
+        if (*h > 360) 
+            *h = *h - 360;
+    } else {
+        *s = 0;
+#ifdef MONO
+	*h = NA_REAL; 
+#else
+	*h = 0;
+#endif
+    }
+}
+
+static double qtrans(double q1, double q2, double hue) {
+    double result = NA_REAL;
+
+    if (hue > 360) 
+        hue = hue - 360;
+    if (hue < 0) 
+        hue = hue + 360;
+
+    if (hue < 60) 
+        result = q1 + (q2 - q1) * hue / 60;
+    else if (hue < 180)
+        result = q2;
+    else if (hue < 240) 
+        result = q1 + (q2 - q1) * (240 - hue) / 60;
+    else 
+        result = q1;
+    return result;
+}
+
+static int HLS_to_RGB(double h, double l, double s,
[…]

### 2007-10-16 pkg/src/colorspace.c
added source acknowledgement
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@12 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -291,6 +291,9 @@ static int HSV_to_RGB(double h, double s, double v,
 /* 
  * rgb all in [0,1] 
  * h in [0, 360], ls in [0,1]
+ *
+ * Form:
+ * http://wiki.beyondunreal.com/wiki/RGB_To_HLS_Conversion
  */
 static int RGB_to_HLS(double r, double g, double b,
 		      double *h, double *l, double *s)

### 2007-10-18 pkg/paper/Palettes/CSDA.Rnw
re-organized paper folder:   - moved files to "Colorspace" folder   - created new "Palettes" folder with CSDA draft   - extended color.bib file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@13 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[shortnames]{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts,rotating}

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\The University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

\Abstract{
  Statistical graphics are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in bar plots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptually based
  and do not introduce optical illusions or systematic bias. Here, we discuss
  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

\begin{document}

<<preliminaries>>=
library("vcd")
library("mvtnorm")
library("KernSmooth")
library("dichromat")

source("functions.R")
source("Nigeria/bnd.R")

generate_png <- TRUE
@

\section{Introduction} \label{sec:intro}

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose a suitable palette for a certain visualization task. Consequently,
sub-optimal color palettes are used quite frequently for color graphics displayed in journal articles
or on presentation slides which could easily be enhanced if the underlying information
were captured appropriately in the color schemes. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

For implementing color palettes in computer programs, typically color spaces are
employed that provide a mapping to Red-Green-Blue (RGB) colors. RGB is an additive
color model which is used for generating colors on computer screens
\citep[by mixing different amounts of the primary colors red, green and blue,
see][for more details]{color:Poynton:2000}. While being eminently useful for generating
colors on computer screens, it is hard for humans to select the coordinates in RGB
space corresponding to a certain color. Therefore, color picker tools in computer
programs are based on different color models---starting from the pioneering introduction
of Hue-Saturation-Value (HSV) colors \citep{color:Smith:1978}---which try to capture the
dimensions of human visual perception. HSV space is a simple transformation of RGB
space which is also implemented in many statistical software packages
\citep[see][for more details on HSV colors and links to further information about the other color
spaces discussed in the following]{color:Wiki+HSV:2007}. Although being easy to use,
HSV space exhibits a number of disadvantages: It is relatively difficult to select sets of
HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{color:Munsell:1905}.
The reason for this is that colors with different hues also differ in saturation and brightness.
For statistical graphics, this is important because it can introduce size distortions in
the perception of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
colors which are good for drawing attention to a plot but hard to look at for a longer time.

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{color:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space.
%% Given a mapping from HCL to RGB coordinates, our formulas are
%% extremely easy to implement in software and we provide such an implementation
%% in the \proglang{R} language \citep{color:R:2007} using the powerful \proglang{R} graphics
%% system \citep[see][]{color:Murrell:2006} and the HCL
%% color implementation from package \pkg{colorspace} \citep{color:Ihaka:2006}. This gives the user
%% both a conceptual and computational tool box for experimenting with color palettes for
%% a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on the implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

\section{Motivation} \label{sec:motivation}

To show what can be gained by selecting appropriate color schemes, we present a collection
of illustrations with typical color graphics based on HSV palettes (always in the left panels)
and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
of the data sets) are taken from recent publications in statistical journals such as
\textit{Journal of the American Statistical Association},
\textit{Journal of Computational and Graphical Statistics} or
\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
an overview of typically colored statistical graphics and their respective pitfalls (concerning
color palette choice).
%% All graphics are in principle very useful but can be even enhanced
%% by using better color schemes.

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{color:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{color:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde>>=
if(generate_png) {
  data("geyser", package = "MASS")
  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
  geyser2 <- cbind(geyser$duration, geyser$waiting)
  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
    if(is.null(col)) col <- rev(grey.colors(33))
    if(legend) {
      filled.contour(x$x1, x$x2, x$fhat, col = col, nlevels = length(col), xlab = xlab, ylab = ylab, ...)
    } else {
      image(x$x1, x$x2, x$fhat,
        breaks = seq(from = min(x$fhat), to = max(x$fhat), length.out = length(col) + 1),
        xlab = xlab, ylab = ylab, col = col, axes = FALSE, cex.lab = 1.5, ...)
      box()
      axis(1, cex.axis = 1.5)
      axis(2, cex.axis = 1.5)
    }
    invisible(x)
  }

  png(file = "bkde.png", height = 810, width = 840)
  par(mfrow = c(2, 2))
  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 160, width = 396)
  par(mfrow = c(4, 1), mar = c(0, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  axis(1, cex.axis = 1.5)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel as done by \citet[Figure~4a]{color:Yang+Buckley+Dudoit:2002}
or \citet[Figures~1--4]{color:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{color:Ihaka:2003}.

%FIXME% examples for sequential palettes:
%%      color:Gelfand+Banerjee+Sirmans:2007

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to add some color to the plot, this could be done by using a better balanced version of the heat
colors (as shown in the bottom right panel). These colors also increase from a yellow to a red
hue while being balanced towards the same gray as in the grayscale palette (i.e., when converted
to a grayscale or printed out on a grayscale printer, the upper and lower right panel would 
essentially look identical). Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p!]
\begin{center}
<<nigeria>>=
if(generate_png) {
  m <- read.bnd("Nigeria/nigeria37.bnd")
  fit <- read.table("Nigeria/reml_purespatial_f_district_spatial.res", header = TRUE)
  fit <- fit[order(fit$district),]

  png(file = "nigeria.png", height = 1050, width = 840)
  par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
  plot(m, z = fit$pmode, fill = rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, fill = dichromat(rev(rainbow(99, start = 0, end = 1/3)), type = "protan"),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 120, width = 396)
  par(mfrow = c(3, 1), mar = c(0, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex = 1.5)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
\end{center}
\end{figure}

A similar example is presented in Figure~\ref{fig:nigeria}
\citep[taken from][Figure~5, left]{color:Kneib:2006}, depicting posterior mode estimates
for childhood mortality in different regions of Nigeria. Deviations from a model for
childhoold mortality are brought out by shading a map according to the corresponding
residuals, revealing decreased mortality in the south-west and increased mortality in
the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the residuals by
the hue, going from green via yellow to red. Our HCL-based palette also employs green
and red hues for negative and postive residuals respectively, but codes neutral values
(around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
a number of advantages: only the interesting areas are highlighted by high-chroma colors;
flashy colors are avoided making it easier to look at the display for a longer time; positive
and negative residuals with the same absolute size receive the same perceptual weight
by being balanced towards the same gray.

%FIXME% other bad examples for diverging
%%      color:Harezlak+Coull+Laird:2007
%%      color:Tenenhaus+Giron+Viennet:2007

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

Although problematic for many tasks, pie charts can be useful for visualizing
whether a set of pie segments constitutes a majority. A typical application is shown
in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
``Bundestag'' following the 2005 election. In this election, five parties were able to
obtain enough votes to enter the Bundestag---however, neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
by using colors as metaphors: red for the social democrats SPD, black for the conservative
CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
found in the (German) media whereas the right panel uses less flashy HCL colors with the
same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
is used). The advantage of the latter is again that they are easier to hold in focus for
a longer time. Furthermore, they are all balanced towards the same gray and have the same
amount of color, resulting in a perceptually balanced palette that does not introduce undesired
graphical distractions. While it could be argued that this pie chart is such a simple
display that a well-balanced palette is not so important and that flashy colors are to
be preferred, the balancing properties of the HCL-based palette are very important in more
complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
colors and shows the distribution of votes in the 2005 election
stratified by province (Bundesland). The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=\textwidth}
\begin{sidewaysfigure}[p]
\begin{center}
<<votes, fig=TRUE, height=6, width=14>>=
data("Bundestag2005", package = "vcd")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hsv),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hcl),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport(2)
@
\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
\end{center}
\end{sidewaysfigure}

Color choice is usually much more important in graphical displays with
shaded areas compared to displays with only points or lines. However, color choice
becomes much more relevant also in scatter plots when there are many points. As an
example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
membership, three different colors (green, yellow, red) are employed. The HSV colors in
the left panel are again very flashy and differ strongly with respect to luminance: the
yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
the amount of color and gray.

As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
Clearly, hair and eye color are not independent and the pattern of association is highlighted
by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
it is shown again that the HCL colors are better balanced (between the red and blue colors)
and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
to read when shaded in a light grey rather than white. More details and extensions to residual-based
shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h!]
\begin{center}
<<scatter>>=
if(generate_png) {
  set.seed(123)
  vc <- function(var1, var2, cov)
    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))

  png(file = "scatter.png", height = 350, width = 840)
  par(mfrow = c(1, 2), mar = rep(0.5, 4))
  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
    col = hsv(c(1/3, 1/6, 0))[id])
  box()
  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
    col = hcl2hex(c(120, 60, 0), 70, 65, fix = FALSE)[id])
  box()
  dev.off()
}
@
\includegraphics{scatter}
\caption{\label{fig:scatter} Scatter plot with three clusters.}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h!]
\begin{center}
<<haireye, fig=TRUE, height=5.5, width=12>>=
he <- margin.table(HairEyeColor, 1:2)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport(2)
@
\caption{\label{fig:haireye} Extended mosaic display for hair and eye color data.}
\end{center}
\end{figure}

\section{Color spaces} \label{sec:spaces}

For choosing color palettes it is helpful to have an idea how
human color vision evolved. It has been hypothesized that it developed in three
distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
2.~\emph{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{color:Ihaka:2003} for more details and references.

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{itemize}
  \item \emph{hue} (dominant wavelength)
  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
  \item \emph{luminance} (brightness, amount of gray)
\end{itemize}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
%% HSV space has the shape of a single regular cone (often inflated to a regular
%% cylinder). Vertical sections through this space are shown in the upper panel
%% of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded:
%% which is most easily seen when converting the vertical sections to 
%% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
The brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{color:CIE:2004}. \cite{color:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$.
%% HCL space has the shape of a distorted double cone: the admissible combinations of
%% chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
%% The lower panel of Figure~\ref{fig:hsv-hcl} shows
%% vertical sections through this space: each of the resulting hue/chroma planes (given
%% luminance) is now properly balanced towards the same gray (going from black to white
%% with increasing luminance) which becomes obvious when converting the colors to
%% a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. 
Given a certain luminance $L$, all colors resulting from different combinations
of hue $H$ and chroma $C$ are balanced towards the same gray (and look the same
when converted to a gray scale). However, the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The reason for this is that some hues lead to light and others to dark colors,
e.g., full chroma yellow is brighter (i.e., has higher luminance) than full chroma blue.

The balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

\section{Color palettes} \label{sec:palettes}

\subsection{Qualitative palettes} \label{sec:qualitative}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-construct>>=
if(generate_png) {
  png(file = "pal-q-construct.png", height = 480, width = 480)
  hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
  rect(0.45, -0.05, 0.55, 0.05, col = hcl2hex(0, 50, 70))
  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl2hex(120, 50, 70))
  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl2hex(240, 50, 70))
  dev.off()
}
@
\includegraphics{pal-q-construct}
\vspace*{-2cm}
\caption{\label{fig:pal-q-construct} Constructing qualitative palettes. In the hue/chroma
plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
$H = 0, 120, 240$.}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic [60, 240]")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold [270, 150]")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm [90, -30]")
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
in different intervals for given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. Usually, these should give the same
perceptual weight to each category so that no group is perceived to be larger or more
important than any other one. Typical applications of qualitative palettes in
statistics would be bar plots \citep[see][]{color:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
or highlighted mosaic displays (see Figure~\ref{fig:votes}).

\cite{color:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
and luminance are kept fixed and only the hue is varied for obtaining different colors
which are consequently all balanced towards the same gray. This effect is illustrated
in Figure~\ref{fig:wheel} which shows a color wheel obtained by varying the hue only in
HSV coordinates $(H, 100, 100)$ and HCL coordinates $(H, 50, 70)$. Clearly, not only the
hue but also the amount of chroma and luminance varies for the HSV wheel.

Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

\subsection{Sequential palettes} \label{sec:sequential}

Sequential palettes are used for coding numerical information that ranges
in a certain interval where low values are considered to be uninteresting and
high values are interesting. Suppose we need to visualize an intensity or
interestingness $i$ which (without loss of generality) is scaled to the unit
interval. A typical application in statistics are heatmaps (see Figure~\ref{fig:bkde}).

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl2hex(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl2hex(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl2hex(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl2hex(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl2hex(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i' \cdot (C_{\max} - C_{\min}),
      L_{\max} - i'' \cdot (L_{\max} - L_{\min})). \]
One application is an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing an appropriate transformation
$i'$ for chroma and a different transformation $i''$ for the luminance.
Such a strategy is adopted for the palettes shown in the lower three rows in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts.

\subsection{Diverging palettes} \label{sec:diverging}

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Examples for this include
residuals or correlations (both with the neutral value 0) or binary classification
probabilities (with neutral value 0.5) that could be visualized in mosaic plots
(see Figure~\ref{fig:haireye}), classification maps or model-based shading in maps
such as Figure~\ref{fig:nigeria}. Analogously to the previous
section, we suppose that we want to visualize an intensity or interestingness $i$
from the interval $[-1, 1]$ (without loss of generality).

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-d-construct>>=
if(generate_png) {
  png(file = "pal-d-construct.png", height = 480, width = 534)
  hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
  hue360 <- hue.slice(360, grid.n = 301, plot = FALSE)
  mychroma <- as.numeric(colnames(hue260))
  luminance <- as.numeric(rownames(hue260))
  nc <- ncol(hue260)
  nl <- nrow(hue260)
  plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
  for(i in 1:(nc-1)) {
    rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
  }
  for(i in 1:(nc-1)) {
    rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
  }
  axis(1, labels = FALSE, at = c(-100, 0, 100))
  axis(2, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
  axis(4, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
  mtext("chroma", side = 3, at = 0, line = 2)
  mtext("luminance", side = 2, at = 50, line = 2)
  mtext("luminance", side = 4, at = 50, line = 2)
  lines(c(-100, 0, 100), c(50, 90, 50), lty = 2)
  rect(-100, 47.5, -90, 52.5, col = hcl2hex(0, 100, 50))
  rect(-55, 67.5, -45, 72.5, col = hcl2hex(0, 50, 70))
  rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(45, 67.5, 55, 72.5, col = hcl2hex(260, 50, 70))
  rect(90, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
  lines(c(-80, 0, 80), c(30, 90, 30), lty = 2)
  rect(-85, 27.5, -75, 32.5, col = hcl2hex(0, 80, 30))
  rect(-45, 57.5, -35, 62.5, col = hcl2hex(0, 40, 60))
  rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
  rect(35, 57.5, 45, 62.5, col = hcl2hex(260, 40, 60))
  rect(75, 27.5, 85, 32.5, col = hcl2hex(260, 80, 30))
  dev.off()
}
@
\includegraphics{pal-d-construct}
\caption{\label{fig:pal-d-construct} Constructing diverging palettes. The chroma/luminance
plane is shown back to back hues $H = 0$ and $H = 260$. Colors are chosen by simultaneously 
varying luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-d-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(4, 1), mar = c(0.25, 0, 0.25, 0))
pal(diverge_hcl(7))
pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
@
\caption{\label{fig:pal-d-examples} Examples for diverging palettes with different
pairs of hues and decreasing luminance contrasts.}
\end{center}
\end{figure}

Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admissible values in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-d-construct} were chosen because they correspond to similar
geometric shapes in the chroma/luminance plane, allowing for both large chroma and
luminance contrasts. 
If potential viewers of the resulting graphic might be color-blind,
the pair of hues should be taken from the yellow/blue axis of the color wheel
rather than the green/red axis as contrasts on the latter axis are more difficult to 
distinguish for color-blind people \citep{color:Lumley:2006}.

Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
hue, chroma and luminance. The first palette uses a broader range on the luminance axis
whereas the others mostly rely on chroma contrasts.

\section{Software} \label{sec:software}

Implementing the different color palettes suggested in the previous section 
is extremely easy if the software environment chosen already provides an implementation
of HCL colors: from the formulas provided above the HCL coordinates for a
palette can be conveniently computed. A bit more work is required if the software
package does not yet provide an HCL implementation. In that case, additional
functionality is needed for translating HCL coordinates to the software package's 
color system which may vary between different packages, but standardized RGB (sRGB)
is often used. The typical way of coordinate conversion is to go first from HCL to
CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
$U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
to sRGB (with the latter conversion depending on the device used for display).
The details of these conversions are somewhat technical and tedious (and hence
omitted here), however the conversion formulas are still straightfoward to
implement and can, for example, be found in \cite{color:Wiki+CIELUV:2007} or
\cite{color:Poynton:2000}.

The \proglang{R} system for statistical computing \citep{color:R:2007} already comes
with an open-source implementation of HCL (and other color spaces) in the package
\pkg{colorspace} \citep{color:Ihaka:2006}. The coordinate transformations mentioned
above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
port to other statistical software systems. Based on \pkg{colorspace} and
the \proglang{R} graphics system \citep[see][]{color:Murrell:2006},
the package \pkg{vcd} \citep{color:Meyer+Zeileis+Hornik:2007} provides implementations
of all palettes discussed above. Qualitative palettes are provided by
\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
\proglang{R}). Sequential palettes based on a single hue are implemented in the function
\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
Technical documentation along with a large collection of examples is available via
\verb/help("rainbow_hcl", package = "vcd")/. \proglang{R} code for reproducing
Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.

\section{Discussion} \label{sec:discussion}

Many statistical graphics---especially when displayed on a computer screen,
e.g., as in interactive usage, electronic papers or presentation slides---employ
colors to code information about a certain variable. Despite this
omnipresence of color, there is often only little guidance in statistical
software packages on how to choose a palette appropriate for a particular
visualization task---auspicious tools such as \pkg{ColorBrewer.org} notwithstanding.
We try to address this problem by suggesting color schemes for coding
categorical information (qualitative palettes) and numerical information
(sequential and diverging palettes) based on the perceptually-based
HCL color space.

We provide paths through HCL space along perceptual axes of the human visual system
so that colors selected along these paths match human perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance.
%% Another parameter in these 
%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
%% Employing a function $|i|^p$ for this, small
%% powers $p$ can be used if palettes with a lot of color should be 
%% constructed (e.g., to clearly separate regions in a classification map). On
%% the other hand, large powers $p$ will result in palettes with less color
%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).

Based on these conceputal guidelines and the computational tools readily
provided in the \proglang{R} system for statistical computing (and easily implemented in other
statistical software packages), users can generate palettes varying these
graphical parameters and thus adapting the colors to their particular graphical display.

%% \section*{Acknowledgments}
%% 
%% We are thankful to David Meyer, Ross Ihaka, Brian~D.~Ripley,
%% Thomas Kneib, Michael H\"ohle, and Thomas Lumley for feedback, suggestions and discussions.

%% Referees:
%% Ross Ihaka, Thomas Lumley, Cynthia Brewer, Martin Maechler

\bibliography{color}

\end{document}

<<barplot, eval=FALSE>>=
x <- rbind(
c(255, 255, 111),
c(255, 255,  47),
c(120,  50,   1),
c(102, 204, 102),
c( 50, 204, 204),
c(203, 153, 204),
c(102, 204, 102),
c(255,  47,  47),
c(102, 204, 102))/255

y <- c(10, 20, 25, 18, 16, 13, 21, 15, 18) - 3

par(mfrow = c(1, 2))
barplot(y, col = hex(RGB(x[,1], x[,2], x[,3])), space = 0)
## barplot(y, col = hcl(attr(as(RGB(x[,1], x[,2], x[,3]), "polarLUV"), "coords")[,"H"],
##  c = 40, l = 80), space = 0)
barplot(y, col = rainbow_hcl(7, c = 40, l = 80)[c(3, 5, 2, 4, 6, 1, 4, 7, 4)], space = 0)
@

<<barplot2, eval=FALSE>>
## from Takasaki & Kwamura (2007, CSDA, Figure 2, Columns 1-6)
x <- cbind(c( 2,  6, 91,  1), c(35, 30, 20, 15), c(41,  9,  8, 42),
           c(10, 40, 40, 10), c(21, 39, 30, 10), c(35, 13, 17, 35))
mybarplot <- function(x, col = NULL, ...) {
  barplot(x, space = 1.5, names = 1:6, las = 1, col = col)
  grid(lty = 1, ny = NULL, nx = 0, col = "black")
  barplot(x, space = 1.5, names = 1:6, las = 1, col = col, add = TRUE)
}

par(mfrow = c(1, 3))
mybarplot(x, col = c("#8995FF", "#AA0DA4", "#F9FFB9", "#B3FFEE"))
mybarplot(x, col = hcl(c(260, 310, 60, 150), c = 40, l = 80))
mybarplot(x, col = grey.colors(4))
@

### 2007-10-18 pkg/paper/Palettes/Nigeria/bnd.R

@@ -1,0 +1,184 @@
+## changes:
+##   - readbndfile() changed to read.bnd()
+##     this returns a suitable object (of class "bnd")
+##     and does *not* yet assign it to some variable
+##   - added quiet argument (as in scan())
+##
+##   - added print() and range() method
+##
+##   - added plot() method (not unsimilar to drawmap())
+##   - changed interface to be similar to plot()
+##     support standard graphical parameters (xlim, xlab, main, ...)
+##     removed specialized color code (replaced by col argument)
+##     removed specialized x11() and postscript() code
+
+## comments:
+##  - better than grey(0:(n-1)/(n-1)) is grey.colors(n)
+##    which (by default) stays away from the extreme colors
+##    black and white and uses a gamma-correction
+##  - T, F should be better TRUE, FALSE
+##  - don't use if(foo == TRUE), just say if(foo)
+##  - don't use return(invisible()) just invisible() is enough
+
+
+read.bnd <- function(file, quiet = TRUE)
+{
+        owarn <- getOption("warn")
+	options(warn = -1)
+
+	data.raw <- scan(file, what = list("", ""), sep = ",", quote = "", quiet = quiet)
+	data.numeric <- list(as.numeric(data.raw[[1]]), as.numeric(data.raw[[2]]))
+
+	npolygon <- sum(is.na(data.numeric[[1]])) - sum(data.raw[[1]] == "is.in")
+	if(!quiet) cat("Note: map consists of", npolygon, "polygons\n")
+	if(!quiet) cat("Reading map ...\n")
+
+	map <- list()
+	i <- 1
+
+	for(k in 1:npolygon) {
+		j <- 1
+		npoints <- data.numeric[[2]][i]
+		if(is.na(data.numeric[[1]][i + 1]) && is.na(data.numeric[[2]][i + 1])) {
+			npoints <- npoints + 1
+		}
+		elem1 <- data.numeric[[1]][i+1:npoints]
+		elem2 <- data.numeric[[2]][i+1:npoints]
+		map[[k]] <- matrix(c(elem1, elem2), ncol = 2)
+		names(map)[k] <- substring(data.raw[[1]][i], 2, nchar(data.raw[[1]][i]) - 1)
+		i <- i + npoints + 1
+	}
+
+	if(sum(is.na(as.numeric(names(map)))) == 0) {
[…]

### 2007-10-18 pkg/paper/Palettes/functions.R

@@ -1,0 +1,96 @@
+pal <- function(col, border = "light gray", ...)
+{
+  n <- length(col)
+  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="", ...)
+  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
+}
+
+hsv.wheel <-
+    function(value = c(0:2, 4:6, 8:10)/10, saturation = 0:100/100, hues = 0:300/300, asp = 1,
+             p.cex = 0.6, do.label = FALSE, cone = TRUE, cex = 1.5)
+{
+    stopifnot(is.numeric(saturation), saturation >= 0, saturation <= 1,
+              is.numeric(hues), hues >= 0, hues <= 1,
+              is.numeric(value), value >= 0, (nch <- length(value)) >= 1)
+    if(is.unsorted(hues)) hues <- sort(hues)
+    if(nch > 1) {
+        op <- par(mfrow= rev(n2mfrow(nch)), mar = c(0,0,0,0))
+        on.exit(par(op))
+    }
+    for(i.c in 1:nch) {
+        plot(-1:1,-1:1, type="n", axes = FALSE, xlab="",ylab="", asp = asp)
+        text(0.4, 0.99, paste("v =", format(100 * value[i.c])), adj = 0, font = 4, cex = cex)
+	l.s <- if(cone) saturation * value[i.c] else saturation
+        for(ang in hues) {
+            a. <- ang * 2 * pi
+            z.a <- exp(1i * a.)
+            cols <- hsv(h = ang, s = saturation, v = value[i.c])
+            points(l.s * z.a, pch = 16, col = cols, cex = p.cex)
+            if(do.label) text(z.a*1.05, labels = ang, col = cols[length(cols)/2], srt = ang)
+        }
+        lines(exp(1i * hues * 2 * pi))
+   }
+   invisible()
+}
+
+hcl.wheel <-
+    function(luminance = 1:9*10, chroma = 0:100, hues = 1:360, asp = 1,
+             p.cex = 0.6, do.label = c(TRUE, FALSE), fixup = FALSE, inner.circle = FALSE, cex = 1.5)
+{
+    do.label <- rep(do.label, length.out = 2)
+    stopifnot(is.numeric(chroma), chroma >= 0, chroma <= 100,
[…]

### 2007-10-18 pkg/paper/Palettes/CSDA.Rnw
colorspace updates:   - added HCL palettes (and diverging HSV palette)     to R code, docs and NAMESPACE   - added PM and Z in DESCRIPTION   - minor doc fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@14 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<preliminaries>>=
-library("vcd")
+library("colorspace")
 library("mvtnorm")
@@
 source("Nigeria/bnd.R")
+
+hcl2hex <- function(h, c, l, gamma = 2.4, fixup = TRUE)
+  hex(polarLUV(L = l, C = c, H = h), gamma = gamma, fixup = fixup)
 
@@
 \begin{center}
-<<votes, fig=TRUE, height=6, width=14>>=
+<<votes, fig=TRUE, height=6, width=14, eval=FALSE>>=
 data("Bundestag2005", package = "vcd")
@@
 \begin{center}
-<<haireye, fig=TRUE, height=5.5, width=12>>=
+<<haireye, fig=TRUE, height=5.5, width=12, eval=FALSE>>=
 he <- margin.table(HairEyeColor, 1:2)

### 2007-10-18 pkg/src/colorspace.c

@@ -292,7 +292,7 @@ static int HSV_to_RGB(double h, double s, double v,
  * rgb all in [0,1] 
  * h in [0, 360], ls in [0,1]
  *
- * Form:
+ * From:
  * http://wiki.beyondunreal.com/wiki/RGB_To_HLS_Conversion
  */
 static int RGB_to_HLS(double r, double g, double b,

### 2007-10-30 pkg/paper/Colorspace/article.Rnw
added an idea for an example/demo to the colorspace article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@15 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2007-10-16 pkg/paper/article.Rnw
@@
 
+\section{An example}
+
+As an example to demonstrate what all of this gives you, implement
+a function that generates colours within a CIEHCL space where the
+colours are as far apart from each other as possible, but limited
+to a narrow L range and a narrow C range.  (The idea is to have
+colours that differ mostly only in hue, but which still are 
+distinguishable when printed in black-and-white.)  Then convert the
+colours to RGB for sending to a screen device.
+
 \end{document}

### 2007-10-31 pkg/paper/Colorspace/article.Rnw
added another idea for an example/demo to the colorspace article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@16 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
minor addition
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@17 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
minor additions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@18 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\textit{Why not use CIE XYZ for storage?  (especially because
+sRGB has limited gamut)  Because CIE storage would
+mean lots more conversions when sending output to graphics devices(?)}
+
 \section{Other color spaces}
 
-HLS (and others) because you may have been given a colour in
+HSV and HLS (and others) because you may have been given a colour in
 such a (crappy) colorspace and want to convert it to something
-better in order to work with it.
+better in order to work with it.  And basically just for completeness!
 
@@
 
+Another example is converting an entire image to greyscale
+(as in the tiger example used in grImport talks).
+
 \end{document}

### 2007-10-31 pkg/paper/Palettes/CSDA.Rnw
possible addition to intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@19 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{amsfonts,rotating}
+\usepackage{paralist}
 
@@
 Color is an integral element of graphical displays in general, and many statistical
-graphics in particular. Statistical software packages typically provide various color
-palettes and allow practitioners to employ these for visualizing data in various types
-of displays. However, more often than not, there is relatively little guidance about
-how to choose a suitable palette for a certain visualization task. Consequently,
-sub-optimal color palettes are used quite frequently for color graphics displayed in journal articles
-or on presentation slides which could easily be enhanced if the underlying information
-were captured appropriately in the color schemes. Here, we address
-this problem by suggesting a color selection strategy for visualizing both categorical
-and numerical information by selecting colors along axes in a color space whose axes
-can be matched with perceptual axes of the human visual system.
+graphics in particular.  Furthermore, 
+an increasing amount of research is being published and viewed online, making
+it easier to include color in published plots.
+However, selecting which colors to use for a statistical graphic is
+not a trivial task.
+
+There are three main obstacles to overcome:
+\begin{compactitem}
+\item
+The colors should not be unappealing.  
+\end{compactitem}
+
+It is not necessary for the 
+colors in a statistical plot to reflect the latest in fashion trends, 
+but basic 
+principles such as avoiding fully saturated colors should be
+adhered to.  The requirement is not that the user should have
+a degree in graphic design, but
+that the software should allow the user to select colors and provide
+control over basic properties such as how bright the colors are.
+
+In order to be able to control basic perceptual features of colors,
+it is necessary to work in a \emph{colorspace} that describes 
+colors in terms of their perceptual properties.
+This article will describe an appropriate colorspace for
+selecting colors.
+
+\begin{compactitem}
+\item 
+The colors in a statistical graphic
+should cooperate with each other.  
+\end{compactitem}
+
+The typical purpose 
+of color in a statistical graphic is to distinguish between different 
+areas or symbols in the plot---to distinguish between different
+groups or between different levels of a variable.  This means
+that there will typically be several colors, or a \emph{palette}
+of colors, used within a plot and 
+that those colours should be related to each other.
+
+Part of the solution to this problem is to work in an
+appropriate colorspace, as in the previous point, so that 
+motion within the colorspace and distances between colors 
+have an intuitive and rational meaning.  However, 
+there still remains the issue of providing basic navigation 
+tools so that users can conveniently generate color
+palettes.  The main part of this article will describe
+algorithms for generating color palettes for common 
+applications in statistical graphics.
+
+\begin{compactitem}
+\item
+The colors should work everywhere.  
+\end{compactitem}
+
+In the ideal situation, colors 
+should be selected so that they continue to work in any 
+context.  For example, different areas of a plot should still be 
+distiguishable when printed in black-and-white, or when the person
+viewing the graphic is colour-blind.
+This article will also briefly address these issues.
 
@@
 whereas long waiting times (around 80 minutes) can be followed by either a long
-or short eruption (around 4 minutes).
+or short eruption (around 2 minutes).
 

### 2007-11-02 pkg/paper/Palettes/CSDA.Rnw
updated Section 2 for CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@22 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
-\usepackage{amsfonts,rotating}
-\usepackage{paralist}
 
+%FIXME% Different/better title?
 \title{Choosing Color Palettes for Statistical Graphics}
@@
 However, selecting which colors to use for a statistical graphic is
-not a trivial task.
+not a trivial task. There are three main obstacles to overcome:
 
-There are three main obstacles to overcome:
-\begin{compactitem}
-\item
-The colors should not be unappealing.  
-\end{compactitem}
-
+\paragraph{The colors should not be unappealing.}
 It is not necessary for the 
@@
 control over basic properties such as how bright the colors are.
-
 In order to be able to control basic perceptual features of colors,
@@
 
-\begin{compactitem}
-\item 
-The colors in a statistical graphic
-should cooperate with each other.  
-\end{compactitem}
-
+\paragraph{The colors in a statistical graphic should cooperate with each other.}
 The typical purpose 
@@
 that those colours should be related to each other.
-
 Part of the solution to this problem is to work in an
@@
 
-\begin{compactitem}
-\item
-The colors should work everywhere.  
-\end{compactitem}
-
+\paragraph{The colors should work everywhere.}
 In the ideal situation, colors 
@@
 This article will also briefly address these issues.
+%FIXME% Also mention that the colors should work on different devices.
+%FIXME% Explain that we discuss how and under which circumstances this
+%FIXME% is achieved.
+
+%FIXME% A sentence or so is needed as a "bridge".
 
@@
 
-To show what can be gained by selecting appropriate color schemes, we present a collection
-of illustrations with typical color graphics based on HSV palettes (always in the left panels)
-and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
-of the data sets) are taken from recent publications in statistical journals such as
-\textit{Journal of the American Statistical Association},
-\textit{Journal of Computational and Graphical Statistics} or
-\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
-an overview of typically colored statistical graphics and their respective pitfalls (concerning
-color palette choice).
-%% All graphics are in principle very useful but can be even enhanced
-%% by using better color schemes.
+To show what can be gained by selecting appropriate color schemes,
+we present two examples for typical color graphics, contrasting
+commonly-used HSV palettes with more suitable HCL palettes.
+Both examples and all HSV palettes are taken from (the electronic
+version of) recent publications in
+\textit{Computational Statistics \& Data Analysis} (Volume~51).
 
 Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
-two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
-Faithful geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications
-include objective functions with two arguments \citep[as in][]{color:Gneiting+Sevcikova+Percival:2006}
-or (physical) measurements on a 2-dimensional grid \citep[as in][]{color:Yang+Buckley+Dudoit:2002}.
-Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
+two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old Faithful
+geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications include 
+classification maps \citep[as in][Figure~2]{color:Tenenhaus+Giron+Viennet:2007},
+function surfaces \citep[as in][Figure~8]{color:Harezlak+Coull+Laird:2007}, or
+measurements on a spatial grid \citep[as in][Figure~3]{color:Gelfand+Banerjee+Sirmans:2007}.
+Figure~\ref{fig:bkde} shows a heatmap bringing out the relationship between the duration
 of an eruption of the Old Faithful geyser in Yellowstone National Park 
@@
   plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
-    if(is.null(col)) col <- rev(grey.colors(33))
+    if(is.null(col)) col <- rev(gray.colors(33))
     if(legend) {
@@
   png(file = "bkde-pal.png", height = 160, width = 396)
-  par(mfrow = c(4, 1), mar = c(0, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
+  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
   plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
@@
   box()
-  axis(1, cex.axis = 1.5)
+  axis(1, cex.axis = 1.8)
   dev.off()
@@
 palettes (as in the two left two panels). These palettes code the variable of interest
-by varying hue in an HSV color wheel as done by \citet[Figure~4a]{color:Yang+Buckley+Dudoit:2002}
-or \citet[Figures~1--4]{color:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
+by varying hue in an HSV color wheel as done by 
+\citet[Figure~8]{color:Harezlak+Coull+Laird:2007}
+or \citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
+The palette in the upper left panel
 codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
@@
 
-%FIXME% examples for sequential palettes:
-%%      color:Gelfand+Banerjee+Sirmans:2007
-
 In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
 coding the variable of interest much better and without flashy colors. If, however, the user wants
-to add some color to the plot, this could be done by using a better balanced version of the heat
-colors (as shown in the bottom right panel). These colors also increase from a yellow to a red
-hue while being balanced towards the same gray as in the grayscale palette (i.e., when converted
-to a grayscale or printed out on a grayscale printer, the upper and lower right panel would 
-essentially look identical). Both palettes have in common that they give increasing perceptual emphasis
+to increase the contrast by adding some color to the plot, this could be done by using a better
+balanced version of the heat colors (as shown in the bottom right panel). These colors also
+increase from a yellow to a red hue while being balanced towards the same gray as in the
+grayscale palette. Thus, when converted to a grayscale or printed out on a grayscale printer
+(as in the printed version of \textit{Computational Statistics \& Data Analysis}),
+the upper and lower right panel would look (virtually) identical.
+Both palettes have in common that they give increasing perceptual emphasis
 to regions with increasing density, resulting in a heatmap that highlights the (small)
 interesting high-density regions and not to the large low-density regions surrounding them.
+By balancing the lower right panel towards light/dark contrasts, it is assured that the graphic
+is intelligible for color-blind readers and that the same graphic works in
+colored electronic versions and printed grayscale versions of a publication.
 
@@
   png(file = "nigeria-pal.png", height = 120, width = 396)
-  par(mfrow = c(3, 1), mar = c(0, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
+  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
   plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
@@
   box()
-  axis(1, cex = 1.5)
+  axis(1, cex.axis = 1.8)
   dev.off()
@@
 
-A similar example is presented in Figure~\ref{fig:nigeria}
+The second example is presented in Figure~\ref{fig:nigeria}
 \citep[taken from][Figure~5, left]{color:Kneib:2006}, depicting posterior mode estimates
-for childhood mortality in different regions of Nigeria. Deviations from a model for
-childhoold mortality are brought out by shading a map according to the corresponding
-residuals, revealing decreased mortality in the south-west and increased mortality in
+for childhood mortality in different regions of Nigeria. Spatial variations (not included
+in the model) in childhood mortality are brought out by shading a map according to the
+corresponding model deviations, 
+revealing decreased mortality in the south-west and increased mortality in
 the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the residuals by
-the hue, going from green via yellow to red. Our HCL-based palette also employs green
+the hue, going from green via yellow to red (see the upper left panel of Figure~\ref{fig:nigeria}).
+Similar approaches are used by \citet[Figure~8]{color:Harezlak+Coull+Laird:2007} and
+\citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
+Our first HCL-based palette in the middle left panel also employs green
 and red hues for negative and postive residuals respectively, but codes neutral values
@@
 
-%FIXME% other bad examples for diverging
-%%      color:Harezlak+Coull+Laird:2007
-%%      color:Tenenhaus+Giron+Viennet:2007
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<seats, fig=TRUE, height=4, width=11>>=
-par(mar = rep(0.8, 4), mfrow = c(1, 2))
-seats <- structure(c(226, 61, 54, 51, 222),
-  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
-parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
-parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
-names(parties_hcl) <- names(parties_hsv) <- names(seats)
-names(seats)[4] <- "Grüne"
-pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
-pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
-@
-\caption{\label{fig:seats} Seats in the German parliament.}
-\end{center}
-\end{figure}
-
-Although problematic for many tasks, pie charts can be useful for visualizing
-whether a set of pie segments constitutes a majority. A typical application is shown
-in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
-``Bundestag'' following the 2005 election. In this election, five parties were able to
-obtain enough votes to enter the Bundestag---however, neither the governing
-coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
-Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
-a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
-by using colors as metaphors: red for the social democrats SPD, black for the conservative
-CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
-for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
-found in the (German) media whereas the right panel uses less flashy HCL colors with the
-same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
-is used). The advantage of the latter is again that they are easier to hold in focus for
-a longer time. Furthermore, they are all balanced towards the same gray and have the same
-amount of color, resulting in a perceptually balanced palette that does not introduce undesired
-graphical distractions. While it could be argued that this pie chart is such a simple
-display that a well-balanced palette is not so important and that flashy colors are to
-be preferred, the balancing properties of the HCL-based palette are very important in more
-complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
-colors and shows the distribution of votes in the 2005 election
-stratified by province (Bundesland). The order of provinces is from 
-north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
-then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
-Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
-Die Linke performed particularly well in the eastern provinces and in Saarland.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{sidewaysfigure}[p]
-\begin{center}
-<<votes, fig=TRUE, height=6, width=14, eval=FALSE>>=
-data("Bundestag2005", package = "vcd")
-votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
-                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
-parties_hcl <- parties_hcl[colnames(votes)]
-parties_hsv <- parties_hsv[colnames(votes)]
-dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
-dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
-
-grid.newpage()
-pushViewport(viewport(layout = grid.layout(1, 2)))
-pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(votes, gp = gpar(fill = parties_hsv),
-  spacing = spacing_highlighting, labeling = labeling_left,
-  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
-  keep_aspect_ratio = FALSE, newpage = FALSE)
-popViewport()
-pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-mosaic(votes, gp = gpar(fill = parties_hcl),
-  spacing = spacing_highlighting, labeling = labeling_left,
-  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
-  keep_aspect_ratio = FALSE, newpage = FALSE)
-popViewport(2)
-@
-\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
-\end{center}
-\end{sidewaysfigure}
-
-Color choice is usually much more important in graphical displays with
-shaded areas compared to displays with only points or lines. However, color choice
-becomes much more relevant also in scatter plots when there are many points. As an
-example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
-the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
-membership, three different colors (green, yellow, red) are employed. The HSV colors in
-the left panel are again very flashy and differ strongly with respect to luminance: the
-yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
-panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
-the amount of color and gray.
-
-As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
-visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
-Clearly, hair and eye color are not independent and the pattern of association is highlighted
-by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
-(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
-with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
-under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
-or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
-it is shown again that the HCL colors are better balanced (between the red and blue colors)
-and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
-to read when shaded in a light grey rather than white. More details and extensions to residual-based
-shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[h!]
-\begin{center}
-<<scatter>>=
-if(generate_png) {
-  set.seed(123)
-  vc <- function(var1, var2, cov)
-    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
-  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
-               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
-               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
-  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))
-
-  png(file = "scatter.png", height = 350, width = 840)
-  par(mfrow = c(1, 2), mar = rep(0.5, 4))
-  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
-    col = hsv(c(1/3, 1/6, 0))[id])
-  box()
-  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
-    col = hcl2hex(c(120, 60, 0), 70, 65, fix = FALSE)[id])
-  box()
-  dev.off()
-}
-@
-\includegraphics{scatter}
-\caption{\label{fig:scatter} Scatter plot with three clusters.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[h!]
-\begin{center}
-<<haireye, fig=TRUE, height=5.5, width=12, eval=FALSE>>=
-he <- margin.table(HairEyeColor, 1:2)
-grid.newpage()
-pushViewport(viewport(layout = grid.layout(1, 2)))
-pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
-popViewport()
-pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
-popViewport(2)
-@
-\caption{\label{fig:haireye} Extended mosaic display for hair and eye color data.}
-\end{center}
-\end{figure}
+Another advantage of the HCL-based palette becomes apparent when we collapse the
+red-green distinctions in the colors to emulate what a person with protanopia
+(a common red-green color blindness) would see. The right panel of Figure~\ref{fig:nigeria}
+displays the respective palettes from the left panel, but transformed for protanopic
+vision using the tools of \cite{color:Lumley:2006}. In the upper right panel, the
+low-mortality regions in the south-west just ``disappear'' because they are transformed
+to virtually the same yellow as the average-mortality regions. In contrast, in the middle
+right panel, regions with increased and decreased mortality remain clearly distinguishable.
+Only the type and amount of color is changed (red turns gray, green turns yellow/brown) but
+regions with large deviations are still easily indentifiable by being balanced towards
+darker grays. To avoid that one of the two ``branches'' of the palette is only gray
+in the protanopic version, we could also employ a purple (rather than a red) hue for
+positive deviations. This is used in the lower panels of Figure~\ref{fig:nigeria}
+showing that after collapsing red-green distinctions, a useful yellow/brown--gray--blue
+palette remains from the original green--gray--purple palette. Thus, the lower left
+panel could be used ensuring that it works for both normal and protanopic vision.
+Similar arguments hold for deuteranopic vision.
 
@@
 3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
-\cite{color:Ihaka:2003} for more details and references.
+\cite{color:Ihaka:2003} and \cite{color:Lumley:2006} for more details and references.
 
@@
 The brightness of
-colors is not uniform over hues and saturations (given value)---therefore,
+colors is not uniform over hues and saturations (given value, see Figure~\ref{fig:wheel})---therefore,
 HSV colors are often not considered to be perceptually based.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[b!]
+\begin{center}
+<<wheel, fig=TRUE, height=4, width=11>>=
+par(mar = rep(0.8, 4), mfrow = c(1, 2))
+par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
+pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
+pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
+@
+\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
+\end{center}
+\end{figure}
 
@@
 Given a certain luminance $L$, all colors resulting from different combinations
-of hue $H$ and chroma $C$ are balanced towards the same gray (and look the same
+of hue $H$ and chroma $C$ are balanced towards the same gray (and thus look virtually identical
 when converted to a gray scale). However, the admissible combinations of
@@
 \subsection{Qualitative palettes} \label{sec:qualitative}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
-\begin{center}
-<<wheel, fig=TRUE, height=4, width=11>>=
-par(mar = rep(0.8, 4), mfrow = c(1, 2))
-par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
-pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
-pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
-@
-\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
-\end{center}
-\end{figure}
 
@@
 important than any other one. Typical applications of qualitative palettes in
-statistics would be bar plots \citep[see][]{color:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
-or highlighted mosaic displays (see Figure~\ref{fig:votes}).
+statistics would be bar plots \citep[see][]{color:Ihaka:2003}, pie charts
+or highlighted mosaic displays.
 
@@
 probabilities (with neutral value 0.5) that could be visualized in mosaic plots
-(see Figure~\ref{fig:haireye}), classification maps or model-based shading in maps
-such as Figure~\ref{fig:nigeria}. Analogously to the previous
+\citep[e.g., as in][Figure~3]{color:Zeileis+Meyer+Hornik:2007},
+classification maps \citep[e.g., as in][Figure~2]{color:Tenenhaus+Giron+Viennet:2007}
+or model-based shading in maps such as Figure~\ref{fig:nigeria}. Analogously to the previous
 section, we suppose that we want to visualize an intensity or interestingness $i$
@@
 hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
-luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
+luminance or chroma contrasts---from a full red over a neutral gray to a full blue.
 As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
@@
 \end{document}
-
-<<barplot, eval=FALSE>>=
-x <- rbind(
-c(255, 255, 111),
-c(255, 255,  47),
-c(120,  50,   1),
-c(102, 204, 102),
-c( 50, 204, 204),
-c(203, 153, 204),
-c(102, 204, 102),
-c(255,  47,  47),
-c(102, 204, 102))/255
-
-y <- c(10, 20, 25, 18, 16, 13, 21, 15, 18) - 3
-
-par(mfrow = c(1, 2))
-barplot(y, col = hex(RGB(x[,1], x[,2], x[,3])), space = 0)
-## barplot(y, col = hcl(attr(as(RGB(x[,1], x[,2], x[,3]), "polarLUV"), "coords")[,"H"],
-##  c = 40, l = 80), space = 0)
-barplot(y, col = rainbow_hcl(7, c = 40, l = 80)[c(3, 5, 2, 4, 6, 1, 4, 7, 4)], space = 0)
-@
-
-<<barplot2, eval=FALSE>>
-## from Takasaki & Kwamura (2007, CSDA, Figure 2, Columns 1-6)
-x <- cbind(c( 2,  6, 91,  1), c(35, 30, 20, 15), c(41,  9,  8, 42),
-           c(10, 40, 40, 10), c(21, 39, 30, 10), c(35, 13, 17, 35))
-mybarplot <- function(x, col = NULL, ...) {
-  barplot(x, space = 1.5, names = 1:6, las = 1, col = col)
-  grid(lty = 1, ny = NULL, nx = 0, col = "black")
-  barplot(x, space = 1.5, names = 1:6, las = 1, col = col, add = TRUE)
-}
-
-par(mfrow = c(1, 3))
-mybarplot(x, col = c("#8995FF", "#AA0DA4", "#F9FFB9", "#B3FFEE"))
-mybarplot(x, col = hcl(c(260, 310, 60, 150), c = 40, l = 80))
-mybarplot(x, col = grey.colors(4))
-@

### 2007-11-06 pkg/paper/Palettes/CSDA.Rnw
hack and slash on the intro to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@23 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
+
+\usepackage{amsfonts}
 
@@
 colors in terms of their perceptual properties.
-This article will describe an appropriate colorspace for
+
+Typically, software allows the user to specify colors as 
+Red-Green-Blue (RGB) triplets.  
+RGB is an additive
+color model which is used for generating colors on computer screens
+\citep[by mixing different amounts of the primary colors red, green and blue,
+see][for more details]{color:Poynton:2000}.
+Unfortunately, it is impossible to control
+the perceptual properties of a color in this colorspace because there
+is no single dimension that corresponds to, for example, the brightness
+of the colour.  In
+this article we will describe a more appropriate colorspace for
 selecting colors.
@@
 of colors, used within a plot and 
-that those colours should be related to each other.
+that those colors should be related to each other.
+
 Part of the solution to this problem is to work in an
@@
 motion within the colorspace and distances between colors 
-have an intuitive and rational meaning.  However, 
-there still remains the issue of providing basic navigation 
-tools so that users can conveniently generate color
-palettes.  The main part of this article will describe
-algorithms for generating color palettes for common 
-applications in statistical graphics.
-
-\paragraph{The colors should work everywhere.}
-In the ideal situation, colors 
-should be selected so that they continue to work in any 
-context.  For example, different areas of a plot should still be 
-distiguishable when printed in black-and-white, or when the person
-viewing the graphic is colour-blind.
-This article will also briefly address these issues.
-%FIXME% Also mention that the colors should work on different devices.
-%FIXME% Explain that we discuss how and under which circumstances this
-%FIXME% is achieved.
-
-%FIXME% A sentence or so is needed as a "bridge".
-
-For implementing color palettes in computer programs, typically color spaces are
-employed that provide a mapping to Red-Green-Blue (RGB) colors. RGB is an additive
-color model which is used for generating colors on computer screens
-\citep[by mixing different amounts of the primary colors red, green and blue,
-see][for more details]{color:Poynton:2000}. While being eminently useful for generating
-colors on computer screens, it is hard for humans to select the coordinates in RGB
-space corresponding to a certain color. Therefore, color picker tools in computer
-programs are based on different color models---starting from the pioneering introduction
-of Hue-Saturation-Value (HSV) colors \citep{color:Smith:1978}---which try to capture the
-dimensions of human visual perception. HSV space is a simple transformation of RGB
-space which is also implemented in many statistical software packages
+have an intuitive and rational meaning.  
+For example, another common software 
+approach \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004} 
+is to allow
+colors to be specified as Hue-Saturation-Value (HSV) 
+triplets \citep{color:Smith:1978},
+where each dimension controls a separate perceptual 
+property of the color.  
+HSV space is a simple transformation of RGB
+space 
 \citep[see][for more details on HSV colors and links to further information about the other color
-spaces discussed in the following]{color:Wiki+HSV:2007}. Although being easy to use,
-HSV space exhibits a number of disadvantages: It is relatively difficult to select sets of
+spaces discussed in the following]{color:Wiki+HSV:2007} and,
+unfortunately, the dimensions in 
+HSV space are not truly independent.  
+For example,
+a change in hue results in a color that may also be
+perceptually brighter and more saturated.  
+This means that it is relatively difficult to select sets of
 HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{color:Munsell:1905}.
-The reason for this is that colors with different hues also differ in saturation and brightness.
 For statistical graphics, this is important because it can introduce size distortions in
@@
 
-Despite these known drawbacks, HSV space is still the predominantly-used color model in
-color picker tools \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004}
-and the basis for many palettes in statistical software packages. A notable
-exception is \pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003},
-an online tool for selecting color schemes for maps. It provides a rich
+A different approach to obtaining a color palette is provided by the 
+\pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003}
+online tool for selecting color schemes for maps. It provides a rich
 collection of prefabricated palettes (with a fixed maximal number
 of colors) and guides its users how to choose a suitable palette for coding
-various types of information. Here, we take a similar approach
-and describe strategies for the choice of color palettes for categorical and numerical
-data. Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
+various types of information.  The drawback to this tool is that
+it only provides a fixed set of colors for each palette and there
+is no way to extend the existing palettes.
+
+Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
 three types of palettes: qualitative, sequential and diverging. The first is tailored
@@
 (or ordinal variables if some numerical coding for its levels is used).
-Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
+Unlike \pkg{ColorBrewer.org}, we not only provide fixed sets of colors but suggest
 a general principle for selecting colors by traversing paths along perceptual axes in
 a suitable color space. Consequently, the user can decide which path exactly should be
-taken and how many colors should be selected. The color space employed by us is the
+taken and how many colors should be selected. The color space we will work in
+ is the
 Hue-Chroma-Luminance (HCL) color space \citep[see][]{color:Ihaka:2003} and we show how
@@
 coordinates in this space.
+
+\paragraph{The colors should work everywhere.}
+The final issue to deal with is that, in an ideal situation, colors 
+should be selected so that they continue to work in any 
+context.  For example, different areas of a plot should still be 
+distiguishable when the person
+viewing the graphic is color-blind
+or when the graphic is printed in black-and-white instead of being
+viewed on screen.
+These goals cannot always be attained, but we will discuss 
+situations where attention should be paid to these issues and
+suggest strategies for resolving any problems.
+%FIXME% Also mention that the colors should work on different devices.
+
+%FIXME% A sentence or so is needed as a "bridge".
+
 %% Given a mapping from HCL to RGB coordinates, our formulas are
@@
 
+\paragraph{}
 The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
@@
 For choosing the hues in a certain palette, various strategies are conceivable.
+%% FIXME:  the figures referred to below no longer exist!!
 A simple and intuitive one is to use colors as metaphors for categories

### 2007-11-06 pkg/paper/Palettes/CSDA.Rnw
yet another revision of the palettes paper intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@24 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%FIXME% The intro is a bit too long and needs to be condensed.
+
 Color is an integral element of graphical displays in general, and many statistical
-graphics in particular.  Furthermore, 
-an increasing amount of research is being published and viewed online, making
-it easier to include color in published plots.
-However, selecting which colors to use for a statistical graphic is
-not a trivial task. There are three main obstacles to overcome:
+graphics in particular. Statistical software packages allow practicioners
+to easily create color graphics which are therefore virtually omnipresent in electronic
+publications such as technical reports, presentation slides, or the electronic
+version of journal articles (e.g., in \emph{Computational Statistics \& Data Analysis})
+and increasingly also in printed journals. However, more often than not,
+color choice in such displays is sub-optimal because there is relatively
+little guidance about how to choose appropriate colors for a particular visualization
+task---and doing so is not a trivial task. There are three main obstacles to overcome:
 
 \paragraph{The colors should not be unappealing.}
-It is not necessary for the 
-colors in a statistical plot to reflect the latest in fashion trends, 
-but basic 
-principles such as avoiding fully saturated colors should be
-adhered to.  The requirement is not that the user should have
-a degree in graphic design, but
-that the software should allow the user to select colors and provide
-control over basic properties such as how bright the colors are.
-In order to be able to control basic perceptual features of colors,
-it is necessary to work in a \emph{colorspace} that describes 
-colors in terms of their perceptual properties.
+It is not necessary for the colors in a statistical plot to reflect the latest
+in fashion trends, but basic principles such as avoiding fully saturated colors
+should be adhered to. The requirement is not that the user should have
+a degree in graphic design, but that the software should provide users with an
+intuitive way to select colors and control their basic properties such as how
+bright the colors are. Thus, it is necessary to employ a \emph{color model}
+or \emph{color space} that describes colors in terms of their perceptual
+properties hue, brightness, and colorfulness.
 
-Typically, software allows the user to specify colors as 
-Red-Green-Blue (RGB) triplets.  
-RGB is an additive
-color model which is used for generating colors on computer screens
-\citep[by mixing different amounts of the primary colors red, green and blue,
-see][for more details]{color:Poynton:2000}.
-Unfortunately, it is impossible to control
-the perceptual properties of a color in this colorspace because there
-is no single dimension that corresponds to, for example, the brightness
-of the colour.  In
-this article we will describe a more appropriate colorspace for
-selecting colors.
+A color model typically supported by software packages is the specification
+of colors as Red-Green-Blue (RGB) triplets. However, this specification
+does not correspond to human color perception but to color generation on
+a computer screen where different color sensations are created by
+by mixing different amounts of the primary colors red, green and blue
+\citep[see][for more details]{color:Poynton:2000}. For humans, it is
+virtually impossible to control the perceptual properties of a color in this
+color space because there is no single dimension that corresponds to, for example,
+the hue or the brightness of the color. Therefore, various perceptually based
+color spaces have been suggested where each dimension of the color space can
+be matched with a perceptual property. One approach widely implemented in
+software packages are Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
+a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
+Unfortunately, the dimensions in HSV space map poorly to perceptual
+properties---furthermore, the use of HSV colors encourages the use of flashy and
+highly saturated colors which are good for drawing attention to a plot but hard
+to look at for a longer time. A perceptually based color model that avoids
+these problems are Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
+a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
+Albeit being less commonly available in software packages, it is extremely
+useful for sepecifying colors intuitively.
 
 \paragraph{The colors in a statistical graphic should cooperate with each other.}
-The typical purpose 
-of color in a statistical graphic is to distinguish between different 
-areas or symbols in the plot---to distinguish between different
-groups or between different levels of a variable.  This means
+The typical purpose of color in a statistical graphic is to distinguish
+between different areas or symbols in the plot---to distinguish between
+different groups or between different levels of a variable.  This means
 that there will typically be several colors, or a \emph{palette}
-of colors, used within a plot and 
-that those colors should be related to each other.
+of colors, used within a plot and that those colors should be related to
+each other.
 
-Part of the solution to this problem is to work in an
-appropriate colorspace, as in the previous point, so that 
-motion within the colorspace and distances between colors 
-have an intuitive and rational meaning.  
-For example, another common software 
-approach \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004} 
-is to allow
-colors to be specified as Hue-Saturation-Value (HSV) 
-triplets \citep{color:Smith:1978},
-where each dimension controls a separate perceptual 
-property of the color.  
-HSV space is a simple transformation of RGB
-space 
-\citep[see][for more details on HSV colors and links to further information about the other color
-spaces discussed in the following]{color:Wiki+HSV:2007} and,
-unfortunately, the dimensions in 
-HSV space are not truly independent.  
-For example,
-a change in hue results in a color that may also be
-perceptually brighter and more saturated.  
-This means that it is relatively difficult to select sets of
-HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{color:Munsell:1905}.
-For statistical graphics, this is important because it can introduce size distortions in
-the perception of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
-Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
-colors which are good for drawing attention to a plot but hard to look at for a longer time.
-
-A different approach to obtaining a color palette is provided by the 
-\pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003}
-online tool for selecting color schemes for maps. It provides a rich
-collection of prefabricated palettes (with a fixed maximal number
-of colors) and guides its users how to choose a suitable palette for coding
-various types of information.  The drawback to this tool is that
-it only provides a fixed set of colors for each palette and there
-is no way to extend the existing palettes.
-
-Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
-three types of palettes: qualitative, sequential and diverging. The first is tailored
-for coding categorical information and the latter two are aimed at numerical variables
-(or ordinal variables if some numerical coding for its levels is used).
-Unlike \pkg{ColorBrewer.org}, we not only provide fixed sets of colors but suggest
-a general principle for selecting colors by traversing paths along perceptual axes in
-a suitable color space. Consequently, the user can decide which path exactly should be
-taken and how many colors should be selected. The color space we will work in
- is the
-Hue-Chroma-Luminance (HCL) color space \citep[see][]{color:Ihaka:2003} and we show how
-the general principles suggested can be turned into formulas describing paths of
-coordinates in this space.
+A natural solution to this task is to vary (at least) one perceptual property
+of the colors, e.g., the hue or the brightness, keeping other properties
+fixed. In a perceptually based color space, this corresponds to selecting
+colors by traversing paths through this space along its dimensions.
+This approach is implemented in many color picker tools
+\citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
+however, most commonly based on the HSV model.
+As argued above, the dimensions in HSV space are not truly independent and
+hence it is not possible to vary just one perceptual property while keeping
+the others fixed. For example, a change in hue results in a color that may also be
+perceptually brighter and more saturated. This means that it is relatively
+difficult to select sets of HSV coordinates that yield colors that are
+``in harmony'' \citep[see][]{color:Munsell:1905}. For statistical graphics,
+this is important because it can introduce size distortions in the perception
+of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
+These problems can again be addressed by employing HCL colors as a more
+appropriate perceptually based color model. Above, we argued that HCL space
+is useful because it allows humans to understand where within the space
+a particular HCL triplet is located. Here, we extend this argument: HCL also allows
+us to understand where we are going, i.e., motion within the color space and
+distances between colors  have an intuitive and rational meaning.
 
@@
 context.  For example, different areas of a plot should still be 
-distiguishable when the person
-viewing the graphic is color-blind
-or when the graphic is printed in black-and-white instead of being
-viewed on screen.
-These goals cannot always be attained, but we will discuss 
-situations where attention should be paid to these issues and
-suggest strategies for resolving any problems.
-%FIXME% Also mention that the colors should work on different devices.
+distinguishable when the graphic is displayed on an LCD projector
+rather than a computer screen, or when it is printed on a grayscale
+printer, or when the person viewing the graphic is color-blind.
+These goals cannot always be attained, but attention should be paid
+to these issues and in many situations it is also possible to
+resolve any problems.
 
-%FIXME% A sentence or so is needed as a "bridge".
+An approach exemplary in this respect is \pkg{ColorBrewer.org}
+\citep{color:Harrower+Brewer:2003}, an online tool for selecting color
+schemes for maps. It provides a rich collection of prefabricated palettes
+(with a fixed maximal number of colors), supplies information about the
+suitability for color-blind viewers or grayscale printing, and guides its
+users how to choose a suitable palette for coding various types of
+information. The drawback to this tool is that it only provides a fixed
+set of colors for each palette and there is no way to extend the
+existing palettes.
+
+Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we 
+take a similar approach and distinguish three types of palettes: qualitative,
+sequential and diverging. The first is tailored for coding categorical
+information and the latter two are aimed at numerical variables (or ordinal
+variables if some numerical coding for its levels is used). Unlike
+\pkg{ColorBrewer.org}, we not only provide fixed sets of colors but suggest
+a general principle for selecting colors by traversing paths along perceptual axes in
+HCL color space. Consequently, the user can decide which path exactly should be
+taken and how many colors should be selected. Furthermore, by matching
+the paths with perceptual meaning, the suitability of a palette for color-blind
+viewers or grayscale printing can be assessed.
 
@@
 
-\paragraph{}
 The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
-several motivating examples, showing how typical HSV-based graphics can be enhanced
+some motivating examples, showing how typical HSV-based graphics can be enhanced
 by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
@@
 Section~\ref{sec:software} offers some general remarks on the implementation in statistical
-software packages as well as some details on the implementation in \proglang{R}.
+software packages as well as some details on the implementation in the \proglang{R}
+system for statistical computing \citep{color:R:2007}.
 Section~\ref{sec:discussion} concludes the paper with a discussion.
@@
 balanced version of the heat colors (as shown in the bottom right panel). These colors also
-increase from a yellow to a red hue while being balanced towards the same gray as in the
-grayscale palette. Thus, when converted to a grayscale or printed out on a grayscale printer
+increase from a yellow to a red hue while using the same brightness levels as in the grayscale
+palette. Thus, when converted to a grayscale or printed out on a grayscale printer
 (as in the printed version of \textit{Computational Statistics \& Data Analysis}),
@@
 interesting high-density regions and not to the large low-density regions surrounding them.
-By balancing the lower right panel towards light/dark contrasts, it is assured that the graphic
-is intelligible for color-blind readers and that the same graphic works in
+By balancing the lower right panel with respect to its brightness levels (i.e., light/dark contrasts),
+it is assured that the graphic is intelligible for color-blind readers and that the same graphic works in
 colored electronic versions and printed grayscale versions of a publication.
@@
 and negative residuals with the same absolute size receive the same perceptual weight
-by being balanced towards the same gray.
+by being balanced with respect to their brightness.
 
@@
 Given a certain luminance $L$, all colors resulting from different combinations
-of hue $H$ and chroma $C$ are balanced towards the same gray (and thus look virtually identical
-when converted to a gray scale). However, the admissible combinations of
+of hue $H$ and chroma $C$ have the same level of brightness, i.e.,
+are balanced towards the same gray and thus look virtually identical
+when converted to a gray scale. However, the admissible combinations of
 chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
@@
 For choosing the hues in a certain palette, various strategies are conceivable.
-%% FIXME:  the figures referred to below no longer exist!!
-A simple and intuitive one is to use colors as metaphors for categories
-(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
+A simple and intuitive one is to use colors as metaphors for categories (e.g., for political parties),
+%% (e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
 another approach would be to use segments from
@@
 Technical documentation along with a large collection of examples is available via
-\verb/help("rainbow_hcl", package = "vcd")/. \proglang{R} code for reproducing
+\verb/help("rainbow_hcl", package = "vcd")/. Furthermore, \proglang{R} code for reproducing
 Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}

### 2007-11-07 pkg/paper/Colorspace/article.Rnw
HLS to HSL in article.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@26 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-HSV and HLS (and others) because you may have been given a colour in
+HSV and HSL (and others) because you may have been given a colour in
 such a (crappy) colorspace and want to convert it to something

### 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
minor changes to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@27 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Color is an integral element of graphical displays in general, and many statistical
-graphics in particular. Statistical software packages allow practicioners
-to easily create color graphics which are therefore virtually omnipresent in electronic
+graphics in particular. 
+It is easy to create color graphics with any statistical software 
+pacakge and color images are therefore virtually omnipresent in electronic
 publications such as technical reports, presentation slides, or the electronic
@@
 little guidance about how to choose appropriate colors for a particular visualization
-task---and doing so is not a trivial task. There are three main obstacles to overcome:
+task---and doing so is not a trivial task. 
+When selecting colors for a statistical graphic,
+there are three main obstacles to overcome:
 
@@
 a degree in graphic design, but that the software should provide users with an
-intuitive way to select colors and control their basic properties such as how
+intuitive way to select colors and control their basic properties, such as how
 bright the colors are. Thus, it is necessary to employ a \emph{color model}
 or \emph{color space} that describes colors in terms of their perceptual
-properties hue, brightness, and colorfulness.
+properties: hue, brightness, and colorfulness.
 
@@
 does not correspond to human color perception but to color generation on
-a computer screen where different color sensations are created by
+a computer screen where different color sensations are created
 by mixing different amounts of the primary colors red, green and blue
@@
 to look at for a longer time. A perceptually based color model that avoids
-these problems are Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
+these problems involves
+Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
+which is
 a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
-Albeit being less commonly available in software packages, it is extremely
+Although this is less commonly available in software packages, it is extremely
 useful for sepecifying colors intuitively.
@@
 
-A natural solution to this task is to vary (at least) one perceptual property
+A natural solution for this task is to vary (at least) one perceptual property
 of the colors, e.g., the hue or the brightness, keeping other properties
@@
 \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
-however, most commonly based on the HSV model.
+however, most are commonly based on the HSV model.
 As argued above, the dimensions in HSV space are not truly independent and
@@
 this is important because it can introduce size distortions in the perception
-of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
+of shaded areas and can produce
+ optical illusions \citep{color:Cleveland+McGill:1983}.
 These problems can again be addressed by employing HCL colors as a more
@@
 
-An approach exemplary in this respect is \pkg{ColorBrewer.org}
+An example of this approach is \pkg{ColorBrewer.org}
 \citep{color:Harrower+Brewer:2003}, an online tool for selecting color

### 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
minor mods to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@28 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %FIXME% Different/better title?
-\title{Choosing Color Palettes for Statistical Graphics}
+\title{Escaping RGBland: selecting colors for \\statistical graphics}
 \author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
@@
 
-Color is an integral element of graphical displays in general, and many statistical
+Color is an integral element of graphical displays in general, 
+and many statistical
 graphics in particular. 
@@
 and increasingly also in printed journals. However, more often than not,
-color choice in such displays is sub-optimal because there is relatively
-little guidance about how to choose appropriate colors for a particular visualization
-task---and doing so is not a trivial task. 
+color choice in such displays is sub-optimal because selecting colors is not
+a trivial task and there is relatively
+little guidance about how to choose appropriate colors for a particular visualization. 
 When selecting colors for a statistical graphic,
@@
 
-\paragraph{The colors should not be unappealing.}
-It is not necessary for the colors in a statistical plot to reflect the latest
-in fashion trends, but basic principles such as avoiding fully saturated colors
-should be adhered to. The requirement is not that the user should have
+\textit{The colors should not be unappealing.}
+It is not necessary for the colors in a statistical plot to reflect 
+fashion trends, but basic principles such as avoiding large areas
+of fully saturated colors \citep{Tufte-1990-EIN}
+should be adhered to. 
+The requirement is not that the user should have
 a degree in graphic design, but that the software should provide users with an
-intuitive way to select colors and control their basic properties, such as how
-bright the colors are. Thus, it is necessary to employ a \emph{color model}
+intuitive way to select colors and control their basic properties. 
+Thus, it is necessary to employ a \emph{color model}
 or \emph{color space} that describes colors in terms of their perceptual
@@
 
-A color model typically supported by software packages is the specification
+A color model typically supported by software packages involves
+ the specification
 of colors as Red-Green-Blue (RGB) triplets. However, this specification
-does not correspond to human color perception but to color generation on
-a computer screen where different color sensations are created
-by mixing different amounts of the primary colors red, green and blue
-\citep[see][for more details]{color:Poynton:2000}. For humans, it is
+corresponds to color generation on
+a computer screen 
+\citep[see][for more details]{color:Poynton:2000}
+rather than
+corresponding to human color perception. For humans, it is
 virtually impossible to control the perceptual properties of a color in this
 color space because there is no single dimension that corresponds to, for example,
-the hue or the brightness of the color. Therefore, various perceptually based
-color spaces have been suggested where each dimension of the color space can
+the hue or the brightness of the color. As a consequence, 
+various perceptually-based
+color spaces have been suggested, where each dimension of the color space can
 be matched with a perceptual property. One approach widely implemented in
-software packages are Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
+software packages involves
+ Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
 a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
 Unfortunately, the dimensions in HSV space map poorly to perceptual
-properties---furthermore, the use of HSV colors encourages the use of flashy and
-highly saturated colors which are good for drawing attention to a plot but hard
-to look at for a longer time. A perceptually based color model that avoids
+properties and the use of HSV colors encourages the use of 
+highly saturated colors. 
+A perceptually based color model that avoids
 these problems involves
@@
 a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
-Although this is less commonly available in software packages, it is extremely
-useful for sepecifying colors intuitively.
+This is the colorspace we advocate in this article.
 
-\paragraph{The colors in a statistical graphic should cooperate with each other.}
+\textit{The colors in a statistical graphic should cooperate with each other.}
 The typical purpose of color in a statistical graphic is to distinguish
@@
 hence it is not possible to vary just one perceptual property while keeping
-the others fixed. For example, a change in hue results in a color that may also be
-perceptually brighter and more saturated. This means that it is relatively
+the others fixed. This means that it is relatively
 difficult to select sets of HSV coordinates that yield colors that are
@@
  optical illusions \citep{color:Cleveland+McGill:1983}.
-These problems can again be addressed by employing HCL colors as a more
-appropriate perceptually based color model. Above, we argued that HCL space
+These problems can again be addressed by employing an HCL colorspace. 
+In addition to
+the earlier point that HCL space
 is useful because it allows humans to understand where within the space
-a particular HCL triplet is located. Here, we extend this argument: HCL also allows
-us to understand where we are going, i.e., motion within the color space and
+a particular HCL triplet is located,  
+HCL also allows
+us to understand motion within the color space because
 distances between colors  have an intuitive and rational meaning.
 
-\paragraph{The colors should work everywhere.}
+\textit{The colors should work everywhere.}
 The final issue to deal with is that, in an ideal situation, colors 
@@
 \citep{color:Harrower+Brewer:2003}, an online tool for selecting color
-schemes for maps. It provides a rich collection of prefabricated palettes
-(with a fixed maximal number of colors), supplies information about the
+schemes for maps. It provides a collection of prefabricated palettes, 
+supplies information about the
 suitability for color-blind viewers or grayscale printing, and guides its
@@
 sequential and diverging. The first is tailored for coding categorical
-information and the latter two are aimed at numerical variables (or ordinal
-variables if some numerical coding for its levels is used). Unlike
-\pkg{ColorBrewer.org}, we not only provide fixed sets of colors but suggest
-a general principle for selecting colors by traversing paths along perceptual axes in
+information and the latter two are aimed at numerical or ordinal
+variables. Unlike
+\pkg{ColorBrewer.org}, we suggest
+a general principle for selecting colors by 
+traversing paths along perceptual axes in
 HCL color space. Consequently, the user can decide which path exactly should be
@@
 \textit{Computational Statistics \& Data Analysis} (Volume~51).
-
-Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
-two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old Faithful
-geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications include 
-classification maps \citep[as in][Figure~2]{color:Tenenhaus+Giron+Viennet:2007},
-function surfaces \citep[as in][Figure~8]{color:Harezlak+Coull+Laird:2007}, or
-measurements on a spatial grid \citep[as in][Figure~3]{color:Gelfand+Banerjee+Sirmans:2007}.
-Figure~\ref{fig:bkde} shows a heatmap bringing out the relationship between the duration
-of an eruption of the Old Faithful geyser in Yellowstone National Park 
-and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
-times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
-whereas long waiting times (around 80 minutes) can be followed by either a long
-or short eruption (around 2 minutes).
 
@@
 \end{figure}
+
+Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
+two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old Faithful
+geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications include 
+classification maps \citep[as in][Figure~2]{color:Tenenhaus+Giron+Viennet:2007},
+function surfaces \citep[as in][Figure~8]{color:Harezlak+Coull+Laird:2007}, or
+measurements on a spatial grid \citep[as in][Figure~3]{color:Gelfand+Banerjee+Sirmans:2007}.
+Figure~\ref{fig:bkde} shows a heatmap bringing out the relationship between the duration
+of an eruption of the Old Faithful geyser in Yellowstone National Park 
+and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
+times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
+whereas long waiting times (around 80 minutes) can be followed by either a long
+or short eruption (around 2 minutes).
 
@@
 par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
+par(lwd=3)
 pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
-pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
+par(lwd=1, new=TRUE)
+pie(rep(1,60), radius = 1, col = rainbow(60), 
+    border=NA, labels = NA)
+par(lwd=3)
+pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), 
+    labels = 360 * 0:5/6)
+par(lwd=1, new=TRUE)
+pie(rep(1,60), radius = 1, col = rainbow_hcl(60, c = 50, l = 70), 
+    border=NA, labels = NA)
 @
@@
 This gives the users the possibility to conveniently experiment with the HCL-based
-palettes by varying several graphical parameters. For qualitative
+palettes by varying several simple and intuitive 
+graphical parameters. For qualitative
 palettes, these are the coordinates on the chroma and luminance axis, respectively,

### 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
added Tufte ref and some more touches
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@29 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%FIXME% Different/better title?
-\title{Escaping RGBland: selecting colors for \\statistical graphics}
+\title{Escaping RGBland: Selecting Colors for Statistical Graphics}
 \author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
@@
 
-\textit{The colors should not be unappealing.}
+\textbf{The colors should not be unappealing.}
 It is not necessary for the colors in a statistical plot to reflect 
 fashion trends, but basic principles such as avoiding large areas
-of fully saturated colors \citep{Tufte-1990-EIN}
+of fully saturated colors \citep{color:Tufte:1990}
 should be adhered to. 
@@
 a computer screen 
-\citep[see][for more details]{color:Poynton:2000}
+\citep[see][]{color:Poynton:2000}
 rather than
@@
 virtually impossible to control the perceptual properties of a color in this
-color space because there is no single dimension that corresponds to, for example,
+color space because there is no single dimension that corresponds to, e.g.,
 the hue or the brightness of the color. As a consequence, 
@@
  Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
-a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
+a simple transformation of RGB triplets \citep[see][]{color:Wiki+HSV:2007}.
 Unfortunately, the dimensions in HSV space map poorly to perceptual
@@
 which is
-a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
+a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
 This is the colorspace we advocate in this article.
 
-\textit{The colors in a statistical graphic should cooperate with each other.}
+\textbf{The colors in a statistical graphic should cooperate with each other.}
 The typical purpose of color in a statistical graphic is to distinguish
@@
 \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
-however, most are commonly based on the HSV model.
+however, these are typically based on the HSV model.
 As argued above, the dimensions in HSV space are not truly independent and
@@
  optical illusions \citep{color:Cleveland+McGill:1983}.
-These problems can again be addressed by employing an HCL colorspace. 
+These problems can again be addressed by employing the HCL colorspace. 
 In addition to
@@
 
-\textit{The colors should work everywhere.}
+\textbf{The colors should work everywhere.}
 The final issue to deal with is that, in an ideal situation, colors 
@@
 schemes for maps. It provides a collection of prefabricated palettes, 
-supplies information about the
-suitability for color-blind viewers or grayscale printing, and guides its
-users how to choose a suitable palette for coding various types of
+information about the suitability for printers or color-blind viewers,
+and guidance on how to choose a suitable palette for coding various types of
 information. The drawback to this tool is that it only provides a fixed
@@
 information and the latter two are aimed at numerical or ordinal
-variables. Unlike
-\pkg{ColorBrewer.org}, we suggest
+variables. Unlike \pkg{ColorBrewer.org}, we suggest
 a general principle for selecting colors by 
@@
 version of) recent publications in
-\textit{Computational Statistics \& Data Analysis} (Volume~51).
+\emph{Computational Statistics \& Data Analysis} (Volume~51).
 
@@
 palette. Thus, when converted to a grayscale or printed out on a grayscale printer
-(as in the printed version of \textit{Computational Statistics \& Data Analysis}),
+(as in the printed version of \emph{Computational Statistics \& Data Analysis}),
 the upper and lower right panel would look (virtually) identical.
@@
 pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
-par(lwd=1, new=TRUE)
-pie(rep(1,60), radius = 1, col = rainbow(60), 
-    border=NA, labels = NA)
+par(lwd=0.5, new=TRUE)
+pie(rep(1, 100), radius = 1, col = rainbow(100), border = NA, labels = NA)
 par(lwd=3)
@@
     labels = 360 * 0:5/6)
-par(lwd=1, new=TRUE)
-pie(rep(1,60), radius = 1, col = rainbow_hcl(60, c = 50, l = 70), 
-    border=NA, labels = NA)
+par(lwd=0.5, new=TRUE)
+pie(rep(1, 100), radius = 1, col = rainbow_hcl(100, c = 50, l = 70), border = NA, labels = NA)
 @
@@
 pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
+pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
 pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
-pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
 @

### 2007-11-08 pkg/paper/Palettes/CSDA.Rnw
fiddle pie chart
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@30 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 pie(rep(1, 100), radius = 1, col = rainbow(100), border = NA, labels = NA)
+par(lwd=3, new=TRUE)
+pie(1, radius=0.5, labels=NA)
+par(lwd=0.5, new=TRUE)
+pie(1, radius=0.5, border=NA, labels=NA)
 par(lwd=3)
@@
 pie(rep(1, 100), radius = 1, col = rainbow_hcl(100, c = 50, l = 70), border = NA, labels = NA)
+par(lwd=3, new=TRUE)
+pie(1, radius=0.5, labels=NA)
+par(lwd=0.5, new=TRUE)
+pie(1, radius=0.5, border=NA, labels=NA)
 @

### 2007-11-08 pkg/paper/Palettes/CSDA.Rnw
last touches, ready to ship
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@31 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
@@
 It is easy to create color graphics with any statistical software 
-pacakge and color images are therefore virtually omnipresent in electronic
+package and color images are therefore virtually omnipresent in electronic
 publications such as technical reports, presentation slides, or the electronic
@@
 a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
-This is the colorspace we advocate in this article.
+This is the color space we advocate in this article.
 
@@
  optical illusions \citep{color:Cleveland+McGill:1983}.
-These problems can again be addressed by employing the HCL colorspace. 
+These problems can again be addressed by employing the HCL color space. 
 In addition to
@@
   dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
-  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
+  plot_bkde2D <- function(x, col = NULL, xlab = "", ylab = "", legend = FALSE, ...) {
     if(is.null(col)) col <- rev(gray.colors(33))
@@
   png(file = "bkde.png", height = 810, width = 840)
-  par(mfrow = c(2, 2))
+  par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
   plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
@@
 \includegraphics{bkde-pal}
-\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
+\caption{\label{fig:bkde} Bivariate density estimation of duration (x-axis) and
+waiting time (y-axis) for Old Faithful geyser eruptions. The palettes employed
+are (counterclockwise) an HSV-based rainbow, HSV-based heat colors, HCL-based heat colors
+and grayscales.}
 \end{center}
@@
 palettes (as in the two left two panels). These palettes code the variable of interest
-by varying hue in an HSV color wheel as done by 
+by varying hue in an HSV color wheel, yielding a ``rainbow'' of colors, as done by 
 \citet[Figure~8]{color:Harezlak+Coull+Laird:2007}
@@
 \includegraphics{nigeria-pal}
-\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
+\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.
+The color palettes employed are an HSV-based rainbow and two HCL-based diverging palettes.
+In the right panels red-green contrasts are collapsed to emulate protanopic vision.}
 \end{center}
@@
 \begin{center}
-<<wheel, fig=TRUE, height=4, width=11>>=
+<<wheel, fig=TRUE, height=3.64, width=10>>=
 par(mar = rep(0.8, 4), mfrow = c(1, 2))
-par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
-par(lwd=3)
+par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), oma = c(0, 0, 0, 0))
+par(lwd = 3)
 pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
-par(lwd=0.5, new=TRUE)
-pie(rep(1, 100), radius = 1, col = rainbow(100), border = NA, labels = NA)
-par(lwd=3)
+par(lwd = 0.5, new = TRUE)
+pie(rep(1, 90), radius = 1, col = rainbow(90), border = NA, labels = NA)
+par(lwd = 3, new = TRUE)
+pie(1, radius = 0.55, labels = NA)
+par(lwd = 0.5, new = TRUE)
+pie(1, radius = 0.55, border = NA, labels = NA)
+par(lwd = 3)
 pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), 
@@
 par(lwd=0.5, new=TRUE)
-pie(rep(1, 100), radius = 1, col = rainbow_hcl(100, c = 50, l = 70), border = NA, labels = NA)
+pie(rep(1, 90), radius = 1, col = rainbow_hcl(90, c = 50, l = 70), border = NA, labels = NA)
+par(lwd = 3, new = TRUE)
+pie(1, radius = 0.55, labels = NA)
+par(lwd = 0.5, new = TRUE)
+pie(1, radius = 0.55, border = NA, labels = NA)
 @

### 2007-11-19 pkg/paper/Palettes/CSDA.Rnw
paper submitted to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@32 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%% only for statreport version:
+%% \fronttitle{Escaping RGBland: Selecting Colors for Statistical Graphics}
+%% \frontauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
+%% \frontnumber{61}
+%% \frontdate{November 2007}
+%% \frontpage
+%% \maketitle
+
 <<preliminaries>>=
@@
 
-%FIXME% The intro is a bit too long and needs to be condensed.
-
 Color is an integral element of graphical displays in general, 
@@
 
-\textbf{The colors should not be unappealing.}
+\textbf{\textit{The colors should not be unappealing.}}
 It is not necessary for the colors in a statistical plot to reflect 
@@
 
-\textbf{The colors in a statistical graphic should cooperate with each other.}
+\textbf{\textit{The colors in a statistical graphic should cooperate with each other.}}
 The typical purpose of color in a statistical graphic is to distinguish
@@
 
-\textbf{The colors should work everywhere.}
+\textbf{\textit{The colors should work everywhere.}}
 The final issue to deal with is that, in an ideal situation, colors 
@@
 
-For choosing color palettes it is helpful to have an idea how
-human color vision evolved. It has been hypothesized that it developed in three
+For choosing color palettes, it is helpful to have a basic idea how
+human color vision works. It has been hypothesized that it evolved in three
 distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
@@
 \cite{color:Ihaka:2003} and \cite{color:Lumley:2006} for more details and references.
+Furthermore, physiological studies have shown that light is captured by three
+distinct receptors (so-called cones) in the retina and hence coded in three dimensions
+by the human visual system. The perception of this coding, i.e., the subjective
+experience of this light, is less well-understood; however, current psychophysical
+theories all describe perception in terms of three dimensions \citep{color:Knoblauch:2002}.
+Therefore, colors are typically described as locations in 3-dimensional spaces.
 
-Due to these three color axes, colors are typically described as
-locations in 3-dimensional spaces. However, human perception of color
-does not correspond to the physiological axes above, but rather to polar
-coordinates in the color plane (yellow/blue vs.~green/red) plus a third
-light/dark axis. Thus, perceptually-based color spaces try to capture the
-perceptual axes
+The three dimensions used by humans to describe colors are typically:
 \begin{itemize}
-  \item \emph{hue} (dominant wavelength)
-  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
-  \item \emph{luminance} (brightness, amount of gray)
+  \item \emph{hue} (dominant wavelength),
+  \item \emph{chroma} (colorfulness, intensity of color as compared to gray),
+  \item \emph{luminance} (brightness, amount of gray).
 \end{itemize}
+Thus, they do not correspond to the axes described above,
+but rather to polar coordinates in the color plane (yellow/blue vs.~green/red) plus a third
+light/dark axis. Color models that attempt to capture these perceptual axes
+are also called perceptually-based color spaces.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[b!]
+\begin{center}
+<<wheel, fig=TRUE, height=3.6, width=10>>=
+par(mar = rep(0.8, 4), mfrow = c(1, 2))
+par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), oma = c(0, 0, 0, 0))
+par(lwd = 3)
+pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6, init.angle = -30)
+par(lwd = 0.5, new = TRUE)
+pie(rep(1, 90), radius = 1, col = rainbow(90), border = NA, labels = NA, init.angle = -2)
+par(lwd = 3, new = TRUE)
+pie(1, radius = 0.55, labels = NA)
+par(lwd = 0.5, new = TRUE)
+pie(1, radius = 0.55, border = NA, labels = NA)
+par(lwd = 3)
+pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6, init.angle = -30)
+par(lwd=0.5, new=TRUE)
+pie(rep(1, 90), radius = 1, col = rainbow_hcl(90, c = 50, l = 70), border = NA, labels = NA, init.angle = -2)
+par(lwd = 3, new = TRUE)
+pie(1, radius = 0.55, labels = NA)
+par(lwd = 0.5, new = TRUE)
+pie(1, radius = 0.55, border = NA, labels = NA)
+@
+\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
+\end{center}
+\end{figure}
 
@@
 statistics software packages, are HSV colors. They are a simple transformation of RGB colors
-and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
+defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
 %% HSV space has the shape of a single regular cone (often inflated to a regular
@@
 HSV colors are often not considered to be perceptually based.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
-\begin{center}
-<<wheel, fig=TRUE, height=3.64, width=10>>=
-par(mar = rep(0.8, 4), mfrow = c(1, 2))
-par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), oma = c(0, 0, 0, 0))
-par(lwd = 3)
-pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
-par(lwd = 0.5, new = TRUE)
-pie(rep(1, 90), radius = 1, col = rainbow(90), border = NA, labels = NA)
-par(lwd = 3, new = TRUE)
-pie(1, radius = 0.55, labels = NA)
-par(lwd = 0.5, new = TRUE)
-pie(1, radius = 0.55, border = NA, labels = NA)
-par(lwd = 3)
-pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), 
-    labels = 360 * 0:5/6)
-par(lwd=0.5, new=TRUE)
-pie(rep(1, 90), radius = 1, col = rainbow_hcl(90, c = 50, l = 70), border = NA, labels = NA)
-par(lwd = 3, new = TRUE)
-pie(1, radius = 0.55, labels = NA)
-par(lwd = 0.5, new = TRUE)
-pie(1, radius = 0.55, border = NA, labels = NA)
-@
-\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
-\end{center}
-\end{figure}
 
@@
 $U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
-to sRGB (with the latter conversion depending on the device used for display).
-The details of these conversions are somewhat technical and tedious (and hence
+to sRGB. The details of these conversions are somewhat technical and tedious (and hence
 omitted here), however the conversion formulas are still straightfoward to
@@
 %% 
-%% We are thankful to David Meyer, Ross Ihaka, Brian~D.~Ripley,
+%% We are thankful to David Meyer, Ross Ihaka, Ken Knoblauch, Brian~D.~Ripley,
 %% Thomas Kneib, Michael H\"ohle, and Thomas Lumley for feedback, suggestions and discussions.
@@
 %% Referees:
-%% Ross Ihaka, Thomas Lumley, Cynthia Brewer, Martin Maechler
+%% Ross Ihaka, Thomas Lumley, Cynthia Brewer, Martin Maechler, Ken Knoblauch
 

### 2008-10-04 pkg/paper/Palettes/CSDA.Rnw
incorporated minor comments from ref 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@35 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
-set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
-$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
+set with $H = 60, 120, 180, 240$. Cold colors (from the blue/green part of the spectrum: 
+$H = 270, 230, 190, 150$) and warm colors (from the yellow/red part of the spectrum:
 $H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

### 2008-10-05 pkg/paper/Palettes/AuthorsResponse.tex
- added new section "Illustrations" - started AuthorsResponse.tex - abstract in third person
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@37 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{a4wide}

\begin{document}

\textbf{Reviewer 1}

\begin{itemize}
  \item \textit{It would be helpful to point out that the CIE Luv colour
        space is only roughly perceptually uniform -- it is a compromise
	between perceptual uniformity and simplicity of conversion formulas.
	According to the authors' reference Poynton (2000) it has a non-uniformity
	of 6:1 compared to 80:1 for the XYZ colour space -- I don't know what the
	non-uniformity is for HSV, but gamma correction presumably means that it
	is smaller than for XYZ.}

        Paul?

  \item \textit{The article really should have better references to the scientific
        literature on colour. Lumley (2006) and Ihaka (2003) are useful, especially
        the latter, but neither is really peer-reviewed literature.  I don't know
        what the journal's policy is on Wikipedia as a reference [although this is
        the sort of topic where Wikipedia typically does quite well]}

	We added some references (Paul, Kurt?) but also kept the Wikipedia
	references because they are quite helpful as overviews.
\end{itemize}

\textbf{Reviewer 2}

\begin{itemize}
  \item \textit{p.7 line 9fb. There is confusion here. ``Warm colors
        (from the blue/green part of the spectrum...) and cold colors
	(from the yellow/red ....)''}

	fixed

  \item \textit{p.13 Harezlak reference. In all other articles, key words in the article
        title are capitalized, except in this reference. I personally would prefer
        if only the lead word in the article title were capitalized.}

	fixed (capitalized in the underlying {\sc Bib}{\TeX} file, the rest
	will be handled by Elsevier's style files)

  \item \textit{p.14 Murrell reference. ``Chapmann" should be ``Chapman". I guess a little
        germaninzation is not too bad :-)}

	Maybe Tschappmann? :-) More seriously: fixed, and thanks for spotting this.
\end{itemize}

\textbf{3rd party review}

\begin{itemize}
  \item \textit{This is a potentially useful paper, although right now it needs more
        work.  The paper is lacking several common statistics examples: a
        scatterplot, barchart or mosaic plot, to introduce and establish the
        credentials of the color choices. Cindy Brewer's work covers maps very
        nicely. But the big problem in using these methods for statistical
        graphics is that we don't have large swatches of adjoining colour.}

	New section: Illustrations.

  \item \textit{Its clear that the HCL is the better approach, as described in Ihaka's
        work. I have a difficult time determining how this paper expands on
        Ihaka's work to make substantial new contributions.}

	Ihaka's work just covered qualitative palettes (as referenced in Section~4.1),
	the other color selection strategies are novel.

  \item \textit{The defaults in statistical graphics packages should use of these
        ideas to create better initial color choices for the user. One would
        expect that the software that these authors have provided will make it
        easier for new developers to build good color defaults into their
        plots. But its not clear that it would be easy to use the vcd
        functions, or that all new packages should depend on vcd. Wouldn't it
        be more accessible to R users if there was a separate color package,
        or part of colorspace, that would make these color scales available?}

	This is a good point! All functions have been moved from vcd to colorspace.

  \item \textit{The color defaults in ggplot2 (http://had.co.nz/ggplot2/) are based on
        some similar ideas as described in this paper. I'm not sure what the
        overlap or connection is, but ggplot2 provides an excellent example of
        great default color schemes, for all sorts of statistical graphics.}

	I think Hadley copied some code (in order not to depend on vcd), following
	my suggestion to do so. We could also cite ggplot2.

\end{itemize}

\end{document}

### 2008-10-05 pkg/paper/Palettes/CSDA.Rnw
- added new section "Illustrations" - started AuthorsResponse.tex - abstract in third person
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@37 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[shortnames]{Z}
+
+\usepackage{rotating}
 %% need no \usepackage{Sweave}
@@
   mosaic displays or heatmaps---it is important that the colors are perceptually based
-  and do not introduce optical illusions or systematic bias. Here, we discuss
+  and do not introduce optical illusions or systematic bias. It is discussed
   how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
@@
 library("dichromat")
+library("vcd")
 
@@
 
+\section{Illustrations} \label{sec:illustrations}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[b!]
+\begin{center}
+<<seats, fig=TRUE, height=4, width=11>>=
+par(mar = rep(0.8, 4), mfrow = c(1, 2))
+seats <- structure(c(226, 61, 54, 51, 222),
+  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
+parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
+parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
+names(parties_hcl) <- names(parties_hsv) <- names(seats)
+names(seats)[4] <- "Grüne" ## FIXME
+pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
+pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
+@
+\caption{\label{fig:seats} Seats in the German parliament.}
+\end{center}
+\end{figure}
+
+Although problematic for many tasks, pie charts can be useful for visualizing
+whether a set of pie segments constitutes a majority. A typical application is shown
+in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
+``Bundestag'' following the 2005 election. In this election, five parties were able to
+obtain enough votes to enter the Bundestag---however, neither the governing
+coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
+Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
+a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
+by using colors as metaphors: red for the social democrats SPD, black for the conservative
+CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
+for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
+found in the (German) media whereas the right panel uses less flashy HCL colors with the
+same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
+is used). The advantage of the latter is again that they are easier to hold in focus for
+a longer time. Furthermore, they are all balanced towards the same gray and have the same
+amount of color, resulting in a perceptually balanced palette that does not introduce undesired
+graphical distractions. While it could be argued that this pie chart is such a simple
+display that a well-balanced palette is not so important and that flashy colors are to
+be preferred, the balancing properties of the HCL-based palette are very important in more
+complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
+colors and shows the distribution of votes in the 2005 election
+stratified by province (Bundesland). The order of provinces is from 
+north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
+then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
+Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
+Die Linke performed particularly well in the eastern provinces and in Saarland.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{sidewaysfigure}[p]
+\begin{center}
+<<votes, fig=TRUE, height=6, width=14>>=
+data("Bundestag2005", package = "vcd")
+votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
+                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
+parties_hcl <- parties_hcl[colnames(votes)]
+parties_hsv <- parties_hsv[colnames(votes)]
+dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
+dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
+
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hsv),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hcl),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport(2)
+@
+\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
+\end{center}
+\end{sidewaysfigure}
+
+Color choice is usually much more important in graphical displays with
+shaded areas compared to displays with only points or lines. However, color choice
+becomes much more relevant also in scatter plots when there are many points. As an
+example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
+the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
+membership, three different colors (green, yellow, red) are employed. The HSV colors in
+the left panel are again very flashy and differ strongly with respect to luminance: the
+yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
+panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
+the amount of color and gray.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<scatter>>=
+if(generate_png) {
+  set.seed(123)
+  vc <- function(var1, var2, cov)
+    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
+  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
+               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
+               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
+  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))
+
+  png(file = "scatter.png", height = 350, width = 840)
+  par(mfrow = c(1, 2), mar = rep(0.5, 4))
+  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
+    col = hsv(c(1/3, 1/6, 0))[id])
+  box()
+  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
+    col = hcl2hex(c(120, 60, 0), 70, 65, fix = FALSE)[id])
+  box()
+  dev.off()
+}
+@
+\includegraphics{scatter}
+\caption{\label{fig:scatter} Scatter plot with three clusters.}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<haireye, fig=TRUE, height=5.5, width=12>>=
+he <- margin.table(HairEyeColor, 1:2)
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
+popViewport(2)
+@
+\caption{\label{fig:haireye} Extended mosaic display for hair and eye color data.}
+\end{center}
+\end{figure}
+
+As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
+visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
+Clearly, hair and eye color are not independent and the pattern of association is highlighted
+by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
+(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
+with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
+under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
+or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
+it is shown again that the HCL colors are better balanced (between the red and blue colors)
+and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
+to read when shaded in a light grey rather than white. More details and extensions to residual-based
+shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.
+
 \section{Software} \label{sec:software}

### 2008-10-05 pkg/paper/Palettes/CSDA.Rnw
updated R ref
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@39 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-A color model typically supported by software packages involves
- the specification
+A color model typically supported by software packages involves the specification
 of colors as Red-Green-Blue (RGB) triplets. However, this specification
-corresponds to color generation on
-a computer screen 
-\citep[see][]{color:Poynton:2000}
-rather than
-corresponding to human color perception. For humans, it is
+corresponds to color generation on a computer screen \citep[see][]{color:Poynton:2000}
+rather than corresponding to human color perception. For humans, it is
 virtually impossible to control the perceptual properties of a color in this
 color space because there is no single dimension that corresponds to, e.g.,
-the hue or the brightness of the color. As a consequence, 
-various perceptually-based
+the hue or the brightness of the color. As a consequence, various perceptually-based
 color spaces have been suggested, where each dimension of the color space can
 be matched with a perceptual property. One approach widely implemented in
-software packages involves
- Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
+software packages involves Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
 a simple transformation of RGB triplets \citep[see][]{color:Wiki+HSV:2007}.
@@
 properties and the use of HSV colors encourages the use of 
-highly saturated colors. 
-A perceptually based color model that avoids
-these problems involves
-Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
-which is
-a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
+highly saturated colors. A perceptually based color model that avoids
+these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
+which is a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
 This is the color space we advocate in this article.
@@
 %% extremely easy to implement in software and we provide such an implementation
-%% in the \proglang{R} language \citep{color:R:2007} using the powerful \proglang{R} graphics
+%% in the \proglang{R} language \citep{color:R:2008} using the powerful \proglang{R} graphics
 %% system \citep[see][]{color:Murrell:2006} and the HCL
@@
 suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
+Some more more illustrations are presented in Section~\ref{sec:illustrations}.
 Section~\ref{sec:software} offers some general remarks on the implementation in statistical
 software packages as well as some details on the implementation in the \proglang{R}
-system for statistical computing \citep{color:R:2007}.
+system for statistical computing \citep{color:R:2008}.
 Section~\ref{sec:discussion} concludes the paper with a discussion.
@@
 
-The \proglang{R} system for statistical computing \citep{color:R:2007} already comes
+The \proglang{R} system for statistical computing \citep{color:R:2008} already comes
 with an open-source implementation of HCL (and other color spaces) in the package

### 2008-10-07 pkg/paper/Palettes/CSDA.Rnw
added a few color research refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@40 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-  png(file = "bkde.png", height = 810, width = 840)
+  png(file = "bkde.png", height = 800, width = 830)
   par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
@@
   ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
-  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
+  plot_bkde2D(dens, col = rev(heat_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
   plot_bkde2D(dens, col = rev(heat.colors(33)))
-  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
+  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
   dev.off()
@@
 if(generate_png) {
-  png(file = "bkde-pal.png", height = 160, width = 396)
-  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
+  png(file = "bkde-pal.png", height = 160, width = 400)
+  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.7, 0, 1.9))
   plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
     hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
     rev(heat.colors(99)))
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
-    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
+    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
   box()
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
-    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
+    rev(heat_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
   box()
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
     rev(rainbow(99, start = 0, end = 1/3)))
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
     rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
@@
     xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
-  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = 
     rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))

### 2008-10-12 pkg/inst/doc/hcl-colors.Rnw
moved hcl-colors vignette from vcd to colorspace
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@42 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[nojss]{jss}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source = TRUE}

\title{HCL-Based Color Palettes in \proglang{R}}
\Plaintitle{HCL-Based Color Palettes in R}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{HCL colors, qualitative palette, sequential palette, diverging palette}

\Abstract{
  The package \pkg{vcd} provides various functions providing perceptually-based
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes). We illustrate how these
  functions can be employed to generate various flavours of these palettes (with
  different emphases) and how they can be plugged into \proglang{R} graphics
  functions.
}

\Address{
  Achim Zeileis, Kurt Hornik\\
  Department of Statistics and Mathematics\\
  Wirtschaftsuniversit\"at Wien\\
  Augasse 2--6\\
  A-1090 Wien, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}, \email{Kurt.Hornik@R-project.org}\\
  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/},\\
  \phantom{URL:} \url{http://statmath.wu-wien.ac.at/~hornik/}
}

\begin{document}

%\VignetteIndexEntry{HCL-Based Color Palettes in R}
%\VignetteDepends{vcd,colorspace,MASS,KernSmooth,mvtnorm,kernlab}
%\VignetteKeywords{HCL colors, qualitative palette, sequential palette, diverging palette}
%\VignettePackage{colorspace}

<<preliminaries>>=
library("vcd")
library("mvtnorm")
library("kernlab")
library("KernSmooth")
@

\section{Introduction} \label{sec:intro}

This is a companion vignette to \cite{color:Zeileis+Hornik:2006} 
and \cite{color:Zeileis+Hornik+Murrell:2007} providing
further technical details on the construction of the palettes as well as
\proglang{R} code illustrating the use of varying palettes in practice.
The palettes as well as some graphical functions demonstrated are contained
in the package \pkg{vcd}.
As a simple convenience function we will use the function
<<pal, echo=TRUE>>=
pal <- function(col, border = "light gray", ...)
{
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
    axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}
@
which displays a set of colors using a sequence of shaded rectangles.

In the remainder of this vignette, we first outline how different types of palettes
can be constructed and generated using the tools from \pkg{vcd}. Subsequently,
we present a collection of examples illustrating the tools in practice.

\section{Color palettes} \label{sec:palettes}

\subsection{Qualitative palettes} \label{sec:qualitative}

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. To give the same
perceptual weight to each category, chroma and luminance are kept 
and only the hue is varied for obtaining different colors
(which are consequently all balanced towards the same gray).

In \pkg{vcd}, qualitative palettes are implemented in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
where \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates.
Figure~\ref{fig:pal-q-examples} depicts
examples for generating qualitative sets of colors $(H, 50, 70)$ produced by
<<pal-q, echo=TRUE, eval=FALSE>>=
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm")
@
From left to right and top to down, this shows a palette from the full spectrum ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors, a `harmonic'
set with $H = 60, 120, 180, 240$, warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$), respectively.

\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p!]
\begin{center}
<<pal-q1, fig=TRUE, height=4, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
<<pal-q>>
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
in different intervals for given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

\subsection{Sequential palettes} \label{sec:sequential}

Sequential palettes are used for coding numerical information that ranges
in a certain interval where low values are considered to be uninteresting and
high values are interesting. Suppose we need to visualize an intensity or
interestingness $i$ which (without loss of generality) is scaled to the unit
interval.

Potentially, all three dimensions of HCL space can be used for coding the intensity,
leading to colors from an interval of hues (i.e., differing types of colors),
an interval of chroma values (i.e., differing colorfulness) and an interval of
luminance (i.e., differing intensity of gray). If we allow chroma and luminance
to increase non-linearly via a function of type $i^p$, the resulting formula is:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]

Two different ``flavors'' of this formula are readily implemented in \pkg{vcd},
employing different defaults. For single hue palettes with $H_1 = H_2$ the 
function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
is provided where the first element of \code{c} and \code{l} give the starting chroma and luminance
coordinate (by default colorful and dark) and the second element the ending coordinate
(by default light gray). The \code{power} argument implements the parameter $p$ 
from the $i^p$ function (and can be a vector of length 2). Sequential palettes using an
interval of hues are provided by
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
named after the HSV-based \proglang{R} function \code{heat.colors()} and
by default starts from a red and going to a yellow hue. The defaults in \code{heat_hcl()}
are set differently compared to \code{sequential_hcl()} as to make the default HCL heat
colors more similar to the HSV version. The defaults of \code{sequential_hcl()}, on the
other hand, are set as to achieve a large contrast on the luminance axis. In addition
\code{terrain_hcl()} is a wrapper for \code{heat_hcl()} producing colors similar to
the HSV-based \code{terrain.colors()} from base \proglang{R}.

Various palettes produced from these functions are shown in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts.
<<pal-s, echo=TRUE, eval=FALSE>>=
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80),
  power = 1)))
@

\setkeys{Gin}{width=.75\textwidth}
\begin{figure}[p!]
\begin{center}
<<pal-s1, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
<<pal-s>>
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

\subsection{Diverging palettes} \label{sec:diverging}

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Analogously to the previous
section, we suppose that we want to visualize an intensity or interestingness $i$
from the interval $[-1, 1]$ (without loss of generality).
Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.

Diverging palettes are implemented in the function
\begin{Soutput}
diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.

Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
hue, chroma and luminance. The first palette uses a broader range on the luminance axis
whereas the others mostly rely on chroma contrasts.
<<pal-d, echo=TRUE, eval=FALSE>>=
pal(diverge_hcl(7))
pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<pal-d1, fig=TRUE, height=4, width=6>>=
par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
<<pal-d>>
@
\caption{\label{fig:pal-d-examples} Examples for diverging palettes with different
pairs of hues and decreasing luminance contrasts.}
\end{center}
\end{figure}

\section{Illustrations} \label{sec:illustrations}

\subsection{Qualitative palettes: Seats and votes in the German Bundestag}

In this section, we show a collection of examples for the various types of palettes applied
to statistical graphics. The first example illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''.
In this election, five parties were able to obtain enough votes to enter the
Bundestag, the numbers of seats are given by

<<seats-data, echo=TRUE>>=
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
seats
@

We choose colors that are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$:

<<seats-colors, echo=TRUE>>=
parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
names(parties) <- names(seats)
@

The distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
showing clearly that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. 
<<seats, echo=TRUE, eval=FALSE>>=
pie(seats, clockwise = TRUE, col = parties, radius = 1)
@

\setkeys{Gin}{width=.6\textwidth}
\begin{figure}[p]
\begin{center}
<<seats1, fig=TRUE, height=5, width=6>>=
par(mar = rep(0.8, 4))
<<seats>>
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

To take a closer look at the regional distribution, we load the \code{Bundestag2005}
data set containing a contingency table with the number of votes for each party
stratified by province (Bundesland). Then, we re-order the provinces from 
north to south, first the 10 western provinces (the former Federal Republic of Germany, FRG),
then the 6 eastern provinces (the former German Democratic Republic, GDR). 
<<votes-data, echo=TRUE>>=
data("Bundestag2005", package = "vcd")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
  c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
@

The data can then be visualized using a highlighted mosaic display via
<<votes, echo=TRUE, eval=FALSE>>=
mosaic(votes, gp = gpar(fill = parties[colnames(votes)]))
@
The annotation for this plot is clearly sub-optimal, hence we use the 
flexible \code{strucplot()} framework provided by \pkg{vcd} and display
the data via
<<votes, echo=TRUE, eval=FALSE>>=
mosaic(votes, gp = gpar(fill = parties[colnames(votes)]),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0),
  varnames = FALSE, pos_labels = "center",
  just_labels = c("center", "center", "center", "right")),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE)
@
The output is shown in Figure~\ref{fig:votes} highlighting that
the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<votes1, fig=TRUE, height=7, width=8.6>>=
<<votes>>
@
\caption{\label{fig:votes} Votes in the German election 2005.}
\end{center}
\end{figure}

\subsection{Sequential palettes: Old Faithful geyser eruptions}

To illustrate sequential palettes, a bivariate density estimation for the Old
Faithful geyser eruptions data \code{geyser} from \pkg{MASS} is visualized.
The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
and it is of some interest to understand the relation ship between the duration
of a geyser eruption and the waiting time for this eruption. To look at the data,
we use a bivariate kernel density estimate provided
by the function \code{bkde2D()} from package \pkg{KernSmooth}.
<<bkde-fit, echo=TRUE>>=
library("KernSmooth")
data("geyser", package = "MASS")
dens <- bkde2D(geyser[,2:1], bandwidth = c(0.2, 3), gridsize = c(201, 201))
@

Subsequently, we look at the estimated kernel density by means of a heatmap
(produced via \code{image()}) using two different sequential palettes: first
with only gray colors
<<bkde1, echo=TRUE, eval=FALSE>>=
image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
  col = rev(heat_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
@
and then using heat colors balanced towards the sam gray levels as above
<<bkde2, echo=TRUE, eval=FALSE>>=
image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
  col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
@
Figure~\ref{fig:bkde} shows the resulting heatmaps revealing a multi-modal
bivariate distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde3, height=6, width=12>>=
png(file = "bkde.png", height = 420, width = 840)
par(mfrow = c(1, 2))
<<bkde1>>
box()
<<bkde2>>
box()
dev.off()
@
\includegraphics{bkde}
\caption{\label{fig:bkde} Bivariate density estimation for duration and waiting time for an eruption.}
\end{center}
\end{figure}

Another interesting question in this data set would be ask how long the waiting
time for the next eruption is following a short or long eruption respectively.
This can be visualized using another bivarate density estimate for the transformed
data set matching the previous duration with the following waiting time:
<<bkde-fit2, echo=TRUE>>=
library("KernSmooth")
geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
dens2 <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
@

Again, we look at this density using two heatmaps generated via
<<bkde4, echo=TRUE, eval=FALSE>>=
image(dens2$x1, dens2$x2, dens2$fhat, xlab = "duration", ylab = "waiting time", 
  col = rev(heat_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
@
and
<<bkde5, echo=TRUE, eval=FALSE>>=
image(dens2$x1, dens2$x2, dens2$fhat, xlab = "duration", ylab = "waiting time", 
  col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
@
Figure~\ref{fig:bkde2} shows the result that illustrates that long and short
waiting times follow long and short eruption durations, resepectively.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde6, height=6, width=12>>=
png(file = "bkde2.png", height = 420, width = 840)
par(mfrow = c(1, 2))
<<bkde4>>
box()
<<bkde5>>
box()
dev.off()
@
\includegraphics{bkde2}
\caption{\label{fig:bkde2} Bivariate density estimation for previous duration and following waiting time for
an eruption.}
\end{center}
\end{figure}

\subsection{Diverging palettes: Arthritis and SVM classification}

Diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. First, we look at the outcome for the female patients from a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. 
<<arthritis-data, echo=TRUE>>=
art <- xtabs(~ Treatment + Improved, data = Arthritis,
  subset = Sex == "Female")
@
For visualizing the data, we use a mosaic display with maximum shading
\citep[as derived by][]{color:Zeileis+Meyer+Hornik:2007} via
<<arthritis, echo=TRUE, eval=FALSE>>=
set.seed(1071)
mosaic(art, gp = shading_max, gp_args = list(n = 5000))
@

The mosaic rectangles in Figure~\ref{fig:arthritis} 
signal that the treatment lead to higher improvement compared to the placebo group; this effect
is shown to be significant by the shading that codes the size of the Pearon residuals.
Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. The palette implicitly used in this plot is
\code{diverge_hcl(5, c = c(100, 0), l = c(50, 90), power = 1)}, it can be modified using
the arguments of \code{shading_max()} which has an interface similar to \code{diverge_hcl()}.

\setkeys{Gin}{width=.7\textwidth}
\begin{figure}[p]
\begin{center}
<<arthritis1, fig=TRUE, height=6, width=7>>=
<<arthritis>>
@
\caption{\label{fig:arthritis} Extended mosaic display for arthritis data.}
\end{center}
\end{figure}

To illustrate the use of diverging palettes in 2-class classification, we generate
some artificial date from a mixture of two bivariate normal distributions with
different means and covariance matrices. The data are generated using \pkg{mvtnorm}
and collected in a data frame \code{ex1}:
<<svm-data, echo=TRUE>>=
library("mvtnorm")
set.seed(123)
x1 <- rmvnorm(75, mean = c(1.5, 1.5),
  sigma = matrix(c(1, 0.8, 0.8, 1), ncol = 2))
x2 <- rmvnorm(75, mean = c(-1, -1),
  sigma = matrix(c(1, -0.3, -0.3, 1), ncol = 2))
X <- rbind(x1, x2)
ex1 <- data.frame(class = factor(c(rep("a", 75),
  rep("b", 75))), x1 = X[,1], x2 = X[,2])
@
We fit a support vector machine (SVM) with a radial basis function kernel to this data
set, using the function \code{ksvm()} from \pkg{kernlab}
<<svm-fit, echo=TRUE>>=
library("kernlab")
fm <- ksvm(class ~ ., data = ex1, C = 0.5)
@
which can subsequently be easily visualized via
<<svm, echo=TRUE, eval=FALSE>>=
plot(fm, data = ex1)
@
The resulting plot in Figure~\ref{fig:class} shows a heatmap with the fit of 
the SVM. The circles and triangles show the original observations, solid symbols correspond to
the support vectors found. The shading underlying the plot visualizes the fitted decision
values: values around 0 are on the decision boundary and are shaded in light gray, while
regions that are firmly classified to one or the other class are shaded in full blue and
red respectively. The palette used by the \code{plot()} method for \code{ksvm} objects
cannot be easily modified---however, the colors employed are equivalent to
\code{diverge_hcl(n, c = c(100, 0), l = c(50, 90), power = 1.3)}.

\setkeys{Gin}{width=.7\textwidth}
\begin{figure}[p]
\begin{center}
<<svm1, height=6, width=8>>=
png(file = "class.png", height = 480, width = 640)
<<svm>>
dev.off()
@
\includegraphics{class}
\caption{\label{fig:class} SVM classification plot.}
\end{center}
\end{figure}

\bibliography{color}

\end{document}

### 2008-10-12 pkg/paper/Palettes/CSDA.Rnw
moved hcl-colors vignette from vcd to colorspace
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@42 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
-The latter works a bit better than the former, however both palettes exhibit several drawbacks.
+The latter works somewhat better than the former, however both palettes exhibit several drawbacks.
 The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
@@
 it is shown again that the HCL colors are better balanced (between the red and blue colors)
-and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
+and less flashy. Even the cells associated with small residuals (below 2) are somewhat easier
 to read when shaded in a light grey rather than white. More details and extensions to residual-based
@@
 of HCL colors: from the formulas provided above the HCL coordinates for a
-palette can be conveniently computed. A bit more work is required if the software
+palette can be conveniently computed. Somewhat more work is required if the software
 package does not yet provide an HCL implementation. In that case, additional
@@
 
-The \proglang{R} system for statistical computing \citep{color:R:2008} already comes
-with an open-source implementation of HCL (and other color spaces) in the package
-\pkg{colorspace} \citep{color:Ihaka:2006}. The coordinate transformations mentioned
-above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
-port to other statistical software systems. Based on \pkg{colorspace} and
-the \proglang{R} graphics system \citep[see][]{color:Murrell:2006},
-the package \pkg{vcd} \citep{color:Meyer+Zeileis+Hornik:2007} provides implementations
-of all palettes discussed above. Qualitative palettes are provided by
+The \proglang{R} system for statistical computing \citep{color:R:2008} provides
+an open-source implementation of HCL (and other color spaces) in the package
+\pkg{colorspace}, originally written by Ross Ihaka. The coordinate transformations mentioned
+above are contained in \proglang{C}~code within \pkg{colorspace} that are easy to
+port to other statistical software systems. Starting from version~0.96, \pkg{colorspace}
+\citep{color:Ihaka+Murrell+Hornik+Zeileis:2008} also includes
+an implementation of all palettes discussed above. 
+\citep[Originally, the code for the palettes was in the \pkg{vcd} package,]%
+[but it was recently moved to \pkg{colorspace} to be more easily acessible.]%
+{color:Meyer+Zeileis+Hornik:2008} Qualitative palettes are provided by
 \code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
@@
 a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
-Technical documentation along with a large collection of examples is available via
-\verb/help("rainbow_hcl", package = "vcd")/. Furthermore, \proglang{R} code for reproducing
+Technical documentation along with a large collection of example palettes is available via
+\verb/help("rainbow_hcl", package = "colorspace")/. Furthermore, \proglang{R}~code for reproducing
 Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
-(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.
+(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "colorspace")}.
 
@@
 it is much more important to have a big difference in luminance.
-%% Another parameter in these 
-%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
-%% Employing a function $|i|^p$ for this, small
-%% powers $p$ can be used if palettes with a lot of color should be 
-%% constructed (e.g., to clearly separate regions in a classification map). On
-%% the other hand, large powers $p$ will result in palettes with less color
-%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).
 
@@
 
-%% Referees:
-%% Ross Ihaka, Thomas Lumley, Cynthia Brewer, Martin Maechler, Ken Knoblauch
-
 \bibliography{color}

### 2008-10-20 pkg/paper/Palettes/CSDA.Rnw
added refs to CSDA paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@43 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   mosaic displays or heatmaps---it is important that the colors are perceptually based
-  and do not introduce optical illusions or systematic bias. It is discussed
+  and do not introduce optical illusions or systematic bias. This article
+discusses
   how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
@@
 properties and the use of HSV colors encourages the use of 
-highly saturated colors. A perceptually based color model that avoids
+highly saturated colors. A perceptually based color model that mitigates
 these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
@@
 3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
-\cite{color:Ihaka:2003} and \cite{color:Lumley:2006} for more details and references.
+\cite{color:Kaiser+Boynton:1996}, 
+\cite{color:Knoblauch:2002}, \cite{color:Ihaka:2003}, 
+and \cite{color:Lumley:2006} for more details and references.
 Furthermore, physiological studies have shown that light is captured by three
@@
 To overcome these drawbacks, various color spaces have been suggested
-that properly map to the perception dimensions, the most prominent of which are
-the CIELUV and CIELAB spaces developed by the \cite{color:CIE:2004}. \cite{color:Ihaka:2003}
+that more closely map to the perception dimensions, the most prominent of which are
+the CIELUV and CIELAB spaces developed by the \cite{color:CIE:2004}. 
+These color spaces are still not perfect---such an ideal is never
+truly attainable because color perception will always
+differ between individuals---but they offer a significant
+improvement over HSV, as Figure~\ref{fig:wheel} illustrates.
+
+\cite{color:Ihaka:2003}
 argues that CIELUV colors are typically preferred for use with emissive technologies
@@
 \section{Illustrations} \label{sec:illustrations}
+
+This section provides several more demonstrations of how 
+HCL-based color palettes can be used to produce statistical
+plots that are less ``flashy'' and more perceptually balanced,
+so that the viewer can
+extract information from the plots more easily and more accurately.
 
@@
 
+The default color palettes in the \pkg{ggplot2} package 
+\citep{color:Wickham:2008} are also based on similar ideas to 
+those discussed in this article.
+
 \section{Discussion} \label{sec:discussion}

### 2008-10-21 pkg/inst/doc/hcl-colors.Rnw
fixed compiler warnings in colorspace.c;  tidied up hcl-colors.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@44 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
-   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
-\Plainauthor{Achim Zeileis, Kurt Hornik}
+   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
+   \And Paul Murrell\\University of Auckland}
+\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
 \Keywords{HCL colors, qualitative palette, sequential palette, diverging palette}
@@
 \Abstract{
-  The package \pkg{vcd} provides various functions providing perceptually-based
+  The package \pkg{colorspace} provides various functions providing perceptually-based
   color palettes for coding categorical data (qualitative palettes) and numerical
@@
   \phantom{URL:} \url{http://statmath.wu-wien.ac.at/~hornik/}
+
+  Paul Murrell\\
+  Department of Statistics\\
+  The University of Auckland\\
+  Private Bag 92019\\
+  Auckland\\
+  New Zealand\\
+  E-mail: \email{paul@stat.auckland.ac.nz}\\
+  URL: \url{http://www.stat.auckland.ac.nz/~paul/}
 }
@@
 %\VignetteIndexEntry{HCL-Based Color Palettes in R}
-%\VignetteDepends{vcd,colorspace,MASS,KernSmooth,mvtnorm,kernlab}
+%\VignetteDepends{colorspace,vcd,MASS,KernSmooth,mvtnorm,kernlab}
 %\VignetteKeywords{HCL colors, qualitative palette, sequential palette, diverging palette}
@@
 <<preliminaries>>=
+library("colorspace")
 library("vcd")
@@
 The palettes as well as some graphical functions demonstrated are contained
-in the package \pkg{vcd}.
+in the package \pkg{colorspace}.
 As a simple convenience function we will use the function
@@
 In the remainder of this vignette, we first outline how different types of palettes
-can be constructed and generated using the tools from \pkg{vcd}. Subsequently,
+can be constructed and generated using the tools from \pkg{colorspace}. Subsequently,
 we present a collection of examples illustrating the tools in practice.
@@
 i.e., for coding a categorical variable. To give the same
-perceptual weight to each category, chroma and luminance are kept 
+perceptual weight to each category, chroma and luminance are kept constant
 and only the hue is varied for obtaining different colors
@@
 
-In \pkg{vcd}, qualitative palettes are implemented in the function
+In \pkg{colorspace}, qualitative palettes are implemented in the function
 \begin{Soutput}
@@
 creating a `dynamic' set of colors, a `harmonic'
-set with $H = 60, 120, 180, 240$, warm colors (from the blue/green part of the spectrum: 
-$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
+set with $H = 60, 120, 180, 240$, cold colors (from the blue/green part of the spectrum: 
+$H = 270, 230, 190, 150$) and warm colors (from the yellow/red part of the spectrum:
 $H = 90, 50, 10, 330$), respectively.
@@
 
-Two different ``flavors'' of this formula are readily implemented in \pkg{vcd},
+Two different ``flavors'' of this formula are readily implemented in \pkg{colorspace},
 employing different defaults. For single hue palettes with $H_1 = H_2$ the 
@@
 coordinate (by default colorful and dark) and the second element the ending coordinate
-(by default light gray). The \code{power} argument implements the parameter $p$ 
+(by default gray and light). The \code{power} argument implements the parameter $p$ 
 from the $i^p$ function (and can be a vector of length 2). Sequential palettes using an

### 2008-10-21 pkg/src/colorspace.c

@@ -1,3 +1,5 @@
+
+#include <ctype.h>
 #include <R.h>
 #include <Rinternals.h>
 #include <Rmath.h>
@@ -38,22 +40,20 @@ static double ftrans(double u, double gamma)
 	return u / 12.92;
 }
 
-static int DEVRGB_to_RGB(double R, double G, double B, double gamma,
+static void DEVRGB_to_RGB(double R, double G, double B, double gamma,
 			 double *r, double *g, double *b)
 {
     *r = ftrans(R, gamma);
     *g = ftrans(G, gamma);
     *b = ftrans(B, gamma);
-    return 1;
 }
 
-static int RGB_to_DEVRGB(double R, double G, double B, double gamma,
+static void RGB_to_DEVRGB(double R, double G, double B, double gamma,
 			 double *r, double *g, double *b)
 {
     *r = gtrans(R, gamma);
     *g = gtrans(G, gamma);
     *b = gtrans(B, gamma);
-    return 1;
 }
 
 /* ----- CIE-XYZ <-> Device independent RGB -----
@@ -64,35 +64,35 @@ static int RGB_to_DEVRGB(double R, double G, double B, double gamma,
  *
  */
 
-static int RGB_to_XYZ(double R, double G, double B,
+static void RGB_to_XYZ(double R, double G, double B,
 		      double XN, double YN, double ZN,
 		      double *X, double *Y, double *Z)
 {
     *X = YN * (0.412453 * R + 0.357580 * G + 0.180423 * B);
     *Y = YN * (0.212671 * R + 0.715160 * G + 0.072169 * B);
     *Z = YN * (0.019334 * R + 0.119193 * G + 0.950227 * B);
-    return 1;
 }
 
-static int XYZ_to_RGB(double X, double Y, double Z,
+static void XYZ_to_RGB(double X, double Y, double Z,
 		      double XN, double YN, double ZN,
 		      double *R, double *G, double *B)
 {
     *R = ( 3.240479 * X - 1.537150 * Y - 0.498535 * Z) / YN;
     *G = (-0.969256 * X + 1.875992 * Y + 0.041556 * Z) / YN;
     *B = ( 0.055648 * X - 0.204043 * Y + 1.057311 * Z) / YN;
-    return 1;
 }
 
 
 /* ----- CIE-XYZ <-> CIE-LAB ----- */
 
+/* UNUSED ?
 static double g(double t)
 {
[…]

### 2008-10-22 pkg/paper/Palettes/CSDA.Rnw
small enhancements to CSDA paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@46 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 truly attainable because color perception will always
-differ between individuals---but they offer a significant
+differ between individuals---but they facilitate color matching for humans
+\citep{color:Zhang+Montag:2006} and offer a significant
 improvement over HSV, as Figure~\ref{fig:wheel} illustrates.
@@
 statistics would be bar plots \citep[see][]{color:Ihaka:2003}, pie charts
-or highlighted mosaic displays.
+or highlighted mosaic displays (see Section~\ref{sec:illustrations}).
 
@@
 
-The simplest solution to this task is to employ light/dark contrasts, i.e., rely
+The simplest solution to this task is to employ light/dark contrasts
+\citep{color:Ware:1988,color:Levkowitz+Herman:1992}, i.e., rely
 on the most basic perceptual axis. The interestingness can be coded
@@
 \end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{sidewaysfigure}[p]
+\begin{center}
+<<votes, fig=TRUE, height=6, width=14>>=
+data("Bundestag2005", package = "vcd")
+votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
+                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
+parties_hcl <- parties_hcl[colnames(votes)]
+parties_hsv <- parties_hsv[colnames(votes)]
+dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
+dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
+
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hsv),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hcl),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport(2)
+@
+\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
+\end{center}
+\end{sidewaysfigure}
 
@@
 
-\setkeys{Gin}{width=\textwidth}
-\begin{sidewaysfigure}[p]
-\begin{center}
-<<votes, fig=TRUE, height=6, width=14>>=
-data("Bundestag2005", package = "vcd")
-votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
-                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
-parties_hcl <- parties_hcl[colnames(votes)]
-parties_hsv <- parties_hsv[colnames(votes)]
-dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
-dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
-
-grid.newpage()
-pushViewport(viewport(layout = grid.layout(1, 2)))
-pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(votes, gp = gpar(fill = parties_hsv),
-  spacing = spacing_highlighting, labeling = labeling_left,
-  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
-  keep_aspect_ratio = FALSE, newpage = FALSE)
-popViewport()
-pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-mosaic(votes, gp = gpar(fill = parties_hcl),
-  spacing = spacing_highlighting, labeling = labeling_left,
-  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
-  keep_aspect_ratio = FALSE, newpage = FALSE)
-popViewport(2)
-@
-\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
-\end{center}
-\end{sidewaysfigure}
-
 Color choice is usually much more important in graphical displays with

### 2008-10-23 pkg/paper/Palettes/CSDA.Rnw
KH feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@47 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
-which is a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
+resulting from a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
 This is the color space we advocate in this article.
@@
 suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
-Some more more illustrations are presented in Section~\ref{sec:illustrations}.
+Some more illustrations are presented in Section~\ref{sec:illustrations}.
 Section~\ref{sec:software} offers some general remarks on the implementation in statistical
@@
 
-%% \section*{Acknowledgments}
-%% 
-%% We are thankful to David Meyer, Ross Ihaka, Ken Knoblauch, Brian~D.~Ripley,
-%% Thomas Kneib, Michael H\"ohle, and Thomas Lumley for feedback, suggestions and discussions.
+\section*{Acknowledgments}
+
+We are thankful to Michael H\"ohle, Ross Ihaka, Thomas Kneib, Ken Knoblauch,
+Thomas Lumley, David Meyer, and Brian~D.~Ripley for feedback, suggestions, and discussions.
 

### 2008-10-24 pkg/paper/Palettes/AuthorsResponse.tex
Updates for re-submission of paper, package ready for new release on CRAN.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@48 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
-\documentclass[a4paper]{article}
-\usepackage{a4wide}
+\documentclass{wuletter}
+\signature{Achim Zeileis, Kurt Hornik, Paul Murrell}
+\usepackage[american]{babel}
+\renewcommand{\rmdefault}{phv}
 
 \begin{document}
+\begin{letter}{
+Erricos~J.~Kontoghiorghes\\
+School of Computer Science and Information Systems\\
+Birkbeck College\\
+University of London\\
+Malet Street, London WC1E 7HX, United Kingdom}
+
+\concerning{Revision of CSDA-D-07-00860}
+
+\opening{Dear Prof.~Kontoghiorghes,}
+
+please find attached the revision of our manuscript
+\begin{quote}
+\textit{Escaping RGBland: Selecting Colors for Statistical Graphics}
+\end{quote}
+for \emph{Computational Statistics \& Data Analysis}.
+
+We would like to thank you, the associate editor and the referees
+for the most helpful reviews of our work. We tried to incorporate
+all of the suggestions. In particular, we have complemented the
+examples from Section~2 (heatmaps, maps) by a set of further illustrations
+in a new Section~5 (including pie charts, highlighted mosaic displays,
+scatter plots, and mosaic displays with residual-based shading).
+Furthermore, we have included some more references to the color and
+computer graphics literature. A detailed point-to-point reply is
+included below.
+
+\closing{Best regards,}
+\end{letter}
+
+\newpage
 
@@
 	non-uniformity is for HSV, but gamma correction presumably means that it
-	is smaller than for XYZ.}
+	is smaller than for XYZ.}\\[-0.3cm]
 
-        Paul?
+        Yes, uniformity is a bit more complicated. Hence, we refrain from 
+	discussing its details explicitly here, and have included some
+	remarks in Section~3 pointing out that the CIELUV model also has
+	limitations.
 
@@
         what the journal's policy is on Wikipedia as a reference [although this is
-        the sort of topic where Wikipedia typically does quite well]}
+        the sort of topic where Wikipedia typically does quite well]}\\[-0.3cm]
 
-	We added some references (Paul, Kurt?) but also kept the Wikipedia
-	references because they are quite helpful as overviews.
+	We have included pointers to Kaiser and Boynton (1996) and Knoblauch (2002)
+	at the beginning of Section~3, and also included some further references
+	to the computer graphics literature in other sections. We have kept the
+	Wikipedia references because we agree with the referee that they are
+	helpful (especially as a first overview).
 \end{itemize}
@@
         (from the blue/green part of the spectrum...) and cold colors
-	(from the yellow/red ....)''}
+	(from the yellow/red ....)''}\\[-0.3cm]
 
@@
         title are capitalized, except in this reference. I personally would prefer
-        if only the lead word in the article title were capitalized.}
+        if only the lead word in the article title were capitalized.}\\[-0.3cm]
 
-	fixed (capitalized in the underlying {\sc Bib}{\TeX} file, the rest
-	will be handled by Elsevier's style files)
+	fixed
 
   \item \textit{p.14 Murrell reference. ``Chapmann" should be ``Chapman". I guess a little
-        germaninzation is not too bad :-)}
+        germaninzation is not too bad :-)}\\[-0.3cm]
 
@@
         nicely. But the big problem in using these methods for statistical
-        graphics is that we don't have large swatches of adjoining colour.}
+        graphics is that we don't have large swatches of adjoining colour.}\\[-0.3cm]
 
-	New section: Illustrations.
+	We have added a new Section~5 ``Illustrations'' that complements
+	the displays from Section~2 (heatmaps, maps) by presenting further
+	applications of the suggested color palettes (pie charts,
+	highlighted mosaic displays, scatter plots, and mosaic displays
+	with residual-based shading).
 
@@
         work. I have a difficult time determining how this paper expands on
-        Ihaka's work to make substantial new contributions.}
+        Ihaka's work to make substantial new contributions.}\\[-0.3cm]
 
@@
         be more accessible to R users if there was a separate color package,
-        or part of colorspace, that would make these color scales available?}
+        or part of colorspace, that would make these color scales available?}\\[-0.3cm]
 
-	This is a good point! All functions have been moved from vcd to colorspace.
+	This is a good point! All functions have been moved from vcd to colorspace,
+	as of vcd~1.2-0 and colorspace~0.97.
 
@@
         overlap or connection is, but ggplot2 provides an excellent example of
-        great default color schemes, for all sorts of statistical graphics.}
+        great default color schemes, for all sorts of statistical graphics.}\\[-0.3cm]
 
-	I think Hadley copied some code (in order not to depend on vcd), following
-	my suggestion to do so. We could also cite ggplot2.
+	ggplot2 also employs HCL color palettes by default, based on similar ideas.
+        ggplot2 has its own code for this---my understanding is that this is 
+	in part to avoid depending on vcd. We have included a pointer to ggplot2
+	in Section~6.
 

### 2008-10-24 pkg/paper/Palettes/CSDA.Rnw
Updates for re-submission of paper, package ready for new release on CRAN.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@48 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\usepackage{amsfonts}
 \usepackage{rotating}
@@
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
-
-\usepackage{amsfonts}
 
@@
   mosaic displays or heatmaps---it is important that the colors are perceptually based
-  and do not introduce optical illusions or systematic bias. This article
-discusses
+  and do not introduce optical illusions or systematic bias. This article discusses
   how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
@@
 \begin{document}
-
-%% only for statreport version:
-%% \fronttitle{Escaping RGBland: Selecting Colors for Statistical Graphics}
-%% \frontauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
-%% \frontnumber{61}
-%% \frontdate{November 2007}
-%% \frontpage
-%% \maketitle
 
@@
 Color is an integral element of graphical displays in general, 
-and many statistical
-graphics in particular. 
+and many statistical graphics in particular. 
 It is easy to create color graphics with any statistical software 
@@
 properties and the use of HSV colors encourages the use of 
-highly saturated colors. A perceptually based color model that mitigates
+highly saturated colors. A perceptually-based color model that mitigates
 these problems involves Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
@@
 of the colors, e.g., the hue or the brightness, keeping other properties
-fixed. In a perceptually based color space, this corresponds to selecting
+fixed. In a perceptually-based color space, this corresponds to selecting
 colors by traversing paths through this space along its dimensions.
@@
 Section~\ref{sec:software} offers some general remarks on the implementation in statistical
-software packages as well as some details on the implementation in the \proglang{R}
+software as well as some details on the implementation in the \proglang{R}
 system for statistical computing \citep{color:R:2008}.
@@
   par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
-  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
-  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
-  plot_bkde2D(dens, col = rev(heat_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
+  plot_bkde2D(dens, col = hsv(h = seq(from = 2/3, to = 0, length = 33), v = 0.9, s = 0.9))
+  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
   plot_bkde2D(dens, col = rev(heat.colors(33)))
-  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
+  plot_bkde2D(dens, col = rev(heat_hcl(33, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
   dev.off()
@@
   abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
-    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
-    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
+    hsv(h = seq(from = 2/3, to = 0, length = 99), v = 0.9, s = 0.9))
   box()
@@
   abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
-    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
+    rev(heat_hcl(99, h = c(0, 90), c = c(80, 30), l = c(30, 90), power = c(1/5, 1.3))))
   box()
@@
   abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 6, col = 
-    rev(heat_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.3))))
+    rev(heat_hcl(99, h = c(0, 90), c = 0,         l = c(30, 90), power = c(1/5, 1.3))))
   box()
@@
 increase from a yellow to a red hue while using the same brightness levels as in the grayscale
-palette. Thus, when converted to a grayscale or printed out on a grayscale printer
+palette \citep[see also][for similar approaches]{color:Ware:1988,color:Levkowitz+Herman:1992}. 
+Thus, when converted to a grayscale or printed out on a grayscale printer
 (as in the printed version of \emph{Computational Statistics \& Data Analysis}),
@@
 revealing decreased mortality in the south-west and increased mortality in
-the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the residuals by
+the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the deviations by
 the hue, going from green via yellow to red (see the upper left panel of Figure~\ref{fig:nigeria}).
@@
 Our first HCL-based palette in the middle left panel also employs green
-and red hues for negative and postive residuals respectively, but codes neutral values
+and red hues for negative and postive deviations respectively, but codes neutral values
 (around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
@@
 flashy colors are avoided making it easier to look at the display for a longer time; positive
-and negative residuals with the same absolute size receive the same perceptual weight
+and negative deviations with the same absolute size receive the same perceptual weight
 by being balanced with respect to their brightness.
@@
 par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
-pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
-pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic [60, 240]")
-pal(rainbow_hcl(4, start = 270, end = 150), main = "cold [270, 150]")
-pal(rainbow_hcl(4, start = 90, end = -30), main = "warm [90, -30]")
+pal(rainbow_hcl(4, start =  30, end = 300, c = 50, l = 70), main = "dynamic [30, 300]")
+pal(rainbow_hcl(4, start =  60, end = 240, c = 50, l = 70), main = "harmonic [60, 240]")
+pal(rainbow_hcl(4, start = 270, end = 150, c = 50, l = 70), main = "cold [270, 150]")
+pal(rainbow_hcl(4, start =  90, end = -30, c = 50, l = 70), main = "warm [90, -30]")
 @
@@
 The simplest solution to this task is to employ light/dark contrasts
-\citep{color:Ware:1988,color:Levkowitz+Herman:1992}, i.e., rely
+\citep{color:Levkowitz+Herman:1992}, i.e., rely
 on the most basic perceptual axis. The interestingness can be coded
@@
 This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
-hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
+hue $H = 260$ with $L_{\max} = 90$
+and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
 respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
@@
 par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
-pal(sequential_hcl(12, c = 0, power = 2.2))
-pal(sequential_hcl(12, power = 2.2))
-pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
-pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
-pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
+pal(sequential_hcl(12, h = 260,        c = 0,         l = c(30, 90), power = 2.2))
+pal(sequential_hcl(12, h = 260,        c = c(80, 0),  l = c(30, 90), power = 2.2))
+pal(      heat_hcl(12, h = c(0, 90),   c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
+pal(rev(  heat_hcl(12, h = c(0, -100), c = c(40, 80), l = c(75, 40), power = 1)))
+pal(   terrain_hcl(12, h = c(130, 0),  c = c(65, 0),  l = c(45, 90), power = c(1/2, 1.5)))
 @
@@
 par(mfrow = c(4, 1), mar = c(0.25, 0, 0.25, 0))
-pal(diverge_hcl(7))
-pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
-pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
-pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
+pal(diverge_hcl(7, h = c(260, 0),   c =  80, l = c(30, 90), power = 1.5))
+pal(diverge_hcl(7, h = c(260, 0),   c = 100, l = c(50, 90), power = 1))
+pal(diverge_hcl(7, h = c(130, 43),  c = 100, l = c(70, 90), power = 1.5))
+pal(diverge_hcl(7, h = c(180, 330), c =  59, l = c(75, 95), power = 1.5))
 @
@@
 names(parties_hcl) <- names(parties_hsv) <- names(seats)
-names(seats)[4] <- "Grüne" ## FIXME
+names(seats)[4] <- "Grüne"
+Encoding(names(seats)) <- "latin1"
 pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
@@
 dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
+Encoding(dimnames(votes)[[1]]) <- "latin1"
 
@@
 above are contained in \proglang{C}~code within \pkg{colorspace} that are easy to
-port to other statistical software systems. Starting from version~0.96, \pkg{colorspace}
-\citep{color:Ihaka+Murrell+Hornik+Zeileis:2008} also includes
+port to other statistical software systems. Starting from version~0.97, %% FIXME
+\pkg{colorspace} \citep{color:Ihaka+Murrell+Hornik+Zeileis:2008} also includes
 an implementation of all palettes discussed above. 
@@
 The default color palettes in the \pkg{ggplot2} package 
-\citep{color:Wickham:2008} are also based on similar ideas to 
-those discussed in this article.
+\citep{color:Wickham:2008} are also based on HCL colors,
+using similar ideas to those discussed in this article.
 
@@
 Thomas Lumley, David Meyer, and Brian~D.~Ripley for feedback, suggestions, and discussions.
+Furthermore, we are indebted to an anonymous associate editor and three referees
+for their reviews that helped improving the manuscript.
 

### 2008-11-19 pkg/paper/Palettes/CSDA.Rnw
Escaping RGBland paper accepted for CSDA. Small touches to the paper, changed references in the package. Prepared release 1.0-0 accompanying the CSDA paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@50 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
-\documentclass[shortnames]{Z}
+\documentclass[shortnames,noheadings]{Z}
 
@@
   mosaic displays or heatmaps---it is important that the colors are perceptually based
-  and do not introduce optical illusions or systematic bias. This article discusses
-  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
-  color palettes for coding categorical data (qualitative palettes) and numerical
+  and do not introduce optical illusions or systematic bias.
+  Based on the perceptually-based Hue-Chroma-Luminance (HCL) color space  
+  suitable color palettes are derived for coding categorical data (qualitative palettes) and numerical
   variables (sequential and diverging palettes).
+}
+
+\usepackage{fancyhdr}
+\setlength{\headheight}{15pt}
+\renewcommand{\headrulewidth}{0pt}
+\pagestyle{fancy}
+\thispagestyle{plain}
+\fancyhf{}
+\fancyhead[RO,LE]{\thepage} 
+\fancyhead[CO]{\textit{\normalsize Achim Zeileis, Kurt Hornik, Paul Murrell}} 
+\fancyhead[CE]{\textit{\normalsize Escaping RGBland: Selecting Colors for Statistical Graphics}} 
+\fancyfoot[LO,LE]{\small Copyright {\copyright} 2008 Elsevier B.V.}
+\fancypagestyle{plain}{
+\fancyhf{}
+\fancyfoot[LO,LE]{\small This is a preprint of an article accepted for publication in
+\textit{Computational Statistics \& Data Analysis}.\\
+ Copyright {\copyright} 2008 Elsevier B.V.} 
 }
@@
 above are contained in \proglang{C}~code within \pkg{colorspace} that are easy to
-port to other statistical software systems. Starting from version~0.97, %% FIXME
+port to other statistical software systems. Version~1.0-0 of
 \pkg{colorspace} \citep{color:Ihaka+Murrell+Hornik+Zeileis:2008} also includes
@@
 [but it was recently moved to \pkg{colorspace} to be more easily acessible.]%
-{color:Meyer+Zeileis+Hornik:2008} Qualitative palettes are provided by
+{color:Meyer+Zeileis+Hornik:2006} Qualitative palettes are provided by
 \code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base

### 2009-02-11 pkg/paper/Palettes/CSDA.Rnw
Re-arranged graphics for CSDA (saving 5 x EUR 473!!).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@53 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[shortnames,noheadings]{Z}
-
-\usepackage{amsfonts}
-\usepackage{rotating}
+\usepackage{amsfonts,thumbpdf}
 %% need no \usepackage{Sweave}
@@
 
-  png(file = "bkde.png", height = 800, width = 830)
+  png(file = "bkde.png", height = 1560, width = 1640, res = 144)
   par(mfrow = c(2, 2), mar = c(4.5, 3.5, 1, 1))
@@
 if(generate_png) {
-  png(file = "bkde-pal.png", height = 160, width = 400)
-  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.7, 0, 1.9))
+  png(file = "bkde-pal.png", height = 320, width = 800, res = 144)
+  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.9, 0, 1.9))
   plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
@@
 waiting time (y-axis) for Old Faithful geyser eruptions. The palettes employed
-are (counterclockwise) an HSV-based rainbow, HSV-based heat colors, HCL-based heat colors
+are (counterclockwise from top left) an HSV-based rainbow, HSV-based heat colors, HCL-based heat colors
 and grayscales.}
@@
 
-  png(file = "nigeria.png", height = 1050, width = 840)
+  png(file = "nigeria.png", height = 2100, width = 1680, res = 144)
   par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
@@
 if(generate_png) {
-  png(file = "nigeria-pal.png", height = 120, width = 396)
-  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
+  png(file = "nigeria-pal.png", height = 240, width = 800, res = 144)
+  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.9, 0, 1.9))
   plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
@@
 \caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.
-The color palettes employed are an HSV-based rainbow and two HCL-based diverging palettes.
-In the right panels red-green contrasts are collapsed to emulate protanopic vision.}
+The color palettes employed are (from top to bottom) an HSV-based rainbow and two
+HCL-based diverging palettes. In the right panels red-green contrasts are collapsed to
+emulate protanopic vision.}
 \end{center}
@@
 \begin{center}
-<<wheel, fig=TRUE, height=3.6, width=10>>=
+<<wheel>>=
+png(file = "wheel.png", height = 500, width = 1400, res = 144)
 par(mar = rep(0.8, 4), mfrow = c(1, 2))
-par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), oma = c(0, 0, 0, 0))
+par(mfrow = c(1, 2), mar = c(1.1, 1, 1.1, 1), oma = c(0, 0, 0, 0))
 par(lwd = 3)
@@
 par(lwd = 0.5, new = TRUE)
-pie(rep(1, 90), radius = 1, col = rainbow(90), border = NA, labels = NA, init.angle = -2)
+pie(rep(1, 180), radius = 1, col = rainbow(180), border = NA, labels = NA, init.angle = -1)
 par(lwd = 3, new = TRUE)
@@
 par(lwd=0.5, new=TRUE)
-pie(rep(1, 90), radius = 1, col = rainbow_hcl(90, c = 50, l = 70), border = NA, labels = NA, init.angle = -2)
+pie(rep(1, 180), radius = 1, col = rainbow_hcl(180, c = 50, l = 70), border = NA, labels = NA, init.angle = -1)
 par(lwd = 3, new = TRUE)
@@
 pie(1, radius = 0.55, border = NA, labels = NA)
-@
+dev.off()
+@
+\includegraphics{wheel}
 \caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
@@
 
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[p]
+\setkeys{Gin}{width=0.55\textwidth}
+\begin{figure}[t!]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-q-construct.png", height = 480, width = 480)
+  png(file = "pal-q-construct.png", height = 960, width = 960, res = 144)
+  par(mar = c(0, 0, 0, 0))
   hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
@@
 \includegraphics{pal-q-construct}
-\vspace*{-2cm}
-\caption{\label{fig:pal-q-construct} Constructing qualitative palettes. In the hue/chroma
-plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
-$H = 0, 120, 240$.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<pal-q-examples, fig=TRUE, height=4, width=6>>=
-par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
+
+\vspace*{1cm}
+
+\setkeys{Gin}{width=0.75\textwidth}
+<<pal-q-examples, fig=TRUE, height=3, width=6>>=
+par(mfrow = c(2, 2), mar = c(1, 0, 3, 2))
 pal(rainbow_hcl(4, start =  30, end = 300, c = 50, l = 70), main = "dynamic [30, 300]")
@@
 @
-\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
-in different intervals for given $C = 50$ and $L = 70$.}
+\caption{\label{fig:pal-q} Qualitative palettes: construction and examples.
+Top: Construction in the hue/chroma
+plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
+$H = 0, 120, 240$. Bottom: Examples by varying hue in different intervals for
+given $C = 50$ and $L = 70$.}
 \end{center}
@@
 
-Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
+Figure~\ref{fig:pal-q} (top panel) depicts how the HCL-based color wheel is constructed.
 It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
@@
 A simple and intuitive one is to use colors as metaphors for categories (e.g., for political parties),
-%% (e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
 another approach would be to use segments from
 the color wheel corresponding to nearby or distant colors. The latter is shown
-in Figure~\ref{fig:pal-q-examples} which depicts
+in Figure~\ref{fig:pal-q} (bottom panel) which depicts
 examples for generating qualitative sets of colors $(H, 50, 70)$.
@@
 $H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
+
+\newpage
 
@@
 
-Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
-  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
-This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
-hue $H = 260$ with $L_{\max} = 90$
-and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
-respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
-The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
-The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
-is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
-the same way, there is a stronger trade-off between using dark colors (with low luminance)
-and colorful colors (with high chroma). Hence, the second path from light gray to 
-full green ends at a much lighter color with $L_{\min} = 75$.
-
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{figure}[t!]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-s-construct.png", height = 420, width = 840)
+  png(file = "pal-s-construct.png", height = 840, width = 1680, res = 144)
   par(mfrow = c(1, 2))
@@
 \includegraphics{pal-s-construct}
-\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
-plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
-either only luminance or both luminance and chroma.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<pal-s-examples, fig=TRUE, height=5, width=6>>=
+
+\vspace*{1cm}
+
+\setkeys{Gin}{width=.75\textwidth}
+<<pal-s-examples, fig=TRUE, height=4, width=6>>=
 par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
@@
 @
-\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
-only luminance (first panel), chroma and luminance (second panel), and hue,
-chroma and luminance (remaining panels).}
+
+\caption{\label{fig:pal-s} Sequential palettes: construction and examples.
+Top: Construction is shown in the chroma/luminance
+plane for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
+either only luminance or both luminance and chroma. Bottom: Examples by varying
+only luminance (first row), chroma and luminance (second row), and hue,
+chroma and luminance (remaining rows).}
 \end{center}
 \end{figure}
+
+Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
+  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
+This strategy is depicted in the top left panel of Figure~\ref{fig:pal-s} for a blue
+hue $H = 260$ with $L_{\max} = 90$
+and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
+respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
+The first two combinations are also shown in the first two rows of the bottom panel in Figure~\ref{fig:pal-s}.
+The top right panel of Figure~\ref{fig:pal-s} shows that the exact same strategy
+is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
+the same way, there is a stronger trade-off between using dark colors (with low luminance)
+and colorful colors (with high chroma). Hence, the second path from light gray to 
+full green ends at a much lighter color with $L_{\min} = 75$.
 
@@
 $i'$ for chroma and a different transformation $i''$ for the luminance.
-Such a strategy is adopted for the palettes shown in the lower three rows in
-Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
+Such a strategy is adopted for the palettes shown in the last three rows in
+Figure~\ref{fig:pal-s} using different pairs of hues as well as different
 chroma and luminance contrasts.
@@
 \setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
+\begin{figure}[t!]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-d-construct.png", height = 480, width = 534)
+  png(file = "pal-d-construct.png", height = 960, width = 1068, res = 144)
   hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
@@
 \includegraphics{pal-d-construct}
-\caption{\label{fig:pal-d-construct} Constructing diverging palettes. The chroma/luminance
-plane is shown back to back hues $H = 0$ and $H = 260$. Colors are chosen by simultaneously 
-varying luminance and chroma.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<pal-d-examples, fig=TRUE, height=4, width=6>>=
+
+\vspace*{0.25cm}
+
+\setkeys{Gin}{width=.65\textwidth}
+<<pal-d-examples, fig=TRUE, height=3.3, width=6>>=
 par(mfrow = c(4, 1), mar = c(0.25, 0, 0.25, 0))
@@
 @
-\caption{\label{fig:pal-d-examples} Examples for diverging palettes with different
+\caption{\label{fig:pal-d} Diverging palettes: construction and examples.
+Top: Construction is shown in the chroma/luminance
+plane for hues $H = 0$ and $H = 260$ (back to back). Colors are chosen by simultaneously 
+varying luminance and chroma. Bottom: Examples with different
 pairs of hues and decreasing luminance contrasts.}
@@
 chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
-Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
+Figure~\ref{fig:pal-d} (top panel) shows the chroma/luminance plane back to back for the
 hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
 luminance or chroma contrasts---from a full red over a neutral gray to a full blue.
-As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
+As the top panels of Figure~\ref{fig:pal-s} illustrates, the pair of hues should be chosen carefully
 because the admissible values in the chrome/luminance plane differ across hues.
@@
 admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
-used in Figure~\ref{fig:pal-d-construct} were chosen because they correspond to similar
+used in Figure~\ref{fig:pal-d} (top panel) were chosen because they correspond to similar
 geometric shapes in the chroma/luminance plane, allowing for both large chroma and
@@
 
-Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
+Figure~\ref{fig:pal-d} (bottom panel) shows various examples of conceivable combinations of
 hue, chroma and luminance. The first palette uses a broader range on the luminance axis
 whereas the others mostly rely on chroma contrasts.
+
+\clearpage
 
@@
 <<seats, fig=TRUE, height=4, width=11>>=
-par(mar = rep(0.8, 4), mfrow = c(1, 2))
+par(mar = c(0.8, 1.6, 0, 0), mfrow = c(1, 2))
 seats <- structure(c(226, 61, 54, 51, 222),
@@
 Encoding(names(seats)) <- "latin1"
-pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
-pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
-@
-\caption{\label{fig:seats} Seats in the German parliament.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{sidewaysfigure}[p]
-\begin{center}
-<<votes, fig=TRUE, height=6, width=14>>=
+pie(seats, clockwise = TRUE, col = parties_hsv, radius = 0.85)
+pie(seats, clockwise = TRUE, col = parties_hcl, radius = 0.85)
+@
+
+<<votes, fig=TRUE, height=5.5, width=13>>=
 data("Bundestag2005", package = "vcd")
@@
   just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  margins = unit(c(2.5, 0.5, 1, 11), "lines"),
   keep_aspect_ratio = FALSE, newpage = FALSE)
@@
 @
-\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
+\caption{\label{fig:election} German election 2005 with HSV-based (left) and HCL-based (right)
+qualitative palette. Top: Pie chart for seats in the parliament.
+Bottom: Mosaic display for votes by province.}
 \end{center}
-\end{sidewaysfigure}
+\end{figure}
 
@@
 whether a set of pie segments constitutes a majority. A typical application is shown
-in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
+in the top row of Figure~\ref{fig:election}, visualizing the distribution of seats in the German parliament
 ``Bundestag'' following the 2005 election. In this election, five parties were able to
@@
 be preferred, the balancing properties of the HCL-based palette are very important in more
-complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
-colors and shows the distribution of votes in the 2005 election
+complex displays such as the mosaic display in the bottom row of Figure~\ref{fig:election}.
+This re-uses the same colors and shows the distribution of votes in the 2005 election
 stratified by province (Bundesland). The order of provinces is from 
@@
 Die Linke performed particularly well in the eastern provinces and in Saarland.
-
-Color choice is usually much more important in graphical displays with
-shaded areas compared to displays with only points or lines. However, color choice
-becomes much more relevant also in scatter plots when there are many points. As an
-example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
-the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
-membership, three different colors (green, yellow, red) are employed. The HSV colors in
-the left panel are again very flashy and differ strongly with respect to luminance: the
-yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
-panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
-the amount of color and gray.
 
@@
 
-  png(file = "scatter.png", height = 350, width = 840)
-  par(mfrow = c(1, 2), mar = rep(0.5, 4))
+  png(file = "scatter.png", height = 700, width = 1680, res = 144)
+  par(mfrow = c(1, 2), oma = c(0, 0.1, 0, 0.1), mar = c(1.2, 3, 1.2, 3.3))
   plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
@@
 \includegraphics{scatter}
-\caption{\label{fig:scatter} Scatter plot with three clusters.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
+
+\vspace*{1cm}
+
 <<haireye, fig=TRUE, height=5.5, width=12>>=
@@
 @
-\caption{\label{fig:haireye} Extended mosaic display for hair and eye color data.}
+\caption{\label{fig:further} Further examples for HSV-based (left) and HCL-based (right)
+palettes. Top: Scatter plot with three clusters and qualitative palette.
+Bottom: Extended mosaic display for hair and eye color data with diverging palette.}
 \end{center}
@@
 
-As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
+Color choice is usually much more important in graphical displays with
+shaded areas compared to displays with only points or lines. However, color choice
+becomes much more relevant also in scatter plots when there are many points. As an
+example, the top row in Figure~\ref{fig:further} depicts a scatter plot with three clusters similar to 
+the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
+membership, three different colors (green, yellow, red) are employed. The HSV colors in
+the left panel are again very flashy and differ strongly with respect to luminance: the
+yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
+panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
+the amount of color and gray.
+
+As our final example, the bottom row in Figure~\ref{fig:further}
+\citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
 visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
@@
 under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
-or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
+or black hair and blue eyes. Comparing the HSV and HCL colors in the bottom row of Figure~\ref{fig:further},
 it is shown again that the HCL colors are better balanced (between the red and blue colors)
@@
 \verb/help("rainbow_hcl", package = "colorspace")/. Furthermore, \proglang{R}~code for reproducing
-Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
+the example palettes in Figures~\ref{fig:pal-q}, \ref{fig:pal-s} and \ref{fig:pal-d}
 (and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "colorspace")}.

### 2013-01-23 pkg/colorspace/paper/Palettes/AuthorsResponse.tex
moved the colorspace package into its own subdirectory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@94 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2008-10-24 pkg/paper/Palettes/AuthorsResponse.tex
@@
         Yes, uniformity is a bit more complicated. Hence, we refrain from 
-	discussing its details explicitly here, and have included some
-	remarks in Section~3 pointing out that the CIELUV model also has
-	limitations.
+	discussing its details explicitly here, and have instead included several
+	remarks in Sections~1 and 3 conveying some limitations of the CIELUV model.
 
@@
 
-	We have included pointers to Kaiser and Boynton (1996) and Knoblauch (2002)
+	We have included additional pointers to Kaiser and Boynton (1996) and Knoblauch (2002)
 	at the beginning of Section~3, and also included some further references
@@
         scatterplot, barchart or mosaic plot, to introduce and establish the
-        credentials of the color choices. Cindy Brewer's work covers maps very
-        nicely. But the big problem in using these methods for statistical
-        graphics is that we don't have large swatches of adjoining colour.}\\[-0.3cm]
+        credentials of the color choices.}\\[-0.3cm]
 
@@
 
-	This is a good point! All functions have been moved from vcd to colorspace,
+	This is a good idea! All functions have been moved from vcd to colorspace,
 	as of vcd~1.2-0 and colorspace~0.97.

### 2013-01-23 pkg/dichromat/R/dichromat.R
added Thomas Lumley's dichromat package (current CRAN version 1.2-4)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@95 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,25 @@
+
+dichromat<- function(colours,type=c("deutan","protan")){
+    ##data(dichromat, envir=environment())
+    require("stats")
+    colours<-col2rgb(colours)
+    colours<-t(colours)
+    colnames(colours)<-c("r","g","b")
+    type<-match.arg(type)
+    
+    if(type=="deutan"){
+        nred<-predict(redd,newdata=colours)
+        ngreen<-predict(greend,newdata=colours)
+        nblue<-predict(blued,newdata=colours)
+    } else if (type=="protan"){
+        nred<-predict(redp,newdata=colours)
+        ngreen<-predict(greenp,newdata=colours)
+        nblue<-predict(bluep,newdata=colours)
+    }
+    
+    nred<-pmax(0,pmin(1,nred/255))
+    ngreen<-pmax(0,pmin(1,ngreen/255))
+    nblue<-pmax(0,pmin(1,nblue/255))
+        rgb(nred,ngreen,nblue)
+    
+}

### 2017-09-24 pkg/colorspace/R/hcl_color_picker.R
Removed dichromat depencency for hclwizard, redrawn shiny app images, added invisible return to hcl_color_picker.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@205 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -6,6 +6,8 @@
 #' hue-chroma or luminance-chroma plane. It is also possible to select individual colors and add them
 #' to a palette for comparison and future reference. 
 #'
+#' @return \code{hcl_color_picker} invisibly returns a vector of colors choosen.
+#'    If no colors have been selected \code{NULL} will be returned.
 #' @examples
 #' \dontrun{
 #' hcl_color_picker()
@@ -34,6 +36,7 @@ color_picker_UI <- function() {
 }
 
 color_picker_sidebarPanel <- function() {
+
   # sidebar with controls to select the color
   shiny::sidebarPanel(
     shiny::sliderInput("H", "Hue",
@@ -53,7 +56,12 @@ color_picker_sidebarPanel <- function() {
     shiny::htmlOutput("colorbox"),
     shiny::actionButton("color_picker", "Pick"),
     shiny::actionButton("color_unpicker", "Unpick"),
-    shiny::actionButton("clear_color_picker", "Clear palette")
+    shiny::actionButton("clear_color_picker", "Clear palette"),
+    # Disable "Return to R" button if running on webserver
+    if ( Sys.getenv('SHINY_PORT') == "" ) {
+      shiny::actionButton("closeapp","Return to R")
+    }
+
   )
 }
 
@@ -161,7 +169,7 @@ color_picker_Server <- function() {
       hexcolor <- hex(polarLUV(as.numeric(input$L), as.numeric(input$C), as.numeric(input$H)))
 
       # only add color if it's not already in the list
-      if (! hexcolor %in% picked_color_list$cl) {
+      if ( ! is.na(hexcolor) && ! hexcolor %in% picked_color_list$cl) {
         picked_color_list$cl <- c(picked_color_list$cl, hexcolor)
       }
     })
@@ -244,6 +252,11 @@ color_picker_Server <- function() {
         'Color list: N/A'
       }
     })
+
+    # Return data to R-console
+    shiny::observeEvent(input$closeapp,
+        shiny::stopApp(invisible( picked_color_list$cl ))
+    );
   })
 }
 

### 2017-09-24 pkg/colorspace/inst/hclwizard/prepareStaticContent.R

@@ -7,11 +7,10 @@
 # -------------------------------------------------------------------
 # - EDITORIAL:   2016-10-23, RS: Created file on pc24-c707.
 # -------------------------------------------------------------------
-# - L@ST MODIFIED: 2016-11-03 17:09 on thinkreto
+# - L@ST MODIFIED: 2017-09-24 12:29 on thinkreto
 # -------------------------------------------------------------------
 
    library("colorspace")
-   library("dichromat")
 
 
 # -------------------------------------------------------------------

### 2017-09-24 pkg/colorspace/inst/hclwizard/server.R

@@ -7,12 +7,11 @@
 # -------------------------------------------------------------------
 # - EDITORIAL:   2015-05-01, RS: Created file on thinkreto.
 # -------------------------------------------------------------------
-# - L@ST MODIFIED: 2016-11-16 10:04 on thinkreto
+# - L@ST MODIFIED: 2017-09-24 12:28 on thinkreto
 # -------------------------------------------------------------------
 
 library("shiny")
 library("colorspace")
-library("dichromat")
 
 #options( shiny.trace = TRUE )
 
@@ -171,7 +170,7 @@ shinyServer(function(input, output, session) {
          # Add desaturation or constraints
          if ( input$desaturate ) colors <- desaturate(colors)
          if ( any(tolower(input$constraint) %in% c("protan","deutan","tritan")) )
-            colors <- dichromat(colors,tolower(input$constraint))
+            colors <- do.call(tolower(input$constraint),list("col"=colors))
          # Reverse if required
          if ( input$reverse ) colors <- rev(colors)
          return(colors)
@@ -262,7 +261,9 @@ shinyServer(function(input, output, session) {
       # RAW
       # --------------------------
       # Generate RGB coordinates
-      RGB <- attr(hex2RGB(colors),"coords")
+      sRGB <- hex2RGB(colors)
+      RGB <- attr( sRGB, "coords" )
+      HCL <- round(attr( as( sRGB, "polarLUV" ), "coords" ))
 
       # Generate output string
       append <- function(x,new) c(x,new)
@@ -270,11 +271,11 @@ shinyServer(function(input, output, session) {
       # RGB 0-1
       raw1 <- append(raw1,"<div style=\"clear: both;\">")
       raw1 <- append(raw1,"<span class=\"output-raw\">")
-      raw1 <- append(raw1,"RGB values [0-1]")
-      for ( i in 1:nrow(RGB) )
+      raw1 <- append(raw1,"HCL values")
+      for ( i in 1:nrow(HCL) )
          raw1 <- append(raw1,ifelse(i %in% colors.na,
                         gsub(" ","&nbsp;",sprintf("<code>%5s %5s %5s</code>","NA","NA","NA")),
-                        sprintf("<code>%5.3f %5.3f %5.3f</code>",RGB[i,1],RGB[i,2],RGB[i,3])))
[…]

### 2017-09-24 pkg/colorspace/inst/hclwizard/ui.R

@@ -7,7 +7,7 @@
 # -------------------------------------------------------------------
 # - EDITORIAL:   2015-05-01, RS: Created file on thinkreto.
 # -------------------------------------------------------------------
-# - L@ST MODIFIED: 2016-11-16 10:03 on thinkreto
+# - L@ST MODIFIED: 2017-09-16 13:53 on thinkreto
 # -------------------------------------------------------------------
 
 

### 2017-09-24 pkg/colorspace/inst/hclconverter/html/appInfo.Rmd
Added hcl_converter, alpha version.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@206 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

## App Information

### Terms and Conditions

Why this app ..

### Terms of Use

It is mutually assumed and understood that uploading and converting images with
this app is done at your own risk.  The creator/maintainer of the
``colorspace`` package and/or the [hclwizard.org](http://hclwizard.org) website
assume no responsibility for the content of the uploaded images and a possible
misuse of the uploaded images.

Uploaded images are stored in the computers temporary directory for the
conversion process only and will be deleted immediately after the results are
shown on the front-end.  Neither the uploaded image files or any of the results
will be kept or stored longer than needed.  However, in case of a script or
server fault or if the process is getting killed by the user, uploaded images
and results might not be deleted as intended and therefore stored in the
temporary folder of the corresponding server for a certain amount of time.

```{r,fig=FALSE,echo=TRUE}
library("colorspace")
print( rnorm(3) )
```

### 2017-09-24 pkg/colorspace/inst/hclconverter/server.R

@@ -1,0 +1,183 @@
+# -------------------------------------------------------------------
+# - NAME:        server.R
+# - AUTHOR:      Reto Stauffer
+# - DATE:        2017-09-16
+# -------------------------------------------------------------------
+# - DESCRIPTION:
+# -------------------------------------------------------------------
+# - EDITORIAL:   2017-09-16, RS: Created file on thinkreto.
+# -------------------------------------------------------------------
+# - L@ST MODIFIED: 2017-09-24 12:47 on thinkreto
+# -------------------------------------------------------------------
+
+
+library("shiny")
+library("colorspace")
+library("png")  # Processing png images
+library("jpeg") # Processing jpeg images
+library("RCurl")
+
+# Restrict upload file size to 1MB (online) and 50MB (local)
+options(shiny.maxRequestSize=1*1024^2)
+if ( Sys.getenv('SHINY_PORT') == "" ) options(shiny.maxRequestSize=50*1024^2)
+
+# Resize the image (neares neighbor interpolation/approximation)
+resizeImage = function(im, w.out=600, h.out) {
+
+   # Dimensions of the input image 
+   din <- list("height"=dim(im)[1],"width"=dim(im)[2],"layers"=dim(im)[3])
+   # If h.out is missing: proportional scaling
+   if ( missing(h.out) ) h.out <- round(din$height/(din$width/w.out))
+   # Create empty array to store resized image
+   out = array(0,c(h.out,w.out,din$layers))
+   # Compute ratios -- final number of indices is n.out, spaced over range of 1:n.in
+   w_ratio = din$width  / w.out
+   h_ratio = din$height / h.out
+   # Do resizing -- select appropriate indices
+   for ( l in 1:din$layers )
+      out[,,l] <- im[ floor(h_ratio* 1:h.out), floor(w_ratio* 1:w.out), l]
+
+   return(out)
+}
+
+# The convert function used
+convert <- function(img, type, target) {
+
+   # - Fast if bw
+   if ( type == 'desaturate' ) {
+      bw <- (img[,,1]*0.2126+img[,,2]*0.7152+img[,,3]*0.0722)
+      writePNG(bw,target=target)
+      return(TRUE)
+   }
+   # - Save original colors
+   if ( type == 'original' ) {
+      writePNG(img,target=target)
+     return(TRUE)
+   }
+   # - Loading RGB values of the original image
+   R <- as.vector(img[,,1])
+   G <- as.vector(img[,,2])
+   B <- as.vector(img[,,3])
[…]

### 2017-09-24 pkg/colorspace/inst/hclconverter/ui.R

@@ -1,0 +1,58 @@
+# -------------------------------------------------------------------
+# - NAME:        ui.R
+# - AUTHOR:      Reto Stauffer
+# - DATE:        2017-09-16
+# -------------------------------------------------------------------
+# - DESCRIPTION:
+# -------------------------------------------------------------------
+# - EDITORIAL:   2017-09-16, RS: Created file on thinkreto.
+# -------------------------------------------------------------------
+# - L@ST MODIFIED: 2017-09-24 12:55 on thinkreto
+# -------------------------------------------------------------------
+library("shiny")
+
+desc <- paste("Please select a file from your disc which you want",
+              "to convert. Only PNG/JPG/JPEG files are allowed.",
+              "Maximum allowed file size is 1 Megabyte.")
+
+# Define UI for data upload app ----
+shinyUI(fluidPage(
+   theme = "hclconvert.css",
+
+   # App title ----
+   titlePanel("HCLwizard Image Converter"),
+
+   fluidRow(
+     column(4,
+         img(src="images/descimage.png",id="descimage"),
+         p(desc)
+     ),
+     column(4, 
+         # Input: Select a file ----
+         fileInput("file", label=h4("Select an image to upload"),
+                   multiple = FALSE,
+                   accept = c("image/png","image/jpeg"))
+     ),
+     column(4,h4("HCL converter status"), textOutput("status"))
+   ),
+
+ 
+   # Horizontal line ----
+   tags$hr(),
+ 
+ 
+   tabsetPanel(
+      tabPanel("Images",
+         withTags( div(class="outputgrid", h3("Original"),    uiOutput("imageorig")) ),
+         withTags( div(class="outputgrid", h3("Protanope"),   uiOutput("imageprotan")) ),
+         withTags( div(class="outputgrid", h3("Deuteranope"), uiOutput("imagedeutan")) ),
+         withTags( div(class="outputgrid", h3("Desaturated"), uiOutput("imagedesaturate")) )
+      ),
+      tabPanel("App Info", 
+         htmlOutput("appInfo"),
+         includeHTML("html/appInfo.html")
+      ),
+      selected = "Images"
+   )
+
+)) # End of shinyUI

### 2017-09-24 pkg/colorspace/R/choose_palette.R
Added hcl_converter functions and shiny app, modified desaturate and simulate_cvd functions (RGB matrix I/O).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@207 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -90,6 +90,11 @@ hclwizard <- function(n = 7L, gui = "shiny", shiny.trace = FALSE) {
    do.call(sprintf("choose_palette_%s", gui), args)
 }
 
+#' @rdname choose_palette
+#' @export
+hcl_wizard <- function(n = 7L, gui = "shiny", shiny.trace = FALSE)
+   hclwizard(n=n,gui=gui,shiny.trace=FALSE)
+
 # hclwizard shiny GUI for selecting color palette
 choose_palette_shiny <- function(pal, shiny.trace = FALSE, n = 7L, ...) {
    # Requirements for shiny application

### 2017-09-24 pkg/colorspace/R/desaturate.R

@@ -3,21 +3,32 @@
 #' Transform a vector of given colors to the corresponding colors with chroma
 #' removed (collapsed to zero) in HCL space.
 #' 
-#' Given colors are first transformed to RGB (either using
-#' \code{\link[colorspace]{hex2RGB}} or \code{\link[grDevices]{col2rgb}}) and
-#' then to HCL (\code{\link[colorspace]{polarLUV}}).  In HCL, chroma is removed
-#' (i.e., collapsed to zero) and then the color is transformed back to a
-#' hexadecimal string.
+#' If input \code{col} is a vector given colors are first transformed to RGB
+#' (either using \code{\link[colorspace]{hex2RGB}} or
+#' \code{\link[grDevices]{col2rgb}}) and then to HCL
+#' (\code{\link[colorspace]{polarLUV}}).  In HCL, chroma is removed (i.e.,
+#' collapsed to zero) and then the color is transformed back to a hexadecimal
+#' string.
+#' 
+#' If input \code{col} is a matrix with three rows named \code{R}, \code{G}, and
+#' \code{B} (top down) they are interpreted as Red-Green-Blue values within the
+#' range \code{[0-255]}. The desaturation takes place in the HCL space as well.
+#' Instead of an (s)RGB color vector a matrix of the same size as the input
+#' \code{col} with desaturated Red-Green-Blue values will be returned.
+#' This can be handy to avoid too many conversions.
 #' 
 #' @param col vector of any of the three kind of R colors, i.e., either a color
 #' name (an element of \code{\link[grDevices]{colors}}), a hexadecimal string
 #' of the form \code{"#rrggbb"} or \code{"#rrggbbaa"} (see
 #' \code{\link[grDevices]{rgb}}), or an integer \code{i} meaning
-#' \code{palette()[i]}.
+#' \code{palette()[i]}. Input \code{col} can also be a matrix with three
+#' rows containing R/G/B (0-255) values, see details.
 #' @param amount numeric specifying the amount of desaturation where \code{1}
 #' corresponds to complete desaturation, \code{0} to no desaturation, and
 #' values in between to partial desaturation.
-#' @return A character vector with (s)RGB codings of the colors in the palette.
+#' @return A character vector with (s)RGB codings of the colors in the palette
[…]

### 2017-09-24 pkg/colorspace/R/hcl_converter.R

@@ -1,0 +1,124 @@
+#' Graphical User Interface to Check Images for Color Constriants
+#'
+#' A graphical user interface (GUI) to check an existing jpg/png image for (possible)
+#' color constraints. The image will be converted to protanope vision, deutanope vision,
+#' and a desaturated version (monochromatic vision).
+#' Allows a rapid check whether the colors used in the image show some
+#' constraints with respect to color deficiency or color blindness.
+#'
+#' @param shiny.trace, \code{logical}. Can be set to \code{TRUE} for more verbose
+#'    output (development). 
+#' @rdname hcl_converter
+#' @export
+hcl_converter <- function( x, overwrite = FALSE, shiny.trace = FALSE) {
+
+   # If input 'x' is missing: start interactive GUI
+   if ( missing(x) ) {
+      # Requirements for shiny application
+      stopifnot(requireNamespace("shiny"))
+      appDir <- system.file("hclconverter", package = "colorspace")
+      if (appDir == "")
+         stop("Could not find hclconverter app directory. Try re-installing `colorspace`.", call. = FALSE)
+      # Start shiny
+      options(shiny.trace=shiny.trace)
+      pal <- shiny::runApp(appDir, display.mode = "normal", quiet = TRUE )
+
+   # Do not start shiny: convert and save images on disc
+   } else {
+      is_img <- check_image_type(x)
+      if ( ! is_img$png & ! is_img$jpg )
+         stop(sprintf("Image \"%s\" is neither png nor jpg/jpeg. Stop.",x))
+      types <- c("deutan","protan","tritan","desaturate")
+      files <- sprintf("%s/%s_%s",dirname(x),types,basename(x))
+      # If overwrite is FALSE: check if files exist and stop if.
+      if ( ! overwrite ) {
+         for ( file in files ) {
+            if ( file.exists(file) )
+               stop(sprintf("File \"%s\" exists. Remove image first, or use hcl_converter( x, overwrite = TRUE)",file))
+         }
+      }
+      # Read image
+      if ( is_img$png ) {
+         in.img <- png::readPNG( x )
+      } else {
+         in.img <- jpeg::readJPEG( x )
+      }
+      # Convert and save image
+      for ( type in types ) {
+         outfile <- sprintf("%s/%s_%s",dirname(x),type,basename(x))
[…]

### 2017-09-24 pkg/colorspace/R/simulate_cvd.R

@@ -14,12 +14,19 @@
 #' Internally, they all call \code{simulate_cvd} along with a (possibly interpolated)
 #' version of the matrices from \code{\link{cvd}}. Matrix interpolation can be carried out with
 #' the function \code{interpolate_cvd_transform} (see Examples).
-
+#'
+#' If input \code{col} is a matrix with three rows named \code{R}, \code{G}, and
+#' \code{B} (top down) they are interpreted as Red-Green-Blue values within the
+#' range \code{[0-255]}. Instead of an (s)RGB color vector a matrix of the same size as the
+#' input \code{col} with the corresponding simulated Red-Green-Blue values will be returned.
+#' This can be handy to avoid too many conversions.
+#'
 #' @param col character. A color or vector of colors, e.g., \code{"#FFA801"} or \code{"blue"}.
+#' Input \code{col} can also be a matrix with three rows containing R/G/B (0-255) values, see details.
 #' @param severity numeric. Severity of the color vision defect, a number between 0 and 1.
 #' @param cvd_transform numeric 3x3 matrix, specifying the color vision deficiency transform matrix.
 #' @param cvd list of cvd transformation matrices. See \code{\link{cvd}} for available options.
-
+#'
 #' @references Machado GM, Oliveira MM, Fernandes LAF (2009).
 #'   A Physiologically-Based Model for Simulation of Color Vision Deficiency.
 #'   \emph{IEEE Transactions on Visualization and Computer Graphics}. \bold{15}(6), 1291--1298.
@@ -39,6 +46,10 @@
 #' # simulate color-vision deficiency by calling `simulate_cvd` with interpolated cvd matrix
 #' simulate_cvd(c("#005000", "blue", "#00BB00"),
 #'              interpolate_cvd_transform(tritanomaly_cvd, severity = 0.6))
+#'
+#' # apply CVD directly on RGB matrix
+#' RGB <- t(hex2RGB(rainbow(3))@coords*255)
+#' deutan(RGB)
 #' 
 #' @importFrom grDevices col2rgb
 simulate_cvd <- function(col, cvd_transform) {
@@ -51,9 +62,12 @@ simulate_cvd <- function(col, cvd_transform) {
     alpha <- substr(col, 8L, 9L)
     col <- substr(col, 1L, 7L)
     col <- grDevices::col2rgb(col)
-  }
[…]

### 2017-09-24 pkg/colorspace/inst/hclconverter/server.R

@@ -7,7 +7,7 @@
 # -------------------------------------------------------------------
 # - EDITORIAL:   2017-09-16, RS: Created file on thinkreto.
 # -------------------------------------------------------------------
-# - L@ST MODIFIED: 2017-09-24 12:47 on thinkreto
+# - L@ST MODIFIED: 2017-09-24 13:50 on thinkreto
 # -------------------------------------------------------------------
 
 
@@ -40,40 +40,40 @@ resizeImage = function(im, w.out=600, h.out) {
    return(out)
 }
 
-# The convert function used
-convert <- function(img, type, target) {
-
-   # - Fast if bw
-   if ( type == 'desaturate' ) {
-      bw <- (img[,,1]*0.2126+img[,,2]*0.7152+img[,,3]*0.0722)
-      writePNG(bw,target=target)
-      return(TRUE)
-   }
-   # - Save original colors
-   if ( type == 'original' ) {
-      writePNG(img,target=target)
-     return(TRUE)
-   }
-   # - Loading RGB values of the original image
-   R <- as.vector(img[,,1])
-   G <- as.vector(img[,,2])
-   B <- as.vector(img[,,3])
-   # - Convert to hex colors 
-   hcols <- hex(sRGB(R=R,G=G,B=B))
-   # - Convert to color blindness  
-   hex <- do.call(type,list("col"=hcols))
-   RGB <- hex2RGB(hex)
- 
-   RGB <- attr(RGB,'coords')
-   img2 <- img
-   img2[,,1] <- matrix( RGB[,'R'] ,dim(img)[1],dim(img)[2])
-   img2[,,2] <- matrix( RGB[,'G'] ,dim(img)[1],dim(img)[2])
-   img2[,,3] <- matrix( RGB[,'B'] ,dim(img)[1],dim(img)[2])
- 
-   # Save image to disc
-   writePNG(img2,target=target)
-   return(TRUE)
-}
+#### The convert function used
+###convert <- function(img, type, target) {
+###
+###   # - Fast if bw
+###   if ( type == 'desaturate' ) {
+###      bw <- (img[,,1]*0.2126+img[,,2]*0.7152+img[,,3]*0.0722)
+###      writePNG(bw,target=target)
+###      return(TRUE)
+###   }
+###   # - Save original colors
+###   if ( type == 'original' ) {
+###      writePNG(img,target=target)
+###     return(TRUE)
+###   }
+###   # - Loading RGB values of the original image
+###   R <- as.vector(img[,,1])
[…]

### 2017-10-15 pkg/colorspace/R/scale_sequential.R
Added continuous version of sequential scale, plus some improvements in accompanying vignette.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@221 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -98,3 +98,93 @@ scale_fill_discrete_sequential <- function(palette = NULL, c1 = NULL, c2 = NULL,
   }
   ggplot2::discrete_scale("fill", "manual", pal, ...)
 }
+
+#' HCL-based continuous sequential color scales for ggplot2
+#'
+#' Continuous ggplot2 color scales using the color palettes generated by \code{\link{sequential_hcl}}. 
+#'
+#' 
+#' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
+#' named palette. This enables easy customization of named palettes.
+#' 
+#' @param c1 Beginning chroma value.
+#' @param c2 Ending chroma value.
+#' @param l1 Beginning luminance value.
+#' @param l2 Ending luminance value.
+#' @param h1 Beginning hue value.
+#' @param h2 Ending hue value. If set to \code{NA}, generates a single-hue scale.
+#' @param p1 Control parameter determining how chroma should vary (1 = linear, 2 = quadratic, etc.).
+#' @param p2 Control parameter determining how luminance should vary (1 = linear, 2 = quadratic, etc.).
+#' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
+#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "seq")} for available options.
+#' @param begin Number in the range of \code{[0, 1]} indicating to which point in the color scale the smallest data value should be mapped.
+#' @param end Number in the range of \code{[0, 1]} indicating to which point in the color scale the largest data value should be mapped.
+#' @param na.value Color to be used for missing data points.
+#' @param guide Type of legend. Use \code{"colourbar"} for continuous color bar. 
+#' @param n_interp Number of discrete colors that should be used to interpolate the continuous color scale. 10 will work fine in most cases.
+#' @param ... common continuous scale parameters: `name`, `breaks`, `labels`, and `limits`. See
+#'  \code{\link[ggplot2]{continuous_scale}} for more details.
+#' @examples
+#' library(ggplot2)
+#' 
+#' nx = 87
+#' ny = 61
+#' df <- data.frame(height = c(volcano), x = rep(1:nx, ny), y = rep(1:ny, each = nx))
+#' ggplot(df, aes(x, y, fill=height)) + 
+#'   geom_raster() + scale_fill_continuous_sequential(palette = "Terrain") +
+#'   coord_fixed(expand = FALSE)
+#' @importFrom stats na.omit
+#' @export
[…]

### 2017-10-16 pkg/colorspace/R/scale_diverging.R
Added diverging scales.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@228 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,195 @@
+#' HCL-based discrete diverging color scales for ggplot2
+#'
+#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{diverge_hcl}}. 
+#'
+#' 
+#' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
+#' named palette. This enables easy customization of named palettes.
+#' 
+#' @param c1 Chroma value at the scale endpoints.
+#' @param l1 Luminance value at the scale endpoints.
+#' @param l2 Luminance value at the scale midpoint.
+#' @param h1 Hue value at the first endpoint.
+#' @param h2 Hue value at the second endpoint.
+#' @param p1 Control parameter determining how chroma should vary (1 = linear, 2 = quadratic, etc.).
+#' @param p2 Control parameter determining how luminance should vary (1 = linear, 2 = quadratic, etc.).
+#' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
+#' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
+#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "div")} for available options.
+#' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
+#'  from the data.
+#' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.
+#' @param ... common discrete scale parameters: \code{name}, \code{breaks}, \code{labels}, \code{na.value}, \code{limits} and \code{guide}. See
+#'  \code{\link[ggplot2]{discrete_scale}} for more details.
+#' @examples
+#' library(ggplot2)
+#' 
+#' # default colors with slightly darkened midpoint
+#' ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
+#'   geom_point() + theme_minimal() + 
+#'   scale_color_discrete_diverging(l2=75)
+#' 
+#' # color scale "Green-Orange"
+#' ggplot(iris, aes(Sepal.Length, fill = Species)) +
+#'   geom_density(alpha = 0.7) + theme_classic() +
+#'     scale_fill_discrete_diverging(palette = "Green-Orange", rev = TRUE)
+#'     
+#' # use `nmax` and `order` to skip some colors
[…]

### 2017-10-16 pkg/colorspace/R/scale_sequential.R

@@ -44,7 +44,7 @@ scale_colour_discrete_sequential <- function(palette = NULL, c1 = NULL, c2 = NUL
                                              nmax = NULL, order = NULL, ...)
 {
   # arguments we want to hand off to function sequential_hcl only if explicitly provided
-  hcl_args <- c("palette", "c1", "c1", "l1", "l2", "h1", "h2", "p1", "p2")
+  hcl_args <- c("palette", "c1", "c2", "l1", "l2", "h1", "h2", "p1", "p2")
   
   # match hcl_args to args provided
   args <- as.list(match.call())
@@ -77,7 +77,7 @@ scale_fill_discrete_sequential <- function(palette = NULL, c1 = NULL, c2 = NULL,
                                            nmax = NULL, order = NULL, ...)
 {
   # arguments we want to hand off to function sequential_hcl only if explicitly provided
-  hcl_args <- c("palette", "c1", "c1", "l1", "l2", "h1", "h2", "p1", "p2")
+  hcl_args <- c("palette", "c1", "c2", "l1", "l2", "h1", "h2", "p1", "p2")
   
   # match hcl_args to args provided
   args <- as.list(match.call())
@@ -146,7 +146,7 @@ scale_colour_continuous_sequential <- function(palette = NULL, c1 = NULL, c2 = N
                                                n_interp = 10, ...)
 {
   # arguments we want to hand off to function sequential_hcl only if explicitly provided
-  hcl_args <- c("palette", "c1", "c1", "l1", "l2", "h1", "h2", "p1", "p2")
+  hcl_args <- c("palette", "c1", "c2", "l1", "l2", "h1", "h2", "p1", "p2")
   
   # match hcl_args to args provided
   args <- as.list(match.call())
@@ -176,7 +176,7 @@ scale_fill_continuous_sequential <- function(palette = NULL, c1 = NULL, c2 = NUL
                                                n_interp = 10, ...)
 {
   # arguments we want to hand off to function sequential_hcl only if explicitly provided
-  hcl_args <- c("palette", "c1", "c1", "l1", "l2", "h1", "h2", "p1", "p2")
+  hcl_args <- c("palette", "c1", "c2", "l1", "l2", "h1", "h2", "p1", "p2")
   
   # match hcl_args to args provided
   args <- as.list(match.call())

### 2017-10-16 pkg/colorspace/R/hcl_palettes.R
qualitative_hcl(), sequential_hcl(), diverge_hcl() only allow palette names of the corresponding type
changed the ggplot2 vignette Blue-Yellow to Purple-Green


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@230 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -196,7 +196,7 @@ qualitative_hcl <- function(n, h = c(0, 360 * (n - 1)/n), c = 50, l = 70,
     ## (1) palette
     if(is.character(h)) palette <- h
     pals <- if(!is.null(palette)) {
-        as.matrix(hcl_palettes(name = palette)[, 2L:9L])[1L, ]
+        as.matrix(hcl_palettes(type = "Qualitative", name = palette)[, 2L:9L])[1L, ]
     } else {
         pals <- structure(c(if(length(h) < 2L) c(h, NA) else rep_len(h, 2L), c[1L], NA, l[1L], NA, NA, NA, 1), .Names = vars.pal)
     }
@@ -250,7 +250,7 @@ sequential_hcl <- function(n, h = 260, c = 80, l = c(30, 90), power = 1.5,
     ## (1) palette
     if(is.character(h)) palette <- h
     pals <- if(!is.null(palette)) {
-        as.matrix(hcl_palettes(name = palette)[, 2L:9L])[1L, ]
+        as.matrix(hcl_palettes(type = "Sequential", name = palette)[, 2L:9L])[1L, ]
     } else {
         pals <- structure(c(
 	    if(length(h) < 2L) c(h, NA) else rep_len(h, 2L),
@@ -328,7 +328,7 @@ diverge_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5,
     ## (1) palette
     if(is.character(h)) palette <- h
     pals <- if(!is.null(palette)) {
-        as.matrix(hcl_palettes(name = palette)[, 2L:9L])[1L, ]
+        as.matrix(hcl_palettes(type = "Diverging", name = palette)[, 2L:9L])[1L, ]
     } else {
         structure(c(rep_len(h, 2L), c(c[1L], NA), rep_len(l, 2L), if(length(power) < 2L) c(power, NA) else rep_len(power, 2L), 1), .Names = vars.pal)
     }

### 2018-09-07 pkg/colorspace/R/hcl_palettes.R
introduced non-monotonic chroma trajectories: new cmax argument for piecewise-linear chroma path (w/o power)
-> added to sequential_hcl() and diverge_hcl() and
hence internally to all stored HCL palettes

using this almost all CARTO, ColorBrewer.org, Viridis sequential
palettes can be approximated very closely

-> added to hcl_palettes()
still need to adapt plot() method to cope with this many palettes



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@271 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,6 +1,6 @@
 #' HCL Color Palettes
 #' 
-#' Qualitative, sequential (single-hue and multi-hue), and diverged
+#' Qualitative, sequential (single-hue and multi-hue), and diverging
 #' color palettes based on the HCL (hue-chroma-luminance) color model.
 #' 
 #' Details are still tbd...
@@ -11,9 +11,12 @@
 #' @param l,l1,l2 luminance value in the HCL color description.
 #' @param power,p1,p2 control parameter determining how chroma and luminance should
 #' be increased (1 = linear, 2 = quadratic, etc.).
+#' @param cmax Maximum chroma value in the HCL color description.
 #' @param gamma Deprecated.
-#' @param fixup logical. Should the color be corrected to a valid RGB value
-#' before correction?
+#' @param fixup logical. Should the color be corrected to a valid RGB value?
+#' @param off numeric. Vector of length 2 indicating horizontal and vertical
+#' offsets between the color rectangles displayed.
+#' @param border character. Color of rectangle borders.
 #' @param alpha numeric vector of values in the range \code{[0, 1]} for alpha
 #' transparency channel (0 means transparent and 1 means opaque).
 #' @param name,palette character. Name of HCL color palette.
@@ -30,9 +33,9 @@
 #'
 #' ## visualize
 #' hcl_palettes("qualitative", plot = TRUE)
-#' hcl_palettes("sequential (single-hue)", plot = TRUE)
-#' hcl_palettes("sequential (multi-hue)", plot = TRUE)
-#' hcl_palettes("diverging", plot = TRUE)
+#' hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE)
+#' hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
+#' hcl_palettes("diverging", n = 7, plot = TRUE)
 #'
 #' ## inspect a specific palette
 #' ## (upper-case, spaces, etc. are ignored for matching)
[…]

### 2018-09-07 pkg/colorspace/tests/palettes.R
reflect small updates to 'Grays' palette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@272 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -46,9 +46,9 @@ identical(qualitative_hcl(6, "Pastel 1"),  qualitative_hcl(6, h1 = 0, c1 = 35, l
 identical(qualitative_hcl(6, "Pastel 1"),  qualitative_hcl(6, h1 = 0, h2 = NA, c1 = 35, l1 = 85))
 identical(qualitative_hcl(6, "Pastel 1"),  qualitative_hcl(6, h1 = 0, h2 = 300, c1 = 35, l1 = 85))
 identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, palette = "Grays"))
-identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, h = 0, c = 0, l = c(15, 95), power = 1.3))
-identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, h = 0, c. = 0, l = c(15, 95), power = 1.3))
-identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, c1 = 0, c2 = 0, l1 = 15, l2 = 95, p1 = 1.3))
+identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, h = 0, c = 0, l = c(15, 98), power = 1.3))
+identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, h = 0, c. = 0, l = c(15, 98), power = 1.3))
+identical(sequential_hcl(7, "Grays"),      sequential_hcl(7, c1 = 0, c2 = 0, l1 = 15, l2 = 98, p1 = 1.3))
 identical(sequential_hcl(7, "Red-Yellow"), sequential_hcl(7, palette = "Red-Yellow"))
 identical(sequential_hcl(7, "Red-Yellow"), sequential_hcl(7, h = c(10, 85), c = c(80, 10), l = c(25, 95), power = c(0.4, 1.3)))
 identical(sequential_hcl(7, "Red-Yellow"), sequential_hcl(7, h = c(10, 85), c. = c(80, 10), l = c(25, 95), power = c(0.4, 1.3)))

### 2018-09-07 pkg/colorspace/R/carto_hcl.R
carto_hcl() renamed to divergex_hcl() as it now only provides fle_x_ible diverging palettes, consisting of two sequential_hcl palettes with all possible parameters (including the new cmax)
ggplot2 scales adapted accordingly, documentation updated



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@275 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,169 +1,0 @@
-#' CARTO Sequential and Diverging Palettes Emulated by HCL Palettes
-#' 
-#' Sequential and diverging CARTO color palettes, emulated by sequential
-#' HCL palettes (combining two sequential palettes for diverging palettes).
-#' 
-#' The \code{carto_hcl} function simply calls \code{\link{sequential_hcl}}
-#' with a prespecified set of hue/chorma/luminance parameters that have
-#' been chosen such that the CARTO color palettes (\url{https://carto.com/carto-colors/})
-#' are approximately emulated. In case of diverging CARTO palettes two
-#' sequential HCL palettes are combined, possibly with rather unbalanced
-#' chroma/luminance specifications.
-#'
-#' Available sequential palettes: RedOr, OrYel, Peach, PinkYl, Mint, BluGrn,
-#' DarkMint, Emrld, ag_GrnYl, BluYl, Teal, TealGrn, PurpOr, Sunset, Magenta.
-#'
-#' Available diverging palettes: ArmyRose, Fall, Geyser, Temps, Tropic, Earth.
-#'
-#' Furthermore some CARTO palettes cannot be emulated well with sequential
-#' HCL palettes, typically due to requiring non-monotonic chroma sequences
-#' or sometimes nonlinear hue sequences. The CARTO palettes not available
-#' in \code{carto_hcl} are: Burg, BurgYl, Purp, SunsetDark, ag_Sunset, BrwnYl, TealRose.
-#' 
-#' @param n the number of colors (\eqn{\ge 1}{>= 1}) to be in the palette.
-#' @param palette character with the name (see details).
-#' @param \dots arguments passed to \code{\link{sequential_hcl}}.
-#' @param rev logical. Should the palette be reversed?
-#' @param h1 numeric. Starting hue coordinate.
-#' @param h2 numeric. Either ending hue coordinate (for sequential palettes)
-#' or center hue coordinate (for diverging palettes).
[…]

### 2018-09-07 pkg/colorspace/R/divergex_hcl.R

@@ -1,0 +1,154 @@
+#' (More) Flexible Diverging HCL Palettes
+#' 
+#' Diverging HCL color palettes generated through combination of two fully
+#' flexible (and possibly unbalanced) multi-hue sequential palettes.
+#'
+#' The \code{divergex_hcl} function simply calls \code{\link{sequential_hcl}}
+#' twice with a prespecified set of hue/chroma/luminance parameters. This is
+#' similar to \code{\link{diverge_hcl}} but allows for more flexibility:
+#' \code{diverge_hcl} employs two \emph{single-hue} sequential palettes,
+#' always uses zero chroma for the neutral/central color, and restricts the
+#' chroma/luminance path to be the same in both \dQuote{arms} of the palette.
+#' In contrast, \code{divergex_hcl} relaxes this to two full \emph{multi-hue}
+#' palettes that can thus go through a non-gray neutral color (typically light
+#' yellow). Consequently, the chroma/luminance paths can be rather unbalanced
+#' between the two arms.
+#'
+#' With this additional flexibility various diverging palettes suggested by
+#' \url{http://ColorBrewer.org/} and CARTO (\url{https://carto.com/carto-colors/})
+#' can be emulated.
+#'
+#' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, and
+#' Tropic (with Tropic also available in \code{diverge_hcl}).
+#'
+#' Available ColorBrewer.org palettes: Spectral, PuOr, RdYlGn, RdYlBu, RdGy
+#' BrBG, PiYG, PRGn, RdBu.
+#' 
+#' @param n the number of colors (\eqn{\ge 1}{>= 1}) to be in the palette.
+#' @param palette character with the name (see details).
+#' @param \dots arguments passed to \code{\link{sequential_hcl}}.
+#' @param rev logical. Should the palette be reversed?
+#' @param h1 numeric. Starting hue coordinate.
+#' @param h2 numeric. Center hue coordinate.
[…]

### 2018-09-07 pkg/colorspace/R/scale_divergex.R

@@ -1,11 +1,14 @@
-#' Discrete CARTO scales for ggplot2, emulated by HCL palettes
+#' Discrete Flexible Diverging Scales for ggplot2
 #'
-#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{carto_hcl}}.
+#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{divergex_hcl}}.
 #' 
-#' Available sequential palettes: RedOr, OrYel, Peach, PinkYl, Mint, BluGrn,
-#' DarkMint, Emrld, ag_GrnYl, BluYl, Teal, TealGrn, PurpOr, Sunset, Magenta.
+#' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps
+#' (all requiring the flexibility of \code{divergex_hcl}), and
+#' Tropic (also available in \code{diverge_hcl}).
 #'
-#' Available diverging palettes: ArmyRose, Fall, Geyser, Temps, Tropic, Earth.
+#' Available ColorBrewer.org palettes: Spectral, RdYlGn, RdYlBu, RdGy
+#' (requiring the flexibility of \code{divergex_hcl}), and BrBG, PiYG,
+#' PRGn, PuOr, RdBu (also available in \code{diverge_hcl}).
 #' 
 #' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
 #' named palette. This enables easy customization of named palettes.
@@ -13,7 +16,7 @@
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
 #' @param palette The name of the palette to be used.
 #' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
-#' @param h1,h2,h3,c1,c2,c3,l1,l2,l3,p1,p2,p3,p4 Parameters to customize the scale. See \code{\link{carto_hcl}} for details.
+#' @param h1,h2,h3,c1,c2,c3,l1,l2,l3,p1,p2,p3,p4,cmax1,cmax2 Parameters to customize the scale. See \code{\link{divergex_hcl}} for details.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
 #' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.
@@ -26,28 +29,28 @@
 #' # default color scale
[…]

### 2018-09-07 pkg/colorspace/R/USSouthPolygon.R
rename diverge_hcl() to diverging_hcl() for consistency all palettes now have the adjective and _hcl: qualitative_hcl(), sequential_hcl(), diverging_hcl()
keep diverge_hcl() for backward compatibility
FIXME: should go only into \alias but not into \usage in .Rd

analogously for diverging_hsv() and diverge_hsv
and divergingx_hcl() and divergex_hcl()

ggplot2 scales only provide the "diverging"
but not the "diverge" variant

also the "diverging" variant is now consistently used in
all documentation, examples, internal calls, vignettes, etc.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/

@@ -15,7 +15,7 @@
 #' @keywords datasets
 #' @examples
 #' ## generate color palette
-#' pal <- diverge_hcl(9)
+#' pal <- diverging_hcl(9)
 #' n <- length(pal)
 #' 
 #' ## draw shaded polygons

### 2018-09-07 pkg/colorspace/R/choose_palette.R

@@ -7,7 +7,7 @@
 #' implemented by \code{\link{polarLUV}}). The GUIs interface the palette
 #' functions \code{\link{qualitative_hcl}} for qualitative palettes,
 #' \code{\link{sequential_hcl}} for sequential palettes with a single or
-#' multiple hues, and \code{\link{diverge_hcl}} for diverging palettes (composed
+#' multiple hues, and \code{\link{diverging_hcl}} for diverging palettes (composed
 #' from two single-hue sequential palettes).
 #' 
 #' Two different GUIs are implemented and can be selected using the function
@@ -74,7 +74,7 @@
 #' @importFrom grDevices dev.cur dev.list dev.new dev.off dev.set
 
 #' @export
-choose_palette <- function(pal = diverge_hcl, n = 7L, parent = NULL, gui = "tcltk") {
+choose_palette <- function(pal = diverging_hcl, n = 7L, parent = NULL, gui = "tcltk") {
    args <- list("pal" = pal, "n" = n, "parent" = parent)
    gui <- match.arg(gui, c("tcltk", "shiny"))
    do.call(sprintf("choose_palette_%s", gui), args)
@@ -109,7 +109,7 @@ choose_palette_shiny <- function(pal, shiny.trace = FALSE, n = 7L, ...) {
 }
 
 # tcltk GUI for selecting a color palette
-choose_palette_tcltk <- function( pal = diverge_hcl, n=7L, parent = NULL, ... ) {
+choose_palette_tcltk <- function( pal = diverging_hcl, n=7L, parent = NULL, ... ) {
 
   # Choose a file interactively
   ChooseFile <- function(cmd, win.title, initialfile=NULL, 
@@ -859,7 +859,7 @@ GetPalette <- function(type, h1, h2, c1, c2, l1, l2, p1, p2, fixup) {
                                     list(d1=c(h1, h2), d2=c(c1, c2),
                                          d3=c(l1, l2), d4=c(p1, p2), d5=fixup)))
    } else if (type %in% c("dive","Diverging")) {
-      f <- diverge_hcl
+      f <- diverging_hcl
       formals(f) <- eval(substitute(alist(n=, h=d1, c=d2, l=d3, power=d4,
                                           gamma=NULL, fixup=d5, alpha=1,
                                           palette=NULL, rev=FALSE, ...=,

### 2018-09-07 pkg/colorspace/R/divergingx_hcl.R

@@ -3,13 +3,13 @@
 #' Diverging HCL color palettes generated through combination of two fully
 #' flexible (and possibly unbalanced) multi-hue sequential palettes.
 #'
-#' The \code{divergex_hcl} function simply calls \code{\link{sequential_hcl}}
+#' The \code{divergingx_hcl} function simply calls \code{\link{sequential_hcl}}
 #' twice with a prespecified set of hue/chroma/luminance parameters. This is
-#' similar to \code{\link{diverge_hcl}} but allows for more flexibility:
-#' \code{diverge_hcl} employs two \emph{single-hue} sequential palettes,
+#' similar to \code{\link{diverging_hcl}} but allows for more flexibility:
+#' \code{diverging_hcl} employs two \emph{single-hue} sequential palettes,
 #' always uses zero chroma for the neutral/central color, and restricts the
 #' chroma/luminance path to be the same in both \dQuote{arms} of the palette.
-#' In contrast, \code{divergex_hcl} relaxes this to two full \emph{multi-hue}
+#' In contrast, \code{divergingx_hcl} relaxes this to two full \emph{multi-hue}
 #' palettes that can thus go through a non-gray neutral color (typically light
 #' yellow). Consequently, the chroma/luminance paths can be rather unbalanced
 #' between the two arms.
@@ -19,9 +19,9 @@
 #' can be emulated.
 #'
 #' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, and
-#' Tropic (with Tropic also available in \code{diverge_hcl}).
+#' Tropic (with Tropic also available in \code{diverging_hcl}).
 #'
-#' Available ColorBrewer.org palettes: Spectral, PuOr, RdYlGn, RdYlBu, RdGy
+#' Available ColorBrewer.org palettes: Spectral, PuOr, RdYlGn, RdYlBu, RdGy,
 #' BrBG, PiYG, PRGn, RdBu.
 #' 
 #' @param n the number of colors (\eqn{\ge 1}{>= 1}) to be in the palette.
@@ -51,7 +51,7 @@
 #' div_pal <- function(palette, border = "light gray")
 #' {
 #'   n <- 7
-#'   col <- divergex_hcl(n, palette)
[…]

### 2018-09-07 pkg/colorspace/R/hcl_palettes.R

@@ -152,7 +152,7 @@ plot.hcl_palettes <- function(x, n = 5L, fixup = TRUE, off = NULL, border = NULL
     dimnames = list(c(typ[3L], rownames(x)[x$type == typ[3L]]), paste("Color", 1L:n)))
 
   dcol <- sapply(which(x$type == typ[4L]), function(i) {
-    diverge_hcl(n = n, h1 = xx[i, "h1"], h2 = xx[i, "h2"], c1 = xx[i, "c1"], l1 = xx[i, "l1"], l2 = xx[i, "l2"],
+    diverging_hcl(n = n, h1 = xx[i, "h1"], h2 = xx[i, "h2"], c1 = xx[i, "c1"], l1 = xx[i, "l1"], l2 = xx[i, "l2"],
       p1 = xx[i, "p1"], p2 = xx[i, "p2"], fixup = fixup)
   })
   dcol <- if(length(dcol) < 1L) NULL else matrix(as.vector(rbind("transparent", t(dcol))), ncol = n,
@@ -339,7 +339,7 @@ sequential_hcl <- function(n, h = 260, c = 80, l = c(30, 90), power = 1.5,
 
 #' @rdname hcl_palettes
 #' @export
-diverge_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5,
+diverging_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5,
   gamma = NULL, fixup = TRUE, alpha = 1, palette = NULL, rev = FALSE, ...,
   h1, h2, c1, l1, l2, p1, p2, cmax)
 {
@@ -422,6 +422,9 @@ diverge_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5,
     return(rval)
 }
 
+#' @rdname hcl_palettes
+#' @export
+diverge_hcl <- diverging_hcl
 
 # -------------------------------------------------------------------
 # Palette specifications

### 2018-09-07 pkg/colorspace/R/palettes.R

@@ -14,7 +14,7 @@
 #' HCL(\code{h}, \code{c[1]}, \code{l[1]}) through to a light color
 #' HCL(\code{h}, \code{c[2]}, \code{l[2]}) by interpolation.
 #' 
-#' \code{diverge_hcl} and \code{diverge_hsv}, compute a set of colors diverging
+#' \code{diverging_hcl} and \code{diverging_hsv}, compute a set of colors diverging
 #' from a neutral center (gray or white, without color) to two different
 #' extreme colors (blue and red by default). This is similar to
 #' \code{\link{cm.colors}}. For the diverging HSV colors, two hues \code{h} are
@@ -84,14 +84,14 @@
 #' 
 #' ## a few useful diverging HCL palettes
 #' par(mar = rep(0, 4), mfrow = c(4, 1))
-#' pal(diverge_hcl(7))
-#' pal(diverge_hcl(7, h = c(246, 40), c = 96, l = c(65, 90)))
-#' pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
-#' pal(diverge_hcl(7, h = c(180, 70), c = 70, l = c(90, 95)))
-#' pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
-#' pal(diverge_hcl(7, h = c(128, 330), c = 98, l = c(65, 90)))
-#' pal(diverge_hcl(7, h = c(255, 330), l = c(40, 90)))
-#' pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
+#' pal(diverging_hcl(7))
+#' pal(diverging_hcl(7, h = c(246, 40), c = 96, l = c(65, 90)))
+#' pal(diverging_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
+#' pal(diverging_hcl(7, h = c(180, 70), c = 70, l = c(90, 95)))
+#' pal(diverging_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
+#' pal(diverging_hcl(7, h = c(128, 330), c = 98, l = c(65, 90)))
+#' pal(diverging_hcl(7, h = c(255, 330), l = c(40, 90)))
+#' pal(diverging_hcl(7, c = 100, l = c(50, 90), power = 1))
 #' 
 #' ## sequential palettes
 #' pal(sequential_hcl(12))
@@ -109,16 +109,16 @@
[…]

### 2018-09-07 pkg/colorspace/R/scale_diverging.R

@@ -1,6 +1,6 @@
 #' HCL-based discrete diverging color scales for ggplot2
 #'
-#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{diverge_hcl}}. 
+#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{diverging_hcl}}. 
 #'
 #' 
 #' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
@@ -45,7 +45,7 @@ scale_colour_discrete_diverging <- function(palette = NULL, c1 = NULL, l1 = NULL
                                              h1 = NULL, h2 = NULL, p1 = NULL, p2 = NULL, alpha = 1, rev = FALSE,
                                              nmax = NULL, order = NULL, aesthetics = "colour", ...)
 {
-  # arguments we want to hand off to function diverge_hcl only if explicitly provided
+  # arguments we want to hand off to function diverging_hcl only if explicitly provided
   hcl_args <- c("palette", "c1", "l1", "l2", "h1", "h2", "p1", "p2")
   
   # match hcl_args to args provided
@@ -63,7 +63,7 @@ scale_colour_discrete_diverging <- function(palette = NULL, c1 = NULL, l1 = NULL
     }
     # set the remaining arguments and call qualitative_hcl
     args <- c(args, list(n = nmax, alpha = alpha, rev = rev))
-    do.call(diverge_hcl, args, envir = parent.frame())[order]
+    do.call(diverging_hcl, args, envir = parent.frame())[order]
   }
   ggplot2::discrete_scale(aesthetics, "manual", pal, ...)
 }
@@ -78,7 +78,7 @@ scale_fill_discrete_diverging <- function(...) scale_colour_discrete_diverging(.
 
 #' HCL-based continuous diverging color scales for ggplot2
 #'
-#' Continuous ggplot2 color scales using the color palettes generated by \code{\link{diverge_hcl}}. 
+#' Continuous ggplot2 color scales using the color palettes generated by \code{\link{diverging_hcl}}. 
 #'
 #' 
 #' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
@@ -121,7 +121,7 @@ scale_colour_continuous_diverging <- function(palette = NULL, c1 = NULL, l1 = NU
                                                mid = 0, na.value = "grey50", guide = "colourbar",
                                                n_interp = 11, aesthetics = "colour", ...)
 {
[…]

### 2018-09-07 pkg/colorspace/R/scale_divergingx.R

@@ -1,14 +1,11 @@
 #' Discrete Flexible Diverging Scales for ggplot2
 #'
-#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{divergex_hcl}}.
+#' Discrete ggplot2 color scales using the color palettes generated by \code{\link{divergingx_hcl}}.
 #' 
-#' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps
-#' (all requiring the flexibility of \code{divergex_hcl}), and
-#' Tropic (also available in \code{diverge_hcl}).
+#' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, Tropic.
 #'
-#' Available ColorBrewer.org palettes: Spectral, RdYlGn, RdYlBu, RdGy
-#' (requiring the flexibility of \code{divergex_hcl}), and BrBG, PiYG,
-#' PRGn, PuOr, RdBu (also available in \code{diverge_hcl}).
+#' Available ColorBrewer.org palettes: Spectral, PuOr, RdYlGn, RdYlBu, RdGy,
+#' BrBG, PiYG, PRGn, RdBu.
 #' 
 #' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
 #' named palette. This enables easy customization of named palettes.
@@ -16,7 +13,7 @@
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
 #' @param palette The name of the palette to be used.
 #' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
-#' @param h1,h2,h3,c1,c2,c3,l1,l2,l3,p1,p2,p3,p4,cmax1,cmax2 Parameters to customize the scale. See \code{\link{divergex_hcl}} for details.
+#' @param h1,h2,h3,c1,c2,c3,l1,l2,l3,p1,p2,p3,p4,cmax1,cmax2 Parameters to customize the scale. See \code{\link{divergingx_hcl}} for details.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
 #' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.
@@ -29,26 +26,26 @@
 #' # default color scale
 #' ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
[…]

### 2018-09-07 pkg/colorspace/R/specplot.R

@@ -58,9 +58,9 @@
 #' specplot(rainbow(100))
 #' 
 #' ## spectrum of HCL-based palettes: qualitative/sequential/diverging
-#' specplot(rainbow_hcl(100))
+#' specplot(qualitative_hcl(100))
 #' specplot(sequential_hcl(100))
-#' specplot(diverge_hcl(100))
+#' specplot(diverging_hcl(100))
 #' 
 #' ## return computed RGB and HCL coordinates
 #' res <- specplot(rainbow(10), plot = FALSE)

### 2018-09-07 pkg/colorspace/tests/palettes.R

@@ -4,14 +4,14 @@ library("colorspace")
 rainbow_hcl(12)
 
 ## a few useful diverging HCL palettes
-diverge_hcl(7)
-diverge_hcl(7, h = c(246, 40), c = 96, l = c(65, 90))
-diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90))
-diverge_hcl(7, h = c(180, 70), c = 70, l = c(90, 95))
-diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95))
-diverge_hcl(7, h = c(128, 330), c = 98, l = c(65, 90))
-diverge_hcl(7, h = c(255, 330), l = c(40, 90))
-diverge_hcl(7, c = 100, l = c(50, 90), power = 1)
+diverging_hcl(7)
+diverging_hcl(7, h = c(246, 40), c = 96, l = c(65, 90))
+diverging_hcl(7, h = c(130, 43), c = 100, l = c(70, 90))
+diverging_hcl(7, h = c(180, 70), c = 70, l = c(90, 95))
+diverging_hcl(7, h = c(180, 330), c = 59, l = c(75, 95))
+diverging_hcl(7, h = c(128, 330), c = 98, l = c(65, 90))
+diverging_hcl(7, h = c(255, 330), l = c(40, 90))
+diverging_hcl(7, c = 100, l = c(50, 90), power = 1)
 
 ## sequential palettes
 sequential_hcl(12)
@@ -22,14 +22,14 @@ rainbow_hcl(12)
 desaturate(rainbow_hcl(12))
 
 ## diverging red-blue colors
-diverge_hsv(7)
-diverge_hcl(7, c = 100, l = c(50, 90))
-desaturate(diverge_hsv(7))
-desaturate(diverge_hcl(7, c = 100, l = c(50, 90)))
+diverging_hsv(7)
+diverging_hcl(7, c = 100, l = c(50, 90))
+desaturate(diverging_hsv(7))
+desaturate(diverging_hcl(7, c = 100, l = c(50, 90)))
 
 ## diverging cyan-magenta colors
-diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95))
-desaturate(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
+diverging_hcl(7, h = c(180, 330), c = 59, l = c(75, 95))
[…]

### 2018-09-07 pkg/colorspace/demo/brewer.R
added demos to compare the HCL spectrum of palettes from RColorBrewer/rcartocolor/viridis with their colorspace approximations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@277 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,16 @@
+if(requireNamespace("RColorBrewer")) {
+
+library("colorspace")
+
+brewer_seq <- c(
+  "YlOrRd", "YlOrBr", "OrRd", "Oranges", "YlGn", "YlGnBu", 
+  "Reds", "RdPu", "PuRd", "Purples", "PuBuGn", "PuBu",
+  "Greens", "BuGn", "GnBu", "BuPu", "Blues"
+)
+for(i in brewer_seq) specplot(RColorBrewer::brewer.pal(9, i), sequential_hcl(9, i, rev = TRUE), main = i, rgb = FALSE)
+
+
+brewer_divx <- c("Spectral", "RdYlGn", "RdYlBu", "RdGy", "RdBu", "PiYG", "PRGn", "PuOr", "BrBG")
+for(i in brewer_divx) specplot(RColorBrewer::brewer.pal(11, i), divergingx_hcl(11, i), main = i, rgb = FALSE)
+
+}

### 2018-09-07 pkg/colorspace/demo/carto.R

@@ -1,0 +1,15 @@
+if(requireNamespace("rcartocolor")) {
+
+library("colorspace")
+
+carto_seq <- c(
+  "DarkMint", "Mint", "BluGrn", "Teal", "TealGrn", "Emrld", "BluYl", "ag_GrnYl",
+  "Peach", "PinkYl", "Burg", "BurgYl", "RedOr", "OrYel", "Purp", "PurpOr",
+  "Sunset", "Magenta", "SunsetDark", "ag_Sunset", "BrwnYl"
+)
+for(i in carto_seq) specplot(rcartocolor::carto_pal(7, i), sequential_hcl(7, i, rev = !grepl("ag_", i)), main = i, rgb = FALSE)
+
+carto_divx <- c("ArmyRose", "Earth", "Fall", "Geyser", "TealRose", "Temps", "Tropic")
+for(i in carto_divx) specplot(rcartocolor::carto_pal(7, i), divergingx_hcl(7, i, rev = i == "ArmyRose"), main = i, rgb = FALSE)
+
+}

### 2018-09-07 pkg/colorspace/demo/viridis.R

@@ -1,0 +1,9 @@
+if(requireNamespace("viridis")) {
+
+library("colorspace")
+
+specplot(viridis::viridis(9), sequential_hcl(9, "Viridis"), main = "Viridis", rgb = FALSE)
+specplot(viridis::plasma(9), sequential_hcl(9, "Plasma"), main = "Plasma", rgb = FALSE)
+specplot(viridis::inferno(9), sequential_hcl(9, "Inferno"), main = "Inferno", rgb = FALSE)
+
+}

### 2018-10-06 pkg/colorspace/R/specplot.R
slightly changed examples
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@319 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -59,9 +59,9 @@
 #' specplot(rainbow(100), rgb = TRUE)
 #' 
 #' ## spectrum of HCL-based palettes: qualitative/sequential/diverging
-#' specplot(qualitative_hcl(100))
-#' specplot(sequential_hcl(100))
-#' specplot(diverging_hcl(100))
+#' specplot(qualitative_hcl(100, "Set 2"))
+#' specplot(sequential_hcl(100, "Blues 2"))
+#' specplot(diverging_hcl(100, "Blue-Red"))
 #' 
 #' ## return computed RGB and HCL coordinates
 #' res <- specplot(rainbow(10), plot = FALSE)

### 2018-10-07 pkg/colorspace/R/hcl_palettes.R
more work on new vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@321 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -19,7 +19,7 @@
 #' Three types of palettes are derived based on the HCL model: \itemize{
 #'   \item Qualitative: Designed for coding categorical information, i.e.,
 #'     where no particular ordering of categories is available and every color
-#'     should receive the same perceptual weights.
+#'     should receive the same perceptual weight.
 #'   \item Sequential: Designed for coding ordered/numeric information, i.e.,
 #'     where colors go from high to low (or vice versa).
 #'   \item Diverging: Designed for coding numeric information around a central
@@ -57,7 +57,7 @@
 #' of increasing (or decreasing) luminance. Thus, the \code{l} argument should provide
 #' a vector of length 2 with starting and ending luminance (equivalently, \code{l1} and
 #' \code{l2} can be used). Without chroma (i.e., \code{c = 0}), this simply corresponds
-#' to a palette like \code{\link[grDevices]{gray.colors}}. For adding chroma, a simple
+#' to a grayscale palette like \code{\link[grDevices]{gray.colors}}. For adding chroma, a simple
 #' strategy would be to pick a single hue (via \code{h} or \code{h1}) and then decrease
 #' chroma from some value (\code{c} or \code{c1}) to zero (i.e., gray) along with
 #' increasing luminance. For bringing out the extremes (a dark high-chroma color vs.
@@ -76,7 +76,7 @@
 #' \code{diverging_hcl} codes the underlying numeric values by a triangular luminance
 #' sequence with different hues in the left and in the right arm of the palette. Thus,
 #' it can be seen as a combination of two sequential palettes with some restrictions:
-#' (a) a single hue for each arm of the palette, (b) chroma and luminance trajectory
+#' (a) a single hue is used for each arm of the palette, (b) chroma and luminance trajectory
 #' are balanced between the two arms, (c) the neutral central value has zero chroma.
 #' To specify such a palette a vector of two hues \code{h} (or equivalently \code{h1}
 #' and \code{h2}), either a single chroma value \code{c} (or \code{c1}) or a vector

### 2018-10-11 pkg/colorspace/R/scale_divergingx.R
consistency touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@327 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,4 +1,4 @@
-#' Discrete Flexible Diverging Scales for ggplot2
+#' HCL-Based Discrete Flexible Diverging Scales for ggplot2
 #'
 #' Discrete ggplot2 color scales using the color palettes generated by \code{\link{divergingx_hcl}}.
 #' 
@@ -79,7 +79,7 @@ scale_fill_discrete_divergingx <- function(...) scale_colour_discrete_divergingx
 
   
   
-#' Continuous Flexible Diverging Scales for ggplot2
+#' HCL-Based Continuous Flexible Diverging Scales for ggplot2
 #'
 #' Continuous ggplot2 color scales using the color palettes generated by \code{\link{divergingx_hcl}}.
 #'

### 2018-10-13 pkg/colorspace/index.Rmd
changed title, added index for pkgdown
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@331 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

### A Toolbox for Manipulating and Assessing Colors and Palettes

```{r hcl, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
library("colorspace")
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

```{r palettes, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
hcl_palettes(name = c(
  "Pastel 1", "Dark 3", "Harmonic",
  "Blues 3", "Greens 3", "Reds 3",
  "Heat", "Purple-Blue", "Viridis",
  "Blue-Red", "Green-Brown", "Tropic"
), plot = TRUE, n = 7)
```

```{r ggplot2, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) + scale_fill_discrete_qualitative()
```

```{r inferno, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 3.7, fig.align = "center", dev = "png", dpi = 48}
par(mar = rep(0, 4))
demoplot(sequential_hcl(99, "Inferno", rev = TRUE), type = "heatmap")
```
```{r deutan, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 3.7, fig.align = "center", dev = "png", dpi = 48}
par(mar = rep(0, 4))
demoplot(deutan(sequential_hcl(99, "Inferno", rev = TRUE)), type = "heatmap")
```

```{r desaturate, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 3.7, fig.align = "center", dev = "png", dpi = 48}
par(mar = rep(0, 4))
demoplot(desaturate(sequential_hcl(99, "Inferno", rev = TRUE)), type = "heatmap")
```

| Color spaces                               | HCL-based palettes                          | ggplot2 scales                                |
|:------------------------------------------:|:-------------------------------------------:|:---------------------------------------------:|
| ![](index_files/figure-html/hcl-1.png)     | ![](index_files/figure-html/palettes-1.png) | ![](index_files/figure-html/ggplot2-1.png)    |

| Demos                                      | Color vision deficiency                     | Utilities                                     |
|:------------------------------------------:|:-------------------------------------------:|:---------------------------------------------:|
| ![](index_files/figure-html/inferno-1.png) | ![](index_files/figure-html/deutan-1.png)   | ![](index_files/figure-html/desaturate-1.png) |

### 2018-10-13 pkg/colorspace/index.Rmd
completed split-up of colorspace.Rmd which itself has been re-written to give a first quick overview of the package's functionality
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@332 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 library("ggplot2")
+theme_set(theme_minimal())
 ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) + scale_fill_discrete_qualitative()

### 2018-10-14 pkg/colorspace/R/specplot.R
tweaked specplot() to be somewhat usable within other layouts (but still requires care)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@337 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -31,6 +31,9 @@
 #' @param type,lwd,lty,pch plotting parameters passed to
 #' \code{\link[graphics]{lines}} for drawing the RGB and HCL coordinates,
 #' respectively. Can be vectors of length 3.
+#' @param mar,oma numeric or logical. Either numeric vectors of length 4 giving
+#' the (outer) margins or a logical indicating whether \code{mar}/\code{oma}
+#' should be set.
 #' @param main character. Main title of the plot.
 #' @param legend logical. Should legends for the coordinates be plotted?
 #' @param palette logical. Should the given palette \code{x} be plotted?
@@ -69,8 +72,8 @@
 #' @export specplot
 #' @importFrom graphics axis image layout legend lines mtext text par plot
 specplot <- function(x, y = NULL, rgb = FALSE, hcl = TRUE, fix = TRUE, cex = 1,
-  type = "l", lwd = 2 * cex, lty = 1, pch = NULL, main = NULL,
-  legend = TRUE, palette = TRUE, plot = TRUE, ...)
+  type = "l", lwd = 2 * cex, lty = 1, pch = NULL, mar = NULL, oma = NULL,
+  main = NULL, legend = TRUE, palette = TRUE, plot = TRUE, ...)
 {
 
   # Replace NA x with white, required for hex2RGB.
@@ -143,14 +146,14 @@ specplot <- function(x, y = NULL, rgb = FALSE, hcl = TRUE, fix = TRUE, cex = 1,
   show_rgb <- !identical(rgb, FALSE)
   show_hcl <- !identical(hcl, FALSE)
 
-  if(plot & (show_rgb | show_hcl) & (length(x.na) == length(x))) {
+  if(plot & (length(x.na) == length(x))) {
     opar <- par(no.readonly = TRUE)
-    on.exit(par(opar))
-
-    par(xaxt = "n", yaxt = "n", bty = "n", mar = rep(0, 4))
+    nam <- names(par(xaxt = "n", yaxt = "n", bty = "n", mar = rep(0, 4)))
+    on.exit(par(opar[nam]))
+    
     plot(0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1))
     text(0, 0, "All colors NA\nCannot draw spectrum", col = 2)
-  } else if(plot & (show_rgb | show_hcl)) {
+  } else if(plot) {
[…]

### 2018-10-14 pkg/colorspace/index.Rmd
renamed hcl_palettes(name = ...) and divergingx_palettes(name = ...) to palette = ... for consistency with actual palette functions and ggplot2 scales
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@338 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ```{r palettes, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
-hcl_palettes(name = c(
+hcl_palettes(palette = c(
   "Pastel 1", "Dark 3", "Harmonic",

### 2018-10-14 pkg/colorspace/R/divergingx_hcl.R

@@ -25,7 +25,7 @@
 #' RdYlGn, Spectral.
 #' 
 #' @param n the number of colors (\eqn{\ge 1}{>= 1}) to be in the palette.
-#' @param palette,name character with the name (see details).
+#' @param palette character with the name (see details).
 #' @param \dots arguments passed to \code{\link{sequential_hcl}}.
 #' @param rev logical. Should the palette be reversed?
 #' @param h1 numeric. Starting hue coordinate.
@@ -123,7 +123,7 @@ divergex_hcl <- divergingx_hcl
 
 #' @rdname divergingx_hcl
 #' @export
-divergingx_palettes <- function(name = NULL, plot = FALSE, n = 7L, ...)
+divergingx_palettes <- function(palette = NULL, plot = FALSE, n = 7L, ...)
 {
   ## collect all divergingx palettex
   pals <- as.data.frame(do.call("rbind", divex_pals))
@@ -134,16 +134,16 @@ divergingx_palettes <- function(name = NULL, plot = FALSE, n = 7L, ...)
   pals <- pals[, c(16L, 1L:15L)]
 
   ## subset by type and name (by flexible matching)
-  if(!is.null(name)) {
+  if(!is.null(palette)) {
     fx <- function(n) tolower(gsub("[-, _, \\,, (, ), \\ , \\.]", "", n))
     namtab <- fx(rownames(pals))
-    name <- sapply(fx(name), function(n) {
+    palette <- sapply(fx(palette), function(n) {
       if(n %in% namtab) return(n)
       n <- startsWith(namtab, n)
-      if(all(!n)) stop("Palette 'name' should be one of: ", paste(rownames(pals), collapse = ", "))
+      if(all(!n)) stop("Named 'palette' should be one of: ", paste(rownames(pals), collapse = ", "))
       namtab[which(n)[1L]]
     })
-    pals <- pals[fx(rownames(pals)) %in% name, , drop = FALSE]
+    pals <- pals[fx(rownames(pals)) %in% palette, , drop = FALSE]
   }
 
   ## add class and show selection

### 2018-10-14 pkg/colorspace/R/hcl_palettes.R

@@ -111,7 +111,7 @@
 #' @param border character. Color of rectangle borders.
 #' @param alpha numeric vector of values in the range \code{[0, 1]} for alpha
 #' transparency channel (0 means transparent and 1 means opaque).
-#' @param name,palette character. Name of HCL color palette.
+#' @param palette character. Name of HCL color palette.
 #' @param rev logical. Should the color palette vector be returned in reverse order?
 #' @param \dots Other arguments passed to \code{\link{hex}}.
 #' @param type character indicating type of HCL palette.
@@ -144,8 +144,8 @@
 #'
 #' ## inspect a specific palette
 #' ## (upper-case, spaces, etc. are ignored for matching)
-#' hcl_palettes(name = "Dark 2")
-#' hcl_palettes(name = "dark2")
+#' hcl_palettes(palette = "Dark 2")
+#' hcl_palettes(palette = "dark2")
 #' 
 #' ## set up actual colors
 #' qualitative_hcl(4, h = c(0, 288), c = 50, l = 60) ## by hand
@@ -182,7 +182,7 @@
 #' ## palettes (for hues 260/blue and 0/red)
 #' hclplot(diverge_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
 #' @export
-hcl_palettes <- function(type = NULL, name = NULL, plot = FALSE, n = 5L, ...)
+hcl_palettes <- function(type = NULL, palette = NULL, plot = FALSE, n = 5L, ...)
 {
   ## subset by type and name (by flexible matching)
   fx <- function(n) tolower(gsub("[-, _, \\,, (, ), \\ , \\.]", "", n))
@@ -201,15 +201,15 @@ hcl_palettes <- function(type = NULL, name = NULL, plot = FALSE, n = 5L, ...)
   } else {
     pals <- hcl_pals
   }
-  if(!is.null(name)) {
+  if(!is.null(palette)) {
     namtab <- fx(rownames(pals))
-    name <- sapply(fx(name), function(n) {
+    palette <- sapply(fx(palette), function(n) {
       if(n %in% namtab) return(n)
       n <- startsWith(namtab, n)
-      if(all(!n)) stop("Palette 'name' should be one of: ", paste(rownames(pals), collapse = ", "))
+      if(all(!n)) stop("Named 'palette' should be one of: ", paste(rownames(pals), collapse = ", "))
[…]

### 2018-10-14 pkg/colorspace/R/lighten.R
first version of manipulation utitlities vignette ready
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@343 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -25,7 +25,7 @@
 #' 
 #' Programmatically lightening and darkening colors can yield unexpected results (see examples). In HCL space,
 #' colors can become either too gray or overly colorful. By contrast, in HLS space it can happen that the
-#' overallamount of lightening or darkening appears to be non-uniform among a group of colors that are 
+#' overall amount of lightening or darkening appears to be non-uniform among a group of colors that are 
 #' lightened or darkened jointly, and again, colors can become either too gray or overly colorful. We
 #' recommend to try different color spaces if the default space for the chosen function (\code{lighten} 
 #' or \code{darken}) does not look right in a specific application.

### 2018-10-15 pkg/colorspace/R/specplot.R
completed first draft of palette visualization vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@347 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -10,7 +10,7 @@
 #' the hues (\code{fix = TRUE}). Also, to avoid jumps from 0 to 360 or vice
 #' versa, the hue coordinates are shifted suitably.
 #' 
-#' By default (\code{plot = TRUE}), the resulting HCL and RGB coordinates are
+#' By default (\code{plot = TRUE}), the resulting HCL and optionally RGB coordinates are
 #' visualized by simple line plots along with the color palette \code{x}
 #' itself.
 #' 

### 2018-10-15 pkg/colorspace/R/swatchplot.R

@@ -3,9 +3,9 @@
 #' Visualization of color palettes in columns of color swatches.
 #' 
 #' The function \code{swatchplot} is a convenience function for displaying
-#' collections of palettes that can be speciefied as lists or matrices of
-#' character color specifications. Essentially, the is just a call to
-#' \code{\link[graphics]{rect}}. The value-added are the heuristics used
+#' collections of palettes that can be specified as lists or matrices of
+#' character color specifications. Essentially, the function just calls
+#' \code{\link[graphics]{rect}} but the value-added are the heuristics used
 #' for choosing default labels, margins, spacings, borders. These are selected
 #' to work well for \code{\link{hcl_palettes}} and might need further tweaking
 #' in future versions.

### 2018-11-02 pkg/colorspace/R/hclplot.R
- situation with C < 0 and L < 1 is treated the same as C > 0 and L < 1 - relaxed conditions for treating a certain coordinate as constant - use median() rather than mean() to compute the representative constant   coordinate - try to better catch if something goes wrong with the lm()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@365 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -162,11 +162,12 @@ hclplot <- function(x, type = NULL, h = NULL, c = NULL, l = NULL,
             nd <- expand.grid(C = 0:maxchroma, L = 0:100)
             if(!is.null(h)) {
                 nd$H <- h
-            } else if(diff(range(HCL[, "H"], na.rm = TRUE)) < 5) {
-                nd$H <- mean(HCL[, "H"], na.rm = TRUE)
+            } else if(n < 3L || diff(range(HCL[, "H"], na.rm = TRUE)) < 12) {
+                nd$H <- median(HCL[, "H"], na.rm = TRUE)
             } else {
                 m <- lm(H ~ C + L, data = as.data.frame(HCL))
-                if(summary(m)$sigma > 7.5) warning("cannot approximate H well as a linear function of C and L")
+		sig <- summary(m)$sigma
+                if(is.na(sig) || sig > 7.5) warning("cannot approximate H well as a linear function of C and L")
                 nd$H <- predict(m, nd)
             }
             if(is.null(main)) {
@@ -212,17 +213,18 @@ hclplot <- function(x, type = NULL, h = NULL, c = NULL, l = NULL,
                 } else {
                     nd$H <- h
                 }
-            } else if(diff(range(HCL[left, "H"]  - min(HCL[ left, "H"], na.rm = TRUE), na.rm = TRUE)) < 5 &
-                     diff(range(HCL[right, "H"] - min(HCL[right, "H"], na.rm = TRUE), na.rm = TRUE)) < 5) {
-                nd$H[nd$left]  <- mean(HCL[ left, "H"] - min(HCL[ left, "H"], na.rm = TRUE), na.rm = TRUE) + min(HCL[ left, "H"], na.rm = TRUE)
-                nd$H[!nd$left] <- mean(HCL[right, "H"] - min(HCL[right, "H"], na.rm = TRUE), na.rm = TRUE) + min(HCL[right, "H"], na.rm = TRUE)
+            } else if(n < 6L || (diff(range(HCL[left, "H"]  - min(HCL[ left, "H"], na.rm = TRUE), na.rm = TRUE)) < 12 &
+                                 diff(range(HCL[right, "H"] - min(HCL[right, "H"], na.rm = TRUE), na.rm = TRUE)) < 12)) {
+                nd$H[nd$left]  <- median(HCL[ left, "H"] - min(HCL[ left, "H"], na.rm = TRUE), na.rm = TRUE) + min(HCL[ left, "H"], na.rm = TRUE)
[…]

### 2018-11-05 pkg/colorspace/R/cvd_image.R
slightly adapted docs of apps for vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@369 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,7 +1,7 @@
-#' Graphical User Interface to Check Images for Color Constriants
+#' Graphical User Interface to Check Images for Color Constraints
 #'
 #' A graphical user interface (GUI) to check an existing jpg/png image for (possible)
-#' color constraints. The image will be converted to protanope vision, deutanope vision,
+#' color constraints. The image will be converted to protanope vision, deuteranope vision,
 #' and a desaturated version (monochromatic vision).
 #' Allows a rapid check whether the colors used in the image show some
 #' constraints with respect to color deficiency or color blindness.

### 2018-11-05 pkg/colorspace/R/hcl_color_picker.R

@@ -1,4 +1,4 @@
-#' Shiny App to Pick Colors in HCL Space
+#' Graphical User Interface to Pick Colors in HCL Space
 #' 
 #' The app visualizes colors either along the hue-chroma plane for a given luminance value or along the
 #' luminance-chroma plane for a given hue. Colors can be entered by specifying the hue (H), chroma (C),

### 2019-01-07 pkg/colorspace/R/scale_diverging.R
spell out hcl_palettes(type = '...') in the documentation, e.g., 'sequential' rather than 'seq'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@385 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -16,7 +16,7 @@
 #' @param p2 Control parameter determining how luminance should vary (1 = linear, 2 = quadratic, etc.).
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
 #' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
-#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "div")} for available options.
+#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "diverging")} for available options.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
 #' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.

### 2019-01-07 pkg/colorspace/R/scale_qualitative.R

@@ -12,7 +12,7 @@
 #' @param h2 Ending hue value.
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
 #' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
-#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "qual")} for available options.
+#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "qualitative")} for available options.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
 #' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.

### 2019-01-07 pkg/colorspace/R/scale_sequential.R

@@ -17,7 +17,7 @@
 #' @param p2 Control parameter determining how luminance should vary (1 = linear, 2 = quadratic, etc.).
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
 #' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
-#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "seq")} for available options.
+#' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "sequential")} for available options.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
 #' @param order Numeric vector listing the order in which the colors should be used. Default is \code{1:nmax}.

### 2019-01-07 pkg/colorspace/R/lighten.R
fix 'space' labeling in example for darken()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@390 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -69,18 +69,18 @@
 #' # darken light colors, example 1
 #' cl <- qualitative_hcl(5, "Pastel 1")
 #' swatchplot(list(
-#'   combined = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "combined"), "30%" = darken(cl, 0.3, space = "combined")),
-#'   HCL      = rbind("0%" = cl, "15%" = darken(cl, 0.15),                     "30%" = darken(cl, 0.3)),
-#'   HLS      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HLS"),      "30%" = darken(cl, 0.3, space = "HLS"))),
+#'   combined = rbind("0%" = cl, "15%" = darken(cl, 0.15),                      "30%" = darken(cl, 0.3)),
+#'   HCL      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HCL"),       "30%" = darken(cl, 0.3, space = "HCL")),
+#'   HLS      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HLS"),       "30%" = darken(cl, 0.3, space = "HLS"))),
 #'   nrow = 4, line = 2.5
 #' )
 #'
 #' # darken light colors, example 2 
 #' cl <- c("#CDE4F3","#E7F3D3","#F7F0C7","#EFCFE5","#D0D1E7")
 #' swatchplot(list(
-#'   combined = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "combined"), "30%" = darken(cl, 0.3, space = "combined")),
-#'   HCL      = rbind("0%" = cl, "15%" = darken(cl, 0.15),                     "30%" = darken(cl, 0.3)),
-#'   HLS      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HLS"),      "30%" = darken(cl, 0.3, space = "HLS"))),
+#'   combined = rbind("0%" = cl, "15%" = darken(cl, 0.15),                      "30%" = darken(cl, 0.3)),
+#'   HCL      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HCL"),       "30%" = darken(cl, 0.3, space = "HCL")),
+#'   HLS      = rbind("0%" = cl, "15%" = darken(cl, 0.15, space = "HLS"),       "30%" = darken(cl, 0.3, space = "HLS"))),
 #'   nrow = 4, line = 2.5
 #' )
 #' @export lighten

### 2019-01-07 pkg/colorspace/R/hcl_palettes.R
'basic' sequential palettes need to have all chroma values < 100 (for hclwizard) -> only affects classification of 'Peach' palette (previously basic -> now advanced)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@391 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -749,10 +749,10 @@ GetPaletteConfig <- function(gui = NULL) {
            "Blues 3", "Purples 3", "Reds 3", "Greens 3", "Oslo",
            # Sequential multiple-hues
            "Purple-Blue", "Purple-Orange", "Red-Blue", "Red-Purple",
-           "Red-Yellow", "Heat", "PinkYl", "Peach", "Green-Yellow", "Terrain
-           2", "Dark Mint", "BluYl", "Blue-Yellow", "Viridis", "Plasma",
+           "Red-Yellow", "Heat", "PinkYl", "Green-Yellow", "Terrain 2",
+	   "Dark Mint", "BluYl", "Blue-Yellow", "Viridis", "Plasma",
            # Sequential multiple-hues advanced
-           "YlGnBu", "Greens", "BuGn", "Teal", "Blues", "BuPu", "Purples",
+           "YlGnBu", "Greens", "BuGn", "Teal", "Peach", "Blues", "BuPu", "Purples",
            "Purp", "Burg", "Reds", "YlOrRd", "Sunset", "RdPu", "Inferno",
            "Lajolla", "Turku",
            # Base color maps (for shiny)
@@ -772,7 +772,7 @@ GetPaletteConfig <- function(gui = NULL) {
        # Extending the type names for use in GUIs
        idx <- which(with(pals, (type == "dive" &  !is.na(CMAX) |
                                (type == "seqs" & (!is.na(CMAX) | !is.na(P2))) |
-                               (type == "seqm" & (!is.na(CMAX))))))
+                               (type == "seqm" & (!is.na(CMAX) | C1 > 100 | C2 > 100)))))
        if ( length(idx) > 0 ) pals$type[idx] <- sprintf("%s_advanced", pals$type[idx])
    }
 

### 2019-01-07 pkg/colorspace/R/specplot.R
various fixes in hue smoothing inside specplot() - better detection of jumps by +/- 360 in hue - use linear instead of spline approximation if fit is good enough - better NA handling - improved comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@392 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -72,6 +72,7 @@
 #' print(res)
 #' @export specplot
 #' @importFrom graphics axis image layout legend lines mtext text par plot
+#' @importFrom stats median sd
 specplot <- function(x, y = NULL, rgb = FALSE, hcl = TRUE, fix = TRUE, cex = 1,
   type = "l", lwd = 2 * cex, lty = 1, pch = NULL, mar = NULL, oma = NULL,
   main = NULL, legend = TRUE, palette = TRUE, plot = TRUE, ...)
@@ -97,39 +98,53 @@ specplot <- function(x, y = NULL, rgb = FALSE, hcl = TRUE, fix = TRUE, cex = 1,
   }
 
   # Fixing hue paths
-  # (1) as(RGB, "polarLUV") returns hue's in the
-  # range of 0-360. A palette from -100 to +100 results in
-  # c(260-360,0,100) - the iterative approach corrects this.
-
   if(fix & nrow(HCL) > 1L) {
+    # (1) as(RGB, "polarLUV") returns hue's in the
+    # range of 0-360. A palette from -100 to +100 results in
+    # c(260-360,0,100) - the iterative approach corrects this.
+
+    ## strategy:
+    ## change H by +360 or -360 if the hue trajectory has a very large jump,
+    ## (i.e., difference > +/- 300) or if the trajectory is essentially linear
+    ## but has a jump by +360 or -360 in additional to the "usual" difference
+    ## (i.e., median difference +/- 360)
+    md <- median(HCL[2L:nrow(HCL), "H"] - HCL[1L:(nrow(HCL) - 1L), "H"], na.rm = TRUE)
     for(i in 2L:nrow(HCL)) {
       if ( any(is.na(HCL[(i-1L):i,])) ) next
       d <- HCL[i, "H"] - HCL[i - 1L, "H"]
-      if (abs(d) > 320) HCL[i, "H"] <- HCL[i, "H"] - sign(d) * 360
+      if (abs(d) > 300 || (abs(d + 360 - md) < 4 || abs(d - 360 - md) < 4) ) HCL[i, "H"] <- HCL[i, "H"] - sign(d) * 360
       if (abs(HCL[i, "H"]) >  360) HCL[1L:i, "H"] <- HCL[1L:i, "H"] - sign(HCL[i, "H"]) * 360
     }
 
-    # (2) Smoothing hue values in batches where chroma is very low
[…]

### 2019-01-07 pkg/colorspace/R/scale_sequential.R
in sequential ggplot2 scales set rev=TRUE by default: - for most palettes this yield "low -> high" = "light -> dark/colorful" - thus matches scale_color_brewer()
adapted vignettes and Rd examples correspondingly



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@394 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -6,6 +6,12 @@
 #' If both a valid palette name and palette parameters are provided then the provided palette parameters overwrite the parameters in the
 #' named palette. This enables easy customization of named palettes.
 #' 
+#' Compared to \code{\link{sequential_hcl}} the ordering of the colors in the sequential ggplot2 scale
+#' are reversed by default (i.e., \code{rev = TRUE}) to be more consistent with ggplot2's own scales such as
+#' \code{\link[ggplot2]{scale_color_brewer}}. For most named palettes this leads to darker and more
+#' colorful colors for larger values on the scale. This is typically the better default on light/white
+#' backgrounds.
+#'
 #' @param c1 Beginning chroma value.
 #' @param c2 Ending chroma value.
 #' @param cmax Maximum chroma value.
@@ -16,7 +22,7 @@
 #' @param p1 Control parameter determining how chroma should vary (1 = linear, 2 = quadratic, etc.).
 #' @param p2 Control parameter determining how luminance should vary (1 = linear, 2 = quadratic, etc.).
 #' @param alpha Numeric vector of values in the range \code{[0, 1]} for alpha transparency channel (0 means transparent and 1 means opaque).
-#' @param rev If \code{TRUE}, reverses the order of the colors in the color scale.
+#' @param rev If \code{TRUE} (default), reverses the order of the colors in the color scale (compared to \code{\link{sequential_hcl}}).
 #' @param palette The name of the palette to be used. Run \code{hcl_palettes(type = "sequential")} for available options.
 #' @param nmax Maximum number of different colors the palette should contain. If not provided, is calculated automatically
 #'  from the data.
@@ -42,7 +48,7 @@
 #' @importFrom stats na.omit
 #' @export
 scale_colour_discrete_sequential <- function(palette = NULL, c1 = NULL, c2 = NULL, cmax = NULL, l1 = NULL, l2 = NULL,
-                                             h1 = NULL, h2 = NULL, p1 = NULL, p2 = NULL, alpha = 1, rev = FALSE,
+                                             h1 = NULL, h2 = NULL, p1 = NULL, p2 = NULL, alpha = 1, rev = TRUE,
                                              nmax = NULL, order = NULL, aesthetics = "colour", ...)
 {
   # arguments we want to hand off to function sequential_hcl only if explicitly provided
[…]

### 2019-01-10 pkg/colorspace/R/choose_palette.R
introduce dedicated environment .colorspace_env that can be used to pass around options (e.g., devloper flags)
this avoids some of the check warnings about <<- assignments

the remaining <<- warnings are avoided by defining the objects in the
enclosing environment within the function


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@411 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -101,9 +101,42 @@ hclwizard <- function(n = 7L, gui = "shiny", ...) {
 hcl_wizard <- function(n = 7L, gui = "shiny", ...)
    hclwizard(n = n, gui = gui, ...)
 
+# -------------------------------------------------------------------
+# Environment for passing around internal information
+# -------------------------------------------------------------------
+
+.colorspace_env <- new.env()
+
+.colorspace_get_info <- function(x = NULL) {
+  if(is.null(x)) return(as.list(.colorspace_env))
+  x <- as.character(x)[1L]
+  return(.colorspace_env[[x]])
+}
+
+.colorspace_set_info <- function(...) {
+  dots <- list(...)
+  if(is.null(names(dots))) {
+    stop("arguments must be named")
+  } else if(any(names(dots) == "")) {
+    warning("ignoring unnamed arguments")
+    dots <- dots[names != ""]
+  }
+  if(length(dots) > 0L) {
+    for(i in names(dots)) {
+      .colorspace_env[[i]] <- dots[[i]]
+    }
+  }
+  invisible(NULL)
+}
+
+.colorspace_set_info(
+  hclwizard_autohclplot = FALSE,
+  hclwizard_ninit       = 7,
+  hclwizard_verbose     = FALSE
+)
 
 # Setting global variables to avoid notes during R CMD check
-utils::globalVariables(c("autohclplot", "verbose", "frame2.cvs.paloffset",
+utils::globalVariables(c("frame2.cvs.paloffset",
                          "frame2.cvs.palwidth", "type",
                          "frame3.scl.1.2", "frame3.scl.2.2", "frame3.scl.3.2", "frame3.scl.4.2",
                          "frame3.scl.5.2", "frame3.scl.6.2", "frame3.scl.7.2", "frame3.scl.8.2",
@@ -124,15 +157,12 @@ choose_palette_shiny <- function(pal, n = 7L, ...) {
    if (appDir == "")
       stop("Could not find hclwizard app directory. Try re-installing `colorspace`.", call. = FALSE)
    # Start shiny
-   Sys.setenv("hclwizard_Ninit" = n)
-   dots         <- list(...)
-   autohclplot <<- ifelse(is.null(dots$autohclplot), FALSE, as.logical(dots$autohclplot))
-   Sys.setenv("hclwizard_autohclplot" = autohclplot)
[…]

### 2019-01-10 pkg/colorspace/R/hcl_palettes.R

@@ -547,6 +547,7 @@ diverging_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5,
 #' @export
 diverge_hcl <- diverging_hcl
 
+
 # -------------------------------------------------------------------
 # Palette specifications
 # -------------------------------------------------------------------

### 2019-01-10 pkg/colorspace/inst/hclwizard/server.R

@@ -17,10 +17,6 @@ library("colorspace")
 
 bpy <- eval(parse(text = "colorspace:::bpy"))
 
-# Create some global variables to avoid notes during R CMD build
-if(getRversion() >= "2.15.1") utils::globalVariables(c("verbose", "autohclplot"))
-
-
 
 # Define server logic required to draw a histogram
 shinyServer(function(input, output, session) {
@@ -283,8 +279,7 @@ shinyServer(function(input, output, session) {
       if ( missing(colors) ) colors <- getColors( input$N )
 
       # dimension to collapse. If autohclplot option has been set to TRUE: take NULL.
-      autohclplot <- as.logical(Sys.getenv("hclwizard_autohclplot"))
-      if ( autohclplot ) {
+      if ( colorspace:::.colorspace_get_info("hclwizard_autohclplot") ) {
           plot_type <- NULL
       } else if ( grepl("^qual", input$typ) ) {
           plot_type <- "qualitative"

### 2019-01-10 pkg/colorspace/inst/hclwizard/ui.R

@@ -33,9 +33,7 @@ default = list("N"    =   7,
                "P1"   = 0.7,
                "P2"   = 1.3)
 # Take default argument if set
-if ( nchar(Sys.getenv("hclwizard_Ninit")) > 0 &
-     is.numeric(as.numeric(Sys.getenv("hclwizard_Ninit"))) )
-   default$N <- as.numeric(Sys.getenv("hclwizard_Ninit"))
+default$N <- colorspace:::.colorspace_get_info("hclwizard_ninit")
 
 
 # Define UI for application that draws a histogram

### 2019-01-10 pkg/colorspace/index.Rmd
improvements in vignettes/www
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@370 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
h2 title on index page (due to new font)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@406 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
employ ag_Synset (rather than Heat) on index page to correspond to logo/favicon
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@412 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2018-10-13 pkg/colorspace/index.Rmd
@@
-### A Toolbox for Manipulating and Assessing Colors and Palettes
+## A Toolbox for Manipulating and Assessing Colors and Palettes
 
@@
 ```{r palettes, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
-hcl_palettes(name = c(
+hcl_palettes(palette = c(
   "Pastel 1", "Dark 3", "Harmonic",
   "Blues 3", "Greens 3", "Reds 3",
-  "Heat", "Purple-Blue", "Viridis",
+  "ag_Sunset", "Purple-Blue", "Viridis",
   "Blue-Red", "Green-Brown", "Tropic"
@@
 
+| Interactive color apps                         |
+|:----------------------------------------------:|
+| ![](articles/screenshots/choose_palette_1.png) |
+
 | Demos                                      | Color vision deficiency                     | Utilities                                     |

### 2019-01-10 pkg/colorspace/R/hcl_palettes.R
added 'register' option for qualitative_hcl(), sequential_hcl(), and diverge_hcl()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@413 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -113,6 +113,8 @@
 #' transparency channel (0 means transparent and 1 means opaque).
 #' @param palette character. Name of HCL color palette.
 #' @param rev logical. Should the color palette vector be returned in reverse order?
+#' @param register character. If set to a non-empty character string, the corresponding
+#' palette is registered with that name for subsequent use (within the same session).
 #' @param \dots Other arguments passed to \code{\link{hex}}.
 #' @param type character indicating type of HCL palette.
 #' @param plot logical. Should the selected HCL color palettes be visualized?
@@ -186,20 +188,21 @@ hcl_palettes <- function(type = NULL, palette = NULL, plot = FALSE, n = 5L, ...)
 {
   ## subset by type and name (by flexible matching)
   fx <- function(n) tolower(gsub("[-, _, \\,, (, ), \\ , \\.]", "", n))
+  pals <- .colorspace_get_info("hcl_pals")
   if(!is.null(type)) {
-    tytab <- c("sequential", fx(levels(hcl_pals$type)))
+    tytab <- c("sequential", fx(levels(pals$type)))
     type <- lapply(type, function(ty) {
       ty <- startsWith(tytab, fx(ty))
-      if(all(!ty)) stop("Palette 'type' should be one of: ", paste(levels(hcl_pals$type), collapse = ", "))
+      if(all(!ty)) stop("Palette 'type' should be one of: ", paste(levels(pals$type), collapse = ", "))
       ty <- tytab[which(ty)[1L]]
       if(ty == "sequential") ty <- c("sequentialsinglehue", "sequentialmultihue")
       return(ty)
     })
     type <- unlist(type)
-    type <- levels(hcl_pals$type)[tytab[-1L] %in% type]
-    pals <- hcl_pals[as.character(hcl_pals$type) %in% type, , drop = FALSE]
+    type <- levels(pals$type)[tytab[-1L] %in% type]
+    pals <- pals[as.character(pals$type) %in% type, , drop = FALSE]
   } else {
-    pals <- hcl_pals
+    pals <- pals
   }
   if(!is.null(palette)) {
     namtab <- fx(rownames(pals))
[…]

### 2019-01-11 pkg/colorspace/R/hcl_palettes.R
better documentation of 'register' functionality in .Rd, vignettes/articles, and NEWS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@417 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -182,7 +182,20 @@
 #' 
 #' ## balanced diverging palette constructed from two simple single-hue sequential
 #' ## palettes (for hues 260/blue and 0/red)
-#' hclplot(diverge_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
+#' hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
+#' 
+#' ## to register a particular adapted palette for re-use in the same session
+#' ## with a new name the register=... argument can be used once, e.g.,
+#' diverging_hcl(7, palette = "Tropic", h2 = 0, register = "mytropic")
+#' 
+#' ## subsequently palete="mytropic" is available in diverging_hcl() and the diverging
+#' ## ggplot2 scales such as scale_color_continuous_diverging() etc.
+#' demoplot(diverging_hcl(11, "mytropic"), type = "map")
+#' 
+#' ## to register this palette in all R sessions you could place the following
+#' ## code in a startup script (e.g., .Rprofile):
+#' ## colorspace::diverging_hcl(7, palette = "Tropic", h2 = 0, register = "mytropic")
+#' 
 #' @export
 hcl_palettes <- function(type = NULL, palette = NULL, plot = FALSE, n = 5L, ...)
 {

### 2019-01-11 pkg/colorspace/R/hclplot.R

@@ -79,7 +79,8 @@
 #' 
 #' ## balanced diverging palette constructed from two simple single-hue sequential
 #' ## palettes (for hues 260/blue and 0/red)
-#' hclplot(diverge_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
+#' hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
+#' 
 #' @export hclplot
 #' @importFrom graphics box lines mtext par plot points rect text
 #' @importFrom stats cor lm median predict

### 2019-02-12 pkg/colorspace/vignettes/screenshots/nigeria.R
new endrainbow vignette with examples from different publications
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@443 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,45 @@
+library("colorspace")
+
+source("../../paper/Palettes/Nigeria/bnd.R")
+m <- read.bnd("../../paper/Palettes/Nigeria/nigeria37.bnd")
+fit <- read.table("../../paper/Palettes/Nigeria/reml_purespatial_f_district_spatial.res", header = TRUE)
+fit <- fit[order(fit$district),]
+
+nigeria_map <- function(col, file) {
+
+  png(file = file, height = 600 + 60, width = 700, res = 72)
+
+  layout(rbind(1, c(3, 2, 4)), heights = c(600, 60), widths = c(300, 800, 300))
+  par(mar = c(0.5, 0, 0, 0), oma = c(2, 1.9, 0, 1.9))
+
+  plot(m, z = fit$pmode, fill = rev(col),
+    zlim = c(-0.65, 0.65), legend = FALSE)
+
+  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 6, col = rev(col))
+  box()
+  axis(1, cex.axis = 2)
+
+  dev.off()
+}
+
+nigeria_rainbow <- rev(rainbow(99, start = 0, end = 1/3))
+nigeria_map(nigeria_rainbow,         "nigeria-rainbow.png")
+nigeria_map(protan(nigeria_rainbow), "nigeria-rainbow-protan.png")
+
+## nigeria_csda <- diverging_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)
+## nigeria_map(nigeria_csda,         "nigeria-csda.png")
+## nigeria_map(protan(nigeria_csda), "nigeria-csda-protan.png")
+
+nigeria_purplegreen <- diverging_hcl(99, "Purple-Green", rev = TRUE)
+nigeria_map(nigeria_purplegreen,         "nigeria-purplegreen.png")
+nigeria_map(protan(nigeria_purplegreen), "nigeria-purplegreen-protan.png")
+
+png(file = "nigeria-rainbow-spectrum.png", height = 600 + 60, width = 700, pointsize = 20)
+specplot(nigeria_rainbow, lwd = 3)
+dev.off()
+
[…]

### 2019-02-12 pkg/colorspace/vignettes/screenshots/noaa.R

@@ -1,0 +1,58 @@
+## packages
+library("colorspace") ## 1.4-1
+library("png")
+
+
+# manipulate the image by adding new color palette
+noaa_pal <- toupper(c("#7dfe00","#00cd00","#008b00", "#104d8b","#1d90fc","#01b1ee",
+  "#00eeee","#8967cd","#912cee", "#8b008b","#8b0000", "#cd0000", "#f04100", "#8b008b"))
+
+manipulate <- function(img, col = sequential_hcl(13, rev = TRUE), repl = noaa_pal)
+{
+  col <- c(col, "#FFFFFF")
+
+  # - Loading RGB values of the original image
+  R <- as.vector(img[,,1])
+  G <- as.vector(img[,,2])
+  B <- as.vector(img[,,3])
+  # - Convert to hex colors
+  hcols <- hex(sRGB(R = R, G = G, B = B))
+
+  # - Now converting
+  for(i in seq_along(repl)) hcols[hcols == repl[i]] <- col[i]
+
+  # - Converting back to sRGB
+  RGB <- hex2RGB(hcols)
+
+  # - Add new RGB values to original image
+  RGB <- attr(RGB,"coords")
+  img[,,1] <- matrix(RGB[, "R"], dim(img)[1], dim(img)[2])
+  img[,,2] <- matrix(RGB[, "G"], dim(img)[1], dim(img)[2])
+  img[,,3] <- matrix(RGB[, "B"], dim(img)[1], dim(img)[2])
+
+  return(img) 
+}
+
+if(FALSE) {
+
+# - Reading the "original" (slightly modified) image of the noaa
+img <- readPNG("noaa-orig.png")
+
+# - Desaturated ones
+writePNG(manipulate(img, col = deutan(noaa_pal)),     target = "noaa-orig-deutan.png")
+writePNG(manipulate(img, col = desaturate(noaa_pal)), target = "noaa-orig-desaturate.png")
+
+bams_pal <- sequential_hcl(13, "Purple-Yellow", rev = TRUE) ## slightly modified compared to Figure 4
+writePNG(manipulate(img, col = bams_pal),             target = "noaa-purpleyellow.png")
+writePNG(manipulate(img, col = deutan(bams_pal)),     target = "noaa-purpleyellow-deutan.png")
+writePNG(manipulate(img, col = desaturate(bams_pal)), target = "noaa-purpleyellow-desaturate.png")
+
[…]

### 2019-02-13 pkg/colorspace/R/simulate_cvd.R
improved docs/refs about CVD
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@446 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -4,7 +4,9 @@
 #' based on a CVD transform matrix.
 #'
 #' Using the physiologically-based model for simulating color vision deficiency (CVD)
-#' of Machado et al. (2009), deutanope, protanope, and tritanope vision can be emulated.
+#' of Machado et al. (2009), different kinds of limitations can be
+#' emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
+#' and tritanope (blue cone cells defective).
 #' The workhorse function to do so is \code{simulate_cvd} which can take any vector
 #' of valid R colors and transform them according to a certain CVD transformation
 #' matrix (see \code{\link{cvd}}) and transformation equation.

### 2019-02-18 pkg/colorspace/paper/Toolbox/slides.Rnw
presentation slides (~30min) for 'toolbox' webpage/paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@447 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
%\usetheme[nototalframenumber, noslidenumber, license, url]{uibk}
\usetheme[license, url]{uibk}
\renewcommand{\headerimage}[1]{%
   \IfStrEqCase{#1}{%
      {1}{%
         \gdef\myheaderimageid{#1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{colorspace.png}%
      }}[%
         \gdef\myheaderimageid{1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{colorspace.png}%
      ]%
}
\headerimage{1}

\title{A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
\subtitle{\phantom{.}\\Achim Zeileis, Jason C. Fisher, Kurt Hornik, Ross Ihaka, Claire D. McWhite, Paul Murrell, Reto Stauffer, Claus O. Wilke}
\author{\vspace*{-6.83cm}\hspace*{12.5cm}\includegraphics[width=1.6cm]{palmfooter.png}\\[5.5cm]}
\URL{http://colorspace.R-Forge.R-project.org/}

%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<preliminaries>>=
library("colorspace")
library("ggplot2")
theme_set(theme_minimal())
@

\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

\begin{document}

\section{colorspace}

\subsection{Color spaces}

\begin{frame}
\frametitle{Color spaces}

\textbf{Origin of the package:} Convert colors between various three-dimensional representations of color.

\medskip

\textbf{In particular:} From the perceptually-based HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB, and corresponding hex codes) space.

\setkeys{Gin}{width=0.8\textwidth}
\begin{center}
<<color-spaces, fig = TRUE, width = 9.5, height = 4.5>>=
par(mar = rep(0, 4))

d <- data.frame(
  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
  stringsAsFactors = FALSE
)

plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
lines(y ~ x, data = d[-c(1, 3, 7, 8),])
lines(y ~ x, data = d[c(1, 3, 5),])
lines(y ~ x, data = d[c(6, 7, 9),])
lines(y ~ x, data = d[c(6, 8, 9),])
points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)

text(d$x, d$y, d$lab)
text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
@
\end{center}

\end{frame}

\subsection{HCL vs. RGB}

\begin{frame}
\frametitle{HCL vs.~RGB}

\begin{minipage}{0.47\textwidth}
\textbf{HCL:} Polar coordinates in CIELUV. Captures perceptual dimensions of the human visual system very well.

\bigskip

\setkeys{Gin}{width=\textwidth}
<<hcl-properties, fig = TRUE, width = 4, height = 2.2>>=
swatchplot(
  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  nrow = 3, line = 5.2, off = 0
)
@
\end{minipage}
\pause
\hspace*{0.6cm}
\begin{minipage}{0.47\textwidth}
\textbf{RGB:} Motivated by how computers/TVs used to generate and still represent color.

\bigskip

\setkeys{Gin}{width=0.9\textwidth}
<<rgb-properties, fig = TRUE, width = 3.6, height = 2.2>>=
swatchplot(
  "Red"       = rgb(0:4/4, 0, 0),
  "Green"     = rgb(0, 0:4/4, 0),
  "Blue"      = rgb(0, 0, 0:4/4),
  nrow = 3, off = 0
)
@
\end{minipage}

\end{frame}

\begin{frame}
\frametitle{HCL vs.~RGB: The End of the Rainbow}

\setkeys{Gin}{width=0.44\textwidth}
<<hcl-rainbow, fig = TRUE, width = 5.2, height = 6>>=
specplot(rainbow_hcl(99, c = 60, l = 75), rgb = TRUE)
@
\hspace*{0.8cm}
\pause
<<rgb-rainbow, fig = TRUE, width = 5.2, height = 6>>=
specplot(rainbow(95), rgb = TRUE)
@

\end{frame}

\subsection{Color palettes}

\begin{frame}
\frametitle{Color palettes: Somewhere over the Rainbow}

\setkeys{Gin}{width=\textwidth}
<<hcl-palettes-principles, fig = TRUE, width = 12, height = 1.6>>=
swatchplot(
  "Qualitative (Set 2)"     = rbind("Color" = qualitative_hcl(5, "Set 2"),     "Desaturated" = desaturate(qualitative_hcl(5, "Set 2"))),
  "Sequential (Blues 3)"    = rbind("Color" = sequential_hcl(7, "Blues 3"),    "Desaturated" = desaturate(sequential_hcl(7, "Blues 3"))),
  "Diverging (Green-Brown)" = rbind("Color" = diverging_hcl(7, "Green-Brown"), "Desaturated" = desaturate(diverging_hcl(7, "Green-Brown"))),
  nrow = 3, line = 7)
@

\bigskip
\pause

\textbf{Qualitative:} For categorical information, i.e., where no particular ordering of categories is available.
Function: \code{qualitative\_hcl()}.

\bigskip
\pause

\textbf{Sequential:} For ordered/numeric information, i.e., where colors go from high to low (or vice versa).
Function: \code{sequential\_hcl()}.

\bigskip
\pause

\textbf{Diverging:} For ordered/numeric information around a central neutral value, i.e.,
where colors diverge from neutral to two extremes. Function: \code{diverging\_hcl()}.

\end{frame}

\begin{frame}[fragile]
\frametitle{Color palettes: Somewhere over the Rainbow}

\only<1>{
\textbf{Sequential:} Luminance contrast is crucial (dark to light or vice versa).
}
\only<2>{
\textbf{Blues 2:} Single hue. Decreasing chroma with increasing luminance.
}
\only<3>{
\textbf{Blues 3:} Single hue. Triangular chroma to achieve higher luminance contrast.
}
\only<4>{
\textbf{Blues:} Multi hue. Triangular chroma. High luminance contrast.
}

\vspace*{-0.4cm}

\begin{minipage}[t]{0.49\textwidth}
\vspace*{0.46cm}

\setkeys{Gin}{width=\textwidth}
<<blues, fig = TRUE, width = 4, height = 2.2>>=
swatchplot(
  "Blues 2" = sequential_hcl(7, "Blues 2"),
  "Blues 3" = sequential_hcl(7, "Blues 3"),
  "Blues"   = sequential_hcl(7, "Blues")
)
@
\end{minipage}
\only<2>{%
\begin{minipage}[t]{0.49\textwidth}
\vspace*{0cm}

\setkeys{Gin}{width=\textwidth}
<<blues2-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
specplot(sequential_hcl(7, "Blues 2"))
@
\end{minipage}%
}%
\only<3>{%
\begin{minipage}[t]{0.49\textwidth}
\vspace*{0cm}

\setkeys{Gin}{width=\textwidth}
<<blues3-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
specplot(sequential_hcl(7, "Blues 3"))
@
\end{minipage}
}%
\only<4>{%
\begin{minipage}[t]{0.49\textwidth}
\vspace*{0cm}

\setkeys{Gin}{width=\textwidth}
<<blues-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
specplot(sequential_hcl(7, "Blues"))
@
\end{minipage}
}

\end{frame}

\begin{frame}[fragile]
\frametitle{Color palettes: Somewhere over the Rainbow}

\textbf{Diverging:} Combine two sequential palettes with balanced chroma/luminance.

\vspace*{-0.4cm}

\begin{minipage}[t]{0.49\textwidth}
\vspace*{0cm}

\setkeys{Gin}{width=\textwidth}
<<red-blue-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
specplot(diverging_hcl(9, "Blue-Red", h2 = 5, rev = TRUE))
@
\end{minipage}
\pause
\begin{minipage}[t]{0.49\textwidth}
\vspace*{0cm}

\setkeys{Gin}{width=\textwidth}
<<green-brown-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
specplot(diverging_hcl(9, "Green-Brown"))
@
\end{minipage}

\end{frame}

\begin{frame}[fragile]
\frametitle{Color palettes: Somewhere over the Rainbow}

\setkeys{Gin}{width=\textwidth}
<<hcl-palettes, fig = TRUE, width = 12, height = 6, echo = TRUE>>=
hcl_palettes(plot = TRUE)
@

\end{frame}

\subsection{Statistical graphics}

\begin{frame}[fragile]
\frametitle{Statistical graphics}

\textbf{Base:}

\begin{itemize}
  \item HCL palette functions return hex color vector.
  \item Typically passed to \code{col = } argument of base plotting functions.
\end{itemize}

\bigskip

\textbf{ggplot2:}

\begin{itemize}
  \item Scales of type \code{scale_<aesthetic>_<datatype>_<colorscale>()}.
  \item \code{<aesthetic>} is \code{fill} or \code{color}/\code{colour}.
  \item \code{<datatype>} is \code{discrete} or \code{continuous}.
  \item \code{<colorscale>} is \code{qualitative}, \code{sequential}, \code{diverging}, or \code{divergingx}.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Statistical graphics: Base}

<<base-graphics, eval = FALSE, echo = TRUE>>=
q4 <- qualitative_hcl(4, palette = "Dark 3")
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
@

\vspace*{-0.6cm}

\setkeys{Gin}{width=0.8\textwidth}
<<base-graphics-plot, fig = TRUE, width = 10, height = 6>>=
<<base-graphics>>
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Statistical graphics: Base}

<<base-graphics2, eval = FALSE, echo = TRUE>>=
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
spineplot(ttnc, col = sequential_hcl(2, palette = "Purples 3"))
@

\vspace*{-0.6cm}

\setkeys{Gin}{width=0.8\textwidth}
<<base-graphics2-plot, fig = TRUE, width = 10, height = 6>>=
<<base-graphics2>>
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Statistical graphics: ggplot2}

\setkeys{Gin}{width=0.8\textwidth}
<<ggplot2-graphics, fig = TRUE, width = 10, height = 5.5, echo = TRUE>>=
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Statistical graphics: ggplot2}

\setkeys{Gin}{width=0.8\textwidth}
<<ggplot2-graphics2, fig = TRUE, width = 10, height = 5.5, echo = TRUE>>=
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
@

\end{frame}

\subsection{Visualization and assessment}

\begin{frame}[fragile]
\frametitle{Visualization and assessment}

\textbf{Visualizations:} Based on vector of colors.

\begin{itemize}
  \item \code{swatchplot()}: Color swatches.
  \item \code{specplot()}: Spectrum of HCL and/or RGB trajectories.
  \item \code{hclplot()}: Trajectories in 2-dimensional HCL space projections.
  \item \code{demoplot()}: Illustrations of typical (and simplified) statistical graphics.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Visualization and assessment: \texttt{hclplot()}}

<<hclplot, eval = FALSE, echo = TRUE>>=
hclplot(qualitative_hcl(7, palette = "Set 2"))
hclplot( sequential_hcl(7, palette = "Blues 3"))
hclplot(  diverging_hcl(7, palette = "Blue-Red"))
@

\medskip

\setkeys{Gin}{width=\textwidth}
<<hclplot-plot, fig = TRUE, width = 15, height = 5, png = TRUE, pdf = FALSE>>=
par(mfrow = c(1, 3))
<<hclplot>>
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Visualization and assessment: \texttt{demoplot()}}

<<demoplot-code, echo = TRUE, eval = FALSE>>=
cl <- sequential_hcl(5, palette = "Heat")
demoplot(cl, type = "...")
@

\medskip

\setkeys{Gin}{width=\textwidth}
<<demoplot, fig = TRUE, width = 15, height = 6>>=
par(mfrow = c(2, 5))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
@

\end{frame}

\subsection{Color vision deficiency}

\begin{frame}[fragile]
\frametitle{Color vision deficiency}

\textbf{Emulate:} Color vision deficies.

\begin{itemize}
  \item \code{deutan()}: Deuteranopia (green deficient).
  \item \code{protan()}: Protanopia (red deficient).
  \item \code{tritan()}: Tritanopia (blue deficient).
\end{itemize}

\bigskip

\textbf{Example:} Maunga Whau \code{volcano} data.

\vspace*{-2.2cm}

\setkeys{Gin}{width=0.6\textwidth}
<<volcano-persp, fig = TRUE, width = 5.5, height = 5.5>>=
z <- 2 * volcano
x <- 10 * (1:nrow(z))
y <- 10 * (1:ncol(z))
par(mar = rep(0, 4))
persp(x, y, z, theta = 10, phi = 30, col = "lightgray", scale = FALSE,
  ltheta = 100, shade = 0.5, border = NA, box = FALSE)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Color vision deficiency}

\setkeys{Gin}{width=\textwidth}
<<volcano-cvd, fig = TRUE, width = 15, height = 7>>=
par(mfrow = c(2, 3), mar = rep(0.5, 4), oma = c(1, 20, 3, 1), las = 1)
rain <- rainbow(11, end = 2/3)
blyl <- sequential_hcl(11, "Blue-Yellow")
demoplot(rain, "heatmap")
mtext("rainbow(11, end = 2/3)", side = 2, line = 1)
mtext("Original", line = 1)
demoplot(desaturate(rain), "heatmap")
mtext("Desaturated", line = 1)
demoplot(deutan(rain), "heatmap")
mtext("Deuteranope", line = 1)
demoplot(blyl, "heatmap")
mtext("sequential_hcl(11, \"Blue-Yellow\")", side = 2, line = 1)
demoplot(desaturate(blyl), "heatmap")
demoplot(deutan(blyl), "heatmap")
@

\end{frame}

\subsection{Approximations of other palettes}

\begin{frame}
\frametitle{Approximations of other palettes}

\vspace*{-0.1cm}

\setkeys{Gin}{width=0.49\textwidth}
<<ylgnbu-spectrum, fig = TRUE, width = 5.3, height = 5.6>>=
specplot(
  rev(RColorBrewer::brewer.pal(9, "YlGnBu")),
  sequential_hcl(9, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
@
\pause
\setkeys{Gin}{width=0.49\textwidth}
<<viridis-spectrum, fig = TRUE, width = 5.3, height = 5.6>>=
specplot(
  viridis::viridis(9),
  sequential_hcl(9, "Viridis"),
  main = "Viridis")
@

\end{frame}

\subsection{Color apps}

\begin{frame}[fragile]
\frametitle{Color apps}

\textbf{Facilitate exploration:} Graphical user interfaces as shiny apps.

\begin{itemize}
  \item \emph{Palette constructor:} \code{choose\_palette()} or \code{hclwizard()} (also in tcltk).
  \item \emph{Color picker:} \code{choose\_color()} or \code{hcl\_color\_picker()}.
  \item \emph{Color vision deficiency emulator:} \code{cvd\_emulator()}.
\end{itemize}

\bigskip

\textbf{Online versions:} \url{http://hclwizard.org/}

\end{frame}

\begin{frame}
\frametitle{Color apps: \texttt{choose\_palette()} / \texttt{hclwizard()}}

\includegraphics[width=0.9\textwidth]{../../vignettes/screenshots/choose_palette_1.png}

\end{frame}

\begin{frame}
\frametitle{Color apps: \texttt{choose\_color()} / \texttt{hcl\_color\_picker()}}

\includegraphics[width=0.65\textwidth]{../../vignettes/screenshots/choose_color.png}

\end{frame}

\begin{frame}
\frametitle{Color apps: \texttt{cvd\_emulator()}}

\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_1.png}
\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_2.png}

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\small

Ihaka R, Murrell P, Hornik K, Fisher JC, Stauffer R, Wilke CO, McWhite CD, Zeileis A (2019).
  \emph{colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
  \url{http://colorspace.R-Forge.R-project.org/}

\bigskip

Zeileis A, Hornik K, Murrell P (2009).
 \dquote{Escaping RGBland: Selecting Colors for Statistical Graphics.}
 \textit{Computational Statistics \& Data Analysis}, \textbf{53}, 3259--3270.
 \doi{10.1016/j.csda.2008.11.033}.

\bigskip

Stauffer R, Mayr GJ, Dabernig M, Zeileis A (2015).
  \dquote{Somewhere over the Rainbow: How to Make Effective Use of Colors in Meteorological Visualizations.}
  \emph{Bulletin of the American Meteorological Society}, \textbf{96}(2), 203--216.
  \doi{10.1175/BAMS-D-13-00155.1}

\end{frame}

\end{document}

### 2019-02-19 pkg/colorspace/paper/Toolbox/slides.Rnw
added recommendation slide
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@449 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\section{colorspace}
+\section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
 
@@
 
+\subsection{Recommendations}
+
+\begin{frame}
+\frametitle{Recommendations}
+
+\begin{itemize}
+  \item Do not overestimate the effectiveness of color.
+  \item Choose type of palette based on the data to be visualized.
+  \item For areas use light colors (higher luminance, lower chroma).
+  \item For points/lines darker colors are needed (lower luminance, higher chroma).
+  \item For palettes with more colors stronger luminance contrasts are needed.
+  \item Triangular chroma trajectories useful for distinguishing central colors.
+\end{itemize}
+
+\end{frame}
+
 \subsection{References}

### 2019-02-23 pkg/colorspace/paper/Toolbox/article.Rmd
started pasting together the vignettes to a JSS article -> already included various tweaks for LaTeX/PDF output
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@453 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

---
author:
  - name: Achim Zeileis
    # primary affiliation
    affiliation: Universität Innsbruck
    # full address
    address: |
      | Department of Statistics
      | Faculty of Economics and Statistics
      | Universitätsstr. 15
      | 6020 Innsbruck, Austria
    email: \email{Achim.Zeileis@R-project.org}
    url: https://eeecon.uibk.ac.at/~zeileis/
  - name: Jason C. Fisher
    affiliation: U.S. Geological Survey
  - name: Kurt Hornik
    affiliation: |
      | WU Wirtschafts-
      | universität Wien
  - name: Ross Ihaka
    affiliation: University of Auckland \AND
  - name: Claire D. McWhite
    affiliation: |
      | The University of
      | Texas at Austin
  - name: Paul Murrell
    affiliation: University of Auckland
  - name: Reto Stauffer
    affiliation: Universität Innsbruck
  - name: Claus O. Wilke
    affiliation: |
      | The University of
      | Texas at Austin
title:
  # title in title case
  formatted: "\\pkg{colorspace}: A Toolbox for Manipulating and Assessing Colors and Palettes"
  # without LaTeX markup (if any)
  plain:     "colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes"
  # with LaTeX markup (if any), used as running title
  short:     "\\pkg{colorspace}: Manipulating and Assessing Colors and Palettes"
abstract: >
  The \proglang{R} package \pkg{colorspace} provides a
  flexible toolbox for selecting individual colors or color palettes, manipulating
  these colors, and employing them in statistical graphics and data
  visualizations. In particular, the package provides a broad range of color
  palettes based on the HCL (Hue-Chroma-Luminance) color space. The three HCL
  dimensions have been shown to match those of the human visual system very well,
  thus facilitating intuitive selection of color palettes through trajectories in
  this space.

  Namely, general strategies for three types of palettes are provided:
  (1)\ Qualitative for coding categorical information, i.e., where no particular
  ordering of categories is available. (2)\ Sequential for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa). (3)\ Diverging for coding ordered/numeric
  information around a central neutral value, i.e., where colors diverge from
  neutral to two extremes.

  To aid selection and application of these palettes the package provides scales
  for use with \pkg{ggplot2}; \pkg{shiny} (and \pkg{tcltk}) apps for interactive
  exploration; visualizations of palette properties;
  accompanying manipulation utilities (like desaturation and lighten/darken), and
  emulation of color vision deficiencies.
keywords:
  # at least one keyword, comma-separated and in sentence case
  formatted: [color, palette, HCL, RGB, hue, color vision deficiency, "\\proglang{R}"]
  plain:     [color, palette, HCL, RGB, hue, color vision deficiency, R]
preamble: >
  \usepackage{thumbpdf,lmodern}
  \usepackage{booktabs,amsmath,longtable,supertabular,framed}
  \newcommand{\tex}[1]{#1}
output: rticles::jss_article
bibliography: ../../vignettes/color.bib
---

\tex{\vspace*{-0.4cm}}

# Introduction {#sec:intro}

The \pkg{colorspace} package provides a broad toolbox for selecting individual
colors or color palettes, manipulating these colors, and employing
them in various kinds of visualizations.

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
library("ggplot2")
theme_set(theme_minimal())
```

At the core of the package there are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in \proglang{R}). For completeness further standard color models are included as well in the package:
`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
`HSV()`.

The HCL space (= polar coordinates in CIELUV) is particularly useful for
specifying individual colors and color palettes as its three axes match those
of the human visual system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.

```{r hcl-properties, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space: Hue (= type of color, dominant wavelength), chroma (= colorfulness), luminance (= brightness). In each line the color coordinates are varied along the indicated axis, keeping the other two coordinates fixed."}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

The \pkg{colorspace} package provides three types of palettes based on the HCL model:

* _Qualitative:_ Designed for coding categorical information, i.e.,
  where no particular ordering of categories is available and every color
  should receive the same perceptual weight. Function: `qualitative_hcl()`.
* _Sequential:_ Designed for coding ordered/numeric information, i.e.,
  where colors go from high to low (or vice versa). Function: `sequential_hcl()`.
* _Diverging:_ Designed for coding ordered/numeric information around a central
  neutral value, i.e., where colors diverge from neutral to two extremes.
  Function: `diverging_hcl()`.

To aid choice and application of these palettes there are: scales for use
with \pkg{ggplot2} [@color:ggplot2]; \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration;
visualizations of palette properties; accompanying manipulation utilities
(like desaturation, lighten/darken, and emulation of color vision deficiencies).

# Getting started {#sec:start}

## Installation

The stable release version of \pkg{colorspace} is hosted on the Comprehensive \proglang{R} Archive Network
(CRAN) at <https://CRAN.R-project.org/package=colorspace> and the development version of \pkg{colorspace}
is hosted on \proglang{R}-Forge.

For \proglang{Python} users a beta re-implementation of the full \pkg{colorspace} package in
\proglang{Python}\ 2/\proglang{Python}\ 3 is also available, see <https://github.com/retostauffer/python-colorspace>.
In the manuscript we focus on the more mature \proglang{R} implementation but
replication materials for most examples are not only provided for \proglang{R} but
also for \proglang{Python}.

## Choosing HCL-based color palettes

The \pkg{colorspace} package ships with a wide range of predefined color palettes,
specified through suitable trajectories in the HCL (hue-chroma-luminance) color space.
A quick overview can be gained easily with `hcl_palettes()` (see Figure\ \ref{fig:hcl-palettes}):

```{r hcl-palettes, message = FALSE, warning = FALSE, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
library("colorspace")
hcl_palettes(plot = TRUE)
```

Using the names from the plot above and a desired number of colors in the palette
a suitable color vector can be easily computed, e.g.,

```{r qualitative-hcl-4}
q4 <- qualitative_hcl(4, palette = "Dark 3")
q4
```

The functions `sequential_hcl()`, and `diverging_hcl()` work analogously. Additionally,
their hue/chroma/luminance parameters can be modified, thus allowing to easily customize
each palette. Moreover, the `choose_palette()`/`hclwizard()` app provide convenient user
interfaces to do the customization interactively. Finally, even more flexible diverging
HCL palettes are provided by `divergingx_hcl()`.

## Usage with base graphics

The color vectors returned by the HCL palette functions can usually be passed directly
to most base graphics function, typically through the `col` argument. Here, the `q4`
vector created above is used in a time series display (see Figure \ref{fig:eustockmarkets-plot}):

```{r eustockmarkets, eval = FALSE}
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
```

```{r eustockmarkets-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
q4 <- qualitative_hcl(4)
par(mar = c(5, 4, 1, 1))
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
```

As another example for a sequential palette a spine plot is created below,
displaying the proportion of Titanic passengers that survived per class.
The `Purples 3` palette is used which is quite similar to the **ColorBrewer.org**
palette `Purples`. Here, only two colors are employed, yielding a dark purple
and light gray.

```{r titanic, eval = FALSE}
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
spineplot(ttnc, col = sequential_hcl(2, palette = "Purples 3"))
```

```{r titanic-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
par(mar = c(5, 4, 1, 1))
spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
```

## Usage with ggplot2

To plug the HCL color palettes into \pkg{ggplot2} graphics suitable discrete and/or
continuous \pkg{gglot2} color scales are provided. The scales are called via the scheme
`scale_<aesthetic>_<datatype>_<colorscale>()`, where `<aesthetic>` is the name
of the aesthetic (`fill`, `color`, `colour`), `<datatype>` is the type of the
variable plotted (`discrete` or `continuous`) and `<colorscale>` sets the type
of the color scale used (`qualitative`, `sequential`, `diverging`,
`divergingx`).

To illustrate their usage two simple examples are shown using the qualitative `Dark 3`
and sequential `Purples 3` palettes that were also employed above. First, semi-transparent
shaded densities of the sepal length from the iris data are shown, grouped by species.

```{r iris-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
```

The sequential palette is used to code the cut levels in a scatter of price by carat
in the diamonds data (or rather a small subsample thereof). The scale function first
generates six colors but then drops the first color because the light gray is too light
in this display. (Alternatively, the chroma and luminance parameters could also be tweaked.)

```{r diamonds-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

## Palette visualization and assessment

The \pkg{colorspace} package also provides a number of functions that aid visualization and
assessment of its palettes.

* `demoplot()` can display a palette (with arbitrary number of colors) in a range of
  typical and somewhat simplified statistical graphics.
* `hclplot()` converts the colors of a palette to the corresponding hue/chroma/luminance
  coordinates and displays them in HCL space with one dimension collapsed. The collapsed
  dimension is the luminance for qualitative palettes and the hue for sequential/diverging palettes.
* `specplot()` also converts the colors to hue/chroma/luminance coordinates but draws
  the resulting spectrum in a line plot.

For the qualitative `Dark 3` palette from above the following plots can be obtained.

```{r visualiation-qualitative, eval = FALSE}
demoplot(q4, "bar")
hclplot(q4)
specplot(q4, type = "o")
```

```{r allplots-qualitative, echo = FALSE, fig.height = 4.5, fig.width = 14, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
allplots <- function(palette, ...) {
  layout(cbind(1, 2, 3:4), heights = c(2, 10))
  par(oma = c(2, 5, 2, 3), mar = rep(0.5, 4))
  demoplot(palette, ...)
  hclplot(palette)
  par(xaxt = "n", yaxt = "n", mar = c(0.2, 3, 0.2, 0), cex = 1)
  image(matrix(seq_along(palette), ncol = 1L), col = palette)
  par(yaxt = "s")
  specplot(palette, type = "o", palette = FALSE, oma = FALSE, mar = c(0.2, 3, 0.2, 0))
}
allplots(q4, "bar")
```

A bar plot would be another typical application for a qualitative palette (instead of the
time series and density plot used above). However, a lighter and less colorful palette
might be preferable in this situation (e.g., `Pastel 1` or `Set 3`).

The other two displays show that luminance is (almost) constant in the palette while
the hue changes linearly along the color "wheel". Ideally, chroma would have also been
constant to completely balance the colors. However, at this luminance the maximum
chroma differs across hues so that the palette is fixed up to use less chroma for
the yellow and green elements.

Subsequently, the same assessment is carried out for the sequential `Purples 3` palette
as employed above.

```{r visualization-sequential, eval = FALSE}
s9 <- sequential_hcl(9, "Purples 3")
demoplot(s9, "heatmap")
hclplot(s9)
specplot(s9, type = "o")
```

```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
s9 <- sequential_hcl(9, "Purples 3")
allplots(s9, "heatmap")
```

Here, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
application for a sequential palette. The elevation of the volcano is brought out clearly,
focusing with the dark colors on the higher elevations.

The other two displays show that hue is constant in the palette while luminance and chroma vary.
Luminance increases monotonically from dark to light (as required for a proper sequential palette).
Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
(compared to a monotonic chroma trajectory).

# Color spaces: S4 classes and utilities  {#sec:color_spaces}

At the core of the \pkg{colorspace} package there are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other, see @color:Ihaka:2003. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in R). For completeness further standard color models are included as well in the package.
The connections are illustrated in the following graph:

```{r color-spaces, fig.width = 9.5, fig.height = 4.5, echo = FALSE, out.width = "0.9\\textwidth", fig.cap = "Foo."}
par(mar = rep(0, 4))

d <- data.frame(
  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
  stringsAsFactors = FALSE
)

plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
lines(y ~ x, data = d[-c(1, 3, 7, 8),])
lines(y ~ x, data = d[c(1, 3, 5),])
lines(y ~ x, data = d[c(6, 7, 9),])
lines(y ~ x, data = d[c(6, 8, 9),])
points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)

text(d$x, d$y, d$lab)
text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
```

Color models that are (or try to be) perceptually-based are displayed with circles,
other color models with rectangles. The corresponding classes and eponymous class
constructors in \pkg{colorspace} are:

* `RGB()` for the classic Red-Green-Blue color model that mixes three primary colors
  with different intensities to obtain a spectrum of colors. The advantage of this
  color model is (or was) that it corresponded to how computer and TV screens generated
  colors, hence it was widely adopted and still is the basis for color specifications
  in many systems. For example, the hex color codes are employed in HTML but also in R.
  However, the RGB model also has some important drawbacks: It does not take into
  account the output device properties, it is not perceptually uniform (a unit step
  within RGB does not produce a constant perceptual change in color), and it is
  unintuitive for humans to specify colors (say brown or pink) in this space.
  [See @color:Wiki+RGB.]
* `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
  correction. Therefore, the gamma-corrected standard RGB (sRGB), as opposed to the
  linearized RGB above, is a good model for specifying colors in software and for
  hardware. But it is still unintuitive for humans to work directly with this.
  Therefore, sRGB is a good place to end up in a color space manipulation but
  it is not a good place to start. [See @color:Wiki+sRGB.]
* `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
  perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
  (colorfulness), and _value_ (brightness, i.e., light vs. dark). Unfortunately,
  the three axes in the HSV model are confounded so that, e.g., brightness changes
  dramaticaly with hue.
  [See @color:Wiki+HSV.]
* `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
  to capture the perceptual axes. It does a somewhat better job but the dimensions
  are still strongly confounded.
  [See @color:Wiki+HSV.]
* `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
  based on experiments with human subjects. It provides a unique triplet of
  XYZ values, coding the standard observer's perception of the color. It is
  device-independent but it is not perceptually uniform and the XYZ coordinates
  have no intuitive meaning.
  [See @color:Wiki+CIEXYZ.]
* `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
  color spaces where the former is typically preferred for emissive technologies
  (such as screens and monitors) whereas the latter is usually preferred when
  working with dyes and pigments. However, the three axes of these two spaces
  still do not correspond to human perceptual axes.
  [See @color:Wiki+CIELUV; @color:Wiki+CIELAB.]
* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
  and AB plane, respectively. Specifically, the polar coordinates of the LUV
  model are known as the HCL (Hue-Chroma-Luminance) model. These capture the
  human perceptual axes very well without confounding effects as in the HSV or
  HLS approaches. (More details follow below.)

All the S4 classes listed above inherit from a virtual class `color` which are
internally always represented by matrices with three columns (corresponding to
the different three dimensions).

Note that since the inception of the color space conversion tools within \pkg{colorspace}
by @color:Ihaka:2003 other R tools for this purpose became available. Notably,
there is `grDevices::convertColor()` in base R [@color:R] as well as
`farver::convert_colour()` [@color:farver] based on a C++ library. For many basic color
conversion purposes the \pkg{colorspace} package and these alternatives are equally suitable.
Some details differ, though, and the color conversion tools in \pkg{colorspace} are
well-embedded into the remaining functionality of the package and make it self-contained.

## Utilities

For working with these color spaces various utilities are available:

* `as()` method: Conversions of a `color` object to the various color spaces,
  e.g., `as(x, "sRGB")`.
* `coords()`: Extract the three-dimensional coordinates pertaining to the current
  color class.
* `hex()`: Convert a color object to sRGB and code in a hex string that can be
  used within R plotting functions.
* `hex2RGB()`: Convert a given hex color string to an sRGB color object which
  can also be coerced to other color spaces.
* `readRGB()` and `readhex()` can read text files into color objects, either
  from RGB coordinates or hex color strings.
* `writehex()`: Writes hex color strings to a text file.
* `whitepoint()`: Query and change the white point employed in conversions
  from CIE XYZ to RGB. Defaults to D65.

## Illustration of basic colorspace functionality

As an example a vector of colors `x` can be specified in the HCL (or polar LUV) model:

```{r polarLUV}
(x <- polarLUV(L = 70, C = 50, H = c(0, 120, 240)))
```

These are pastel red (hue = 0), green (hue = 120), and blue (hue = 240) with moderate chroma and luminance.
For display in other systems an sRGB representation might be needed:

```{r sRGB}
(y <- as(x, "sRGB"))
```

With `coords(x)` or `coords(y)` the displayed coordinates can also be extracted
as numeric matrices. And from sRGB we can also coerce to HSV for example:

```{r HSV}
as(y, "HSV")
```

For display in many systems (including R itself) hex color codes based on the
sRGB coordinates can be created:

```{r hex}
hex(x)
```

# HCL-based color palettes {#sec:hcl_palettes}

As motivated in the previous article ([Color Spaces: S4 Classes and Utilities](color_spaces.html)),
the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
of the human visual system very well. Therefore, the \pkg{colorspace} package provides
three types of palettes based on the HCL model:

* _Qualitative:_ Designed for coding categorical information, i.e.,
  where no particular ordering of categories is available and every color
  should receive the same perceptual weight.
* _Sequential:_ Designed for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa).
* _Diverging:_ Designed for coding ordered/numeric information around a central
  neutral value, i.e., where colors diverge from neutral to two extremes.

The corresponding functions are `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. Their construction principles are exemplified in the
following color swatches and explained in more detail below. The desaturated
palettes bring out clearly that luminance differences (= light-dark contrasts),
or the lack thereof, are crucial for coding the different kinds of palettes.

```{r hcl-palettes-principles, echo = FALSE, fig.width = 12, fig.height = 1.6, out.width = "\\textwidth", fig.cap = "Foo."}
swatchplot(
  "Qualitative (Set 2)"     = rbind("Color" = qualitative_hcl(5, "Set 2"),     "Desaturated" = desaturate(qualitative_hcl(5, "Set 2"))),
  "Sequential (Blues 3)"    = rbind("Color" = sequential_hcl(7, "Blues 3"),    "Desaturated" = desaturate(sequential_hcl(7, "Blues 3"))),
  "Diverging (Green-Brown)" = rbind("Color" = diverging_hcl(7, "Green-Brown"), "Desaturated" = desaturate(diverging_hcl(7, "Green-Brown"))),
  nrow = 3, line = 7)
```

More details about the construction of such palettes is provided in the following
while the article on [Palette Visualization and Assessment](palette_visualization.html)
introduces further tools to better understand the properties of color palettes.

To facilitate obtaining good sets of colors, HCL parameter combinations that yield
useful palettes are accessible by name. These can be listed using the function
`hcl_palettes()`:

```{r hcl-palettes-print}
hcl_palettes()
```

To inspect the HCL parameter combinations for a specific palette simply include the `name`
where upper- vs. lower-case, spaces, etc. are ignored for matching the label, i.e.,
`"set2"` matches `"Set 2"`:

```{r hcl-palette}
hcl_palettes(palette = "set2")
```

To compute the actual color hex codes (representing sRGB coordinates), the functions
`qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`, respectively, can be used.
Either all parameters can be specified "by hand" through the HCL parameters, an entire
palette can be specified "by name", or the name-based specification can be modified by a
few HCL parameters. In case of the HCL parameters, either a vector-based specification
such as `h = c(0, 270)` or individual parameters `h1 = 0` and `h2 = 270`
can be used.

The first three of the following commands lead to equivalent output. The fourth command
yields a modified set of colors (lighter due to a luminance of 80 instead of 70).

```{r qualitative-hcl}
qualitative_hcl(4, h = c(0, 270), c = 60, l = 70)
qualitative_hcl(4, h1 = 0, h2 = 270, c1 = 60, l1 = 70)
qualitative_hcl(4, palette = "set2")
qualitative_hcl(4, palette = "set2", l = 80)
```

## Qualitative palettes

`qualitative_hcl()` distinguishes the underlying categories by a sequence of
hues while keeping both chroma and luminance constant to give each color in the
resulting palette the same perceptual weight. Thus, `h` should be a pair of
hues (or equivalently `h1` and `h2` can be used) with the starting and
ending hue of the palette. Then, an equidistant sequence between these hues is
employed, by default spanning the full color wheel (i.e, the full 360 degrees).
Chroma `c` (or equivalently `c1`) and luminance `l` (or equivalently
`l1`) are constants.

In the following graphic the available named palettes are shown. The first five
palettes are close to the _ColorBrewer.org_ palettes of the same name [@color:Harrower+Brewer:2003].
They employ different levels of chroma and luminance and, by default, span the full
hue range. The remaining four palettes are taken from @color:Ihaka:2003. They are
based on the same chroma (= 50) and luminance (= 70) but the hue is restricted to
different intervals.

```{r hcl-palettes-qualitative, fig.width = 5.5, fig.height = 5.0, out.width = "0.6\\textwidth", fig.cap = "Foo."}
hcl_palettes("qualitative", plot = TRUE)
```

When qualtitative palettes are employed for shading areas in statistical displays
(e.g., in bar plots, pie charts, or regions in maps), then it is typically
less distracting to use lighter colors (with moderate chroma and high luminance)
such as "Pastel 1" or "Set 3". When coloring points or lines then more flashy
colors (with high chroma) are often required: On a white background a moderate
luminance as in "Dark 2" or "Dark 3" usually works better while on a black/dark
background the luminance should be higher as in "Set 2" for example.

## Sequential palettes (single-hue)

`sequential_hcl()` codes the underlying numeric values by a monotonic sequence
of increasing (or decreasing) luminance. Thus, the `l` argument should provide
a vector of length 2 with starting and ending luminance (equivalently, `l1` and
`l2` can be used). Without chroma (i.e., `c = 0`), this simply corresponds
to a grayscale palette like `gray.colors()`, see "Grays" and "Light Grays" below.

For adding chroma, a simple strategy would be to pick a single hue (via `h` or `h1`)
and then decrease chroma from some value (`c` or `c1`) to zero (i.e., gray) along with
increasing luminance. For bringing out the extremes (a dark high-chroma color vs.
a light gray) this is already very effective, see "Blues 2", "Purples 2", "Reds 2",
and "Greens 2".

For distinguishing also colors in the middle two strategies can be employed:
(a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
hue `h1` and ending hue `h2`). More details are provided in the next section.
(b) Instead of a decreasing chroma a triangular chroma trajectory can be employed
from `c1` over `cmax` to `c2` (or equivalently a vector `c` of
length 3). This yields high-chroma colors in the middle of the palette that are
more easily distinguished from the dark and light extremes. See "Blues 3", "Purples 3",
"Reds 3", and "Greens 3" below.

Instead of employing linear trajectories in the chroma or luminance coordinates
some palettes employ a power transformation of the chroma and/or luminance
trajectory. Either a vector `power` of length 2 or separate `p1` (for chroma)
and `p2` (for luminance) can be specified. If missing, the latter defaults to the
former.

```{r hcl-palettes-sequentials, fig.width = 6.5, fig.height = 6.0, out.width = "0.7\\textwidth", fig.cap = "Foo."}
hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE)
```

All except the last palette are inspired by the _ColorBrewer.org_ palettes with
the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
They are intended for a white/light background. The last palette is taken from
the scientific color maps of @color:Crameri:2018 and is intended for a black/dark
background and hence the order is reversed starting from a light blue (not a light gray).

To distinguish many colors in a sequential palette it is important to have a strong
contrast on the luminance axis, possibly enhanced by an accompanying pronounced
variation in chroma. When only a few colors are needed (e.g., for coding an ordinal
categorical variable with few levels) then a lower luminance contrast may suffice.

## Sequential palettes (multi-hue)

To not only bring out extreme colors in a sequential palette but also better distinguish
middle colors it is a common strategy to employ a sequence of hues. Thus, the basis
of such a palette is still a monotonic luminance sequence as above (combined with a
monotonic or triangular chroma sequence). But, instead of a single hue, an interval
of hues in `h` (or beginning hue `h1` and ending hue `h2`) is specified.

`sequential_hcl()` allows to combine variations in hue (`h` and `h1`/`h2`, respectively),
chroma (`c` and `c1`/`c2`/`cmax`, respectively),
luminance (`l` and `l1`/`l2`, respectively), and
power transformations for the chroma and luminance trajectories
(`power` and `p1`/`p2`, respectively). This yields a broad variety of sequential
palettes, including many that closely match other well-known color palettes.
The plot below shows all the named multi-hue sequential palettes in \pkg{colorspace}:

* "Purple-Blue" to "Terrain 2" are various palettes created during the development
  of \pkg{colorspace}, e.g., by @color:Zeileis+Hornik+Murrell:2009 or @color:Stauffer+Mayr+Dabernig:2015
  among others.
* "Viridis" to "Inferno" closely match the palettes that @color:Smith+VanDerWalt:2015
  developed for _matplotlib_ and that gained popularity recently.
* "Dark Mint" to "BrwnYl" closely match palettes provided in _CARTO_ [@color:CARTO].
* "YlOrRd" to "Blues" closely match _ColorBrewer.org_ palettes [@color:Harrower+Brewer:2003].
* "Lajolla" and "Turku" closely match the scientific color maps of the same name by
  @color:Crameri:2018 and are intended for a black/dark background.

```{r hcl-palettes-sequentialm, fig.width = 8.5, fig.height = 9.0, out.width = "\\textwidth", fig.cap = "Foo."}
hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
```

Note that the palettes differ substantially in the amount of chroma and luminance
contrasts, respectively. For example, many palettes go from a dark high-chroma color
to a neutral low-chroma color (e.g., "Reds", "Purples", "Greens", "Blues") or even
light gray (e.g., "Purple-Blue"). But some palettes also employ relatively high chroma
throughout the palette (e.g., the viridis and many _CARTO_ palettes). To emphasize
the extremes the former strategy is typically more suitable while the latter works
better if all values along the sequence should receive some more perceptual weight.

## Diverging palettes

`diverging_hcl()` codes the underlying numeric values by a triangular luminance
sequence with different hues in the left and in the right "arm" of the palette. Thus,
it can be seen as a combination of two sequential palettes with some restrictions:
(a) a single hue is used for each arm of the palette, (b) chroma and luminance trajectory
are balanced between the two arms, (c) the neutral central value has zero chroma.
To specify such a palette a vector of two hues `h` (or equivalently `h1`
and `h2`), either a single chroma value `c` (or `c1`) or a vector
of two chroma values `c` (or `c1` and `cmax`), a vector of two
luminances `l` (or `l1` and `l2`), and power parameter(s) `power`
(or `p1` and `p2`) are used. For more flexible diverging palettes without
the restrictrictions above (and consequently more parameters) see the `divergingx_hcl()`
palettes introduced below.

The plot below shows all such diverging palettes that have been named in \pkg{colorspace}:

* "Blue-Red" to "Cyan-Magenta" have been developed for \pkg{colorspace} starting from
  @color:Zeileis+Hornik+Murrell:2009, taking inspiration from various other palettes,
  including more balanced and simplified versions of several _ColorBrewer.org_
  palettes [@color:Harrower+Brewer:2003].
* "Tropic" closely matches the palette of the same name from _CARTO_ [@color:CARTO].
* "Broc" to "Vik" and "Berlin" to "Tofino" closely match the scientific color maps of the same name by
  @color:Crameri:2018, where the first three are intended for a white/light background and
  the other three for a black/dark background.

```{r hcl-palettes-diverging, fig.width = 6.5, fig.height = 7.5, out.width = "0.7\\textwidth", fig.cap = "Foo."}
hcl_palettes("diverging", n = 7, plot = TRUE)
```

When choosing a particular palette for a display similar considerations apply as
for the sequential palettes. Thus, large luminance differences are important when
many colors are used while smaller luminance contrasts may suffice for palettes with
fewer colors etc.

## Construction details

The three different types of palettes (_qualitative_, _sequential_, and _diverging_)
are all constructed by combining three different types of trajectories (_constant_, _linear_, _triangular_)
for the three different coordinates (hue _H_, chroma _C_, luminance _L_):

\begin{table}[t!]
\centering
\begin{tabular}{llll}
\hline
Type          & H                                 & C                          & L                \\ \hline
Qualitative   & Linear  		          & Constant	               & Constant	  \\
Sequential    & Constant (= single-hue) \emph{or} & Linear (+ power) \emph{or} & Linear (+ power) \\
	      & Linear (= multi-hue)	          & Triangular (+ power)       & 		  \\
Diverging     & Constant ($2\times$)	          & Linear (+ power) \emph{or} & Linear (+ power) \\
	      & 			          & Triangular (+ power)       & 		  \\
\hline
\end{tabular}
\caption{\label{tab:hcl} Foo, see Equations~\ref{eq:constant}--\ref{eq:triangular}.}
\end{table}

As pointed out initially in this article, the luminance is probably the most important
property for defining the type of palette. It is constant for qualitative palettes, monotonic
for sequential palettes (linear or a power transformation), and uses two monotonic trajectories
(linear or a power transformation) diverging from the same neutral value.

The trajectories for the hue are also rather intuitive and straightforward for the three
different types of palettes. However, the chroma trajectories are probably most complicated
and least obvious from the examples above. Hence, the exact mathematical equations underlying
the chroma trajectories are given in the following (i.e., using the parameters `c1`, `c2`, `cmax`,
and `p1`, respectively). Analogous equations apply for the other two coordinates, though.

The trajectories are functions of the _intensity_ $i \in [0, 1]$ where $1$ corresponds to
the full intensity:

\begin{align}
\text{Constant: }   & c_1 \label{eq:constant} \\[0.2cm]
\text{Linear: }     & c_2 - (c_2 - c_1) \cdot i \label{eq:linear} \\[0.2cm]
\text{Triangular: } & \left\{ \begin{array}{lcl}
                      c_2 - (c_2 - c_{\max}) \cdot \frac{i}{j}            & \text{if } i & \le j \\
                      c_{\max} - (c_{\max} - c_1) \cdot \frac{i - j}{1 - j} &              & >   j
                     \end{array} \right. \label{eq:triangular}
\end{align}

where $j$ is the intensity at which $c_{\max}$ is assumed. It is constructed such that
the slope to the left is minus the slope to the right of $j$:

$$
j = \left(1 + \frac{|c_{\max} - c_1|}{|c_{\max} - c_2|} \right)^{-1}
$$

Instead of using a linear intensity $i$ going from $1$ to $0$, one can replace $i$ with $i^{p_1}$
in the equations above. This then leads to power-transformed curves that add or remove chroma
more slowly or more quickly depending on whether the power parameter $p_1$ is $< 1$ or $> 1$.

The three types of trajectories are also depicted below. Note that full intensity $i = 1$ is on
the left and zero intensity $i = 0$ on the right of each panel.

```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, out.width = "\\textwidth", fig.cap = "Foo."}
mylab <- function(x, y, text, ratio = 1, col = "black", ...) {
  rect(x - 0.045/ratio, y - 3.5, x + 0.04/ratio, y + 4, border = col, ...)
  text(x, y, text, col = col, ...)
}

par(mfrow = c(1, 3))
i <- 0:100/100
c1 <- 80
plot(i, rep(c1, length(i)), xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Constant",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
mylab(0.06, c1 + 6, "c1")

c2 <- 10
plot(1 - i, c2 - (c2 - c1) * i, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Linear",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, c2 - (c2 - c1) * i^1.6, lty = 2, lwd = 2)
mylab(0.06, c1 + 5, "c1")
mylab(0.94, c2 - 5, "c2")
mylab(0.68, 40, "p1", col = gray(0.35))
text(0.78, 40, "= 1", col = gray(0.35))
mylab(0.18, 40, "p1", col = gray(0.35))
text(0.30, 40, "= 1.6", col = gray(0.35))

c1 <- 60
cmax <- 80
c2 <- 10
maxat <- 1/(1 + abs(cmax - c1) / abs(cmax - c2))
p1 <- 1
ci1 <- ifelse(i^p1 <= maxat, c2 - (c2 - cmax) * (i^p1)/maxat, cmax - (cmax - c1) * ((i^p1 - maxat)/(1 - maxat)))
p2 <- 1.6
ci2 <- ifelse(i^p2 <= maxat, c2 - (c2 - cmax) * (i^p2)/maxat, cmax - (cmax - c1) * ((i^p2 - maxat)/(1 - maxat)))

plot(1 - i, ci1, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Triangular",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, ci2, lwd = 2, lty = 2)
mylab(0.06, c1 - 5, "c1")
mylab(1 - maxat + 0.1, cmax + 4, "cmax", ratio = 0.5)
mylab(0.94, c2 - 5, "c2")
mylab(0.77, 40, "p1", col = gray(0.35))
text(0.87, 40, "= 1", col = gray(0.35))
mylab(0.27, 40, "p1", col = gray(0.35))
text(0.39, 40, "= 1.6", col = gray(0.35))
```

The concrete parameters in the plot above are:

* Constant: `c1 = 80`.
* Linear: `c1 = 80`, `c2 = 10`, `p1 = 1` (solid) vs. `p1 = 1.6` (dashed).
* Triangular: `c1 = 60`, `cmax = 80`, `c2 = 10`, `p1 = 1` (solid) vs. `p1 = 1.6` (dashed).

Further discussion of these trajectories and how they can be visualized and assessed
for a given color palette is provided in the article:
[Palette Visualization and Assessment](palette_visualization.html).

## Registering your own palettes

The `hcl_palettes()` already come with a wide range of pre-defined palettes to which
customizations can be easily added. However, it might also be convenient to register
a custom palette so that it can subsequently be re-used with a new dedicated name.
This is supported by adding a `register = "..."` argument once to a call to
`qualitative_hcl()`, `sequential_hcl()`, or `diverging_hcl()`:

```{r register1}
qualitative_hcl(3, palette = "set2", l = 80, register = "myset")
```

The new palette is then included in `hcl_palettes()`:

```{r register2}
hcl_palettes("Qualitative")
```

And can be used subsequently in `qualitative_hcl()` as well as the
qualitative [ggplot2 color scales](ggplot2_color_scales.html), e.g.,

```{r register3}
qualitative_hcl(4, palette = "myset")
```

Remarks:

* The number of colors in the palette that was used during registration is not
  actually stored and can be modified subsequently. The same holds for arguments
  `alpha` and `rev`.
* When registering a new palette with an old name that was already available
  previously, the old palette gets overwritten. We recommend not to overwrite
  the palettes that are pre-defined in the package (albeit it is technically possible).
* The registration of a palette is only stored for the current session. When
  R is restarted and/or the `colorspace` package reloaded, only the pre-defined
  palettes from the package are available. Thus, to make a palette permanently
  available a registration R code like
  `colorspace::qualitative_hcl(3, palette = "set2", l = 80, register = "myset")`
  can be placed in your `.Rprofile` or similar startup scripts.

## Flexible diverging palettes

The `divergingx_hcl()` function provides more flexible diverging palettes by
simply calling `sequential_hcl()` twice with prespecified sets of hue, chroma,
and luminance parameters. Thus, it does not pose any restrictions that
the two "arms" of the palette need to be balanced and also allows to go
through a non-gray neutral color (typically light yellow).
Consequently, the chroma/luminance paths can be rather unbalanced.

The plot below shows all such flexible diverging palettes that have been named in \pkg{colorspace}:

* "ArmyRose" to "Tropic" closely match the palettes of the same name from _CARTO_ [@color:CARTO].
* "PuOr" to "Spectral" closely match the palettes of the same name from _ColorBrewer.org_ [@color:Harrower+Brewer:2003].

```{r divergingx-palettes, fig.width = 6.5, fig.height = 8.0, out.width = "0.7\\textwidth", fig.cap = "Foo."}
divergingx_palettes(n = 7, plot = TRUE)
```

Typically, the more restricted `diverging_hcl()` palettes should be preferred because they are
more balanced. However, by being able to go through light yellow as the neutral
color warmer diverging palettes are available.

## HCL (and HSV) color palettes corresponding to base R palettes

To facilitate switching from base R palette functions to the HCL-based palettes
above, \pkg{colorspace} provides a few convenience interfaces:

* `rainbow_hcl()`: Convenience interface to `qualitative_hcl()` for a HCL-based "rainbow"
  palette to replace the (in)famous `rainbow()` palette.
* `heat_hcl()`: Convenience interface to `sequential_hcl()` with default parameters
  chosen to generate more balanced heat colors than the basic `heat.colors()` function.
* `terrain_hcl()`: Convenience interface to `sequential_hcl()` with default parameters
  chosen to generate more balanced terrain colors than the basic `terrain.colors()` function.
* `diverging_hsv()`: Diverging palettes generated in HSV space rather than HCL space as
  in `diverging_hcl()`. This is provided for didactic purposes to contrast the more balanced
  HCL palettes with the more flashy and unbalanced HSV palettes.

# HCL-based color scales for ggplot2 {#sec:ggplot2_color_scales}

All [HCL-based color palettes](hcl_palettes.html) in the \pkg{colorspace} package [@color:colorspace]
are also provided as discrete and/or continuous color scales for the use with the _ggplot2_ package
[@color:ggplot2; @color:ggplot2pkg].

The scales are called via the scheme

    scale_<aesthetic>_<datatype>_<colorscale>()

where

- `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`).
- `<datatype>` is the type of the variable plotted (`discrete` or `continuous`).
- `<colorscale>` sets the type of the color scale used (`qualitative`, `sequential`, `diverging`, `divergingx`).

A few examples of these scales are illustrated in the following sections.

## Using the scales in ggplot2

A discrete qualitative scale applied to a fill aesthetic corresponds to the function `scale_fill_discrete_qualitative()`:

```{r density-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative()
```

Similarly, a color aesthetic for a discrete qualitative scale corresponds to the function `scale_color_discrete_qualitative()`:

```{r point-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point() +
  scale_color_discrete_qualitative(palette = "Set 2")
```

A continuous sequential scale applied to a color aesthetic corresponds to the function `scale_color_continuous_sequential()`:

```{r jitter-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Heat")
```

A continuous sequential scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_sequential()`:

```{r raster-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
df <- data.frame(height = c(volcano), x = c(row(volcano)), y = c(col(volcano)))
ggplot(df, aes(x, y, fill = height)) + 
  geom_raster() + 
  coord_fixed(expand = FALSE) +
  scale_fill_continuous_sequential(palette = "Blues")
```

A continuous diverging scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_diverging()`:

```{r tile-diverging, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
cm <- cor(mtcars)
df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
levels(df$var1) <- levels(df$var2) <- names(mtcars)
ggplot(df, aes(var1, var2, fill = cor)) + 
  geom_tile() + 
  coord_fixed() +
  ylab("variable") +
  scale_x_discrete(position = "top", name = "variable") +
  scale_fill_continuous_diverging("Blue-Red 3")
```

## Customizing the scales

All scale functions accept a `palette` argument which allows you to pick a specific color palette out of a selection of different options. All available palettes are listed at the end of this document. For example, we could use the "Harmonic" palette when we need a qualitative color scale:

```{r density-qualitative2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Harmonic")
```

The color palettes are calculated on the fly depending on the number of different colors needed. But sometimes, it may be desireable to pick specific colors out of a larger set, e.g., when we are making two plots where one contains a subset of the data of the other, or when the default order of colors is not ideal. Therefore, all discrete scales provide parameters `nmax` to set the total number of colors requested and `order` (a vector of integers) to reorder the color palette.

Applied to the previous plot, we could for example do the following:

```{r density-qualitative3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Harmonic", nmax = 5, order = c(5, 1, 2))
```

The `nmax` option is also convenient to remove some colors from a scale that may not be appropriate for the plot. For example, the `scale_color_brewer()` scale that comes with _ggplot2_ tends to produce points that are too light:

```{r point-brewer, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
gg <- ggplot(dsamp, aes(carat, price, color = cut)) + geom_point()
gg + scale_color_brewer(palette = "Blues")
```

Similar problems can arise with the HCL palettes, but there we have the option of creating additional colors that we then do not use:

```{r point-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
gg + scale_color_discrete_sequential(palette = "Blues", nmax = 6, order = 2:6)
```

(We use `order = 2:6` to pick the five darkest colors and omit the lightest color.)

All [HCL-based color palettes](hcl_palettes.html) are defined via sets of hue (H), chroma (C), and luminance (L) values. For example, the qualitative scales vary hue from a start value to an end value while keeping chroma and luminance fixed. Similarly, single-hue sequential scales vary chroma and luminance while keeping the hue fixed. We can override these settings by specifying the corresponding H, C, or L values in addition to the palette name. As an example, consider the following plot:

```{r jitter-sequential2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Terrain")
```

Now assume we generally like the color scale but find it a bit too pink at the end. We can fix this issue by specifying an alternative final hue value, e.g., `h2 = 60` corresponding to yellow:

```{r jitter-sequential3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Terrain", h2 = 60)
```

The next example uses a diverging scale. First consider the plot with the unmodified "Blue-Yellow 2" palette:

```{r tile-diverging2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
cm <- cor(mtcars)
df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
levels(df$var1) <- levels(df$var2) <- names(mtcars)
gg <- ggplot(df, aes(var1, var2, fill = cor)) + 
  geom_tile() + 
  coord_fixed() +
  ylab("variable") +
  scale_x_discrete(position = "top", name = "variable")

gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2")
```

And now the same plot with some palette customizations: The ordering is reversed so that blue is used for
positive correlations and yellow for negative ones. Moreover, the power parameter `p2` for the luminance
is increased so that only correlations close to an absolute value of 1 have dark colors while intermediate
correlations have relatively light colors.

```{r tile-diverging3, fig.width = 6.5, fig.asp = .7, fig.cap = "Foo."}
gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2", rev = TRUE, p2 = 2)
```

See the reference manual for the exact set of customization parameters that are available for each scale.

The continuous scales also provide the option to limit the scale range to which data are mapped, via the parameters `begin` and `end`. As an example, assume we are using the approximation of the viridis scale provided by `scale_color_continuous_sequential()`:

```{r jitter-sequential4, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Viridis")
```

If we want to remove some of the darkest blues and some of the brightest yellows from this scale, we can write:

```{r jitter-sequential5, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Viridis", begin = 0.15, end = 0.9)
```

## Available palettes

In the following, we are visualizing all scales currently available via
pre-defined names. These visualizations are generated by the function
`hcl_palettes()` with option `plot = TRUE`.

The discrete qualitative scales are all called via `scale_*_discrete_qualitative(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Pastel 1"`. There are no continuous qualitative
scales.

The discrete sequential scales are all called via `scale_*_discrete_sequential(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Grays"`. Continuous approximations to the discrete
scales exist and can be called via `scale_*_continuous_sequential(palette = "name")`

The discrete diverging scales are all called via `scale_*_discrete_diverging(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Blue-Red"`. Continuous approximations to the discrete
scales exist and can be called via `scale_*_continuous_diverging(palette = "name")`

# Palette visualization and assessment {#sec:palette_visualization}

The \pkg{colorspace} package provides several visualization functions for depicting
one or more color palettes and their underlying properties. Color palettes
can be visualized by:

* `swatchplot()`: Color swatches.
* `specplot()`: Spectrum of HCL and/or RGB trajectories.
* `hclplot()`: Trajectories in 2-dimensional HCL space projections.
* `demoplot()`: Illustrations of typical (and simplified) statistical graphics.

## Color swatches

The function `swatchplot()` is a convenience function for displaying
collections of palettes that can be specified as lists or matrices of
hex color codes. Essentially, it is just a call to the base graphics
`rect()` function but with heuristics for choosing default labels, margins,
spacings, borders, etc. These are selected to work well for `hcl_palettes()`
and might need further tweaking in future versions.

As a first simple example, the properties of the three HCL axes are brought out by varying
only one coordinate while keeping the other two fixed:

```{r swatch-hcl, fig.width = 4, fig.height = 2.2, fig.cap = "Foo."}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

This shows the following:

- Hue: Only the hue (= type of color) changes from H = 0 (red) via 60 (yellow), etc. to 300 (purple)
  while chroma and luminance are fixed to moderate values of C = 60 and L = 65, respectively.
- Chroma: Only the chroma (= colorfulness) changes from C = 0 (gray) to 100 (colorful)
  while hue and luminance are fixed to H = 0 (red) and L = 65, respectively.
- Luminance: Only the luminance (= brightness) changes from L = 90 (light) to 25 (dark)
  while hue and chroma are fixed to H = 260 (blue) and C = 25 (low, close to gray), respectively.

As a somewhat more complex example a `swatchplot()` is created with three matrices
of sequential color palettes of blues, purples, reds, and greens. For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

- Single-hue: In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs. light gray).
- Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
- Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, out.width = "\\textwidth", fig.cap = "Foo."}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5
)
```

## HCL (and RGB) spectrum

As the properties of a palette in terms of the perceptual dimensions _hue_,
_chroma_, and _luminance_ are not always clear from looking just at color
swatches or (statistical) graphics based on these palettes, the `specplot()`
function provides an explicit display for the coordinates of the HCL trajectory associated
with a palette. This can bring out clearly various aspects, e.g.,
whether hue is constant, chroma is monotonic or triangular, and whether
luminance is approximately constant (as in many qualitative palettes),
monotonic (as in sequential palettes), or diverging.

The function first transforms a given color palette to its HCL (`polarLUV()`)
coordinates. As the hues for low-chroma colors are not (or only poorly) identified,
by default a smoothing is applied to the hues. Also, to avoid jumps from 0 to 360 or vice
versa, the hue coordinates are shifted suitably.

By default, the resulting trajectories in the HCL spectrum are visualized by a simple line plot:

- Hue is drawn in red and coordinates are indicated on the axis on the right
  with range [-360, 360].
- Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
  is used unless the palette necessitates higher chroma values.
- Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].

Additionally, a color swatch for the palette is included. Optionally, a
second spectrum for the corresponding trajectories of RGB coordinates can be included. However,
this is usually just of interest for palettes created in RGB space (or simple
transformations of RGB).

The illustrations below show how basic qualitative, sequential, and diverging
palettes are constructed in HCL space (the corresponding mathematical equations
are provided in the [construction details](hcl_palettes.html#construction-details)).
In the qualitative "Set 2" palette below, the
hue traverses the entire color "wheel" (from 0 to 360 degrees) while keeping
chroma and luminance (almost) constant (C = 60 and L = 70).

```{r spec-qualitative, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(qualitative_hcl(100, "Set 2"))
```

Note that due to the restrictions of the HCL color space, some of the green/blue
colors have a slightly smaller maximum chroma resulting in a small dip in the chroma
curve. This is fixed automatically (by default) and is hardly noticable in visualizations,
though.

The sequential "Blues 2" palette below employs a single hue (H = 260) and a monotonically
increasing luminance trajectory (from dark to light). Chroma simply decreases monotonically
along with increasing luminance.

```{r spec-sequential, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(sequential_hcl(100, "Blues 2"))
```

Finally, the diverging "Blue-Red" palette is depicted below. It simply combines a blue
and a red sequential single-hue palette (similar to the "Blues 2" palette discussed above).
Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral
value.

```{r spec-diverging, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(diverging_hcl(100, "Blue-Red"))
```

To contrast these well-balanced HCL-based palettes with a poorly-balanced palette, the
spectrum of the (in)famous RGB rainbow palette is depicted in both RGB and HCL space.

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, fig.cap = "Foo."}
specplot(rainbow(100), rgb = TRUE)
```

The RGB spectrum shows that the trajectories are quite simple in RGB space but the
HCL spectrum shows that the result is very unbalanced with shifts and kinks in both
chroma and (more importantly) luminance. This is why this palette is not suitable
for encoding underlying data in statistical graphics. See also the related discussion
of [color vision deficiency](color_vision_deficiency.html).

## Trajectories in HCL space

While the `specplot()` function above works well for bringing out the HCL coordinates
associated with a given palette, it does not bring out how the palette fits into
the HCL space. For example, it is not so clear whether high chroma values are
close to the maximum possible for a given hue. Thus, it cannot be judged so easily
how the parameters of the hue, chroma, and luminance trajectories can be
modified to obtain another palette.

Therefore, the `hclplot()` is another visualization of the HCL coordinates associated
with a palette. It does so by collapsing over one of the coordinates (either the hue H
or the luminance L) and displays a heatmap of colors combining the remaining
two dimensions. The coordinates for the given color palette are highlighted
to bring out its trajectory. In case the hue is really fixed (as in single-hue
sequential palettes) or the luminance is really fixed (as in the qualitative
palettes), collapsing is straightforward. However, when the coordinate that is
collapsed over is actually not constant in the palette, a simple bivariate linear
model is used to capture how the collapsed coordinate varies along with the
two displayed coordinates.

The function `hclplot()` has been designed to work well with the
`hcl_palettes()` in this package. While it is possible to apply
it to other color palettes as well, the results might look weird or confusing
if these palettes are constructed very differently (e.g., like the
highly-saturated base R palettes). To infer the default `type` of projection
`hclplot()` assesses the luminance trajectory and sets the default correspondingly:

- `type = "qualitative"` if luminance is approximately constant.
- `type = "sequential"` if luminance is monotonic.
- `type = "diverging"` if luminance is diverging with two monotonic "arms" in the trajectory.

Thus, for qualitative palettes - where luminance and chroma are fixed - the varying
hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance:

```{r hcl-qualitative, fig.width = 4, fig.height = 4, fig.cap = "Foo.", dev = "png", dpi = 150, out.width = "0.5\\textwidth"}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

The display below compares three single-hue sequential palettes by projection
to the luminance-chroma plane for the given fixed hue. In the left panel
the hue 260 is used with a simple linear chroma trajectory. The other two
panels employ a triangular chroma trajectory for hue 245, either with
a piecewise-linear (center) or power-transformed (right) trajectory.

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
```

Note that for H = 260 it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for H = 245. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

Finally, the display below compares two multi-hue sequential palettes along
with a diverging palette.

```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
```

The multi-hue palette on the left employs a small
hue range, resulting in a palette of "blues" just with slightly more distinction
of the middle colors in the palette. In contrast, the multi-hue "blue-yellow"
palette in the center panel uses a large hue range, resulting in more color
contrasts throughout the palette. Finally, the balanced diverging palette
in the right panel is constructed from two simple single-hue sequential
palettes (for hues 260/blue and 0/red) that are completely balanced between
the two "arms" of the palette.

## Demonstration of statistical graphics

To demonstrate how different kinds of color palettes work in different
kinds of statistical displays, `demoplot()` provides a simple convenience
interface to some base graphics with (mostly artificial) data sets.
As a first overview, all built-in demos are displayed with the same
sequential heat colors palette: `sequential_hcl(5, "Heat")`.

```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

All types of demos can, in principle, deal with arbitrarily many colors from any palette
but clearly the graphics differ in various respects such as:

- Work best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs. many colors (e.g., heatmap, perspective, ...).
- Intended for categorical data (e.g., bar, pie, ...) vs. continuous numeric data (e.g., heatmap, perspective, ...).
- Shading areas (e.g., map, bar, pie, ...) vs. shading points or lines (scatter, lines).

Hence, in the following some further illustrations are organized by type of
palette, using suitable demos for the particular palettes.

_Qualitative palettes:_ Light pastel colors typically work better for
shading areas (pie, left) while darker and more colorful palettes are usually
preferred for points (center) or lines (right).

```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"),    type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"),   type = "lines")
```

_Sequential palettes:_ Heatmaps (left) or perspective plots (center)
often employ almost continuous gradients with strong luminance contrasts.
In contrast, when only a few ordered categories are to be displayed
(e.g., in a spine plot, right) more colorful sequential palettes 
like the viridis palette can be useful.

```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"),        type = "perspective")
demoplot(sequential_hcl( 4, "Viridis"),      type = "spine")
```

_Diverging palettes:_ In some displays (such as the map, left),
it is useful to employ an almost continuous gradient with
strong luminance contrast to bring out the extremes. Here, this
contrast is amplified by a larger power transformation emphasizing
the extremes even further. In contrast, when fewer colors are needed
more colorful palettes with lower luminance contrasts can be desired.
This is exemplified by a mosaic (center) and bar plot (right).

```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl( 5, "Green-Orange"),        type = "mosaic")
demoplot(diverging_hcl( 5, "Blue-Red 2"),          type = "bar")
```

All displays above focused on palettes designed for light/white backgrounds.
Therefore, to conclude, some palettes are highlighted that work well on
dark/black backgrounds.

```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```

# Apps for choosing colors and palettes interactively {#sec:hclwizard}

To facilitate exploring the package and employing it when working with colors,
several graphical user interfaces (GUIs) are provided within the package as
_shiny_ apps [@color:shiny]. All of these GUIs or apps can either be run locally
from within R but are also provided at <http://hclwizard.org/>.

* **Palette constructor:** `choose_palette()` or `hclwizard()` or `hcl_wizard()`.  
* **Color picker:** `choose_color()` or equivalently `hcl_color_picker()`.
* **Color vision deficiency emulator:** `cvd_emulator()`.

In addition to the _shiny_ app, the **palette constructor** app is also available as a Tcl/Tk GUI via R package
_tcltk_ shipped with base R [@color:R]. The _tcltk_ version can only be run locally
and is considerably faster while the _shiny_ version has a nicer interface with more
features  and can be run online. The `choose_palette()` function by default starts the
_tcltk_ version while `hclwizard()`/`hcl_wizard()` by default start the _shiny_
version.

## Choose palettes with the HCL color model

This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
by default starts the _shiny_ version:

![](../../vignettes/screenshots/choose_palette_1.png)

The _tcltk_ version is started by default with `choose_palette()`:

![](../../vignettes/screenshots/choose_palette_3.png)

However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.

The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
`sequential_hcl()` for sequential palettes with single or
multiple hues, and `diverging_hcl()` for diverging palettes (composed
from two single-hue sequential palettes). See the discussion of
[HCL-based color palettes](hcl_palettes.html) for more details.

The GUIs
allow for interactive modification of the arguments of the respective
palette-generating functions, i.e., starting/ending hue (wavelength, type of
color), minimal/maximal chroma (colorfulness), minimal maximal luminance
(brightness, amount of gray), and a power transformations that control how
quickly/slowly chroma and/or luminance are changed through the palette.
Subsets of the parameters may not be applicable depending on the type of
palette chosen. See `qualitative_hcl()` and @color:Zeileis+Hornik+Murrell:2009 for
a more detailed explanation of the different arguments. 
@color:Stauffer+Mayr+Dabernig:2015 provide more examples and guidance.

Optionally, the active palette can be illustrated by using a
[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
such as a map, heatmap, scatter plot, perspective 3D surface etc.
To demonstrate different types of deficiencies, the active palette may be
desaturated (emulating printing on a grayscale printer) and collapsed to
emulate different types of color-blindness (without red-green or green-blue
contrasts) using the `simulate_cvd()` functions. To facilitate generation of
palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available:

![](../../vignettes/screenshots/choose_palette_2.png)

## Choose individual colors with the HCL color model

This GUI can be started with either `choose_color()` or equivalently `hcl_color_picker()`.

![](../../vignettes/screenshots/choose_color.png)

It shows the HCL color space either as a hue-chroma plane for a given luminance value or
as a luminance-chroma plane for a given hue. Colors can be entered by:

* Clicking on a color coordinate in the hue-chroma or luminance-chroma plane.
* Specifying the hue/chroma/luminance values via sliders.
* Entering an RGB hex code.

By repeating the selection a palette of colors can be constructed and returned
within R for subsequent usage in visualizations.

## Emulate color vision deficiencies

This GUI can be started with `cvd_emulator()`.

![](../../vignettes/screenshots/cvd_emulator_1.png){width=49%}
![](../../vignettes/screenshots/cvd_emulator_2.png){width=49%}

It allows to upload a raster image in
JPG or PNG format which is then check for various kinds of color vision deficiencies
at the selected severity. By default the severity is set to 100% and all supported
kinds of color vision deficiency are checked for, i.e.,

* Monochromatic (desaturated grayscale).
* Deuteranope vision (green deficient).
* Protanope vision (red deficient).
* Tritanope vision (blue deficient).

# Color vision deficiency emulation {#sec:color_vision_deficiency}

Using the physiologically-based model for simulating color vision deficiency (CVD)
of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
and tritanope (blue cone cells defective).
The workhorse function to do so is `simulate_cvd()` which can take any vector
of valid R colors and transform them according to a certain CVD transformation
matrix and transformation equation. The transformation matrices have been established
by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
`deutanomaly_cvd`, and `tritanomaly_cvd`.

The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
simulating the corresponding kind of color blindness with a given severity (calling
`simulate_cvd()` internally).

For further guidance on color blindness in relation to statistical graphics
see @color:Lumley:2006 which accompanies the R package _dichromat_ [@color:dichromat]
and is based on earlier emulation techniques
[@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].

## Illustration: Heatmap with sequential palette

To illustrate that poor color choices can severely reduce the usefulness of a statistical
graphic for readers with color vision deficiencies, we employ the infamous RGB rainbow
color palette in a heatmap. In base R this can be generated by `rainbow(11, end = 2/3)`
ranging from red (for high values) to blue (for low values).

The poor results for the RGB rainbow palette are contrasted with a proper sequential
palette ranging dark blue to light yellow: `sequential_hcl(11, "Blue-Yellow")`.

The statistical graphic employed for illustration is a heatmap of the well-known
Maunga Whau `volcano` data from base R. This heatmap is easily available as
`demoplot(..., "heatmap")` where `...` is the color vector to be used, e.g.,

```{r colors}
rainbow(11, end = 2/3)
deutan(rainbow(11, end = 2/3))
```

and so on. To aid the interpretation of the heatmap a perspective display using
only gray shades is included first, providing another intuitive display of what
the terrain around Maunga Whau looks like.

```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100}
z <- 2 * volcano
x <- 10 * (1:nrow(z))
y <- 10 * (1:ncol(z))
par(mar = rep(0, 4))
persp(x, y, z, theta = 10, phi = 30, col = "lightgray", scale = FALSE,
  ltheta = 100, shade = 0.5, border = NA, box = FALSE)
```

Subsequently, all combinations of palette and color vision deficiency are visualized.
Additionally, a grayscale version is created with `desaturate()`.

```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
rain <- rainbow(11, end = 2/3)
blyl <- sequential_hcl(11, "Blue-Yellow")
demoplot(rain, "heatmap")
mtext("Original", side = 2, line = 1)
mtext("rainbow(11, end = 2/3)", line = 1)
demoplot(blyl, "heatmap")
mtext("sequential_hcl(11, \"Blue-Yellow\")", line = 1)
demoplot(desaturate(rain), "heatmap")
mtext("Desaturated", side = 2, line = 1)
demoplot(desaturate(blyl), "heatmap")
demoplot(deutan(rain), "heatmap")
mtext("Deuteranope", side = 2, line = 1)
demoplot(deutan(blyl), "heatmap")
demoplot(protan(rain), "heatmap")
mtext("Protanope", side = 2, line = 1)
demoplot(protan(blyl), "heatmap")
demoplot(tritan(rain), "heatmap")
mtext("Tritanope", side = 2, line = 1)
demoplot(tritan(blyl), "heatmap")
```

This clearly shows how poorly the RGB rainbow performs, often giving quite
misleading impressions of what the terrain around Maunga Whau looks like.
In contrast, the HCL-based blue-yellow palette works reasonably well in all settings. The most important
problem of the RGB rainbow is that it is not monotonic in luminance, making
correct interpretation quite hard. Moreover, the red-green contrasts deteriorate
substantially in the dichromatic emulations.

## Illustration: Map with diverging palette

As another example for the poor performance of the RGB rainbow we employ a
shaded map. This is available as `demoplot(..., "map")` and is based on county
polygons for Alabama, Georgia, and South Carolina along with an artifical variable
used for coloring.

Often the red-yellow-green RGB spectrum is used for a diverging palette with 
yellow as the neutral value. In base R this can be generated by
`rainbow(11, end = 1/3)`. However, this palette has again a number of weaknesses,
especially that the green-yellow part of the palette almost collapses to the same
color when desaturated or when color blindness is emulated.

To illustrate that much more balanced palettes for the same purpose are available
the `Geyser` palette [mimicked from @color:CARTO] is adopted:
`divergingx_hcl(11, "Geyser", rev = TRUE)`. This also goes from red via yellow to
a green/blue and hence needs the `divergingx_hcl()` function (as opposed to
`diverging_hcl()` which always uses a neutral gray).

```{r map-diverging, echo = FALSE, fig.height = 13, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
rain <- rainbow(11, end = 1/3)
gysr <- divergingx_hcl(11, "Geyser", rev = TRUE)
demoplot(rain, "map")
mtext("Original", side = 2, line = 1)
mtext("rainbow(11, end = 1/3)", line = 1)
demoplot(gysr, "map")
mtext("divergingx_hcl(11, \"Geyser\", rev = TRUE)", line = 1)
demoplot(desaturate(rain), "map")
mtext("Desaturated", side = 2, line = 1)
demoplot(desaturate(gysr), "map")
demoplot(deutan(rain), "map")
mtext("Deuteranope", side = 2, line = 1)
demoplot(deutan(gysr), "map")
demoplot(protan(rain), "map")
mtext("Protanope", side = 2, line = 1)
demoplot(protan(gysr), "map")
demoplot(tritan(rain), "map")
mtext("Tritanope", side = 2, line = 1)
demoplot(tritan(gysr), "map")
```

While many versions of the RGB rainbow displays are hard to read because
they do not bring out any differences in the green-yellow arm of the palette,
the HCL-based palette works reasonably well in all settings. Only the grayscale
version cannot bring out the different arms of the palette. However, at least
both directions of deviation are visible even if they cannot be distinguished.
This is preferable to the RGB rainbow which hides all differences in the
green-yellow arm of the palette. (However, if grayscale printing is desired
a sequential rather than a diverging palette is probably necessary.)

# Color manipulation and utilities {#sec:manipulation_utilities}

The \pkg{colorspace} package provides several color manipulation utilities that are
useful for creating, assessing, or transforming color palettes, namely:

* `desaturate()`: Desaturate colors by chroma removal in HCL space.
* `darken()` and `lighten()`: Algorithmically lighten or darken colors in HCL and/or HLS space.
* `max_chroma()`: Compute maximum chroma for given hue and luminance in HCL space.
* `mixcolor()`: Additive mixing of two colors by computing the convex combination.

## Desaturation in HCL space

Desaturation should map a given color to the gray with the same "brightness".
In principle, any perceptually-based color model (HCL, HLS, HSV, ...) could be
employed for this but HCL works particularly well because its coordinates
capture the perceptual properties better than most other color models.

Thus, the `desaturate()` function converts any given hex color code or named
R color to the corresponding HCL coordinates and sets the chroma to zero.
Thus, only the luminance matters which captures the "brightness" mentioned above.
Finally, the resulting HCL coordinates are transformed back to hex color codes
for use in R.

For illustration, a few simple examples are presented below. More examples in the
context of palettes for statistical graphics are discussed along with the
[color vision deficiency](color_vision_deficiency.html) article.

First, `desaturate()` is used to desaturate a vector of R color names:

```{r desaturate-named}
desaturate(c("white", "orange", "blue", "black"))
```

Notice that the hex codes corresponding to three coordinates in sRGB space
are always the same, indicating gray colors.

Analogously, hex color codes can also be transformed - in this case RGB rainbow
colors from the base R function `rainbow()`:

```{r desaturate-rainbow}
rainbow(3)
desaturate(rainbow(3))
```

Already this simple example shows that the three RGB rainbow colors have very
different grayscale levels. This can be brought even more clearly when using 
a full color wheel (of colors with hues in [0, 360] degrees). While the RGB
`rainbow()` is very unbalanced the HCL `rainbow_hcl()` (or also `qualitative_hcl()`)
is (by design) balanced with respect to luminance.

```{r desaturate-wheel, fig.height = 6, fig.width = 6, fig.cap = "Foo."}
wheel <- function(col, radius = 1, ...)
  pie(rep(1, length(col)), col = col, radius = radius, ...) 
par(mar = rep(0.5, 4), mfrow = c(2, 2))
wheel(rainbow(8))
wheel(desaturate(rainbow(8)))
wheel(rainbow_hcl(8))
wheel(desaturate(rainbow_hcl(8)))
```

## Lighten or darken colors

In principle, a similar approach for lightening and darkening colors can be employed as for desaturation
above. The colors can simply be transformed to HCL space and then the luminance can either be decreased
(turning the color darker) or increased (turning it lighter) while preserving the hue and chroma coordinates.

This strategy typically works well for lightening colors, although in some situations the result can be
rather colorful. Conversely, when darkening rather light colors with little chroma, this can result
in rather gray colors.

In these situations, an alternative might be to apply the analogous strategy in HLS space which is frequently
used in HTML style sheets. However, this strategy may also yield colors that are either too gray or too
colorful. A compromise that sometimes works well is to adjust the luminance coordinate in HCL space but
to take the chroma coordinate corresponding the HLS transformation.

We have found that typically the HCL-based transformation performs best for lightening colors and this is
hence the default in `lighten()`. For darkening colors, the combined strategy often works best and is hence
the default in `darken()`. In either case it is recommended to try the other available strategies in case
the default yields unexpected results.

Regardless of the chosen color space, the adjustment of the `L` component can occur by two methods, relative
(the default) and absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
`L * (1 - amount)` (relative). See `lighhten()` and `darken()` for more details.

For illustration a qualitative palette (Okabe-Ito) is transformed by two levels of both lightening and
darkening, respectively.

```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, fig.cap = "Foo."}
oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
swatchplot(
  "-40%" = lighten(oi, 0.4),
  "-20%" = lighten(oi, 0.2),
  "  0%" = oi,
  " 20%" =  darken(oi, 0.2),
  " 40%" =  darken(oi, 0.4),
  off = c(0, 0)
)
```

## Maximum chroma for given hue and luminance

As the possible combinations of chroma and luminance in HCL space depend on hue,
it is not obvious which trajectories through HCL space are possible prior to
trying a specific HCL coordinate by calling `polarLUV()`. To avoid having to
`fixup` the color upon conversion to RGB `hex` color codes, the
`max_chroma()` function computes (approximately) the maximum chroma
possible.

For illustration we show that for given luminance (here: L = 50) the maximum chroma
varies substantially with hue:

```{r max-chroma-L50}
max_chroma(seq(0, 360, by = 60), 50)
```

Similarly, maximum chroma also varies substantially across luminance values for
a given hue (here: H = 120, green):

```{r max-chroma-H120}
max_chroma(120, seq(0, 100, by = 20))
```

In the plots below more combinations are visualized: In the left panel for maximum
chroma across hues given luminance and in the right panel with increasing luminance
given hue.

```{r max-chroma-plot, echo = FALSE, fig.height = 4.5, fig.width = 10, out.width = "\\textwidth", fig.cap = "Foo."}
H <- seq(0, 360, by = 10)
L <- seq(30, 90, by = 20)
C <- outer(H, L, max_chroma)

par(mfrow = c(1, 2), xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 5), las = 1)
plot(0, 0, type = "n", xlim = c(0, 360), ylim = c(0, 165),
  xlab = "Hue (H)", ylab = "Maximum chroma (C)")
axis(4, at = C[37, ], labels = paste("L =", L))
for(i in seq_along(L)) {
  lines(H, C[, i])
  points(H, C[, i], col = hex(polarLUV(L[i], C[, i], H)), pch = 19, cex = 1.5, xpd = TRUE)
}

L <- seq(0, 100, by = 5)
H <- c(0, 60, 120, 250, 330)
C <- outer(H, L, max_chroma)

par(xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 1), las = 1)
plot(0, 0, type = "n", xlim = c(0, 175), ylim = c(0, 100),
  xlab = "Maximum chroma (C)", ylab = "Luminance (L)")
for(i in rev(seq_along(H))) {
  lines(C[i, ], L)
  points(C[i, ], L, col = hex(polarLUV(L, C[i, ], H[i])), pch = 19, cex = 1.5, xpd = TRUE)
  j <- which.max(C[i,])
  text(C[i, j], L[j], paste("H =", H[i]), pos = 4, col = hex(polarLUV(L[j], C[i, j], H[i])))
}
points(0, 100, cex = 1.6, xpd = TRUE)
```

## Additive mixing of two colors

In additive color models like `RGB()` or `XYZ()` it can be useful to combine
colors by additive mixing. Below a fully saturated red and green are mixed,
yielding a medium brownish yellow.

```{r mixcolor, fig.height = 1.5, fig.width = 5, fig.cap = "Foo."}
R <- RGB(1, 0, 0)
G <- RGB(0, 1, 0)
Y <- mixcolor(0.5, R, G)
Y
par(mar = rep(0, 4))
swatchplot(c(hex(R), hex(Y), hex(G)))
```

# Approximating palettes from other packages {#sec:approximations}

The flexible specification of HCL-based color palettes in \pkg{colorspace} allows
to closely approximate color palettes from various other packages:

- _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
- _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
  as provided by R package _viridis_ [@color:viridis].
- The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].

See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
following `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).

## Approximations of ColorBrewer.org palettes

```{r brewer, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
specplot(
  rev(RColorBrewer::brewer.pal(9, "YlGnBu")),
  sequential_hcl(9, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
```

```{r viridis, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
specplot(
  viridis::viridis(9),
  sequential_hcl(9, "Viridis"),
  main = "Viridis"
)
```

# Summary and discussion {#sec:summary}

As usual...

# Computational details {#computational-details .unnumbered}

The results in this paper were obtained using
\proglang{R} `r paste(R.Version()[6:7], collapse = ".")` [@color:R]
with the packages
\pkg{colorspace} `r packageVersion("colorspace")` [@color:colorspace],
\pkg{ggplot2} `r packageVersion("ggplot2")` [@color:ggplot2pkg],
\pkg{RColorBrewer} `r packageVersion("RColorBrewer")` [@color:RColorBrewer],
\pkg{rcartocolor} `r packageVersion("rcartocolor")` [@color:rcartocolor],
\pkg{viridis} `r packageVersion("viridis")` [@color:viridis],
\pkg{scico} `r packageVersion("scico")` [@color:scico]. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
<https://CRAN.R-project.org/>.

# Acknowledgments {#acknowledgments .unnumbered}

All acknowledgments (note the AE spelling) should be collected in this
unnumbered section before the references. It may contain the usual
information about funding and feedback from colleagues/reviewers/etc.
Furthermore, information such as relative contributions of the authors
may be added here (if any).

### 2019-02-24 pkg/colorspace/paper/Toolbox/article.Rmd
removed some details and redundancies
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@454 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r preliminaries, echo=FALSE, message=FALSE}
+```{r preliminaries, echo = FALSE, message = FALSE}
 library("colorspace")
@@
 theme_set(theme_minimal())
+knitr::opts_chunk$set(message = FALSE, warning = FALSE,
+  dev = "pdf", dpi = 150,
+  fig.height = 6, fig.width = 6, fig.pos = "t!",
+  fig.align = "center", fig.show = "hold", fig.path = "Figures/")
 ```
@@
 
-```{r hcl-properties, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space: Hue (= type of color, dominant wavelength), chroma (= colorfulness), luminance (= brightness).
-In each line the color coordinates are varied along the indicated axis, keeping the other two coordinates fixed."}
+```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space.
+Top: Hue $H$ changes from 0 (red) via 60 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$.
+Center: Chroma $C$ changes from 0 (gray) to 100 (colorful) with fixed $H = 0$ (red) and $L = 65$.
+Bottom: Luminance $L$ changes from 90 (light) to 25 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
 swatchplot(
-  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
-  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
-  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
-  off = 0
+  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
+  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
+  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
+  nrow = 3, line = 5.2, off = 0
 )
@@
 
-## Installation
-
 The stable release version of \pkg{colorspace} is hosted on the Comprehensive \proglang{R} Archive Network
@@
 
-For \proglang{Python} users a beta re-implementation of the full \pkg{colorspace} package in
-\proglang{Python}\ 2/\proglang{Python}\ 3 is also available, see <https://github.com/retostauffer/python-colorspace>.
-In the manuscript we focus on the more mature \proglang{R} implementation but
-replication materials for most examples are not only provided for \proglang{R} but
-also for \proglang{Python}.
-
 ## Choosing HCL-based color palettes
@@
 
-```{r hcl-palettes, message = FALSE, warning = FALSE, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
+```{r hcl-palettes, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
 library("colorspace")
@@
 
-```{r eustockmarkets-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
+```{r eustockmarkets-plot, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
 q4 <- qualitative_hcl(4)
@@
 
-```{r titanic-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r titanic-plot, echo = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
 ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
@@
 
-To plug the HCL color palettes into \pkg{ggplot2} graphics suitable discrete and/or
-continuous \pkg{gglot2} color scales are provided. The scales are called via the scheme
-`scale_<aesthetic>_<datatype>_<colorscale>()`, where `<aesthetic>` is the name
-of the aesthetic (`fill`, `color`, `colour`), `<datatype>` is the type of the
-variable plotted (`discrete` or `continuous`) and `<colorscale>` sets the type
-of the color scale used (`qualitative`, `sequential`, `diverging`,
-`divergingx`).
+To plug the HCL color palettes into \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
+suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
+The scales are called via the scheme
+
+    scale_<aesthetic>_<datatype>_<colorscale>()
+
+where
+
+- `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`).
+- `<datatype>` is the type of the variable plotted (`discrete` or `continuous`).
+- `<colorscale>` sets the type of the color scale used (`qualitative`, `sequential`, `diverging`, `divergingx`).
 
@@
 
-```{r iris-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r iris-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
 library("ggplot2")
@@
 
-```{r diamonds-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r diamonds-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
 dsamp <- diamonds[1 + 1:1000 * 50, ]
@@
 
-As motivated in the previous article ([Color Spaces: S4 Classes and Utilities](color_spaces.html)),
+As motivated in the previous section,
 the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
 of the human visual system very well. Therefore, the \pkg{colorspace} package provides
-three types of palettes based on the HCL model:
-
-* _Qualitative:_ Designed for coding categorical information, i.e.,
-  where no particular ordering of categories is available and every color
-  should receive the same perceptual weight.
-* _Sequential:_ Designed for coding ordered/numeric information, i.e.,
-  going from high to low (or vice versa).
-* _Diverging:_ Designed for coding ordered/numeric information around a central
-  neutral value, i.e., where colors diverge from neutral to two extremes.
-
-The corresponding functions are `qualitative_hcl()`, `sequential_hcl()`,
-and `diverging_hcl()`. Their construction principles are exemplified in the
-following color swatches and explained in more detail below. The desaturated
+three palette functions based on the HCL model: `qualitative_hcl()`, `sequential_hcl()`,
+and `diverging_hcl()`. Their construction principles are exemplified in
+Figure\ \ref{fig:hcl-palettes-principles} and explained in more detail below. The desaturated
 palettes bring out clearly that luminance differences (= light-dark contrasts),
@@
 
+## Approximating palettes from other packages
+
+The flexible specification of HCL-based color palettes in \pkg{colorspace} allows
+to closely approximate color palettes from various other packages:
+
+- _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
+- _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
+- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
+  as provided by R package _viridis_ [@color:viridis].
+- The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].
+
+See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
+following `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
+of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
+
+```{r brewer-carto-viridis, results = "hide", echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+specplot(
+  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
+  sequential_hcl(7, "YlGnBu"),
+  main = "ColorBrewer.org: YlGnBu"
+)
+specplot(
+  viridis::viridis(7),
+  sequential_hcl(7, "Viridis"),
+  main = "viridis: Viridis"
+)
+specplot(
+  rcartocolor::carto_pal(7, "ag_Sunset"),
+  sequential_hcl(7, "ag_Sunset"),
+  main = "CARTO: ag_Sunset"
+)
+specplot(
+  viridis::plasma(7),
+  sequential_hcl(7, "Plasma"),
+  main = "viridis: Plasma"
+)
+```
+
 ## HCL (and HSV) color palettes corresponding to base R palettes
@@
 
-# HCL-based color scales for ggplot2 {#sec:ggplot2_color_scales}
-
-All [HCL-based color palettes](hcl_palettes.html) in the \pkg{colorspace} package [@color:colorspace]
-are also provided as discrete and/or continuous color scales for the use with the _ggplot2_ package
-[@color:ggplot2; @color:ggplot2pkg].
-
-The scales are called via the scheme
-
-    scale_<aesthetic>_<datatype>_<colorscale>()
-
-where
-
-- `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`).
-- `<datatype>` is the type of the variable plotted (`discrete` or `continuous`).
-- `<colorscale>` sets the type of the color scale used (`qualitative`, `sequential`, `diverging`, `divergingx`).
-
-A few examples of these scales are illustrated in the following sections.
-
-## Using the scales in ggplot2
-
-A discrete qualitative scale applied to a fill aesthetic corresponds to the function `scale_fill_discrete_qualitative()`:
-
-```{r density-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
-  scale_fill_discrete_qualitative()
-```
-
-Similarly, a color aesthetic for a discrete qualitative scale corresponds to the function `scale_color_discrete_qualitative()`:
-
-```{r point-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point() +
-  scale_color_discrete_qualitative(palette = "Set 2")
-```
-
-A continuous sequential scale applied to a color aesthetic corresponds to the function `scale_color_continuous_sequential()`:
-
-```{r jitter-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
-  scale_color_continuous_sequential(palette = "Heat")
-```
-
-A continuous sequential scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_sequential()`:
-
-```{r raster-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-df <- data.frame(height = c(volcano), x = c(row(volcano)), y = c(col(volcano)))
-ggplot(df, aes(x, y, fill = height)) + 
-  geom_raster() + 
-  coord_fixed(expand = FALSE) +
-  scale_fill_continuous_sequential(palette = "Blues")
-```
-
-A continuous diverging scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_diverging()`:
-
-```{r tile-diverging, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-cm <- cor(mtcars)
-df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
-levels(df$var1) <- levels(df$var2) <- names(mtcars)
-ggplot(df, aes(var1, var2, fill = cor)) + 
-  geom_tile() + 
-  coord_fixed() +
-  ylab("variable") +
-  scale_x_discrete(position = "top", name = "variable") +
-  scale_fill_continuous_diverging("Blue-Red 3")
-```
-
-## Customizing the scales
-
-All scale functions accept a `palette` argument which allows you to pick a specific color palette out of a selection of different options.
-All available palettes are listed at the end of this document.
-For example, we could use the "Harmonic" palette when we need a qualitative color scale:
-
-```{r density-qualitative2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
-  scale_fill_discrete_qualitative(palette = "Harmonic")
-```
-
-The color palettes are calculated on the fly depending on the number of different colors needed.
-But sometimes, it may be desireable to pick specific colors out of a larger set, e.g., when we are making two plots where one contains a subset of the data of the other, or when the default order of colors is not ideal.
-Therefore, all discrete scales provide parameters `nmax` to set the total number of colors requested and `order` (a vector of integers) to reorder the color palette.
-
-Applied to the previous plot, we could for example do the following:
-
-```{r density-qualitative3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
-  scale_fill_discrete_qualitative(palette = "Harmonic", nmax = 5, order = c(5, 1, 2))
-```
-
-The `nmax` option is also convenient to remove some colors from a scale that may not be appropriate for the plot.
-For example, the `scale_color_brewer()` scale that comes with _ggplot2_ tends to produce points that are too light:
-
-```{r point-brewer, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-dsamp <- diamonds[1 + 1:1000 * 50, ]
-gg <- ggplot(dsamp, aes(carat, price, color = cut)) + geom_point()
-gg + scale_color_brewer(palette = "Blues")
-```
-
-Similar problems can arise with the HCL palettes, but there we have the option of creating additional colors that we then do not use:
-
-```{r point-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-gg + scale_color_discrete_sequential(palette = "Blues", nmax = 6, order = 2:6)
-```
-
-(We use `order = 2:6` to pick the five darkest colors and omit the lightest color.)
-
-All [HCL-based color palettes](hcl_palettes.html) are defined via sets of hue (H), chroma (C), and luminance (L) values.
-For example, the qualitative scales vary hue from a start value to an end value while keeping chroma and luminance fixed.
-Similarly, single-hue sequential scales vary chroma and luminance while keeping the hue fixed.
-We can override these settings by specifying the corresponding H, C, or L values in addition to the palette name.
-As an example, consider the following plot:
-
-```{r jitter-sequential2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
-  scale_color_continuous_sequential(palette = "Terrain")
-```
-
-Now assume we generally like the color scale but find it a bit too pink at the end. We can fix this issue by specifying an alternative final hue value, e.g., `h2 = 60` corresponding to yellow:
-
-```{r jitter-sequential3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
-  scale_color_continuous_sequential(palette = "Terrain", h2 = 60)
-```
-
-The next example uses a diverging scale. First consider the plot with the unmodified "Blue-Yellow 2" palette:
-
-```{r tile-diverging2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-cm <- cor(mtcars)
-df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
-levels(df$var1) <- levels(df$var2) <- names(mtcars)
-gg <- ggplot(df, aes(var1, var2, fill = cor)) + 
-  geom_tile() + 
-  coord_fixed() +
-  ylab("variable") +
-  scale_x_discrete(position = "top", name = "variable")
-
-gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2")
-```
-
-And now the same plot with some palette customizations: The ordering is reversed so that blue is used for
-positive correlations and yellow for negative ones. Moreover, the power parameter `p2` for the luminance
-is increased so that only correlations close to an absolute value of 1 have dark colors while intermediate
-correlations have relatively light colors.
-
-```{r tile-diverging3, fig.width = 6.5, fig.asp = .7, fig.cap = "Foo."}
-gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2", rev = TRUE, p2 = 2)
-```
-
-See the reference manual for the exact set of customization parameters that are available for each scale.
-
-The continuous scales also provide the option to limit the scale range to which data are mapped, via the parameters `begin` and `end`.
-As an example, assume we are using the approximation of the viridis scale provided by `scale_color_continuous_sequential()`:
-
-```{r jitter-sequential4, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
-  scale_color_continuous_sequential(palette = "Viridis")
-```
-
-If we want to remove some of the darkest blues and some of the brightest yellows from this scale, we can write:
-
-```{r jitter-sequential5, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
-ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
-  scale_color_continuous_sequential(palette = "Viridis", begin = 0.15, end = 0.9)
-```
-
-## Available palettes
-
-In the following, we are visualizing all scales currently available via
-pre-defined names. These visualizations are generated by the function
-`hcl_palettes()` with option `plot = TRUE`.
-
-The discrete qualitative scales are all called via `scale_*_discrete_qualitative(palette = "name")`,
-where `name` is the name of the palette, e.g., `palette = "Pastel 1"`. There are no continuous qualitative
-scales.
-
-The discrete sequential scales are all called via `scale_*_discrete_sequential(palette = "name")`,
-where `name` is the name of the palette, e.g., `palette = "Grays"`. Continuous approximations to the discrete
-scales exist and can be called via `scale_*_continuous_sequential(palette = "name")`
-
-The discrete diverging scales are all called via `scale_*_discrete_diverging(palette = "name")`,
-where `name` is the name of the palette, e.g., `palette = "Blue-Red"`. Continuous approximations to the discrete
-scales exist and can be called via `scale_*_continuous_diverging(palette = "name")`
-
 # Palette visualization and assessment {#sec:palette_visualization}
@@
 
-As a first simple example, the properties of the three HCL axes are brought out by varying
-only one coordinate while keeping the other two fixed:
-
-```{r swatch-hcl, fig.width = 4, fig.height = 2.2, fig.cap = "Foo."}
-swatchplot(
-  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
-  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
-  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
-  off = 0
-)
-```
-
-This shows the following:
-
-- Hue: Only the hue (= type of color) changes from H = 0 (red) via 60 (yellow), etc. to 300 (purple)
-  while chroma and luminance are fixed to moderate values of C = 60 and L = 65, respectively.
-- Chroma: Only the chroma (= colorfulness) changes from C = 0 (gray) to 100 (colorful)
-  while hue and luminance are fixed to H = 0 (red) and L = 65, respectively.
-- Luminance: Only the luminance (= brightness) changes from L = 90 (light) to 25 (dark)
-  while hue and chroma are fixed to H = 260 (blue) and C = 25 (low, close to gray), respectively.
+As a first simple example:
+`swatchplot("Label" = sequential_hcl(5))`.
 
@@
 
-```{r spec-qualitative, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
+```{r spec-qualitative, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 specplot(qualitative_hcl(100, "Set 2"))
@@
 
-```{r spec-sequential, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
+```{r spec-sequential, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 specplot(sequential_hcl(100, "Blues 2"))
@@
 
-```{r spec-diverging, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
+```{r spec-diverging, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 specplot(diverging_hcl(100, "Blue-Red"))
@@
 
-```{r spec-rainbow, fig.width = 5, fig.height = 7.5, fig.cap = "Foo."}
+```{r spec-rainbow, fig.width = 5, fig.height = 7.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 specplot(rainbow(100), rgb = TRUE)
@@
 
-# Apps for choosing colors and palettes interactively {#sec:hclwizard}
-
-To facilitate exploring the package and employing it when working with colors,
-several graphical user interfaces (GUIs) are provided within the package as
-_shiny_ apps [@color:shiny]. All of these GUIs or apps can either be run locally
-from within R but are also provided at <http://hclwizard.org/>.
-
-* **Palette constructor:** `choose_palette()` or `hclwizard()` or `hcl_wizard()`.  
-* **Color picker:** `choose_color()` or equivalently `hcl_color_picker()`.
-* **Color vision deficiency emulator:** `cvd_emulator()`.
-
-In addition to the _shiny_ app, the **palette constructor** app is also available as a Tcl/Tk GUI via R package
-_tcltk_ shipped with base R [@color:R]. The _tcltk_ version can only be run locally
-and is considerably faster while the _shiny_ version has a nicer interface with more
-features  and can be run online. The `choose_palette()` function by default starts the
-_tcltk_ version while `hclwizard()`/`hcl_wizard()` by default start the _shiny_
-version.
-
-## Choose palettes with the HCL color model
-
-This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
-by default starts the _shiny_ version:
-
-![](../../vignettes/screenshots/choose_palette_1.png)
-
-The _tcltk_ version is started by default with `choose_palette()`:
-
-![](../../vignettes/screenshots/choose_palette_3.png)
-
-However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.
-
-The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
-`sequential_hcl()` for sequential palettes with single or
-multiple hues, and `diverging_hcl()` for diverging palettes (composed
-from two single-hue sequential palettes). See the discussion of
-[HCL-based color palettes](hcl_palettes.html) for more details.
-
-The GUIs
-allow for interactive modification of the arguments of the respective
-palette-generating functions, i.e., starting/ending hue (wavelength, type of
-color), minimal/maximal chroma (colorfulness), minimal maximal luminance
-(brightness, amount of gray), and a power transformations that control how
-quickly/slowly chroma and/or luminance are changed through the palette.
-Subsets of the parameters may not be applicable depending on the type of
-palette chosen. See `qualitative_hcl()` and @color:Zeileis+Hornik+Murrell:2009 for
-a more detailed explanation of the different arguments. 
-@color:Stauffer+Mayr+Dabernig:2015 provide more examples and guidance.
-
-Optionally, the active palette can be illustrated by using a
-[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
-such as a map, heatmap, scatter plot, perspective 3D surface etc.
-To demonstrate different types of deficiencies, the active palette may be
-desaturated (emulating printing on a grayscale printer) and collapsed to
-emulate different types of color-blindness (without red-green or green-blue
-contrasts) using the `simulate_cvd()` functions. To facilitate generation of
-palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available:
-
-![](../../vignettes/screenshots/choose_palette_2.png)
-
-## Choose individual colors with the HCL color model
-
-This GUI can be started with either `choose_color()` or equivalently `hcl_color_picker()`.
-
-![](../../vignettes/screenshots/choose_color.png)
-
-It shows the HCL color space either as a hue-chroma plane for a given luminance value or
-as a luminance-chroma plane for a given hue. Colors can be entered by:
-
-* Clicking on a color coordinate in the hue-chroma or luminance-chroma plane.
-* Specifying the hue/chroma/luminance values via sliders.
-* Entering an RGB hex code.
-
-By repeating the selection a palette of colors can be constructed and returned
-within R for subsequent usage in visualizations.
-
-## Emulate color vision deficiencies
-
-This GUI can be started with `cvd_emulator()`.
-
-![](../../vignettes/screenshots/cvd_emulator_1.png){width=49%}
-![](../../vignettes/screenshots/cvd_emulator_2.png){width=49%}
-
-It allows to upload a raster image in
-JPG or PNG format which is then check for various kinds of color vision deficiencies
-at the selected severity. By default the severity is set to 100% and all supported
-kinds of color vision deficiency are checked for, i.e.,
-
-* Monochromatic (desaturated grayscale).
-* Deuteranope vision (green deficient).
-* Protanope vision (red deficient).
-* Tritanope vision (blue deficient).
-
 # Color vision deficiency emulation {#sec:color_vision_deficiency}
@@
 
-```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100}
+```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100, out.width = "0.5\\textwidth"}
 z <- 2 * volcano
@@
 
-```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "0.85\\textwidth", fig.pos = "p!", fig.cap = "Foo."}
 par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
@@
 
-## Illustration: Map with diverging palette
-
-As another example for the poor performance of the RGB rainbow we employ a
-shaded map. This is available as `demoplot(..., "map")` and is based on county
-polygons for Alabama, Georgia, and South Carolina along with an artifical variable
-used for coloring.
-
-Often the red-yellow-green RGB spectrum is used for a diverging palette with 
-yellow as the neutral value. In base R this can be generated by
-`rainbow(11, end = 1/3)`. However, this palette has again a number of weaknesses,
-especially that the green-yellow part of the palette almost collapses to the same
-color when desaturated or when color blindness is emulated.
-
-To illustrate that much more balanced palettes for the same purpose are available
-the `Geyser` palette [mimicked from @color:CARTO] is adopted:
-`divergingx_hcl(11, "Geyser", rev = TRUE)`. This also goes from red via yellow to
-a green/blue and hence needs the `divergingx_hcl()` function (as opposed to
-`diverging_hcl()` which always uses a neutral gray).
-
-```{r map-diverging, echo = FALSE, fig.height = 13, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
-par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
-rain <- rainbow(11, end = 1/3)
-gysr <- divergingx_hcl(11, "Geyser", rev = TRUE)
-demoplot(rain, "map")
-mtext("Original", side = 2, line = 1)
-mtext("rainbow(11, end = 1/3)", line = 1)
-demoplot(gysr, "map")
-mtext("divergingx_hcl(11, \"Geyser\", rev = TRUE)", line = 1)
-demoplot(desaturate(rain), "map")
-mtext("Desaturated", side = 2, line = 1)
-demoplot(desaturate(gysr), "map")
-demoplot(deutan(rain), "map")
-mtext("Deuteranope", side = 2, line = 1)
-demoplot(deutan(gysr), "map")
-demoplot(protan(rain), "map")
-mtext("Protanope", side = 2, line = 1)
-demoplot(protan(gysr), "map")
-demoplot(tritan(rain), "map")
-mtext("Tritanope", side = 2, line = 1)
-demoplot(tritan(gysr), "map")
-```
-
-While many versions of the RGB rainbow displays are hard to read because
-they do not bring out any differences in the green-yellow arm of the palette,
-the HCL-based palette works reasonably well in all settings. Only the grayscale
-version cannot bring out the different arms of the palette. However, at least
-both directions of deviation are visible even if they cannot be distinguished.
-This is preferable to the RGB rainbow which hides all differences in the
-green-yellow arm of the palette. (However, if grayscale printing is desired
-a sequential rather than a diverging palette is probably necessary.)
+# Apps for choosing colors and palettes interactively {#sec:hclwizard}
+
+To facilitate exploring the package and employing it when working with colors,
+several graphical user interfaces (GUIs) are provided within the package as
+_shiny_ apps [@color:shiny]. All of these GUIs or apps can either be run locally
+from within R but are also provided at <http://hclwizard.org/>.
+
+* **Palette constructor:** `choose_palette()` or `hclwizard()` or `hcl_wizard()`.  
+* **Color picker:** `choose_color()` or equivalently `hcl_color_picker()`.
+* **Color vision deficiency emulator:** `cvd_emulator()`.
+
+In addition to the _shiny_ app, the **palette constructor** app is also available as a Tcl/Tk GUI via R package
+_tcltk_ shipped with base R [@color:R]. The _tcltk_ version can only be run locally
+and is considerably faster while the _shiny_ version has a nicer interface with more
+features  and can be run online. The `choose_palette()` function by default starts the
+_tcltk_ version while `hclwizard()`/`hcl_wizard()` by default start the _shiny_
+version.
+
+## Choose palettes with the HCL color model
+
+This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
+by default starts the _shiny_ version.
+The _tcltk_ version is started by default with `choose_palette()`.
+However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.
+
+```{r choose_palette, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
+knitr::include_graphics("../../vignettes/screenshots/choose_palette_1.png")
+```
+
+The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
+`sequential_hcl()` for sequential palettes with single or
+multiple hues, and `diverging_hcl()` for diverging palettes (composed
+from two single-hue sequential palettes).
+
+The GUIs
+allow for interactive modification of the arguments of the respective
+palette-generating functions, i.e., starting/ending hue (wavelength, type of
+color), minimal/maximal chroma (colorfulness), minimal maximal luminance
+(brightness, amount of gray), and a power transformations that control how
+quickly/slowly chroma and/or luminance are changed through the palette.
+Subsets of the parameters may not be applicable depending on the type of
+palette chosen. See `qualitative_hcl()` and @color:Zeileis+Hornik+Murrell:2009 for
+a more detailed explanation of the different arguments. 
+@color:Stauffer+Mayr+Dabernig:2015 provide more examples and guidance.
+
+Optionally, the active palette can be illustrated by using a
+[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
+such as a map, heatmap, scatter plot, perspective 3D surface etc.
+To demonstrate different types of deficiencies, the active palette may be
+desaturated (emulating printing on a grayscale printer) and collapsed to
+emulate different types of color-blindness (without red-green or green-blue
+contrasts) using the `simulate_cvd()` functions. To facilitate generation of
+palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available.
+
+## Choose individual colors with the HCL color model
+
+This GUI can be started with either `choose_color()` or equivalently `hcl_color_picker()`.
+
+```{r choose_color, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
+knitr::include_graphics("../../vignettes/screenshots/choose_color.png")
+```
+
+It shows the HCL color space either as a hue-chroma plane for a given luminance value or
+as a luminance-chroma plane for a given hue. Colors can be entered by:
+
+* Clicking on a color coordinate in the hue-chroma or luminance-chroma plane.
+* Specifying the hue/chroma/luminance values via sliders.
+* Entering an RGB hex code.
+
+By repeating the selection a palette of colors can be constructed and returned
+within R for subsequent usage in visualizations.
+
+## Emulate color vision deficiencies
+
+This GUI can be started with `cvd_emulator()`.
+
+\begin{figure}[t!]
+\centering 
+\includegraphics[width=0.49\textwidth]{../../vignettes/screenshots/cvd_emulator_1} 
+\includegraphics[width=0.49\textwidth]{../../vignettes/screenshots/cvd_emulator_2}
+\caption[Foo]{Foo.}\label{fig:cvd_emulator}
+\end{figure}
+
+It allows to upload a raster image in
+JPG or PNG format which is then checked for various kinds of color vision deficiencies
+at the selected severity. By default the severity is set to 100% and all supported
+kinds of color vision deficiency are checked for.
 
@@
 
-```{r desaturate-wheel, fig.height = 6, fig.width = 6, fig.cap = "Foo."}
+```{r desaturate-wheel, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 wheel <- function(col, radius = 1, ...)
@@
 
-```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, fig.cap = "Foo."}
+```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
 oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
@@
 
-In the plots below more combinations are visualized: In the left panel for maximum
-chroma across hues given luminance and in the right panel with increasing luminance
-given hue.
-
-```{r max-chroma-plot, echo = FALSE, fig.height = 4.5, fig.width = 10, out.width = "\\textwidth", fig.cap = "Foo."}
-H <- seq(0, 360, by = 10)
-L <- seq(30, 90, by = 20)
-C <- outer(H, L, max_chroma)
-
-par(mfrow = c(1, 2), xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 5), las = 1)
-plot(0, 0, type = "n", xlim = c(0, 360), ylim = c(0, 165),
-  xlab = "Hue (H)", ylab = "Maximum chroma (C)")
-axis(4, at = C[37, ], labels = paste("L =", L))
-for(i in seq_along(L)) {
-  lines(H, C[, i])
-  points(H, C[, i], col = hex(polarLUV(L[i], C[, i], H)), pch = 19, cex = 1.5, xpd = TRUE)
-}
-
-L <- seq(0, 100, by = 5)
-H <- c(0, 60, 120, 250, 330)
-C <- outer(H, L, max_chroma)
-
-par(xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 1), las = 1)
-plot(0, 0, type = "n", xlim = c(0, 175), ylim = c(0, 100),
-  xlab = "Maximum chroma (C)", ylab = "Luminance (L)")
-for(i in rev(seq_along(H))) {
-  lines(C[i, ], L)
-  points(C[i, ], L, col = hex(polarLUV(L, C[i, ], H[i])), pch = 19, cex = 1.5, xpd = TRUE)
-  j <- which.max(C[i,])
-  text(C[i, j], L[j], paste("H =", H[i]), pos = 4, col = hex(polarLUV(L[j], C[i, j], H[i])))
-}
-points(0, 100, cex = 1.6, xpd = TRUE)
-```
-
 ## Additive mixing of two colors
@@
 
-```{r mixcolor, fig.height = 1.5, fig.width = 5, fig.cap = "Foo."}
+```{r mixcolor}
 R <- RGB(1, 0, 0)
@@
 Y
-par(mar = rep(0, 4))
-swatchplot(c(hex(R), hex(Y), hex(G)))
-```
-
-# Approximating palettes from other packages {#sec:approximations}
-
-The flexible specification of HCL-based color palettes in \pkg{colorspace} allows
-to closely approximate color palettes from various other packages:
-
-- _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
-- _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
-- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
-  as provided by R package _viridis_ [@color:viridis].
-- The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].
-
-See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
-following `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
-of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
-
-## Approximations of ColorBrewer.org palettes
-
-```{r brewer, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
-specplot(
-  rev(RColorBrewer::brewer.pal(9, "YlGnBu")),
-  sequential_hcl(9, "YlGnBu"),
-  main = "ColorBrewer.org: YlGnBu"
-)
-```
-
-```{r viridis, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
-specplot(
-  viridis::viridis(9),
-  sequential_hcl(9, "Viridis"),
-  main = "Viridis"
-)
 ```
@@
 As usual...
+
+For \proglang{Python} users a beta re-implementation of the full \pkg{colorspace} package in
+\proglang{Python}\ 2/\proglang{Python}\ 3 is also available, see <https://github.com/retostauffer/python-colorspace>.
+In the manuscript we focus on the more mature \proglang{R} implementation but
+replication materials for most examples are not only provided for \proglang{R} but
+also for \proglang{Python}.
 

### 2019-02-25 pkg/colorspace/paper/Toolbox/article.Rmd
further streamlining
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@455 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r eustockmarkets-plot, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
-q4 <- qualitative_hcl(4)
-par(mar = c(5, 4, 1, 1))
-plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
-legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
-```
-
 As another example for a sequential palette a spine plot is created below,
@@
 
-```{r titanic-plot, echo = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r eustockmarkets-titanic, echo = FALSE, fig.height = 4, fig.width = 10, out.width = "\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
+par(mar = c(5, 4, 1, 1), mfrow = c(1, 2))
+plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
+legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
 ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
-par(mar = c(5, 4, 1, 1))
 spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
@@
 
-```{r iris-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r iris-ggplot, eval = FALSE}
 library("ggplot2")
@@
 
-```{r diamonds-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
+```{r diamonds-ggplot, eval = FALSE}
+dsamp <- diamonds[1 + 1:1000 * 50, ]
+ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
+  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
+```
+
+```{r iris-diamonds, echo = FALSE, fig.height = 4, fig.width = 5, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+library("ggplot2")
+ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
+  scale_fill_discrete_qualitative(palette = "Dark 3")
+set.seed(1)
 dsamp <- diamonds[1 + 1:1000 * 50, ]
@@
   "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
-  nrow = 5
+  nrow = 5, line = 5
 )
@@
 
-The illustrations below show how basic qualitative, sequential, and diverging
-palettes are constructed in HCL space (the corresponding mathematical equations
-are provided in the [construction details](hcl_palettes.html#construction-details)).
-In the qualitative "Set 2" palette below, the
-hue traverses the entire color "wheel" (from 0 to 360 degrees) while keeping
-chroma and luminance (almost) constant (C = 60 and L = 70).
-
-```{r spec-qualitative, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
-specplot(qualitative_hcl(100, "Set 2"))
-```
-
-Note that due to the restrictions of the HCL color space, some of the green/blue
-colors have a slightly smaller maximum chroma resulting in a small dip in the chroma
-curve. This is fixed automatically (by default) and is hardly noticable in visualizations,
-though.
-
-The sequential "Blues 2" palette below employs a single hue (H = 260) and a monotonically
-increasing luminance trajectory (from dark to light). Chroma simply decreases monotonically
-along with increasing luminance.
-
-```{r spec-sequential, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
-specplot(sequential_hcl(100, "Blues 2"))
-```
-
-Finally, the diverging "Blue-Red" palette is depicted below. It simply combines a blue
-and a red sequential single-hue palette (similar to the "Blues 2" palette discussed above).
+FIXME: Point to previous plots
+
+Finally, the diverging "Green-Brown" palette is depicted below. It simply combines a green
+and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
 Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
@@
 
-```{r spec-diverging, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
-specplot(diverging_hcl(100, "Blue-Red"))
-```
-
 To contrast these well-balanced HCL-based palettes with a poorly-balanced palette, the
@@
 
-```{r spec-rainbow, fig.width = 5, fig.height = 7.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
+```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+specplot(diverging_hcl(100, "Green-Brown"), rgb = TRUE)
 specplot(rainbow(100), rgb = TRUE)

### 2019-02-25 pkg/colorspace/paper/Toolbox/article.Rmd
text improvements and additional references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@456 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 knitr::opts_chunk$set(message = FALSE, warning = FALSE,
-  dev = "pdf", dpi = 150,
+  dev = "pdf", dpi = 150, out.width = "\\textwidth", fig.cap = "FIXME.",
   fig.height = 6, fig.width = 6, fig.pos = "t!",
@@
 
-# Getting started {#sec:start}
+# A quick tour {#sec:tour}
 
@@
 
-```{r hcl-palettes, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
+```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
 library("colorspace")
@@
 to most base graphics function, typically through the `col` argument. Here, the `q4`
-vector created above is used in a time series display (see Figure \ref{fig:eustockmarkets-plot}):
+vector created above is used in a time series display (see the left panel of
+Figure\ \ref{fig:eustockmarkets-titanic}):
 
@@
 
-As another example for a sequential palette a spine plot is created below,
+As another example for a sequential palette a spine plot is created (see the right
+panel of Figure\ \ref{fig:eustockmarkets-titanic}),
 displaying the proportion of Titanic passengers that survived per class.
-The `Purples 3` palette is used which is quite similar to the **ColorBrewer.org**
-palette `Purples`. Here, only two colors are employed, yielding a dark purple
-and light gray.
+The `"Purples 3"` palette is used which is quite similar to the \pkg{ColorBrewer.org}
+[@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
+employed, yielding a dark purple and light gray.
 
@@
 
-```{r eustockmarkets-titanic, echo = FALSE, fig.height = 4, fig.width = 10, out.width = "\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
-par(mar = c(5, 4, 1, 1), mfrow = c(1, 2))
+```{r eustockmarkets-titanic, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "Using \\pkg{colorspace} with base \\proglang{R} graphics.
+Left: Time series plot of log-prices from \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette.
+Right: Spine plot with survival proportions across passenger classes in the \\code{titanice} data with \\code{sequential\\_hcl(2, \"Purples 3\")} palette."}
+par(mar = c(5, 4, 1, 2), mfrow = c(1, 2))
 plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
@@
 
-To illustrate their usage two simple examples are shown using the qualitative `Dark 3`
-and sequential `Purples 3` palettes that were also employed above. First, semi-transparent
-shaded densities of the sepal length from the iris data are shown, grouped by species.
+To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
+and sequential `"Purples 3"` palettes that were also employed above. First, semi-transparent
+shaded densities of the sepal length from the `iris` data are shown, grouped by species.
 
@@
 library("ggplot2")
-ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
+ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
+  geom_density(alpha = 0.6) +
   scale_fill_discrete_qualitative(palette = "Dark 3")
@@
 The sequential palette is used to code the cut levels in a scatter of price by carat
-in the diamonds data (or rather a small subsample thereof). The scale function first
+in the `diamonds` data (or rather a small subsample thereof). The scale function first
 generates six colors but then drops the first color because the light gray is too light
@@
 ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
-  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
-```
-
-```{r iris-diamonds, echo = FALSE, fig.height = 4, fig.width = 5, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+  scale_color_discrete_sequential(palette = "Purples 3",
+    nmax = 6, order = 2:6)
+```
+
+```{r iris-diamonds, echo = FALSE, fig.height = 4, fig.width = 5, out.width = "0.49\\textwidth", fig.cap = "Using \\pkg{colorspace} with \\pkg{ggplot2} graphics.
+Left: Kernel density of sepal length, grouped and shaded by species, in the \\code{iris} data with semi-transparent \\code{scale\\_fill\\_discrete\\_qualitative(palette = \"Dark 3\")} color scale.
+Right: Scatter plot of price by carat, shaded by cut levels, in a subsample of the \\code{diamonds} data with the \\code{scale\\_color\\_discrete\\_sequential(palette = \"Purples 3\", nmax = 6, order = 2:6)} color scale."}
 library("ggplot2")
@@
 
-For the qualitative `Dark 3` palette from above the following plots can be obtained.
+For the qualitative `"Dark 3"` palette from above the following plots can be obtained
+(see Figure\ \ref{fig:allplots-qualitative}).
 
@@
 
-```{r allplots-qualitative, echo = FALSE, fig.height = 4.5, fig.width = 14, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r allplots-qualitative, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for \\code{qualitative\\_hcl(4, \"Dark 3\")} palette.
+Left: Demo bar plot.
+Center: Hue-chroma plane at fixed $L = 60$ in HCL space.
+Right: HCL spectrum with linearly changing hue (around color wheel), almost constant chroma, and constant luminance."}
 allplots <- function(palette, ...) {
@@
 
-A bar plot would be another typical application for a qualitative palette (instead of the
-time series and density plot used above). However, a lighter and less colorful palette
-might be preferable in this situation (e.g., `Pastel 1` or `Set 3`).
-
-The other two displays show that luminance is (almost) constant in the palette while
-the hue changes linearly along the color "wheel". Ideally, chroma would have also been
-constant to completely balance the colors. However, at this luminance the maximum
-chroma differs across hues so that the palette is fixed up to use less chroma for
-the yellow and green elements.
-
-Subsequently, the same assessment is carried out for the sequential `Purples 3` palette
+The bar plot is used as a typical application for a qualitative palette (in addition to the
+time series and density plots used above). The other two displays show that luminance
+is (almost) constant in the palette while the hue changes linearly along the color
+"wheel". Ideally, chroma would have also been constant to completely balance the colors.
+However, at this luminance the maximum chroma differs across hues so that the palette
+is fixed up to use less chroma for the yellow and green elements.
+
+Note also that in a bar plot areas are shaded (and not just points or lines) so that
+lighter colors would be preferable. In the density plot in Figure\ \ref{fig:iris-diamonds}
+this was achieved through semi-transparency. Alternatively, luminance could be increased
+as is done in the `"Pastel 1"` or `"Set 3"` palettes.
+
+Subsequently, the same types of assessment are carried out for the sequential `"Purples 3"` palette
 as employed above.
@@
 
-```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for \\code{sequential\\_hcl(4, \"Purples 3\")} palette.
+Left: Demo heatmap.
+Center: Chroma-luminance plane at fixed $H = 270$ in HCL space.
+Right: HCL spectrum with constant hue, triangular chroma, and increasting luminance."}
 s9 <- sequential_hcl(9, "Purples 3")
@@
 
-At the core of the \pkg{colorspace} package there are various utilities for computing with
-color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
-representations of color to each other, see @color:Ihaka:2003. A particularly important
-mapping is the one from the perceptually-based and device-independent color model
-HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
-specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
-in R). For completeness further standard color models are included as well in the package.
-The connections are illustrated in the following graph:
-
-```{r color-spaces, fig.width = 9.5, fig.height = 4.5, echo = FALSE, out.width = "0.9\\textwidth", fig.cap = "Foo."}
+```{r color-spaces, fig.width = 9.5, fig.height = 4.5, echo = FALSE, out.width = "0.9\\textwidth", fig.cap = "Relationship between three-dimensional color spaces implemented in \\pkg{colorspace}.
+Color models that are (or try to be) perceptually-based are displayed with circles, other color models with rectangles."}
 par(mar = rep(0, 4))
@@
 
+At the core of the \pkg{colorspace} package there are various utilities for computing with
+color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
+representations of color to each other, see @color:Ihaka:2003. A particularly important
+mapping is the one from the perceptually-based and device-independent color model
+HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
+specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
+in \proglang{R}). For completeness further standard color models are included as well in the package.
+The connections are illustrated in Figure\ \ref{fig:color-spaces}.
 Color models that are (or try to be) perceptually-based are displayed with circles,
-other color models with rectangles. The corresponding classes and eponymous class
-constructors in \pkg{colorspace} are:
+other color models with rectangles.
+
+## Implemented color spaces
+
+The color spaces, implemented in \pkg{colorspace}, along with corresponding \proglang{S}4 classes
+and eponymous class constructors are:
 
@@
   colors, hence it was widely adopted and still is the basis for color specifications
-  in many systems. For example, the hex color codes are employed in HTML but also in R.
+  in many systems. For example, the hex color codes are employed in HTML but also in \proglang{R}.
   However, the RGB model also has some important drawbacks: It does not take into
@@
   unintuitive for humans to specify colors (say brown or pink) in this space.
-  [See @color:Wiki+RGB.]
+  See @color:Wiki+RGB for more details.
 * `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
@@
   Therefore, sRGB is a good place to end up in a color space manipulation but
-  it is not a good place to start. [See @color:Wiki+sRGB.]
+  it is not a good place to start. See @color:Wiki+sRGB for more details.
 * `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
   perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
-  (colorfulness), and _value_ (brightness, i.e., light vs. dark). Unfortunately,
+  (colorfulness), and _value_ (brightness, i.e., light vs.\ dark). Unfortunately,
   the three axes in the HSV model are confounded so that, e.g., brightness changes
   dramaticaly with hue.
-  [See @color:Wiki+HSV.]
+  See @color:Wiki+HSV for more details.
 * `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
@@
   are still strongly confounded.
-  [See @color:Wiki+HSV.]
+  See @color:Wiki+HSV for more details.
 * `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
-  based on experiments with human subjects. It provides a unique triplet of
+  based on psychophysical experiments with human subjects. It provides a unique triplet of
   XYZ values, coding the standard observer's perception of the color. It is
@@
   have no intuitive meaning.
-  [See @color:Wiki+CIEXYZ.]
+  See @color:Wiki+CIEXYZ for more details.
 * `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
@@
   still do not correspond to human perceptual axes.
-  [See @color:Wiki+CIELUV; @color:Wiki+CIELAB.]
+  See @color:Wiki+CIELUV; @color:Wiki+CIELAB for more details.
 * `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
@@
 
-All the S4 classes listed above inherit from a virtual class `color` which are
+All the \proglang{S}4 classes listed above inherit from a virtual class `color` which are
 internally always represented by matrices with three columns (corresponding to
@@
 
-Note that since the inception of the color space conversion tools within \pkg{colorspace}
-by @color:Ihaka:2003 other R tools for this purpose became available. Notably,
-there is `grDevices::convertColor()` in base R [@color:R] as well as
-`farver::convert_colour()` [@color:farver] based on a C++ library. For many basic color
-conversion purposes the \pkg{colorspace} package and these alternatives are equally suitable.
-Some details differ, though, and the color conversion tools in \pkg{colorspace} are
-well-embedded into the remaining functionality of the package and make it self-contained.
+Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
+\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
+Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
+[computed in high-level \proglang{R}, @color:R] and
+`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
+conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
+though [see the discussion in @color:convertColor]. Moreover, base \proglang{R} also provides
+`grDevices::hcl()` for mapping HCL representations to hex codes.
+
+To make the package self-contained and exactly backward compatible the \proglang{C}
+code in \pkg{colorspace} is still used as the basis for all color space conversions.
 
@@
 
-For working with these color spaces various utilities are available:
+For working with the implemented \proglang{S}4 color spaces various utilities are available:
 
@@
 * `hex()`: Convert a color object to sRGB and code in a hex string that can be
-  used within R plotting functions.
+  used within \proglang{R} plotting functions.
 * `hex2RGB()`: Convert a given hex color string to an sRGB color object which
@@
 
-For display in many systems (including R itself) hex color codes based on the
+For display in many systems (including \proglang{R} itself) hex color codes based on the
 sRGB coordinates can be created:
@@
 
-```{r hcl-palettes-principles, echo = FALSE, fig.width = 12, fig.height = 1.6, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r hcl-palettes-principles, echo = FALSE, fig.width = 12, fig.height = 1.6, fig.cap = "Examples of palette types in \\pkg{colorspace}.
+Qualitative palettes are balanced towards the same luminance level while sequential and diverging palettes go from dark to light and/or vice versa, respectively."}
 swatchplot(
@@
 
-More details about the construction of such palettes is provided in the following
-while the article on [Palette Visualization and Assessment](palette_visualization.html)
-introduces further tools to better understand the properties of color palettes.
-
 To facilitate obtaining good sets of colors, HCL parameter combinations that yield
@@
 
-To inspect the HCL parameter combinations for a specific palette simply include the `name`
-where upper- vs. lower-case, spaces, etc. are ignored for matching the label, i.e.,
+To inspect the HCL parameter combinations for a specific palette simply include the `palette` name
+where upper- vs.\ lower-case, spaces, etc.\ are ignored for matching the label, i.e.,
 `"set2"` matches `"Set 2"`:
@@
 
-To compute the actual color hex codes (representing sRGB coordinates), the functions
+To compute the actual color hex codes (representing sRGB coordinates) based on these parameters, the functions
 `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`, respectively, can be used.
@@
 
-`qualitative_hcl()` distinguishes the underlying categories by a sequence of
+As suggested by @color:Ihaka:2003 `qualitative_hcl()` distinguishes the underlying categories by a sequence of
 hues while keeping both chroma and luminance constant to give each color in the
@@
 ending hue of the palette. Then, an equidistant sequence between these hues is
-employed, by default spanning the full color wheel (i.e, the full 360 degrees).
+employed, by default spanning the full color wheel (i.e., the full 360 degrees).
 Chroma `c` (or equivalently `c1`) and luminance `l` (or equivalently
@@
 
-In the following graphic the available named palettes are shown. The first five
-palettes are close to the _ColorBrewer.org_ palettes of the same name [@color:Harrower+Brewer:2003].
+In Figure\ \ref{fig:hcl-palettes-qualitative} the available named palettes are shown. The first five
+palettes are close to the \pkg{ColorBrewer.org} palettes of the same name [@color:Harrower+Brewer:2003].
 They employ different levels of chroma and luminance and, by default, span the full
@@
 
-```{r hcl-palettes-qualitative, fig.width = 5.5, fig.height = 5.0, out.width = "0.6\\textwidth", fig.cap = "Foo."}
+```{r hcl-palettes-qualitative, fig.width = 5.5, fig.height = 5.0, out.width = "0.4\\textwidth", fig.cap = "Prespecified qualitative HCL palettes available in \\code{qualitative\\_hcl()} in \\pkg{colorspace}."}
 hcl_palettes("qualitative", plot = TRUE)
@@
 less distracting to use lighter colors (with moderate chroma and high luminance)
-such as "Pastel 1" or "Set 3". When coloring points or lines then more flashy
+such as `"Pastel 1"` or `"Set 3"`. When coloring points or lines then more flashy
 colors (with high chroma) are often required: On a white background a moderate
-luminance as in "Dark 2" or "Dark 3" usually works better while on a black/dark
-background the luminance should be higher as in "Set 2" for example.
+luminance as in `"Dark 2"` or `"Dark 3"` usually works better while on a black/dark
+background the luminance should be higher as in `"Set 2"` for example.
+Some examples with demo graphics are provided in Section\ \ref{sec:palette_visualization}.
 
@@
 
-`sequential_hcl()` codes the underlying numeric values by a monotonic sequence
+As suggested by @color:Zeileis+Murrell+Hornik:2009 `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
 of increasing (or decreasing) luminance. Thus, the `l` argument should provide
@@
 `l2` can be used). Without chroma (i.e., `c = 0`), this simply corresponds
-to a grayscale palette like `gray.colors()`, see "Grays" and "Light Grays" below.
+to a grayscale palette like `gray.colors()`, see `"Grays"` and `"Light Grays"`
+in Figure\ \ref{fig:hcl-palettes-sequentials}.
 
@@
 and then decrease chroma from some value (`c` or `c1`) to zero (i.e., gray) along with
-increasing luminance. For bringing out the extremes (a dark high-chroma color vs.
-a light gray) this is already very effective, see "Blues 2", "Purples 2", "Reds 2",
-and "Greens 2".
+increasing luminance. For bringing out the extremes (a dark high-chroma color vs.\ a
+light gray) this is already very effective, see `"Blues 2"`, `"Purples 2"`, `"Reds 2"`,
+and `"Greens 2"`.
 
@@
 length 3). This yields high-chroma colors in the middle of the palette that are
-more easily distinguished from the dark and light extremes. See "Blues 3", "Purples 3",
-"Reds 3", and "Greens 3" below.
+more easily distinguished from the dark and light extremes. See `"Blues 3"`, `"Purples 3"`,
+`"Reds 3"`, and `"Greens 3"` in Figure\ \ref{fig:hcl-palettes-sequentials}.
 
@@
 
-```{r hcl-palettes-sequentials, fig.width = 6.5, fig.height = 6.0, out.width = "0.7\\textwidth", fig.cap = "Foo."}
+```{r hcl-palettes-sequentials, fig.width = 6.5, fig.height = 6.0, out.width = "0.5\\textwidth", fig.cap = "Prespecified sequential single-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
 hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE)
@@
 
-All except the last palette are inspired by the _ColorBrewer.org_ palettes with
+All except the last palette are inspired by the \pkg{ColorBrewer.org} palettes with
 the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
@@
 palettes, including many that closely match other well-known color palettes.
-The plot below shows all the named multi-hue sequential palettes in \pkg{colorspace}:
-
-* "Purple-Blue" to "Terrain 2" are various palettes created during the development
+Figure\ \ref{fig:hcl-palettes-sequentialm} shows all the named multi-hue sequential palettes in \pkg{colorspace}:
+
+```{r hcl-palettes-sequentialm, fig.width = 8.5, fig.height = 9.0, fig.cap = "Prespecified sequential multi-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
+```
+
+* `"Purple-Blue"` to `"Terrain 2"` are various palettes created during the development
   of \pkg{colorspace}, e.g., by @color:Zeileis+Hornik+Murrell:2009 or @color:Stauffer+Mayr+Dabernig:2015
   among others.
-* "Viridis" to "Inferno" closely match the palettes that @color:Smith+VanDerWalt:2015
-  developed for _matplotlib_ and that gained popularity recently.
-* "Dark Mint" to "BrwnYl" closely match palettes provided in _CARTO_ [@color:CARTO].
-* "YlOrRd" to "Blues" closely match _ColorBrewer.org_ palettes [@color:Harrower+Brewer:2003].
-* "Lajolla" and "Turku" closely match the scientific color maps of the same name by
+* `"Viridis"` to `"Inferno"` closely match the palettes that @color:Smith+VanDerWalt:2015
+  developed for \pkg{matplotlib} and that gained popularity recently.
+* `"Dark Mint"` to `"BrwnYl"` closely match palettes provided in \pkg{CARTO} [@color:CARTO].
+* `"YlOrRd"` to `"Blues"` closely match \pkg{ColorBrewer.org} palettes [@color:Harrower+Brewer:2003].
+* `"Lajolla"` and `"Turku"` closely match the scientific color maps of the same name by
   @color:Crameri:2018 and are intended for a black/dark background.
-
-```{r hcl-palettes-sequentialm, fig.width = 8.5, fig.height = 9.0, out.width = "\\textwidth", fig.cap = "Foo."}
-hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
-```
 
@@
 contrasts, respectively. For example, many palettes go from a dark high-chroma color
-to a neutral low-chroma color (e.g., "Reds", "Purples", "Greens", "Blues") or even
-light gray (e.g., "Purple-Blue"). But some palettes also employ relatively high chroma
-throughout the palette (e.g., the viridis and many _CARTO_ palettes). To emphasize
+to a neutral low-chroma color (e.g., `"Reds"`, `"Purples"`, `"Greens"`, `"Blues"`) or even
+light gray (e.g., `"Purple-Blue"`). But some palettes also employ relatively high chroma
+throughout the palette (e.g., the viridis and many \pkg{CARTO} palettes). To emphasize
 the extremes the former strategy is typically more suitable while the latter works
@@
 
-The plot below shows all such diverging palettes that have been named in \pkg{colorspace}:
-
-* "Blue-Red" to "Cyan-Magenta" have been developed for \pkg{colorspace} starting from
+Figure\ \ref{fig:hcl-palettes-diverging} shows all such diverging palettes that have been named in \pkg{colorspace}:
+
+```{r hcl-palettes-diverging, fig.width = 6.5, fig.height = 7.5, out.width = "0.7\\textwidth", fig.cap = "Prespecified diverging HCL palettes available in \\code{diverging\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("diverging", n = 7, plot = TRUE)
+```
+
+* `"Blue-Red"` to `"Cyan-Magenta"` have been developed for \pkg{colorspace} starting from
   @color:Zeileis+Hornik+Murrell:2009, taking inspiration from various other palettes,
-  including more balanced and simplified versions of several _ColorBrewer.org_
+  including more balanced and simplified versions of several \pkg{ColorBrewer.org}
   palettes [@color:Harrower+Brewer:2003].
-* "Tropic" closely matches the palette of the same name from _CARTO_ [@color:CARTO].
-* "Broc" to "Vik" and "Berlin" to "Tofino" closely match the scientific color maps of the same name by
+* `"Tropic"` closely matches the palette of the same name from \pkg{CARTO} [@color:CARTO].
+* `"Broc"` to `"Vik"` and `"Berlin"` to `"Tofino"` closely match the scientific color maps of the same name by
   @color:Crameri:2018, where the first three are intended for a white/light background and
   the other three for a black/dark background.
-
-```{r hcl-palettes-diverging, fig.width = 6.5, fig.height = 7.5, out.width = "0.7\\textwidth", fig.cap = "Foo."}
-hcl_palettes("diverging", n = 7, plot = TRUE)
-```
 
@@
 
-The three different types of palettes (_qualitative_, _sequential_, and _diverging_)
-are all constructed by combining three different types of trajectories (_constant_, _linear_, _triangular_)
-for the three different coordinates (hue _H_, chroma _C_, luminance _L_):
+Table\ \ref{tab:hcl} summarizes which types of trajectories (_constant_, _linear_, _triangular_)
+are used for the three HCL coordinates (hue _H_, chroma _C_, luminance _L_) to construct
+the different types of palettes (_qualitative_, _sequential_, and _diverging_).
 
@@
 \hline
-Type          & H                                 & C                          & L                \\ \hline
+Type          & $H$                               & $C$                        & $L$              \\ \hline
 Qualitative   & Linear  		          & Constant	               & Constant	  \\
@@
 \end{tabular}
-\caption{\label{tab:hcl} Foo, see Equations~\ref{eq:constant}--\ref{eq:triangular}.}
+\caption{\label{tab:hcl} Types of trajectories used for the HCL coordinates to construct qualitative,
+sequential, and diverging palettes, see Equations~\ref{eq:constant}--\ref{eq:triangular}.}
 \end{table}
 
-As pointed out initially in this article, the luminance is probably the most important
+As emphasized in Figure\ \ref{fig:hcl-palettes-principles}, the luminance is probably the most important
 property for defining the type of palette. It is constant for qualitative palettes, monotonic
-for sequential palettes (linear or a power transformation), and uses two monotonic trajectories
+for sequential palettes (linear or a power transformation), or uses two monotonic trajectories
 (linear or a power transformation) diverging from the same neutral value.
@@
 The trajectories for the hue are also rather intuitive and straightforward for the three
-different types of palettes. However, the chroma trajectories are probably most complicated
+different types of palettes (constant vs.\ linear). However, the chroma trajectories are probably most complicated
 and least obvious from the examples above. Hence, the exact mathematical equations underlying
 the chroma trajectories are given in the following (i.e., using the parameters `c1`, `c2`, `cmax`,
-and `p1`, respectively). Analogous equations apply for the other two coordinates, though.
+and `p1`, respectively) and are depicted in Figure\ \ref{fig:hcl-trajectories}.
+Analogous equations apply for the other two coordinates, though.
 
@@
 Instead of using a linear intensity $i$ going from $1$ to $0$, one can replace $i$ with $i^{p_1}$
-in the equations above. This then leads to power-transformed curves that add or remove chroma
+in Equations\ \ref{eq:constant}--\ref{eq:triangular}. This then leads to power-transformed curves that add or remove chroma
 more slowly or more quickly depending on whether the power parameter $p_1$ is $< 1$ or $> 1$.
 
-The three types of trajectories are also depicted below. Note that full intensity $i = 1$ is on
-the left and zero intensity $i = 0$ on the right of each panel.
-
-```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, fig.cap = "Foo."}
 mylab <- function(x, y, text, ratio = 1, col = "black", ...) {
@@
 
-The concrete parameters in the plot above are:
+The three types of trajectories are also depicted in Figure\ \ref{fig:hcl-trajectories}. Note that full intensity $i = 1$ is on
+the left and zero intensity $i = 0$ on the right of each panel. The concrete parameters are:
 
 * Constant: `c1 = 80`.
-* Linear: `c1 = 80`, `c2 = 10`, `p1 = 1` (solid) vs. `p1 = 1.6` (dashed).
-* Triangular: `c1 = 60`, `cmax = 80`, `c2 = 10`, `p1 = 1` (solid) vs. `p1 = 1.6` (dashed).
+* Linear: `c1 = 80`, `c2 = 10`, `p1 = 1` (solid) vs.\ `p1 = 1.6` (dashed).
+* Triangular: `c1 = 60`, `cmax = 80`, `c2 = 10`, `p1 = 1` (solid) vs.\ `p1 = 1.6` (dashed).
 
 Further discussion of these trajectories and how they can be visualized and assessed
-for a given color palette is provided in the article:
-[Palette Visualization and Assessment](palette_visualization.html).
+for a given color palette is provided in Section\ \ref{sec:palette_visualization}.
 
@@
 * The registration of a palette is only stored for the current session. When
-  R is restarted and/or the `colorspace` package reloaded, only the pre-defined
+  \proglang{R} is restarted and/or the `colorspace` package reloaded, only the pre-defined
   palettes from the package are available. Thus, to make a palette permanently
-  available a registration R code like
+  available a registration \proglang{R} code like
   `colorspace::qualitative_hcl(3, palette = "set2", l = 80, register = "myset")`
@@
 
-The plot below shows all such flexible diverging palettes that have been named in \pkg{colorspace}:
-
-* "ArmyRose" to "Tropic" closely match the palettes of the same name from _CARTO_ [@color:CARTO].
-* "PuOr" to "Spectral" closely match the palettes of the same name from _ColorBrewer.org_ [@color:Harrower+Brewer:2003].
+Figure\ \ref{fig:divergingx-palettes} shows all such flexible diverging palettes that have been named in \pkg{colorspace}:
 
@@
 ```
+
+* `"ArmyRose"` to `"Tropic"` closely match the palettes of the same name from \pkg{CARTO} [@color:CARTO].
+* `"PuOr"` to `"Spectral"` closely match the palettes of the same name from \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].
 
@@
 
-- _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
-- _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
-- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
-  as provided by R package _viridis_ [@color:viridis].
-- The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].
+- \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
+- \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
+- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
+  as provided by \proglang{R} package \pkg{viridis} [@color:viridis].
+- The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].
 
@@
 
-To facilitate switching from base R palette functions to the HCL-based palettes
+To facilitate switching from base \proglang{R} palette functions to the HCL-based palettes
 above, \pkg{colorspace} provides a few convenience interfaces:
@@
   (along with increasing luminance). This is typically sufficient to clearly bring
-  out the extreme colors (dark/colorful vs. light gray).
+  out the extreme colors (dark/colorful vs.\ light gray).
 - Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
@@
 
-```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, fig.cap = "Foo."}
 bprg <- c("Blues", "Purples", "Reds", "Greens")
@@
 - Hue is drawn in red and coordinates are indicated on the axis on the right
-  with range [-360, 360].
+  with range $[-360, 360]$.
 - Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
@@
 
-Finally, the diverging "Green-Brown" palette is depicted below. It simply combines a green
-and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
+Finally, the diverging `"Green-Brown"` palette is depicted in the left panel of Figure\ \ref{fig:specplot}.
+It simply combines a green and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
 Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
-In the center the palette passes through a light gray (with zero chroma) as the neutral
-value.
-
-To contrast these well-balanced HCL-based palettes with a poorly-balanced palette, the
-spectrum of the (in)famous RGB rainbow palette is depicted in both RGB and HCL space.
+In the center the palette passes through a light gray (with zero chroma) as the neutral value.
+By including the corresponding RGB spectrum in the top panel, it also becomes apparent that
+choosing such well-balanced palettes through trajectories in RGB color space is not straightforward.
+This balanced palette -- based on relatively simple HCL trajectories -- is contrasted
+with a poorly-balanced palette -- based on simple linear RGB trajectories in the right panel of
+Figure\ \ref{fig:specplot}. This depicts the RGB and HCL spectrum of the 
+(in)famous RGB rainbow palette. [See @color:Hawkins+McNeall+Stephenson:2014 for a
+plea why the RGB rainbow palette should be avoided in almost all scientific graphics.]
 
@@
 if these palettes are constructed very differently (e.g., like the
-highly-saturated base R palettes). To infer the default `type` of projection
+highly-saturated base \proglang{R} palettes). To infer the default `type` of projection
 `hclplot()` assesses the luminance trajectory and sets the default correspondingly:
@@
 Thus, for qualitative palettes - where luminance and chroma are fixed - the varying
-hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance:
+hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance
+(Figure\ \ref{fig:hcl-qualitative}):
 
@@
 
-The display below compares three single-hue sequential palettes by projection
+Figure\ \ref{fig:hcl-sequential} compares three single-hue sequential palettes by projection
 to the luminance-chroma plane for the given fixed hue. In the left panel
@@
 
-```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
 par(mfrow = c(1, 3))
@@
 
-Finally, the display below compares two multi-hue sequential palettes along
+Finally, Figure\ \ref{fig:hcl-multi} compares two multi-hue sequential palettes along
 with a diverging palette.
 
-```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, out.width = "\\textwidth", fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.cap = "Foo."}
 par(mfrow = c(3, 3))
@@
 
-- Work best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs. many colors (e.g., heatmap, perspective, ...).
-- Intended for categorical data (e.g., bar, pie, ...) vs. continuous numeric data (e.g., heatmap, perspective, ...).
-- Shading areas (e.g., map, bar, pie, ...) vs. shading points or lines (scatter, lines).
+- Work best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs.\ many colors (e.g., heatmap, perspective, ...).
+- Intended for categorical data (e.g., bar, pie, ...) vs.\ continuous numeric data (e.g., heatmap, perspective, ...).
+- Shading areas (e.g., map, bar, pie, ...) vs.\ shading points or lines (scatter, lines).
 
@@
 
-```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
 par(mfrow = c(1, 3))
@@
 
-All displays above focused on palettes designed for light/white backgrounds.
-Therefore, to conclude, some palettes are highlighted that work well on
+Figures\ \ref{fig:demoplot-qualitative}--\ref{fig:demoplot-diverging} focused on palettes designed for light/white backgrounds.
+Therefore, to conclude, some palettes are highlighted in Figure\ \ref{fig:demoplot-dark} that work well on
 dark/black backgrounds.
 
-```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, out.width = "\\textwidth", fig.cap = "Foo."}
+```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.cap = "Foo."}
 par(mfrow = c(2, 3), bg = "black")
@@
 The workhorse function to do so is `simulate_cvd()` which can take any vector
-of valid R colors and transform them according to a certain CVD transformation
+of valid \proglang{R} colors and transform them according to a certain CVD transformation
 matrix and transformation equation. The transformation matrices have been established
@@
 For further guidance on color blindness in relation to statistical graphics
-see @color:Lumley:2006 which accompanies the R package _dichromat_ [@color:dichromat]
+see @color:Lumley:2006 which accompanies the \proglang{R} package \pkg{dichromat} [@color:dichromat]
 and is based on earlier emulation techniques
@@
 graphic for readers with color vision deficiencies, we employ the infamous RGB rainbow
-color palette in a heatmap. In base R this can be generated by `rainbow(11, end = 2/3)`
+color palette in a heatmap. In base \proglang{R} this can be generated by `rainbow(11, end = 2/3)`
 ranging from red (for high values) to blue (for low values).
@@
 The statistical graphic employed for illustration is a heatmap of the well-known
-Maunga Whau `volcano` data from base R. This heatmap is easily available as
+Maunga Whau `volcano` data from base \proglang{R}. This heatmap is easily available as
 `demoplot(..., "heatmap")` where `...` is the color vector to be used, e.g.,
@@
 several graphical user interfaces (GUIs) are provided within the package as
-_shiny_ apps [@color:shiny]. All of these GUIs or apps can either be run locally
-from within R but are also provided at <http://hclwizard.org/>.
+\pkg{shiny} apps [@color:shiny]. All of these GUIs or apps can either be run locally
+from within \proglang{R} but are also provided at <http://hclwizard.org/>.
 
@@
 
-In addition to the _shiny_ app, the **palette constructor** app is also available as a Tcl/Tk GUI via R package
-_tcltk_ shipped with base R [@color:R]. The _tcltk_ version can only be run locally
-and is considerably faster while the _shiny_ version has a nicer interface with more
+In addition to the \pkg{shiny} app, the **palette constructor** app is also available as a Tcl/Tk GUI via \proglang{R} package
+\pkg{tcltk} shipped with base \proglang{R} [@color:R]. The \pkg{tcltk} version can only be run locally
+and is considerably faster while the \pkg{shiny} version has a nicer interface with more
 features  and can be run online. The `choose_palette()` function by default starts the
-_tcltk_ version while `hclwizard()`/`hcl_wizard()` by default start the _shiny_
+\pkg{tcltk} version while `hclwizard()`/`hcl_wizard()` by default start the \pkg{shiny}
 version.
@@
 This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
-by default starts the _shiny_ version.
-The _tcltk_ version is started by default with `choose_palette()`.
+by default starts the \pkg{shiny} version.
+The \pkg{tcltk} version is started by default with `choose_palette()`.
 However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.
 
-```{r choose_palette, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
+```{r choose_palette, echo = FALSE}
 knitr::include_graphics("../../vignettes/screenshots/choose_palette_1.png")
@@
 
-```{r choose_color, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
+```{r choose_color, echo = FALSE, fig.cap="Foo."}
 knitr::include_graphics("../../vignettes/screenshots/choose_color.png")
@@
 By repeating the selection a palette of colors can be constructed and returned
-within R for subsequent usage in visualizations.
+within \proglang{R} for subsequent usage in visualizations.
 
@@
 Finally, the resulting HCL coordinates are transformed back to hex color codes
-for use in R.
-
-For illustration, a few simple examples are presented below. More examples in the
+for use in \proglang{R}.
+
+For illustration, a few simple examples are presented in Figure\ \ref{fig:desaturate-named}. More examples in the
 context of palettes for statistical graphics are discussed along with the
@@
 
-First, `desaturate()` is used to desaturate a vector of R color names:
+First, `desaturate()` is used to desaturate a vector of \proglang{R} color names:
 
@@
 Analogously, hex color codes can also be transformed - in this case RGB rainbow
-colors from the base R function `rainbow()`:
+colors from the base \proglang{R} function `rainbow()`:
 

### 2019-02-26 pkg/colorspace/paper/Toolbox/article.Rmd
first draft of paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-## Usage with ggplot2
+## Usage with ggplot2 {#sec:ggplot2}
 
@@
 
-Subsequently, the same types of assessment are carried out for the sequential `"Purples 3"` palette
-as employed above.
+Subsequently, the same types of assessment are carried out in Figure\ \ref{fig:allplots-sequential}
+for the sequential `"Purples 3"` palette as employed above.
 
@@
 
-```{r hcl-palettes-qualitative, fig.width = 5.5, fig.height = 5.0, out.width = "0.4\\textwidth", fig.cap = "Prespecified qualitative HCL palettes available in \\code{qualitative\\_hcl()} in \\pkg{colorspace}."}
-hcl_palettes("qualitative", plot = TRUE)
+```{r hcl-palettes-qualitative, fig.width = 8.0, fig.height = 2.5, out.width = "0.8\\textwidth", fig.cap = "Prespecified qualitative HCL palettes available in \\code{qualitative\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("qualitative", plot = TRUE, nrow = 5)
 ```
@@
 
-```{r hcl-palettes-sequentials, fig.width = 6.5, fig.height = 6.0, out.width = "0.5\\textwidth", fig.cap = "Prespecified sequential single-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
-hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE)
+```{r hcl-palettes-sequentials, fig.width = 8.0, fig.height = 3.0, out.width = "0.8\\textwidth", fig.cap = "Prespecified sequential single-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE, nrow = 6)
 ```
@@
 
-```{r hcl-palettes-diverging, fig.width = 6.5, fig.height = 7.5, out.width = "0.7\\textwidth", fig.cap = "Prespecified diverging HCL palettes available in \\code{diverging\\_hcl()} in \\pkg{colorspace}."}
-hcl_palettes("diverging", n = 7, plot = TRUE)
+```{r hcl-palettes-diverging, fig.width = 8.0, fig.height = 4.75, out.width = "0.8\\textwidth", fig.cap = "Prespecified diverging HCL palettes available in \\code{diverging\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("diverging", n = 7, plot = TRUE, nrow = 10)
 ```
@@
 
-```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, fig.cap = "Foo."}
+```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, fig.cap = "Types of trajectories to construct HCL color palettes, exemplified for the chroma coordinates, see Equations~\\ref{eq:constant}--\\ref{eq:triangular}."}
 mylab <- function(x, y, text, ratio = 1, col = "black", ...) {
@@
 box()
-lines(1 - i, c2 - (c2 - c1) * i^1.6, lty = 2, lwd = 2)
+lines(1 - i, c2 - (c2 - c1) * i^1.6, lwd = 2, col = gray(0.5))
 mylab(0.06, c1 + 5, "c1")
 mylab(0.94, c2 - 5, "c2")
-mylab(0.68, 40, "p1", col = gray(0.35))
-text(0.78, 40, "= 1", col = gray(0.35))
-mylab(0.18, 40, "p1", col = gray(0.35))
-text(0.30, 40, "= 1.6", col = gray(0.35))
+mylab(0.68, 40, "p1", col = gray(0.5))
+text(0.78, 40, "= 1", col = gray(0.5))
+mylab(0.18, 40, "p1", col = gray(0.5))
+text(0.30, 40, "= 1.6", col = gray(0.5))
 
@@
 box()
-lines(1 - i, ci2, lwd = 2, lty = 2)
+lines(1 - i, ci2, lwd = 2, col = gray(0.5))
 mylab(0.06, c1 - 5, "c1")
@@
 mylab(0.94, c2 - 5, "c2")
-mylab(0.77, 40, "p1", col = gray(0.35))
-text(0.87, 40, "= 1", col = gray(0.35))
-mylab(0.27, 40, "p1", col = gray(0.35))
-text(0.39, 40, "= 1.6", col = gray(0.35))
+mylab(0.77, 40, "p1", col = gray(0.5))
+text(0.87, 40, "= 1", col = gray(0.5))
+mylab(0.27, 40, "p1", col = gray(0.5))
+text(0.39, 40, "= 1.6", col = gray(0.5))
 ```
@@
 
-The `hcl_palettes()` already come with a wide range of pre-defined palettes to which
+The `hcl_palettes()` already come with a wide range of predefined palettes to which
 customizations can be easily added. However, it might also be convenient to register
-a custom palette so that it can subsequently be re-used with a new dedicated name.
+a custom palette so that it can subsequently be reused with a new dedicated name.
 This is supported by adding a `register = "..."` argument once to a call to
@@
 And can be used subsequently in `qualitative_hcl()` as well as the
-qualitative [ggplot2 color scales](ggplot2_color_scales.html), e.g.,
+qualitative \pkg{ggplot2} color scales (see Section\ \ref{sec:ggplot2}), e.g.,
 
@@
   previously, the old palette gets overwritten. We recommend not to overwrite
-  the palettes that are pre-defined in the package (albeit it is technically possible).
+  the palettes that are predefined in the package (albeit it is technically possible).
 * The registration of a palette is only stored for the current session. When
-  \proglang{R} is restarted and/or the `colorspace` package reloaded, only the pre-defined
+  \proglang{R} is restarted and/or the \pkg{colorspace} package reloaded, only the predefined
   palettes from the package are available. Thus, to make a palette permanently
@@
 
-```{r divergingx-palettes, fig.width = 6.5, fig.height = 8.0, out.width = "0.7\\textwidth", fig.cap = "Foo."}
-divergingx_palettes(n = 7, plot = TRUE)
+```{r divergingx-palettes, fig.width = 8.0, fig.height = 4.25, out.width = "0.8\\textwidth", fig.cap = "Prespecified flexible diverging HCL palettes available in \\code{divergingx\\_hcl()} in \\pkg{colorspace}."}
+divergingx_palettes(n = 7, plot = TRUE, nrow = 9)
 ```
@@
 
-- \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
-- \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
-- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
+* \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
+  See `demo("brewer", package = "colorspace")`.
+* \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
+  See `demo("carto", package = "colorspace")`.
+* The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
   as provided by \proglang{R} package \pkg{viridis} [@color:viridis].
-- The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].
-
-See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
-following `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
+  See `demo("viridis", package = "colorspace")`.
+* The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].
+  See `demo("scico", package = "colorspace")`.
+
+The graphics resulting from the demos can also be viewed online at
+<http://colorspace.R-Forge.R-project.org/articles/approximations.html>.
+
+In Figure\ \ref{fig:brewer-carto-viridis} `specplot()` panels (see also
+Section\ \ref{sec:specplot}) are shown for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
+and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
+Each panel compares the hue, chroma, and luminance trajectories
 of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
-
-```{r brewer-carto-viridis, results = "hide", echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
+palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
+admissible RGB colors.
+
+Furthermore, Figure\ \ref{fig:brewer-carto-viridis} illustrates what sets the viridis palettes apart from
+other sequential palettes. While the hue and luminance trajectories of `"Viridis"` and `"YlGnBu"` are very similar,
+the chroma trajectories differ: While lighter colors (with high luminance) have low chroma for `"YlGnBu"`, they have
+increasing chroma for `"Viridis"`. Similarly, `"ag_Sunset"` and `"Plasma"` have similar hue and luminance trajectories
+but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
+does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
+However, they work better on a dark/black background (see Figure\ \ref{fig:demoplot-dark}). Also, they might be a reasonable
+alternative for qualitative palettes when grayscale printing should also work.
+
+```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatch, solid lines) along with their HCL-based approximations (bottom swatch, dashed lines)."}
 specplot(
@@
 
-## Color swatches
+## Color swatches {#sec:swatchplot}
 
@@
 spacings, borders, etc. These are selected to work well for `hcl_palettes()`
-and might need further tweaking in future versions.
-
-As a first simple example:
-`swatchplot("Label" = sequential_hcl(5))`.
-
-As a somewhat more complex example a `swatchplot()` is created with three matrices
+and might need further tweaking in future versions. Thus,
+Figures\ \ref{fig:hcl-properties}--\ref{fig:hcl-palettes} as well as
+Figures\ \ref{fig:hcl-palettes-principles}--\ref{fig:hcl-palettes-diverging}
+all use `swatchplot()` internally. For a simple stand-alone illustration
+consider: `swatchplot("Palette" = sequential_hcl(5))`.
+
+As a somewhat more complex example, a `swatchplot()` is created with three matrices
 of sequential color palettes of blues, purples, reds, and greens. For all palettes,
@@
 
-- Single-hue: In each palette the hue is fixed and chroma decreases monotonically
+* Single-hue: In each palette the hue is fixed and chroma decreases monotonically
   (along with increasing luminance). This is typically sufficient to clearly bring
   out the extreme colors (dark/colorful vs.\ light gray).
-- Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
+* Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
   triangular. Compared to the basic single-hue palette above this allows to better
   distinguish the colors in the middle and not only the extremes.
-- Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
+* Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
   is triangular but additionally the hue varies slightly. This can further enhance
@@
 
-```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, fig.cap = "Foo."}
+```{r swatch-brpg, fig.width = 11.5, fig.height = 3, fig.cap = "Variations of blue, purple, red, and green palettes with single hue and monotonic chroma (left), single hue and triangular chroma (center), and multiple hues and triangular chroma (right)."}
 bprg <- c("Blues", "Purples", "Reds", "Greens")
@@
 "Multi-hue (advanced)"  = t(sapply(bprg,	   sequential_hcl, n = 7)),
-  nrow = 5, line = 5
-)
-```
-
-## HCL (and RGB) spectrum
+nrow = 5, line = 5)
+```
+
+## HCL (and RGB) spectrum {#sec:specplot}
 
@@
 versa, the hue coordinates are shifted suitably.
-
 By default, the resulting trajectories in the HCL spectrum are visualized by a simple line plot:
@@
 
-FIXME: Point to previous plots
-
-Finally, the diverging `"Green-Brown"` palette is depicted in the left panel of Figure\ \ref{fig:specplot}.
+As spectrum plots have already been used for illustration in Figures\ \ref{fig:allplots-qualitative}
+(for a qualitative palette) as well as \ref{fig:allplots-sequential} and \ref{brewer-carto-viridis}
+(for sequential palettes), this section only provides a couple of additional illustrations.
+The diverging `"Green-Brown"` palette is depicted in the left panel of Figure\ \ref{fig:specplot}.
 It simply combines a green and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
-Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
+Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are rather balanced between both arms.
 In the center the palette passes through a light gray (with zero chroma) as the neutral value.
@@
 
-```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "Foo."}
+```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of the balanced diverging \\code{\"Green-Brown\"} palette (left panel) and the (in)famous and rather unbalanced \\code{rainbow()} palette (right panel)."}
 specplot(diverging_hcl(100, "Green-Brown"), rgb = TRUE)
@@
 
-The RGB spectrum shows that the trajectories are quite simple in RGB space but the
-HCL spectrum shows that the result is very unbalanced with shifts and kinks in both
-chroma and (more importantly) luminance. This is why this palette is not suitable
-for encoding underlying data in statistical graphics. See also the related discussion
-of [color vision deficiency](color_vision_deficiency.html).
-
-## Trajectories in HCL space
+The RGB spectrum of the rainbow palette shows that the trajectories are quite simple in RGB space but
+lead to substantial variations in chroma and (more importantly) luminance.
+This is why this palette is not suitable for encoding underlying data in statistical graphics.
+See also the related discussion of color vision deficiency in Section\ \ref{sec:color_vision_deficiency}.
+
+## Trajectories in HCL space {#sec:hclplot}
 
@@
 Therefore, the `hclplot()` is another visualization of the HCL coordinates associated
-with a palette. It does so by collapsing over one of the coordinates (either the hue H
-or the luminance L) and displays a heatmap of colors combining the remaining
+with a palette. It does so by collapsing over one of the coordinates (either the hue $H$
+or the luminance $L$) and displays a heatmap of colors combining the remaining
 two dimensions. The coordinates for the given color palette are highlighted
@@
 
-- `type = "qualitative"` if luminance is approximately constant.
-- `type = "sequential"` if luminance is monotonic.
-- `type = "diverging"` if luminance is diverging with two monotonic "arms" in the trajectory.
-
-Thus, for qualitative palettes - where luminance and chroma are fixed - the varying
+* `type = "qualitative"` if luminance is approximately constant.
+* `type = "sequential"` if luminance is monotonic.
+* `type = "diverging"` if luminance is diverging with two monotonic "arms" in the trajectory.
+
+Thus, for qualitative palettes -- where luminance and chroma are fixed -- the varying
 hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance
@@
 
-```{r hcl-qualitative, fig.width = 4, fig.height = 4, fig.cap = "Foo.", dev = "png", dpi = 150, out.width = "0.5\\textwidth"}
+```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with a qualitative palette with varying hue $H$ and chroma fixed at $C = 50$."}
 hclplot(qualitative_hcl(9, c = 50, l = 70))
@@
 
-```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes.
+Left: Linear chroma for $H = 260$.
+Center: Triangular chroma for $H = 245$.
+Right: Power-transformed triangular chroma for $H = 245$."}
 par(mfrow = c(1, 3))
 hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
-hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
-hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
-```
-
-Note that for H = 260 it is possible to go to dark colors (= low luminance) with
-high chroma while this is not possible to the same extent for H = 245. Hence,
+hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
+  power = 1))
+hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
+  power = c(0.8, 1.4)))
+```
+
+Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
+high chroma while this is not possible to the same extent for $H = 245$. Hence,
 chroma has to be decreased when proceeding to the dark low-luminance colors.
@@
 
-```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
+```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with blue multi-hue palette and triangular chroma (left), blue-yellow multi-hue palette and linear chroma (center), and diverging blue-red palette with balanced linear chroma."}
 par(mfrow = c(1, 3))
-hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
-hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
-hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
+hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0),
+  l = c(30, 95), power = 1))
+hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95),
+  power = 1))
+hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95),
+  power = 1))
 ```
@@
 
-## Demonstration of statistical graphics
-
-To demonstrate how different kinds of color palettes work in different
-kinds of statistical displays, `demoplot()` provides a simple convenience
-interface to some base graphics with (mostly artificial) data sets.
-As a first overview, all built-in demos are displayed with the same
-sequential heat colors palette: `sequential_hcl(5, "Heat")`.
-
-```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.cap = "Foo."}
+## Demonstration of statistical graphics {#sec:demoplot}
+
+```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.cap = "All built-in \\code{demoplot} types with the same \\code{sequential\\_hcl(5, \"Heat\")} palette."}
 par(mfrow = c(3, 3))
@@
 
+To demonstrate how different kinds of color palettes work in different
+kinds of statistical displays, `demoplot()` provides a simple convenience
+interface to some base graphics with (mostly artificial) data sets.
+As a first overview, Figure\ \ref{fig:demoplot-all} displays all built-in
+demos with the same sequential heat colors palette: `sequential_hcl(5, "Heat")`.
 All types of demos can, in principle, deal with arbitrarily many colors from any palette
@@
 
-```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
+```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{qualitative\\_hcl()} palettes."}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
+```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{sequential\\_hcl()} palettes."}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
+```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{diverging\\_hcl()} palettes."}
 par(mfrow = c(1, 3))
@@
 
-```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.cap = "Foo."}
+```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.cap = "Examples for \\code{demoplot()} with different palettes that work well on a black/dark background."}
 par(mfrow = c(2, 3), bg = "black")
@@
 and tritanope (blue cone cells defective).
+
+Below we briefly describe our \proglang{R} interface to these emulation techniques
+and show them in practice for a heatmap with sequential palette. Another example
+with diverging palette is available at
+<http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
+Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
+palette is almost always a bad choice in scientific displays. See
+<http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
+illustrations.
+
+## R functions
+
 The workhorse function to do so is `simulate_cvd()` which can take any vector
@@
 `deutanomaly_cvd`, and `tritanomaly_cvd`.
-
 The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
@@
 ranging from red (for high values) to blue (for low values).
-
 The poor results for the RGB rainbow palette are contrasted with a proper sequential
-palette ranging dark blue to light yellow: `sequential_hcl(11, "Blue-Yellow")`.
+palette ranging from dark blue to light yellow: `sequential_hcl(11, "Blue-Yellow")`.
 
@@
 and so on. To aid the interpretation of the heatmap a perspective display using
-only gray shades is included first, providing another intuitive display of what
+only gray shades is provided in Figure\ \ref{fig:volcano}, providing another intuitive display of what
 the terrain around Maunga Whau looks like.
 
-```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100, out.width = "0.5\\textwidth"}
+```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Perspective visualization of Maunga Whau \\code{volcano} data (Mount Eden, Auckland, New Zealand)."}
 z <- 2 * volcano
@@
 
-Subsequently, all combinations of palette and color vision deficiency are visualized.
-Additionally, a grayscale version is created with `desaturate()`.
-
-```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "0.85\\textwidth", fig.pos = "p!", fig.cap = "Foo."}
+```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "0.8\\textwidth", fig.pos = "p!", fig.cap = "Heatmap of Maunga Whau \\code{volcano} data with RGB rainbow (left) and HCL-based blue-yellow palette (right).
+The first row shows the original color palettes while subsequent rows emulate various color deficiencies."}
 par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
@@
 
+Subsequently, all combinations of palette and color vision deficiency are visualized.
+Additionally, a grayscale version is created with `desaturate()`.
 This clearly shows how poorly the RGB rainbow performs, often giving quite
@@
 several graphical user interfaces (GUIs) are provided within the package as
-\pkg{shiny} apps [@color:shiny]. All of these GUIs or apps can either be run locally
-from within \proglang{R} but are also provided at <http://hclwizard.org/>.
+\pkg{shiny} apps [@color:shiny]. All of these GUIs/apps can either be run locally
+from within \proglang{R} but are also provided online at <http://hclwizard.org/>.
 
@@
 
-In addition to the \pkg{shiny} app, the **palette constructor** app is also available as a Tcl/Tk GUI via \proglang{R} package
+In addition to the \pkg{shiny} version, the **palette constructor** app is also available
+as a \proglang{Tcl}/\proglang{Tk} GUI via \proglang{R} package
 \pkg{tcltk} shipped with base \proglang{R} [@color:R]. The \pkg{tcltk} version can only be run locally
@@
 
+```{r choose_palette, echo = FALSE, fig.cap = "App for interactively choosing HCL-based color palettes: \\code{choose\\_color()}/\\code{hclwizard()}."}
+knitr::include_graphics("../../vignettes/screenshots/choose_palette_1.png")
+```
+
 This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
-by default starts the \pkg{shiny} version.
+by default starts the \pkg{shiny} version (see Figure\ \ref{fig:choose_palette}).
 The \pkg{tcltk} version is started by default with `choose_palette()`.
@@
 
-```{r choose_palette, echo = FALSE}
-knitr::include_graphics("../../vignettes/screenshots/choose_palette_1.png")
-```
-
-The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
-`sequential_hcl()` for sequential palettes with single or
-multiple hues, and `diverging_hcl()` for diverging palettes (composed
-from two single-hue sequential palettes).
-
-The GUIs
+The GUIs interface the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
+palettes from Section\ \ref{sec:hcl_palettes}. They
 allow for interactive modification of the arguments of the respective
-palette-generating functions, i.e., starting/ending hue (wavelength, type of
-color), minimal/maximal chroma (colorfulness), minimal maximal luminance
-(brightness, amount of gray), and a power transformations that control how
+palette-generating functions, i.e., starting/ending hue, minimal/maximal chroma,
+minimal maximal luminance, and a power transformations that control how
 quickly/slowly chroma and/or luminance are changed through the palette.
 Subsets of the parameters may not be applicable depending on the type of
-palette chosen. See `qualitative_hcl()` and @color:Zeileis+Hornik+Murrell:2009 for
-a more detailed explanation of the different arguments. 
-@color:Stauffer+Mayr+Dabernig:2015 provide more examples and guidance.
-
-Optionally, the active palette can be illustrated by using a
-[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
-such as a map, heatmap, scatter plot, perspective 3D surface etc.
-To demonstrate different types of deficiencies, the active palette may be
-desaturated (emulating printing on a grayscale printer) and collapsed to
-emulate different types of color-blindness (without red-green or green-blue
-contrasts) using the `simulate_cvd()` functions. To facilitate generation of
-palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available.
+palette chosen.
+
+Optionally, the active palette can be illustrated by using a `specplot()` (see Section\ \ref{sec:specplot}),
+`hclplot()` (see Section\ \ref{sec:hclplot}), or `demoplot()`
+(see Section\ \ref{sec:demoplot}), and assessed using emulation
+of color vision deficiencies (see Section\ \ref{sec:color_vision_deficiency}).
+To facilitate generation of palettes for black/dark backgrounds, a "dark mode"
+of the GUIs is also available:
 
@@
 
+```{r choose_color, echo = FALSE, fig.cap = "App for interactively choosing individual colors in HCL space: \\code{choose\\_color()}/\\code{hcl\\_color\\_picker()}."}
+knitr::include_graphics("../../vignettes/screenshots/choose_color.png")
+```
+
 This GUI can be started with either `choose_color()` or equivalently `hcl_color_picker()`.
-
-```{r choose_color, echo = FALSE, fig.cap="Foo."}
-knitr::include_graphics("../../vignettes/screenshots/choose_color.png")
-```
-
 It shows the HCL color space either as a hue-chroma plane for a given luminance value or
@@
 ## Emulate color vision deficiencies
-
-This GUI can be started with `cvd_emulator()`.
 
@@
 \includegraphics[width=0.49\textwidth]{../../vignettes/screenshots/cvd_emulator_2}
-\caption[Foo]{Foo.}\label{fig:cvd_emulator}
+\caption{App for emulating color vision deficiencies for uploaded raster images: \code{cvd\_emulator()}.}\label{fig:cvd_emulator}
 \end{figure}
 
+This GUI can be started with `cvd_emulator()`.
 It allows to upload a raster image in
@@
 Thus, the `desaturate()` function converts any given hex color code or named
-R color to the corresponding HCL coordinates and sets the chroma to zero.
+\proglang{R} color to the corresponding HCL coordinates and sets the chroma to zero.
 Thus, only the luminance matters which captures the "brightness" mentioned above.
 Finally, the resulting HCL coordinates are transformed back to hex color codes
-for use in \proglang{R}.
-
-For illustration, a few simple examples are presented in Figure\ \ref{fig:desaturate-named}. More examples in the
-context of palettes for statistical graphics are discussed along with the
-[color vision deficiency](color_vision_deficiency.html) article.
-
-First, `desaturate()` is used to desaturate a vector of \proglang{R} color names:
+for use in \proglang{R}. First, `desaturate()` is used to desaturate a vector of \proglang{R} color names:
 
@@
 are always the same, indicating gray colors.
-
-Analogously, hex color codes can also be transformed - in this case RGB rainbow
+Analogously, hex color codes can also be transformed -- in this case RGB rainbow
 colors from the base \proglang{R} function `rainbow()`:
@@
 a full color wheel (of colors with hues in [0, 360] degrees). While the RGB
-`rainbow()` is very unbalanced the HCL `rainbow_hcl()` (or also `qualitative_hcl()`)
+`rainbow()` is very unbalanced, the HCL `rainbow_hcl()` (or also `qualitative_hcl()`)
 is (by design) balanced with respect to luminance.
 
-```{r desaturate-wheel, out.width = "0.5\\textwidth", fig.cap = "Foo."}
+```{r desaturate-wheel, out.width = "0.5\\textwidth", fig.cap = "Color wheels in RGB (left) and HCL (right) space in color (top) and desaturated grayscale (bottom)."}
 wheel <- function(col, radius = 1, ...)
@@
 wheel(rainbow(8))
+wheel(rainbow_hcl(8))
 wheel(desaturate(rainbow(8)))
-wheel(rainbow_hcl(8))
 wheel(desaturate(rainbow_hcl(8)))
@@
 (turning the color darker) or increased (turning it lighter) while preserving the hue and chroma coordinates.
-
 This strategy typically works well for lightening colors, although in some situations the result can be
@@
 (the default) and absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
-`L * (1 - amount)` (relative). See `lighhten()` and `darken()` for more details.
-
-For illustration a qualitative palette (Okabe-Ito) is transformed by two levels of both lightening and
-darkening, respectively.
-
-```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
+`L * (1 - amount)` (relative). See `?lighten` and `?darken` for more details.
+
+For illustration the qualitative palette suggested by @color:Okabe+Ito:2008 is transformed by two
+levels of both lightening and darkening, respectively.
+
+```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, out.width = "0.5\\textwidth", fig.cap = "Okabe-Ito palette (0\\%) along with two levels of both lightening and darkening, respectively."}
 oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
@@
 trying a specific HCL coordinate by calling `polarLUV()`. To avoid having to
-`fixup` the color upon conversion to RGB `hex` color codes, the
+`fixup` the color upon conversion to RGB `hex()` color codes, the
 `max_chroma()` function computes (approximately) the maximum chroma
 possible. 
-
-For illustration we show that for given luminance (here: L = 50) the maximum chroma
+For illustration we show that for given luminance (here: $L = 50$) the maximum chroma
 varies substantially with hue:
@@
 Similarly, maximum chroma also varies substantially across luminance values for
-a given hue (here: H = 120, green):
+a given hue (here: $H = 120$, green):
 
@@
 
-As usual...
-
-For \proglang{Python} users a beta re-implementation of the full \pkg{colorspace} package in
-\proglang{Python}\ 2/\proglang{Python}\ 3 is also available, see <https://github.com/retostauffer/python-colorspace>.
+This manuscript provides an overview of the broad capabilities of the
+\pkg{colorspace} package for selecting individual colors or color palettes,
+manipulating these colors, and employing them in various kinds of visualizations.
+
+In particular, the package provides various qualitative, sequential, and
+diverging palettes derived by relatively simple trajectories in HCL
+(Hue-Chroma-Luminance) space. In contrast to many other packages providing
+modern balanced color palettes (such as \pkg{ColorBrewer.org}, \pkg{CARTO},
+\pkg{viridis}, or \pkg{scico}) special emphasis is given to flexibility of
+the palettes that can be adjusted to the particular needs of a given
+data visulization. The manuscript already provided various tips and tricks
+regarding factors that are relevant in choosing an effective palette
+in a given situation. Further useful is provided in many sources, including
+@color:Ware:1988, @color:Okabe+Ito:2008, @color:Aigner:2010,
+@color:Stauffer+Mayr+Dabernig:2015, @color:Zhang:2015, @color:Rost:2018,
+@color:Wilke:2019, and @color:Ciechanowski:2019.
+
+In addition to the \proglang{R} version of \pkg{colorspace}, a
+\proglang{Python}\ 2/\proglang{Python}\ 3 re-implementation is in beta
+and available at <https://github.com/retostauffer/python-colorspace>.
 In the manuscript we focus on the more mature \proglang{R} implementation but
@@
 <https://CRAN.R-project.org/>.
-
-# Acknowledgments {#acknowledgments .unnumbered}
-
-All acknowledgments (note the AE spelling) should be collected in this
-unnumbered section before the references. It may contain the usual
-information about funding and feedback from colleagues/reviewers/etc.
-Furthermore, information such as relative contributions of the authors
-may be added here (if any).

### 2019-02-26 pkg/colorspace/paper/Toolbox/article.Rmd
minor edits
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@458 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   To aid selection and application of these palettes the package provides scales
-  for use with \pkg{ggplot2}; \pkg{shiny} (and \pkg{tcltk}) apps for interactive
-  exploration; visualizations of palette properties;
+  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
+  exploration, visualizations of palette properties,
   accompanying manipulation utilities (like desaturation and lighten/darken), and
@@
 their hue/chroma/luminance parameters can be modified, thus allowing to easily customize
-each palette. Moreover, the `choose_palette()`/`hclwizard()` app provide convenient user
+each palette. Moreover, the `choose_palette()`/`hclwizard()` app provides convenient user
 interfaces to do the customization interactively. Finally, even more flexible diverging
@@
 To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
-and sequential `"Purples 3"` palettes that were also employed above. First, semi-transparent
+and sequential `"Purples 3"` palettes that were also employed above
+(see Figure\ \ref{fig:iris-diamonds}). First, semi-transparent
 shaded densities of the sepal length from the `iris` data are shown, grouped by species.
@@
 As spectrum plots have already been used for illustration in Figures\ \ref{fig:allplots-qualitative}
-(for a qualitative palette) as well as \ref{fig:allplots-sequential} and \ref{brewer-carto-viridis}
+(for a qualitative palette) as well as \ref{fig:allplots-sequential} and \ref{fig:brewer-carto-viridis}
 (for sequential palettes), this section only provides a couple of additional illustrations.

### 2019-02-26 pkg/colorspace/paper/Toolbox/article.Rmd
added more references to other color-related packages -> more intro/discussion still needed (marked with FIXME)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@460 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+FIXME: More comparison with/discussion of other palette packages.
+the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
+as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
+the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
+package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
+and can be approximated reasonably well in \pkg{colorspace}
+(see Section\ \ref{sec:approximations}).
+A broad range of packages that provide color palettes that are "in harmony" as
+they are extracted and clustered from movies, paintings, etc. However, many of these
+are not so suitable for scientific visualization as they are not clearly 
+qualitative/sequential/diverging. An influential package following such an
+approach is \pkg{weanderson} [@color:wesanderson]. The \pkg{paletteer}
+[@color:paletteer] package aggregates these and provides them in a unified
+user interface.
+Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
+coordinates, along with visualization and assessment of palettes (beyond
+color swatches).
+
 # A quick tour {#sec:tour}
@@
 
-## Approximating palettes from other packages
+## Approximating palettes from other packages {#sec:approximations}
 
@@
 
+FIXME: Discuss relation ship to other packages:
+
+* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories".
+* \pkg{roloc} [@color:roloc]: naming palettes.
+* \pkg{colorscience} [@color:colorscience]: More color mappings.
+* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.
+
 In addition to the \proglang{R} version of \pkg{colorspace}, a

### 2019-03-04 pkg/colorspace/paper/Toolbox/article.Rmd
added civids palette to divergingx_hcl and vignettes/article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@461 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 qualitative/sequential/diverging. An influential package following such an
-approach is \pkg{weanderson} [@color:wesanderson]. The \pkg{paletteer}
+approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
 [@color:paletteer] package aggregates these and provides them in a unified
@@
 
-## Flexible diverging palettes
+## Flexible diverging palettes {#sec:divergingx}
 
@@
 * `"PuOr"` to `"Spectral"` closely match the palettes of the same name from \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].
+* `"Cividis"` closely matches the palette of the same name from the \pkg{viridis} family [@color:viridis].
+  Note that despite having two "arms" with blue vs.\ yellow colors and a low-chroma center color,
+  this is probably better classified as a sequential palette due to the monotonic chroma going from dark to light.
+  (See Section\ \ref{sec:approximations} for more details.)
 
@@
 
-In Figure\ \ref{fig:brewer-carto-viridis} `specplot()` panels (see also
-Section\ \ref{sec:specplot}) are shown for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
-and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
-Each panel compares the hue, chroma, and luminance trajectories
-of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
-The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
-palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
-admissible RGB colors.
-
-Furthermore, Figure\ \ref{fig:brewer-carto-viridis} illustrates what sets the viridis palettes apart from
-other sequential palettes. While the hue and luminance trajectories of `"Viridis"` and `"YlGnBu"` are very similar,
-the chroma trajectories differ: While lighter colors (with high luminance) have low chroma for `"YlGnBu"`, they have
-increasing chroma for `"Viridis"`. Similarly, `"ag_Sunset"` and `"Plasma"` have similar hue and luminance trajectories
-but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
-does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
-However, they work better on a dark/black background (see Figure\ \ref{fig:demoplot-dark}). Also, they might be a reasonable
-alternative for qualitative palettes when grayscale printing should also work.
-
-```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatch, solid lines) along with their HCL-based approximations (bottom swatch, dashed lines)."}
+```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatches, solid lines) along with their HCL-based approximations (bottom swatches, dashed lines)."}
 specplot(
@@
 ```
+
+Figure\ \ref{fig:brewer-carto-viridis} shows a selection of such approximations using `specplot()` 
+(see also Section\ \ref{sec:specplot}) for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
+and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
+Each panel compares the hue, chroma, and luminance trajectories
+of the original palettes (top swatches, solid lines) and their HCL-based approximations (bottom swatches, dashed lines).
+The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
+palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
+admissible RGB colors.
+
+Furthermore, Figure\ \ref{fig:brewer-carto-viridis} illustrates what sets the viridis palettes apart from
+other sequential palettes. While the hue and luminance trajectories of `"Viridis"` and `"YlGnBu"` are very similar,
+the chroma trajectories differ: While lighter colors (with high luminance) have low chroma for `"YlGnBu"`, they have
+increasing chroma for `"Viridis"`. Similarly, `"ag_Sunset"` and `"Plasma"` have similar hue and luminance trajectories
+but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
+does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
+However, they work better on a dark/black background (see Figure\ \ref{fig:demoplot-dark}). Also, they might be a reasonable
+alternative for qualitative palettes when grayscale printing should also work.
+
+```{r cividis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of \\texttt{viridis::cividis} (top swatch, solid lines) along with an HCL-based approximation (bottom swatch, dashed lines)."}
+specplot(viridis::cividis(9), divergingx_hcl(9, "Cividis"), main = "viridis: Cividis")
+```
+
+Another somewhat nonstandard palette from the viridis family is the
+cividis palette based on blue and yellow hues and hence safe for red-green deficient viewers.
+Figure\ \ref{fig:cividis} shows the corresponding `specplot()` along with an HCL-based approximation.
+What is unusual about this palette: The hue and chroma trajectories would suggest a diverging palette,
+as there are two "arms" wth different hues and a zero-chroma point in the center. However,
+the luminance trajectory clearly indicates a sequential palette as colors go monotonically from
+dark to light. Due to this unusual mixture the palette cannot be composed using the trajectories from
+Table\ \ref{tab:hcl}.
+
+However, the tools in \pkg{colorspace} can still be employed to easily reconstruct the palette.
+One strategy would be to set up the trajectories manually, using a linear luminance, piecewise
+linear chroma, and piecewise constant hue:
+
+```{r cividis_hcl}
+cividis_hcl <- function(n) {
+  i <- seq(1, 0, length.out = n)
+  hex(polarLUV(
+    L = 92 - (92 - 13) * i,
+    C = approx(c(1, 0.9, 0.5, 0), c(30, 50, 0, 95), xout = i)$y,
+    H = c(255, 75)[1 + (i < 0.5)]
+  ), fix = TRUE)
+}
+```
+
+Instead of constructing the hex code from the HCL coordinates via \pkg{colorspace}'s `hex(polarLUV(L, C, H))`,
+the base \proglang{R} function `hcl(H, C, L)` from \pkg{grDevices} could also be used.
+
+In addition to manually setting up a dedicated function `cividis_hcl()`, it is possible to approximate the
+palette using `divergingx_hcl()` (see Section\ \ref{sec:divergingx}), e.g.,
+
+```{r diverging_cividis, eval = FALSE}
+divergingx_hcl(n,
+  h1 = 255, h2 = NA, h3 = 75,
+  c1 = 30, cmax1 = 47, c2 = 0, c3 = 95,
+  l1 = 13, l2 = 52, l3 = 92,
+  p1 = 1.1, p3 = 1.0
+)
+```
+
+This uses a slight power transformation with `p1 = 1` in the blue arm of the palette but otherwise essentially
+corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
+using the above parameters.
 

### 2019-03-04 pkg/colorspace/R/divergingx_hcl.R

@@ -15,8 +15,8 @@
 #' between the two arms.
 #'
 #' With this additional flexibility various diverging palettes suggested by
-#' \url{http://ColorBrewer.org/} and CARTO (\url{https://carto.com/carto-colors/})
-#' can be emulated.
+#' \url{http://ColorBrewer.org/}, CARTO (\url{https://carto.com/carto-colors/}),
+#' or \pkg{viridis} can be emulated.
 #'
 #' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, and
 #' Tropic (with Tropic also available in \code{diverging_hcl}).
@@ -177,3 +177,6 @@ divex_pals[["BrBG"]]     <- c( 55,  NA, 180,  40,   0,  30,  25,  97,  20, 0.8, 
 divex_pals[["RdYlBu"]]   <- c( 10,  85, 260, 105,  45,  70,  35,  98,  35, 1.5, 1.2, 0.6, 1.2, 150, 10)
 divex_pals[["RdYlGn"]]   <- c( 10,  85, 140, 105,  45,  50,  35,  98,  35, 1.5, 1.2, 0.8, 1.2, 150, 75)
 divex_pals[["Spectral"]] <- c(  0,  85, 270,  90,  45,  65,  37,  98,  37, 1.0, 1.2, 1.0, 1.2, 120, NA)
+
+## Cividis
+divex_pals[["Cividis"]]  <- c(255,  NA,  75,  30,   0,  95,  13,  52,  92, 1.1, 1.0, 1.0,  NA,  47, NA)

### 2019-03-04 pkg/colorspace/R/hcl_palettes.R

@@ -91,8 +91,8 @@
 #' provided in the package and can be specified by a name instead of a set of
 #' parameters/coordinates. The examples have been taken from the literature and many
 #' approximate color palettes from other software packages such as ColorBrewer.org
-#' (\pkg{RColorBrewer}), CARTO colors (\pkg{rcartocolor}), or \pkg{scico}. The function
-#' \code{hcl_palettes} can be used to query the available pre-specified palettes. It
+#' (\pkg{RColorBrewer}), CARTO colors (\pkg{rcartocolor}), \pkg{scico}, or \pkg{virids}. The
+#' function \code{hcl_palettes} can be used to query the available pre-specified palettes. It
 #' comes with a \code{print} method (listing names and types), a \code{summary} method
 #' (additionally listing the underlying parameters/coordinates), and a \code{plot}
 #' method that creates a \code{\link{swatchplot}} with suitable labels.

### 2019-03-06 pkg/colorspace/paper/Toolbox/article.Rmd
added Zissou 1 palette to divergingx_hcl() along with some docs and discussion in vignettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@465 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ```{r divergingx-palettes, fig.width = 8.0, fig.height = 4.25, out.width = "0.8\\textwidth", fig.cap = "Prespecified flexible diverging HCL palettes available in \\code{divergingx\\_hcl()} in \\pkg{colorspace}."}
-divergingx_palettes(n = 7, plot = TRUE, nrow = 9)
+divergingx_palettes(n = 7, plot = TRUE, nrow = 10)
 ```
@@
 * `"PuOr"` to `"Spectral"` closely match the palettes of the same name from \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].
+* `"Zissou 1"` closely matches the palette of the same name from \pkg{wesanderson} [@color:wesanderson].
 * `"Cividis"` closely matches the palette of the same name from the \pkg{viridis} family [@color:viridis].

### 2019-03-06 pkg/colorspace/R/divergingx_hcl.R

@@ -15,8 +15,9 @@
 #' between the two arms.
 #'
 #' With this additional flexibility various diverging palettes suggested by
-#' \url{http://ColorBrewer.org/}, CARTO (\url{https://carto.com/carto-colors/}),
-#' or \pkg{viridis} can be emulated.
+#' \url{http://ColorBrewer.org/} and CARTO (\url{https://carto.com/carto-colors/}),
+#' can be emulated along with the Zissou 1 palette from \pkg{wesanderson} and
+#' Cividis from \pkg{viridis}.
 #'
 #' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, and
 #' Tropic (with Tropic also available in \code{diverging_hcl}).
@@ -177,6 +178,9 @@ divex_pals[["BrBG"]]     <- c( 55,  NA, 180,  40,   0,  30,  25,  97,  20, 0.8, 
 divex_pals[["RdYlBu"]]   <- c( 10,  85, 260, 105,  45,  70,  35,  98,  35, 1.5, 1.2, 0.6, 1.2, 150, 10)
 divex_pals[["RdYlGn"]]   <- c( 10,  85, 140, 105,  45,  50,  35,  98,  35, 1.5, 1.2, 0.8, 1.2, 150, 75)
 divex_pals[["Spectral"]] <- c(  0,  85, 270,  90,  45,  65,  37,  98,  37, 1.0, 1.2, 1.0, 1.2, 120, NA)
+
+## wesanderson
+divex_pals[["Zissou 1"]] <- c(218,  71,  12,  46,  88, 165,  59,  82,  52, 0.2, 1.0, 3.0, 1.0,  33, NA)
 
 ## Cividis
 divex_pals[["Cividis"]]  <- c(255,  NA,  75,  30,   0,  95,  13,  52,  92, 1.1, 1.0, 1.0,  NA,  47, NA)

### 2019-03-07 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from Claire
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@466 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Using the physiologically-based model for simulating color vision deficiency (CVD)
-of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
-emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
-and tritanope (blue cone cells defective).
-
-Below we briefly describe our \proglang{R} interface to these emulation techniques
-and show them in practice for a heatmap with sequential palette. Another example
-with diverging palette is available at
+Different kinds of limitations can be emulated using the physiologically-based model
+for simulating color vision deficiency (CVD) of @color:Machado+Oliveira+Fernandes:2009:
+deuteranomaly (green cone cells defective),
+protanomaly (red cone cells defective), and
+tritanomaly (blue cone cells defective).
+While most other CVD simulations handle only dichromacy, where one of three cones is
+non-functional, @color:Machado+Oliveira+Fernandes:2009 provides a unified model of
+both dichromacy and anomalous trichromacy, where one cone has shifted spectral
+sensitivity. As anomalous trichromacy is the most common form of color vision
+deficiency, it is important to emulate along with the rarer, but more severe
+dichromacy. Below we briefly describe our \proglang{R} interface to these emulation
+techniques and show them in practice for a heatmap with sequential palette.
+Another example with diverging palette is available at
 <http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
-Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
-palette is almost always a bad choice in scientific displays. See
+Finally, CVD emulation is particularly useful for bringing out why the RGB
+rainbow palette is almost always a bad choice in scientific displays. See
 <http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
@@
 
-The workhorse function to do so is `simulate_cvd()` which can take any vector
-of valid \proglang{R} colors and transform them according to a certain CVD transformation
-matrix and transformation equation. The transformation matrices have been established
-by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
-`deutanomaly_cvd`, and `tritanomaly_cvd`.
-The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
-simulating the corresponding kind of color blindness with a given severity (calling
-`simulate_cvd()` internally).
-
-For further guidance on color blindness in relation to statistical graphics
-see @color:Lumley:2006 which accompanies the \proglang{R} package \pkg{dichromat} [@color:dichromat]
-and is based on earlier emulation techniques
+The workhorse function to emulate color vision deficiencies is `simulate_cvd()`
+which can take any vector of valid \proglang{R} colors and transform them according to a
+certain CVD transformation matrix and transformation equation. The
+transformation matrices have been established by @color:Machado+Oliveira+Fernandes:2009 and are
+provided in objects `protanomaly_cvd`, `deutanomaly_cvd`, and `tritanomaly_cvd`.
+The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level
+functions for simulating the corresponding kind of color blindness with a given
+`severity` (calling `simulate_cvd()` internally). A severity of `1` corresponds to
+dichromacy, `0` to normal color vision, and intermediate values to varying
+severities of anomalous trichromacy. For further guidance on color blindness in
+relation to statistical graphics see @color:Lumley:2006 which accompanies the \proglang{R}
+package \pkg{dichromat} [@color:dichromat] and is based on earlier emulation techniques
 [@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].

### 2019-03-08 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from JCF
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@468 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 (CRAN) at <https://CRAN.R-project.org/package=colorspace> and the development version of \pkg{colorspace}
-is hosted on \proglang{R}-Forge.
+is hosted on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/projects/colorspace/}.
 
@@
 specified through suitable trajectories in the HCL (hue-chroma-luminance) color space.
-A quick overview can be gained easily with `hcl_palettes()` (see Figure\ \ref{fig:hcl-palettes}):
+A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes}):
 
@@
 
-Using the names from the plot above and a desired number of colors in the palette
-a suitable color vector can be easily computed, e.g.,
+A suitable vector of colors can be easily computed by specifying the desired number
+of colors and palette name (see Figure\ \ref{fig:hcl-palettes} for possible palette names), e.g.,
 
@@
 The functions `sequential_hcl()`, and `diverging_hcl()` work analogously. Additionally,
-their hue/chroma/luminance parameters can be modified, thus allowing to easily customize
+a palette's hue/chroma/luminance parameters can be modified, thus allowing to easily customize
 each palette. Moreover, the `choose_palette()`/`hclwizard()` app provides convenient user
-interfaces to do the customization interactively. Finally, even more flexible diverging
+interfaces to perform palette customization interactively. Finally, even more flexible diverging
 HCL palettes are provided by `divergingx_hcl()`.
@@
 and sequential `"Purples 3"` palettes that were also employed above
-(see Figure\ \ref{fig:iris-diamonds}). First, semi-transparent
-shaded densities of the sepal length from the `iris` data are shown, grouped by species.
+(see Figure\ \ref{fig:iris-diamonds}). For the first example, semi-transparent
+shaded densities of the sepal length from the `iris` data are shown, grouped by species
+ (see the left panel of Figure\ \ref{fig:iris-ggplot}).
 
@@
 
-The sequential palette is used to code the cut levels in a scatter of price by carat
-in the `diamonds` data (or rather a small subsample thereof). The scale function first
+And for the second example the sequential palette is used to code the cut levels
+in a scatter of price by carat in the `diamonds` data (or rather a small subsample
+thereof, see the right panel of Figure\ \ref{fig:iris-ggplot}). The scale function first
 generates six colors but then drops the first color because the light gray is too light
-in this display. (Alternatively, the chroma and luminance parameters could also be tweaked.)
+here. (Alternatively, the chroma and luminance parameters could also be tweaked.)
 
@@
 
-Here, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
+In Figure\ \ref{fig:applots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
 application for a sequential palette. The elevation of the volcano is brought out clearly,
 focusing with the dark colors on the higher elevations.
-
 The other two displays show that hue is constant in the palette while luminance and chroma vary.
@@
 Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
-(compared to a monotonic chroma trajectory).
+when compared to a monotonic chroma trajectory.
 
@@
 
-At the core of the \pkg{colorspace} package there are various utilities for computing with
+At the core of the \pkg{colorspace} package are various utilities for computing with
 color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
-representations of color to each other, see @color:Ihaka:2003. A particularly important
+representations of color to each other [@color:Ihaka:2003]. A particularly important
 mapping is the one from the perceptually-based and device-independent color model
@@
 The connections are illustrated in Figure\ \ref{fig:color-spaces}.
-Color models that are (or try to be) perceptually-based are displayed with circles,
-other color models with rectangles.
+Color models that are (or try to be) perceptually-based are displayed with circles
+and models that are not are displayed with rectangles.
 
@@
 
-To make the package self-contained and exactly backward compatible the \proglang{C}
+To make the \pkg{colorspace} package self-contained and exactly backward compatible, the \proglang{C}
 code in \pkg{colorspace} is still used as the basis for all color space conversions.
@@
 
-These are pastel red (hue = 0), green (hue = 120), and blue (hue = 240) with moderate chroma and luminance.
+The resulting three colors are pastel red (hue = 0), green (hue = 120), and blue (hue = 240) with moderate chroma and luminance.
 For display in other systems an sRGB representation might be needed:
@@
 Figure\ \ref{fig:hcl-palettes-principles} and explained in more detail below. The desaturated
-palettes bring out clearly that luminance differences (= light-dark contrasts),
-or the lack thereof, are crucial for coding the different kinds of palettes.
+palettes in the second row of Figure\ \ref{fig:hcl-palettes-principles}
+bring out clearly that luminance differences (= light-dark contrasts)
+are crucial for sequential and diverging palettes while qualitative palettes
+are balanced at the same luminance.
 
@@
 
-To compute the actual color hex codes (representing sRGB coordinates) based on these parameters, the functions
-`qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`, respectively, can be used.
+To compute the actual color hex codes (representing sRGB coordinates) based on these HCL parameters, the functions
+`qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()` can be used.
 Either all parameters can be specified "by hand" through the HCL parameters, an entire
@@
 
-In Figure\ \ref{fig:hcl-palettes-qualitative} the available named palettes are shown. The first five
+Figure\ \ref{fig:hcl-palettes-qualitative} shows the named palettes available in the
+`qualitative_hcl()` function. The first five
 palettes are close to the \pkg{ColorBrewer.org} palettes of the same name [@color:Harrower+Brewer:2003].
@@
 luminance as in `"Dark 2"` or `"Dark 3"` usually works better while on a black/dark
-background the luminance should be higher as in `"Set 2"` for example.
+background the luminance should be higher as in `"Set 2"`.
 Some examples with demo graphics are provided in Section\ \ref{sec:palette_visualization}.
@@
 
-As suggested by @color:Zeileis+Murrell+Hornik:2009 `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
-of increasing (or decreasing) luminance. Thus, the `l` argument should provide
+As suggested by @color:Zeileis+Hornik+Murrell:2009 `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
+of increasing (or decreasing) luminance. Thus, the function's `l` argument should provide
 a vector of length 2 with starting and ending luminance (equivalently, `l1` and
@@
 
-For distinguishing also colors in the middle two strategies can be employed:
+For distinguishing colors in the middle, two strategies can be employed:
 (a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
@@
 
-Instead of employing linear trajectories in the chroma or luminance coordinates
+Instead of employing linear trajectories in the chroma or luminance coordinates,
 some palettes employ a power transformation of the chroma and/or luminance
@@
 the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
-They are intended for a white/light background. The last palette is taken from
+They are intended for a white/light background. The last palette (`"Oslo"`) is taken from
 the scientific color maps of @color:Crameri:2018 and is intended for a black/dark
@@
 of such a palette is still a monotonic luminance sequence as above (combined with a
-monotonic or triangular chroma sequence). But, instead of a single hue, an interval
-of hues in `h` (or beginning hue `h1` and ending hue `h2`) is specified.
-
-`sequential_hcl()` allows to combine variations in hue (`h` and `h1`/`h2`, respectively),
+monotonic or triangular chroma sequence). But rather than using a single hue, an interval
+of hues in `h` (or beginning hue `h1` and ending hue `h2`) can be specified.
+
+`sequential_hcl()` allows combined variations in hue (`h` and `h1`/`h2`, respectively),
 chroma (`c` and `c1`/`c2`/`cmax`, respectively),
@@
 but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
-does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
+does not work as well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
 However, they work better on a dark/black background (see Figure\ \ref{fig:demoplot-dark}). Also, they might be a reasonable
@@
 `rect()` function but with heuristics for choosing default labels, margins,
-spacings, borders, etc. These are selected to work well for `hcl_palettes()`
-and might need further tweaking in future versions. Thus,
+spacings, borders, etc. These heuristics are selected to work well for `hcl_palettes()`
+and might need further tweaking in future versions of the package. Thus,
 Figures\ \ref{fig:hcl-properties}--\ref{fig:hcl-palettes} as well as
@@
 regarding factors that are relevant in choosing an effective palette
-in a given situation. Further useful is provided in many sources, including
+in a given situation. Further useful guidance is provided in many sources, including
 @color:Ware:1988, @color:Okabe+Ito:2008, @color:Aigner:2010,

### 2019-03-08 pkg/colorspace/paper/Toolbox/article.Rmd
added a little more detail about some other colour packages
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@469 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 user interface.
+The \pkg{pals} package also provides a large number of palettes,
+but some are just pre-existing palettes, and where the palettes are new
+there is no description of a specific algorithm that was used to generate
+them.  
 Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
@@
 
-Different kinds of limitations can be emulated using the physiologically-based model
-for simulating color vision deficiency (CVD) of @color:Machado+Oliveira+Fernandes:2009:
-deuteranomaly (green cone cells defective),
-protanomaly (red cone cells defective), and
-tritanomaly (blue cone cells defective).
-While most other CVD simulations handle only dichromacy, where one of three cones is
-non-functional, @color:Machado+Oliveira+Fernandes:2009 provides a unified model of
-both dichromacy and anomalous trichromacy, where one cone has shifted spectral
-sensitivity. As anomalous trichromacy is the most common form of color vision
-deficiency, it is important to emulate along with the rarer, but more severe
-dichromacy. Below we briefly describe our \proglang{R} interface to these emulation
-techniques and show them in practice for a heatmap with sequential palette.
-Another example with diverging palette is available at
+Using the physiologically-based model for simulating color vision deficiency (CVD)
+of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
+emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
+and tritanope (blue cone cells defective).
+
+Below we briefly describe our \proglang{R} interface to these emulation techniques
+and show them in practice for a heatmap with sequential palette. Another example
+with diverging palette is available at
 <http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
-Finally, CVD emulation is particularly useful for bringing out why the RGB
-rainbow palette is almost always a bad choice in scientific displays. See
+Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
+palette is almost always a bad choice in scientific displays. See
 <http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
@@
 
-The workhorse function to emulate color vision deficiencies is `simulate_cvd()`
-which can take any vector of valid \proglang{R} colors and transform them according to a
-certain CVD transformation matrix and transformation equation. The
-transformation matrices have been established by @color:Machado+Oliveira+Fernandes:2009 and are
-provided in objects `protanomaly_cvd`, `deutanomaly_cvd`, and `tritanomaly_cvd`.
-The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level
-functions for simulating the corresponding kind of color blindness with a given
-`severity` (calling `simulate_cvd()` internally). A severity of `1` corresponds to
-dichromacy, `0` to normal color vision, and intermediate values to varying
-severities of anomalous trichromacy. For further guidance on color blindness in
-relation to statistical graphics see @color:Lumley:2006 which accompanies the \proglang{R}
-package \pkg{dichromat} [@color:dichromat] and is based on earlier emulation techniques
+The workhorse function to do so is `simulate_cvd()` which can take any vector
+of valid \proglang{R} colors and transform them according to a certain CVD transformation
+matrix and transformation equation. The transformation matrices have been established
+by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
+`deutanomaly_cvd`, and `tritanomaly_cvd`.
+The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
+simulating the corresponding kind of color blindness with a given severity (calling
+`simulate_cvd()` internally).
+
+For further guidance on color blindness in relation to statistical graphics
+see @color:Lumley:2006 which accompanies the \proglang{R} package \pkg{dichromat} [@color:dichromat]
+and is based on earlier emulation techniques
 [@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].
@@
 
-* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories".
-* \pkg{roloc} [@color:roloc]: naming palettes.
-* \pkg{colorscience} [@color:colorscience]: More color mappings.
+* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories". This can generate 
+  palettes with the (a large number of) 
+  colours spaced far apart from each other in Luv space.
+  It also has functions for visualising colour palettes in different ways 
+  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
+* \pkg{roloc} [@color:roloc]: colour conversions, but not between
+  numeric colour spaces;  instead, from numeric colour spaces to
+  colour names.  
+  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
+* \pkg{colorscience} [@color:colorscience]: This package is for Colour
+Science.  It has data sets, does more complex conversions (including
+different chromatic adaptation algorithms).  
+So more sophisticated than \pkg{colorspace}
+in terms of flexibility and low-level detail, BUT it does not do
+palette-generation.
 * \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.

### 2019-03-10 pkg/colorspace/R/divergingx_hcl.R
sequential_hcl(), diverging_hcl(), and divergingx_hcl() now all use the same internal seqhcl() workhorse function.
In turn, seqhcl() calls the internal lintrj() and tritrj() functions
implementing the linear and triangular trajectories, respectively.

divergingx_hcl(n) with even n erroneously returned the same central
color twice, fixed now.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@471 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -104,15 +104,12 @@ divergingx_hcl <- function(n, palette = "Geyser", ..., rev = FALSE,
 
     ## call sequential_hcl() once or twice
     n2 <- ceiling(n/2)    
-    rval <- sequential_hcl(n2, rev = TRUE,
-    	h1 = pals["h1"], h2 = pals["h2"], c1 = pals["c1"], c2 = pals["c2"],
-    	l1 = pals["l1"], l2 = pals["l2"], p1 = pals["p1"], p2 = pals["p2"],
-	cmax = pals["cmax1"], ...)
-    if(floor(n/2) < n2) rval <- rval[-1L]
-    rval <- c(rev(rval), sequential_hcl(n2, rev = TRUE,
-    	h1 = pals["h3"], h2 = pals["h2"], c1 = pals["c3"], c2 = pals["c2"],
-    	l1 = pals["l3"], l2 = pals["l2"], p1 = pals["p3"], p2 = pals["p4"],
-	cmax = pals["cmax2"], ...))
+    rval <- seq.int(1, by = -2/(n - 1), length.out = n2)
+    rval <- c(seqhcl(rval, pals["h1"], if(is.na(pals["h2"])) pals["h1"] else pals["h2"], pals["c1"], pals["c2"],
+                     pals["l1"], pals["l2"], pals["p1"], pals["p2"], pals["cmax1"], as.logical(pals["fixup"]), ...),
+    	  rev(seqhcl(rval, pals["h3"], if(is.na(pals["h2"])) pals["h3"] else pals["h2"], pals["c3"], pals["c2"],
+	             pals["l3"], pals["l2"], pals["p3"], pals["p4"], pals["cmax2"], as.logical(pals["fixup"]), ...)))
+    if(floor(n/2) < n2) rval <- rval[-n2]
     if(rev) rval <- rev(rval)
     return(rval)   
 }

### 2019-03-10 pkg/colorspace/R/hcl_palettes.R

@@ -390,6 +390,25 @@ qualitative_hcl <- function(n, h = c(0, 360 * (n - 1)/n), c = 80, l = 60,
     if(register) invisible(rval) else return(rval)
 }
 
+## trajectories
+lintrj <- function(i, p1, p2) p2 - (p2 - p1) * i
+tritrj <- function(i, j, p1, p2, pm) ifelse(i <= j,
+  p2 - (p2 - pm) * i/j,
+  pm - (pm - p1) * abs((i - j)/(1 - j)))
+
+## HCL sequence
+seqhcl <- function(i, h1, h2, c1, c2, l1, l2, p1, p2, cmax, fixup, ...) {
+    j <- 1/(1 + abs(cmax - c1) / abs(cmax - c2))
+    if (!is.na(j) && (j <= 0 | j >= 1)) j <- NA
+    hex(polarLUV(
+      L = lintrj(i^p2, l1, l2),
+      C = if(is.na(j)) lintrj(i^p1, c1, c2) else tritrj(i^p1, j, c1, c2, cmax),
+      H = lintrj(i, h1, h2)),
+      fixup = fixup, ...
+    )
+}
+
+
 #' @rdname hcl_palettes
 #' @export
 sequential_hcl <- function(n, h = 260, c = 80, l = c(30, 90), power = 1.5,
@@ -467,21 +486,8 @@ sequential_hcl <- function(n, h = 260, c = 80, l = c(30, 90), power = 1.5,
     if(is.na(pals["p2"])) pals["p2"] <- pals["p1"]
 
     ## HCL trajectory
-    cmaxat <- 1/(1 + abs(pals["cmax"] - pals["c1"]) / abs(pals["cmax"] - pals["c2"]))
-    if ( ! is.na(cmaxat) && (cmaxat <= 0 | cmaxat >= 1) ) cmaxat <- NA
     rval <- seq(1, 0, length = n)
-    rval <- hex(polarLUV(
-        L = pals["l2"] - (pals["l2"] - pals["l1"]) * rval^pals["p2"],
-        C = if(is.na(cmaxat)) {
-              pals["c2"] - (pals["c2"] - pals["c1"]) * rval^pals["p1"]
-            } else {
-              ifelse(rval^pals["p1"] <= cmaxat,
-                     pals["c2"] - (pals["c2"] - pals["cmax"]) * (rval^pals["p1"])/cmaxat,
[…]

### 2019-03-10 pkg/colorspace/R/divergingx_hcl.R
make 'alpha' and 'fixup' arguments explicit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@473 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -27,7 +27,10 @@
 #' 
 #' @param n the number of colors (\eqn{\ge 1}{>= 1}) to be in the palette.
 #' @param palette character with the name (see details).
-#' @param \dots arguments passed to \code{\link{sequential_hcl}}.
+#' @param \dots arguments passed to \code{\link{hex}}.
+#' @param fixup logical. Should the color be corrected to a valid RGB value?
+#' @param alpha numeric vector of values in the range \code{[0, 1]} for alpha
+#' transparency channel (0 means transparent and 1 means opaque).
 #' @param rev logical. Should the palette be reversed?
 #' @param h1 numeric. Starting hue coordinate.
 #' @param h2 numeric. Center hue coordinate.
@@ -59,16 +62,15 @@
 #' @rdname divergingx_hcl
 
 #' @export
-divergingx_hcl <- function(n, palette = "Geyser", ..., rev = FALSE,
+divergingx_hcl <- function(n, palette = "Geyser", ...,
+  fixup = TRUE, alpha = 1, rev = FALSE,
   h1, h2, h3, c1, c2, c3, l1, l2, l3, p1, p2, p3, p4, cmax1, cmax2)
 {
     ## empty palette
     if(n < 1L) return(character(0L))
 
-    ## obtained stored coordinates    
-    palette <- match.arg(palette, names(divex_pals))
-    pals <- divex_pals[[palette]]
-    names(pals) <- c("h1", "h2", "h3", "c1", "c2", "c3", "l1", "l2", "l3", "p1", "p2", "p3", "p4", "cmax1", "cmax2")
+    ## obtained stored coordinates
+    pals <- as.matrix(divergingx_palettes(palette = palette)[, -1L])[1L, ]
 
     ## replace coordinates (if specified)
     if(!missing(h1)) pals["h1"] <- h1
@@ -106,10 +108,19 @@ divergingx_hcl <- function(n, palette = "Geyser", ..., rev = FALSE,
     n2 <- ceiling(n/2)    
     rval <- seq.int(1, by = -2/(n - 1), length.out = n2)
     rval <- c(seqhcl(rval, pals["h1"], if(is.na(pals["h2"])) pals["h1"] else pals["h2"], pals["c1"], pals["c2"],
-                     pals["l1"], pals["l2"], pals["p1"], pals["p2"], pals["cmax1"], as.logical(pals["fixup"]), ...),
[…]

### 2019-03-10 pkg/colorspace/paper/Toolbox/slides.Rnw
typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@450 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
incorporate improved influenza images -> also in slides now
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@479 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
+
+\subsection{Motivation}
+
+\begin{frame}
+\frametitle{Motivation}
+
+\vspace*{-0.3cm}
+
+\begin{minipage}{0.54\textwidth}
+\only<1,4>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow.png}}%
+\only<2>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow-deutan.png}}%
+\only<3>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow-gray.png}}%
+\only<5>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow.png}}%
+\only<6>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow-deutan.png}}%
+\only<7>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow-gray.png}}
+\end{minipage}
+\hspace*{0.6cm}
+\begin{minipage}{0.4\textwidth}
+
+\textbf{Map:} Influenza severity in Germany (week 8, 2019).\\
+
+\textbf{Source:} Arbeitsgemeinschaft Influenza, Robert-Koch-Institut.\\
+
+\textbf{Reported in:} SPIEGEL Online, Tagesschau, \dots\\
+
+\visible<2->{\textbf{Problems:} Color vision deficiency.}
+\visible<3->{Grayscale.}
+\visible<4->{Flashy.\\}
+
+\visible<5->{\textbf{Alternative:} HCL-based sequential palette.}
+
+\end{minipage}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation}
+
+\textbf{Need tools for:}
+\begin{itemize}
+  \item Construction of palettes with better perceptual properties.
+  \item Assessment of color palettes.
+  \item Manipulation of colors.
+\end{itemize}
+
+\pause
+\bigskip
+
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=\textwidth]{../../vignettes/screenshots/bobross.jpg}
+\end{minipage}
+\hspace*{0.6cm}
+\begin{minipage}{0.36\textwidth}
+\vspace*{-4cm}
+Because Bob Ross would not approve of this!
+\end{minipage}
+
+\end{frame}
 
@@
 
-\textbf{Emulate:} Color vision deficies.
+\textbf{Emulate:} Color vision deficiencies.
 

### 2019-03-11 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from KH
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@481 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-  Namely, general strategies for three types of palettes are provided:
+  General strategies for three types of palettes are provided:
   (1)\ Qualitative for coding categorical information, i.e., where no particular
@@
 At the core of the package there are various utilities for computing with
-color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
+color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
 representations of color to each other. A particularly important
@@
 
+This paper is organized as follows...
+
 # A quick tour {#sec:tour}
@@
 HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
-specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
-in \proglang{R}). For completeness further standard color models are included as well in the package.
+specifications in many systems based on the corresponding hex codes
+[@color:Wiki+Webcolors], e.g., in HTML but also in \proglang{R}.
+For completeness further standard color models are included as well in the package.
 The connections are illustrated in Figure\ \ref{fig:color-spaces}.
@@
 
-All the \proglang{S}4 classes listed above inherit from a virtual class `color` which are
+All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which are
 internally always represented by matrices with three columns (corresponding to
@@
 conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
-though [see the discussion in @color:convertColor]. Moreover, base \proglang{R} also provides
+though [see the discussion in @color:convertColor]. For more complex conversions, including
+different chromatic adaptation algorithms, a more comprehensive color science approach
+is implemented in the \proglang{R} package \pkg{colorscience} [@color:colorscience].
+Finally, base \proglang{R} also provides
 `grDevices::hcl()` for mapping HCL representations to hex codes.
@@
 and `p1`, respectively) and are depicted in Figure\ \ref{fig:hcl-trajectories}.
-Analogous equations apply for the other two coordinates, though.
+Analogous equations apply for the other two coordinates.
 
@@
 palette-generating functions, i.e., starting/ending hue, minimal/maximal chroma,
-minimal maximal luminance, and a power transformations that control how
+minimal maximal luminance, and power transformations that control how
 quickly/slowly chroma and/or luminance are changed through the palette.
@@
 
-* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories". This can generate 
+* \pkg{Polychrome} [@color:Polychrome, @color:Polychrome2]: qualitative palettes with more "categories". This can generate 
   palettes with the (a large number of) 
@@
   (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
-* \pkg{roloc} [@color:roloc]: colour conversions, but not between
+* \pkg{roloc} [@color:roloc,@color:roloc2]: colour conversions, but not between
   numeric colour spaces;  instead, from numeric colour spaces to

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
some feedback from RS, and started rewriting intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@483 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   \usepackage{booktabs,amsmath,longtable,supertabular,framed}
+  \shortcites{color:colorspace,color:ggplot2pkg}
   \newcommand{\tex}[1]{#1}
@@
 # Introduction {#sec:intro}
-
-The \pkg{colorspace} package provides a broad toolbox for selecting individual
-colors or color palettes, manipulating these colors, and employing
-them in various kinds of visualizations.
 
@@
 
-At the core of the package there are various utilities for computing with
-color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
-representations of color to each other. A particularly important
-mapping is the one from the perceptually-based and device-independent color model
-HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
-specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
-in \proglang{R}). For completeness further standard color models are included as well in the package:
-`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
-`HSV()`.
-
-The HCL space (= polar coordinates in CIELUV) is particularly useful for
-specifying individual colors and color palettes as its three axes match those
-of the human visual system very well: Hue (= type of color, dominant wavelength),
+Color is an integral and omnipresent element of many statistical graphics and data
+visualizations. Therefore, colors should be carefully chosen to support the information
+displayed [@color:Tufte:1990, @color:Brewer:1999, @color:Ware:2004, @color:Wilkinson:2005, @color:Wilke:2019].
+However, until relatively recently many software packages have been using color
+palettes derived from simple RGB (red-green-blue)
+color combinations such as the RGB "rainbow" (or "jet") color palette
+with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
+and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.
+
+To address these problems, many improved color palettes with better perceptual
+properties have been suggested in the literature
+[@color:Harrower+Brewer:2003, @color:Zeileis+Hornik+Murrell:2009, @color:Smith+VanDerWalt:2015, @color:Crameri:2018]
+and also made available in standard software packages including various
+packages for \proglang{R} [@color:R] such as 
+\pkg{RColorBrewer} [@color:RColorBrewer],
+\pkg{rcartocolor} [@color:rcartocolor],
+\pkg{viridis} [@color:viridis],
+\pkg{wesanderson} [@color:wesanderson],
+or \pkg{scico} [@color:scico] among many others.
+In packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
+many of these palettes are collected and provided with a unified interface.
+Most of these palettes, however, are pre-existing palettes, stored as a
+limited set of colors and interpolated as necessary. And even if specific
+algorithms have been used in the first construction of the palettes, these
+are often not reflected in the software implementations.
+
+In contrast, the \pkg{colorspace} package [@color:colorspace] adopts a perceptually-based
+approach -- the so-called HCL (hue--chroma--luminance) space -- and
+establishes a broad range of color palettes through general and adaptable
+trajectories in HCL space. In addition to constructing the palettes the HCL
+space is also employed to visualize and assess the properties of a given
+palette (beyond simple color swatches). The reason for choosing the HCL space
+for all of these tools is that its three axes match those of the human visual system very well:
+Hue (= type of color, dominant wavelength),
 chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
@@
 
+The \pkg{colorspace} package provides a broad toolbox for selecting individual
+colors or color palettes, manipulating these colors, and employing
+them in various kinds of visualizations.
+
+At the core of the package there are various utilities for computing with
+color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
+representations of color to each other. A particularly important
+mapping is the one from the perceptually-based and device-independent color model
+HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
+specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
+in \proglang{R}). For completeness further standard color models are included as well in the package:
+`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
+`HSV()`.
+
 # A quick tour {#sec:tour}
@@
 shaded densities of the sepal length from the `iris` data are shown, grouped by species
- (see the left panel of Figure\ \ref{fig:iris-ggplot}).
+(see the left panel of Figure\ \ref{fig:iris-diamonds}).
 
@@
 in a scatter of price by carat in the `diamonds` data (or rather a small subsample
-thereof, see the right panel of Figure\ \ref{fig:iris-ggplot}). The scale function first
+thereof, see the right panel of Figure\ \ref{fig:iris-diamonds}). The scale function first
 generates six colors but then drops the first color because the light gray is too light
@@
 
-In Figure\ \ref{fig:applots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
+In Figure\ \ref{fig:allplots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
 application for a sequential palette. The elevation of the volcano is brought out clearly,
@@
 
-For distinguishing colors in the middle, two strategies can be employed:
+For distinguishing colors in the center of the palette, two strategies can be employed:
 (a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
@@
 
-* Single-hue: In each palette the hue is fixed and chroma decreases monotonically
+* _Single-hue:_ In each palette the hue is fixed and chroma decreases monotonically
   (along with increasing luminance). This is typically sufficient to clearly bring
   out the extreme colors (dark/colorful vs.\ light gray).
-* Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
+* _Single-hue (advanced):_ The hue is fixed (as above) but the chroma trajectory is
   triangular. Compared to the basic single-hue palette above this allows to better
   distinguish the colors in the middle and not only the extremes.
-* Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
+* _Multi-hue (advanced):_ As in the advanced single-hue palette the chroma trajectory
   is triangular but additionally the hue varies slightly. This can further enhance
@@
 
-- Hue is drawn in red and coordinates are indicated on the axis on the right
-  with range $[-360, 360]$.
-- Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
+* Hue is drawn in red and coordinates are indicated on the axis on the right
+  with range $[0, 360]$ or (if necessary) $[-360, 360]$.
+* Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
   is used unless the palette necessitates higher chroma values.
-- Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].
+* Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].
 

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
re-wrote intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@484 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Color is an integral and omnipresent element of many statistical graphics and data
-visualizations. Therefore, colors should be carefully chosen to support the information
-displayed [@color:Tufte:1990, @color:Brewer:1999, @color:Ware:2004, @color:Wilkinson:2005, @color:Wilke:2019].
+visualizations. Therefore, colors should be carefully chosen to support all viewers in
+accessing the information displayed [@color:Tufte:1990; @color:Brewer:1999; @color:Ware:2004; @color:Wilkinson:2005; @color:Wilke:2019].
 However, until relatively recently many software packages have been using color
@@
 To address these problems, many improved color palettes with better perceptual
-properties have been suggested in the literature
-[@color:Harrower+Brewer:2003, @color:Zeileis+Hornik+Murrell:2009, @color:Smith+VanDerWalt:2015, @color:Crameri:2018]
-and also made available in standard software packages including various
-packages for \proglang{R} [@color:R] such as 
+properties have been suggested in the literature over the last one or two decades
+[@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
+Many systems for systems for statistical and scientific computing provide
+infrastructure for such color palettes, e.g., for \proglang{R} [@color:R] the
+list of useful packages encompasses:
 \pkg{RColorBrewer} [@color:RColorBrewer],
+\pkg{viridis} [@color:viridis],
 \pkg{rcartocolor} [@color:rcartocolor],
-\pkg{viridis} [@color:viridis],
 \pkg{wesanderson} [@color:wesanderson],
 or \pkg{scico} [@color:scico] among many others.
-In packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
-many of these palettes are collected and provided with a unified interface.
+Furthermore, packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
+collect many of the proposed palettes in combination with a unified interface.
 Most of these palettes, however, are pre-existing palettes, stored as a
 limited set of colors and interpolated as necessary. And even if specific
-algorithms have been used in the first construction of the palettes, these
+algorithms have been used in the initial construction of the palettes, these
 are often not reflected in the software implementations.
-
-In contrast, the \pkg{colorspace} package [@color:colorspace] adopts a perceptually-based
-approach -- the so-called HCL (hue--chroma--luminance) space -- and
-establishes a broad range of color palettes through general and adaptable
-trajectories in HCL space. In addition to constructing the palettes the HCL
-space is also employed to visualize and assess the properties of a given
-palette (beyond simple color swatches). The reason for choosing the HCL space
-for all of these tools is that its three axes match those of the human visual system very well:
-Hue (= type of color, dominant wavelength),
-chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
 
@@
 
-The \pkg{colorspace} package provides three types of palettes based on the HCL model:
+The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
+approach, utiliting the perceptually-based HCL (hue--chroma--luminance) color space
+[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
+system very well: Hue (= type of color, dominant wavelength),
+chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
+Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
+strategies for color palettes; manipulate individual colors and color palettes;
+and assess and visualize the properties of color palettes (beyond simple color swatches).
+Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:
 
@@
 
-To aid choice and application of these palettes there are: scales for use
-with \pkg{ggplot2} [@color:ggplot2]; \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration;
-visualizations of palette properties; accompanying manipulation utilities
-(like desaturation, lighten/darken, and emulation of color vision deficiencies).
-
-FIXME: More comparison with/discussion of other palette packages.
-the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
-as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
-the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
-package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
-and can be approximated reasonably well in \pkg{colorspace}
-(see Section\ \ref{sec:approximations}).
-A broad range of packages that provide color palettes that are "in harmony" as
-they are extracted and clustered from movies, paintings, etc. However, many of these
-are not so suitable for scientific visualization as they are not clearly 
-qualitative/sequential/diverging. An influential package following such an
-approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
-[@color:paletteer] package aggregates these and provides them in a unified
-user interface.
-The \pkg{pals} package also provides a large number of palettes,
-but some are just pre-existing palettes, and where the palettes are new
-there is no description of a specific algorithm that was used to generate
-them.  
-Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
-coordinates, along with visualization and assessment of palettes (beyond
-color swatches).
-
-This paper is organized as follows...
-
-The \pkg{colorspace} package provides a broad toolbox for selecting individual
-colors or color palettes, manipulating these colors, and employing
-them in various kinds of visualizations.
-
-At the core of the package there are various utilities for computing with
-color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
-representations of color to each other. A particularly important
-mapping is the one from the perceptually-based and device-independent color model
-HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
-specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
-in \proglang{R}). For completeness further standard color models are included as well in the package:
-`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
-`HSV()`.
+A broad collection of prespecified palettes are shipped in the package, existing
+palettes can be easily tweaked, and new or adapted palettes registered. The prespecified
+palettes include suitable HCL color choices that closely approximate of most palettes
+from packages \pkg{RColorBrewer}, \pkg{rcartocolor}, and \pkg{viridis} by using only a
+small set of hue, chroma, and luminance parameters.
+
+To aid choice and application of these palettes there are
+(a) scales for use with \pkg{ggplot2} [@color:ggplot2],
+(b) \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration,
+(c) visualizations of palette properties, and
+(d) accompanying manipulation utilities (like desaturation, lighten/darken, and emulation of color vision deficiencies).
+
+The remainder of the manuscript is organized as follows:
+Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
+Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
+Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
+Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
+that help assess the suitability of colors for colorblind viewers.
+Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
+Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
+before Section\ \ref{sec:summary} concludes the manuscript.
 
@@
 At the core of the \pkg{colorspace} package are various utilities for computing with
-color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
+color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
 representations of color to each other [@color:Ihaka:2003]. A particularly important
@@
 Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
-high chroma while this is not possible to the same extent for $H = 245$. Hence,
+high chroma while this is not possible to the same extent for $H = 245$ because of
+the distorted shape of the HCL space. Hence,
 chroma has to be decreased when proceeding to the dark low-luminance colors.
@@
 
-* **Palette constructor:** `choose_palette()` or `hclwizard()` or `hcl_wizard()`.
-* **Color picker:** `choose_color()` or equivalently `hcl_color_picker()`.
-* **Color vision deficiency emulator:** `cvd_emulator()`.
+* _Palette constructor:_ `choose_palette()` or `hclwizard()` or `hcl_wizard()`.
+* _Color picker:_ `choose_color()` or equivalently `hcl_color_picker()`.
+* _Color vision deficiency emulator:_ `cvd_emulator()`.
 
@@
 To facilitate generation of palettes for black/dark backgrounds, a "dark mode"
-of the GUIs is also available:
+of the GUIs is also available.
 
@@
 This strategy typically works well for lightening colors, although in some situations the result can be
-rather colorful. Conversely, when darkening rather light colors with little chroma, this can result
+somewhat too colorful. Conversely, when darkening rather light colors with little chroma, this can result
 in rather gray colors.
@@
 
-Regardless of the chosen color space, the adjustment of the `L` component can occur by two methods, relative
-(the default) and absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
+Regardless of the chosen color space, the adjustment of the `L` component by a certain `amount` can occur by two methods, relative
+(the default) or absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
 `L * (1 - amount)` (relative). See `?lighten` and `?darken` for more details.

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
first draft of summary/discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@485 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
-approach, utiliting the perceptually-based HCL (hue--chroma--luminance) color space
+approach, utiliting the perceptually-based HCL (Hue--Chroma--Luminance) color space
 [@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
@@
 The \pkg{colorspace} package ships with a wide range of predefined color palettes,
-specified through suitable trajectories in the HCL (hue-chroma-luminance) color space.
+specified through suitable trajectories in the HCL (Hue-Chroma-Luminance) color space.
 A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes}):
@@
 diverging palettes derived by relatively simple trajectories in HCL
-(Hue-Chroma-Luminance) space. In contrast to many other packages providing
+(Hue--Chroma--Luminance) space. In contrast to many other packages providing
 modern balanced color palettes (such as \pkg{ColorBrewer.org}, \pkg{CARTO},
@@
 the palettes that can be adjusted to the particular needs of a given
-data visulization. The manuscript already provided various tips and tricks
-regarding factors that are relevant in choosing an effective palette
-in a given situation. Further useful guidance is provided in many sources, including
+data visulization. The manuscript also provides various tips and tricks
+for choosing an effective palette in a given situation. Further useful
+guidance is provided in many sources, including:
 @color:Ware:1988, @color:Okabe+Ito:2008, @color:Aigner:2010,
 @color:Stauffer+Mayr+Dabernig:2015, @color:Zhang:2015, @color:Rost:2018,
-@color:Wilke:2019, and @color:Ciechanowski:2019.
-
-FIXME: Discuss relation ship to other packages:
-
-* \pkg{Polychrome} [@color:Polychrome, @color:Polychrome2]: qualitative palettes with more "categories". This can generate 
-  palettes with the (a large number of) 
-  colours spaced far apart from each other in Luv space.
-  It also has functions for visualising colour palettes in different ways 
-  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
-* \pkg{roloc} [@color:roloc,@color:roloc2]: colour conversions, but not between
-  numeric colour spaces;  instead, from numeric colour spaces to
-  colour names.  
-  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
-* \pkg{colorscience} [@color:colorscience]: This package is for Colour
-Science.  It has data sets, does more complex conversions (including
-different chromatic adaptation algorithms).  
-So more sophisticated than \pkg{colorspace}
-in terms of flexibility and low-level detail, BUT it does not do
-palette-generation.
-* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.
+@color:Wilke:2019, and @color:Ciechanowski:2019, among many others.
+
+Further \proglang{R} packages that can complement the palettes provided
+by \pkg{colorspace} include:
+\pkg{Polychrome} [@color:Polychrome; @color:Polychrome2] implements strategies
+for qualitative palettes with many "categories". While the qualitative palettes
+in Section\ \ref{sec:hcl_palettes} yield only about 6--8 clearly distinguishable
+colors due to the fixed chroma and luminance, \pkg{Polychrome} relaxes this
+restriction and can thus find a larger number of colors in CIELUV space that
+are spaced as far apart as possible. It also has several functions for visualizing
+color palettes in different ways. 
+The palette collection packages \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
+also provide a wide range of prespecified palettes, including some qualitative
+schemes with many categories. Note that the palettes are quite diverse, though,
+and not all of them are equally suitable for coding qualitative information.
+The visualization functions in \pkg{colorspace} from Section\ \ref{sec:palette_visualization}
+may be helpful in assessing their properties.
+\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions, but
+not between numeric colour spaces; instead, from numeric color spaces to
+English color names.
 

### 2019-03-14 pkg/colorspace/paper/Toolbox/article.Rmd
COW feedback integrated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@486 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-  General strategies for three types of palettes are provided:
+  Using the HCL color model general strategies for three types of palettes are implemented:
   (1)\ Qualitative for coding categorical information, i.e., where no particular
@@
 
-  To aid selection and application of these palettes the package provides scales
+  To aid selection and application of these palettes the package also contains scales
   for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
@@
 To address these problems, many improved color palettes with better perceptual
-properties have been suggested in the literature over the last one or two decades
+properties have been receiving increasing attention in the literature
 [@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
@@
 
-```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space.
-Top: Hue $H$ changes from 0 (red) via 60 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$.
+```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.4, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space.
+Top: Hue $H$ changes from 0 (red) via 75 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$.
 Center: Chroma $C$ changes from 0 (gray) to 100 (colorful) with fixed $H = 0$ (red) and $L = 65$.
-Bottom: Luminance $L$ changes from 90 (light) to 25 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
-swatchplot(
-  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
-  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
-  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
-  nrow = 3, line = 5.2, off = 0
-)
+Bottom: Luminance $L$ changes from 95 (light) to 15 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
+par(mar = c(0.1, 6.1, 0, 0))
+plot(0, 0, type = "n", xlim = c(0, 5), ylim = c(0, 3), axes = FALSE, xlab = "", ylab = "")
+hcl_prop <- rbind(
+  "Hue\n(Type of color)"    = hcl(0:4 * 75, 60, 65),
+  "Chroma\n(Colorfulness)"  = hcl(0, 0:4 * 25, 65),
+  "Luminance\n(Brightness)" = hcl(260, 25, 95 - 0:4 * 20))
+mtext(rownames(hcl_prop), side = 2, at = 2:0 + 0.65, las = 1, line = 5.2, adj = 0, cex = 1)
+rect(rep(0:4, each = 3), rep(2:0, 5) + 0.3, rep(1:5, each = 3), rep(3:1, 5), col = hcl_prop, border = "lightgray")
+text(rep(0:4, each = 3) + 0.5, rep(2:0, 5) + 0.2, rbind(0:4 * 75, 0:4 * 25, 95 - 0:4 * 20), cex = 0.7)
 ```
@@
 The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
-approach, utiliting the perceptually-based HCL (Hue--Chroma--Luminance) color space
-[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
+approach that gives the user direct access to the construction principles underlying
+its palettes. These are based on simple trajectories in the perceptually-based
+HCL (Hue--Chroma--Luminance) color space [@color:Wiki+HCL] whose axes match those of the human visual
 system very well: Hue (= type of color, dominant wavelength),
 chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
-Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
+Thus, utilizing this color model the \pkg{colorspace} package can derive general and adaptable
 strategies for color palettes; manipulate individual colors and color palettes;
@@
 
-A broad collection of prespecified palettes are shipped in the package, existing
-palettes can be easily tweaked, and new or adapted palettes registered. The prespecified
+A broad collection of prespecified palettes are shipped in the package. In addition, existing
+palettes can be easily tweaked and new or adapted palettes registered. The prespecified
 palettes include suitable HCL color choices that closely approximate of most palettes
@@
 
-To aid choice and application of these palettes there are
+To aid choice and application of these palettes the package provides
 (a) scales for use with \pkg{ggplot2} [@color:ggplot2],
-(b) \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration,
+(b) \pkg{shiny} [@color:shiny] and \pkg{tcltk} [@color:R] apps for interactive exploration,
 (c) visualizations of palette properties, and
@@
 Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
+Section\ \ref{sec:color_spaces} summarizes the \proglang{S}4 color space classes and methods in the package.
 Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
@@
 
-```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
+```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Brief overview of available predefined palettes in \\pkg{colorspace}."}
 library("colorspace")
@@
 A suitable vector of colors can be easily computed by specifying the desired number
-of colors and palette name (see Figure\ \ref{fig:hcl-palettes} for possible palette names), e.g.,
+of colors and the palette name (see Figure\ \ref{fig:hcl-palettes} for possible palette names), e.g.,
 
@@
 The functions `sequential_hcl()`, and `diverging_hcl()` work analogously. Additionally,
-a palette's hue/chroma/luminance parameters can be modified, thus allowing to easily customize
+a palette's hue/chroma/luminance parameters can be modified, thus allowing for easy customization of
 each palette. Moreover, the `choose_palette()`/`hclwizard()` app provides convenient user
@@
 
-As another example for a sequential palette a spine plot is created (see the right
-panel of Figure\ \ref{fig:eustockmarkets-titanic}),
+As another example for a sequential palette, we demonstrate how to create a spine plot (see the right
+panel of Figure\ \ref{fig:eustockmarkets-titanic})
 displaying the proportion of Titanic passengers that survived per class.
-The `"Purples 3"` palette is used which is quite similar to the \pkg{ColorBrewer.org}
+The `"Purples 3"` palette is used, which is quite similar to the \pkg{ColorBrewer.org}
 [@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
@@
 
-To plug the HCL color palettes into \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
+To provide access to the HCL color palettes from within \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
 suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
-The scales are called via the scheme
+The scales are named via the scheme
 
@@
 To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
-and sequential `"Purples 3"` palettes that were also employed above
-(see Figure\ \ref{fig:iris-diamonds}). For the first example, semi-transparent
+and sequential `"Purples 3"` palettes that were also employed above. For the first example, semi-transparent
 shaded densities of the sepal length from the `iris` data are shown, grouped by species
@@
 application for a sequential palette. The elevation of the volcano is brought out clearly,
-focusing with the dark colors on the higher elevations.
+using dark colors to give emphasis to higher elevations.
 The other two displays show that hue is constant in the palette while luminance and chroma vary.
@@
 
-The color spaces, implemented in \pkg{colorspace}, along with corresponding \proglang{S}4 classes
-and eponymous class constructors are:
+The color spaces, implemented in \pkg{colorspace}, along with their corresponding \proglang{S}4 classes
+and eponymous class constructors, are:
 
@@
   linearized RGB above, is a good model for specifying colors in software and for
-  hardware. But it is still unintuitive for humans to work directly with this.
+  hardware. But it is still unintuitive for humans to work directly with this color space.
   Therefore, sRGB is a good place to end up in a color space manipulation but
@@
   See @color:Wiki+CIELUV; @color:Wiki+CIELAB for more details.
-* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
+* `polarLUV()` and `polarLAB()` take polar coordinates in the UV plane
   and AB plane, respectively. Specifically, the polar coordinates of the LUV
@@
 
-All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which are
+All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which is
 internally always represented by matrices with three columns (corresponding to
@@
 `farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
-conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
-though [see the discussion in @color:convertColor]. For more complex conversions, including
+conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable
+[see the discussion in @color:convertColor]. For more complex conversions, including
 different chromatic adaptation algorithms, a more comprehensive color science approach
@@
 As motivated in the previous section,
-the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
+the HCL space is particularly useful for specifying individual colors and color palettes, as its three axes match those
 of the human visual system very well. Therefore, the \pkg{colorspace} package provides
@@
 As suggested by @color:Ihaka:2003 `qualitative_hcl()` distinguishes the underlying categories by a sequence of
-hues while keeping both chroma and luminance constant to give each color in the
+hues while keeping both chroma and luminance constant, to give each color in the
 resulting palette the same perceptual weight. Thus, `h` should be a pair of
@@
 When qualtitative palettes are employed for shading areas in statistical displays
-(e.g., in bar plots, pie charts, or regions in maps), then it is typically
-less distracting to use lighter colors (with moderate chroma and high luminance)
-such as `"Pastel 1"` or `"Set 3"`. When coloring points or lines then more flashy
+(e.g., in bar plots, pie charts, or regions in maps), lighter colors (with moderate
+chroma and high luminance) such as `"Pastel 1"` or `"Set 3"` are typically less
+distracting. By contrast, when coloring points or lines, more flashy
 colors (with high chroma) are often required: On a white background a moderate
@@
 
-As suggested by @color:Zeileis+Hornik+Murrell:2009 `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
+As suggested by @color:Zeileis+Hornik+Murrell:2009, `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
 of increasing (or decreasing) luminance. Thus, the function's `l` argument should provide
@@
 hue `h1` and ending hue `h2`). More details are provided in the next section.
-(b) Instead of a decreasing chroma a triangular chroma trajectory can be employed
-from `c1` over `cmax` to `c2` (or equivalently a vector `c` of
+(b) Instead of a decreasing chroma, a triangular chroma trajectory can be employed
+from `c1` over `cmax` to `c2` (equivalently specified as a vector `c` of
 length 3). This yields high-chroma colors in the middle of the palette that are
@@
 trajectory. Either a vector `power` of length 2 or separate `p1` (for chroma)
-and `p2` (for luminance) can be specified. If missing, the latter defaults to the
+and `p2` (for luminance) can be specified. If the latter is missing, it defaults to the
 former.
@@
 
-All except the last palette are inspired by the \pkg{ColorBrewer.org} palettes with
+All except the last one are inspired by the \pkg{ColorBrewer.org} palettes with
 the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
@@
 
-This uses a slight power transformation with `p1 = 1` in the blue arm of the palette but otherwise essentially
+This uses a slight power transformation with `p1 = 1.1` in the blue arm of the palette but otherwise essentially
 corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
@@
 
-As a somewhat more complex example, a `swatchplot()` is created with three matrices
-of sequential color palettes of blues, purples, reds, and greens. For all palettes,
+Next, we demonstrate a more complex example of a `swatchplot()` with three matrices
+of sequential color palettes of blues, purples, reds, and greens
+(see Figure\ \ref{fig:swatch-brpg}). For all palettes,
 luminance increases monotonically to yield a proper sequential palette. However,
@@
 While the `specplot()` function above works well for bringing out the HCL coordinates
-associated with a given palette, it does not bring out how the palette fits into
+associated with a given palette, it does not show how the palette fits into
 the HCL space. For example, it is not so clear whether high chroma values are
@@
 
-```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with a qualitative palette with varying hue $H$ and chroma fixed at $C = 50$."}
-hclplot(qualitative_hcl(9, c = 50, l = 70))
+```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with the qualitative \\texttt{\"Dynamic\"} palette with varying hue $H$ and chroma fixed at $C = 50$."}
+hclplot(qualitative_hcl(9, "Dynamic"))
 ```
@@
 
-```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes.
+```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes (similar to \\texttt{\"Blues 2\"} and \\texttt{\"Blues 3\"}).
 Left: Linear chroma for $H = 260$.
@@
 
-The workhorse function to do so is `simulate_cvd()` which can take any vector
+The workhorse function to emulate color-vision deficiency is `simulate_cvd()`, which can take any vector
 of valid \proglang{R} colors and transform them according to a certain CVD transformation
@@
 ranging from red (for high values) to blue (for low values).
-The poor results for the RGB rainbow palette are contrasted with a proper sequential
+The poor results for the RGB rainbow palette are contrasted in Figure\ \ref{fig:heatmap-sequential} with a proper sequential
 palette ranging from dark blue to light yellow: `sequential_hcl(11, "Blue-Yellow")`.
@@
 
-In addition to the \pkg{shiny} version, the **palette constructor** app is also available
+In addition to the \pkg{shiny} version, the _palette constructor_ app is also available
 as a \proglang{Tcl}/\proglang{Tk} GUI via \proglang{R} package
@@
 
-This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
-by default starts the \pkg{shiny} version (see Figure\ \ref{fig:choose_palette}).
-The \pkg{tcltk} version is started by default with `choose_palette()`.
-However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.
-
-The GUIs interface the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
-palettes from Section\ \ref{sec:hcl_palettes}. They
+The \pkg{shiny} version of this GUI is shown in Figure\ \ref{fig:choose_palette}.
+It interfaces the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
+palettes from Section\ \ref{sec:hcl_palettes}. The GUIs
 allow for interactive modification of the arguments of the respective
@@
 * `max_chroma()`: Compute maximum chroma for given hue and luminance in HCL space.
-* `mixcolor()`: Additive mixing of two colors by computing the convex combination.
+* `mixcolor()`: Additively mix two colors by computing their convex combination.
 
@@
 
-Thus, the `desaturate()` function converts any given hex color code or named
+The `desaturate()` function converts any given hex color code or named
 \proglang{R} color to the corresponding HCL coordinates and sets the chroma to zero.
@@
 trying a specific HCL coordinate by calling `polarLUV()`. To avoid having to
-`fixup` the color upon conversion to RGB `hex()` color codes, the
+fix up the color upon conversion to RGB `hex()` color codes, the
 `max_chroma()` function computes (approximately) the maximum chroma

### 2019-03-18 pkg/colorspace/R/divergingx_hcl.R
paper now on arxiv -> references & citation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@487 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -51,6 +51,9 @@
 #'
 #' @return A character vector with (s)RGB codings of the colors in the palette.
 #' @seealso \code{\link[colorspace]{sequential_hcl}}, \code{\link[colorspace]{diverging_hcl}}
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
+#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
 #' @keywords color
 #' @examples
 #' ## show emulated CARTO/ColorBrewer.org palettes

### 2019-03-18 pkg/colorspace/R/hcl_palettes.R

@@ -133,6 +133,10 @@
 #' Visualizations.  \emph{Bulletin of the American Meteorological Society},
 #' \bold{96}(2), 203--216.
 #' \doi{10.1175/BAMS-D-13-00155.1}
+#'
+#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
+#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
 #' @keywords color
 #' @examples
 #' ## overview of all _named_ HCL palettes

### 2019-03-18 pkg/colorspace/R/simulate_cvd.R

@@ -35,6 +35,10 @@
 #'   \doi{10.1109/TVCG.2009.113}
 #'   Online version with supplements at
 #'   \url{http://www.inf.ufrgs.br/~oliveira/pubs_files/CVD_Simulation/CVD_Simulation.html}.
+#'
+#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
+#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
 #' @keywords colors cvd colorblind
 #' @seealso \code{\link{cvd}}
 #' @export

### 2019-03-18 pkg/colorspace/R/specplot.R

@@ -57,6 +57,10 @@
 #' Visualizations.  \emph{Bulletin of the American Meteorological Society},
 #' \bold{96}(2), 203--216.
 #' \doi{10.1175/BAMS-D-13-00155.1}
+#'
+#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
+#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
 #' @keywords hplot
 #' @examples
 #' ## spectrum of the (in)famous RGB rainbow palette (in both RGB and HCL)

### 2019-03-18 pkg/colorspace/paper/Toolbox/slides.Rnw
arxiv paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@488 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Ihaka R, Murrell P, Hornik K, Fisher JC, Stauffer R, Wilke CO, McWhite CD, Zeileis A (2019).
-  \emph{colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-  \url{http://colorspace.R-Forge.R-project.org/}
+Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+  \dquote{colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
+  arXiv:1903.06490, arXiv.org E-Print Archive.
+  \url{http://arxiv.org/abs/1903.06490}
 

### 2019-03-18 pkg/colorspace/paper/Toolbox/article.Rmd
re-introduce changes from Claire from r466 that were erroneously undone in r469
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@492 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Using the physiologically-based model for simulating color vision deficiency (CVD)
-of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
-emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
-and tritanope (blue cone cells defective).
-
-Below we briefly describe our \proglang{R} interface to these emulation techniques
-and show them in practice for a heatmap with sequential palette. Another example
-with diverging palette is available at
+Different kinds of limitations can be emulated using the physiologically-based model
+for simulating color vision deficiency (CVD) of @color:Machado+Oliveira+Fernandes:2009:
+deuteranomaly (green cone cells defective),
+protanomaly (red cone cells defective), and
+tritanomaly (blue cone cells defective).
+While most other CVD simulations handle only dichromacy, where one of three cones is
+non-functional, @color:Machado+Oliveira+Fernandes:2009 provide a unified model of
+both dichromacy and anomalous trichromacy, where one cone has shifted spectral
+sensitivity. As anomalous trichromacy is the most common form of color vision
+deficiency, it is important to emulate along with the rarer, but more severe
+dichromacy. Below we briefly describe our \proglang{R} interface to these emulation
+techniques and show them in practice for a heatmap with sequential palette.
+Another example with diverging palette is available at
 <http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
-Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
-palette is almost always a bad choice in scientific displays. See
+Finally, CVD emulation is particularly useful for bringing out why the RGB
+rainbow palette is almost always a bad choice in scientific displays. See
 <http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
@@
 
-The workhorse function to emulate color-vision deficiency is `simulate_cvd()`, which can take any vector
-of valid \proglang{R} colors and transform them according to a certain CVD transformation
-matrix and transformation equation. The transformation matrices have been established
-by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
-`deutanomaly_cvd`, and `tritanomaly_cvd`.
-The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
-simulating the corresponding kind of color blindness with a given severity (calling
-`simulate_cvd()` internally).
-
-For further guidance on color blindness in relation to statistical graphics
-see @color:Lumley:2006 which accompanies the \proglang{R} package \pkg{dichromat} [@color:dichromat]
-and is based on earlier emulation techniques
+The workhorse function to emulate color vision deficiencies is `simulate_cvd()`
+which can take any vector of valid \proglang{R} colors and transform them according to a
+certain CVD transformation matrix and transformation equation. The
+transformation matrices have been established by @color:Machado+Oliveira+Fernandes:2009 and are
+provided in objects `protanomaly_cvd`, `deutanomaly_cvd`, and `tritanomaly_cvd`.
+The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level
+functions for simulating the corresponding kind of color blindness with a given
+`severity` (calling `simulate_cvd()` internally). A severity of `1` corresponds to
+dichromacy, `0` to normal color vision, and intermediate values to varying
+severities of anomalous trichromacy. For further guidance on color blindness in
+relation to statistical graphics see @color:Lumley:2006 which accompanies the \proglang{R}
+package \pkg{dichromat} [@color:dichromat] and is based on earlier emulation techniques
 [@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].

### 2019-06-25 pkg/colorspace/paper/Toolbox/slides.Rnw
add info about grDevices::hcl.colors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@495 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\subsection{Base R}
+
+\begin{frame}
+\frametitle{Base R}
+
+\textbf{In 3.6.0:} All prespecified palettes also via \code{grDevices::hcl.colors()}.
+
+\vspace*{-1.8cm}
+
+\setkeys{Gin}{width=\textwidth}
+<<hcl-colors, fig = TRUE, width = 11, height = 8>>=
+hcl.swatch <- function(palette = hcl.pals(), n = 5, nrow = 11,
+  border = if (n < 15) "black" else NA)
+{
+    cols <- sapply(palette, hcl.colors, n = n)
+    ncol <- ncol(cols)
+    nswatch <- min(ncol, nrow)
+
+    par(mar = rep(0.1, 4),
+        mfrow = c(1, min(10, ncol %/% nrow + 1)),
+        pin = c(1, 0.5 * nswatch),
+        cex = 0.7)
+
+    while (length(palette)) {
+        subset <- 1:min(nrow, ncol(cols))
+        plot.new()
+        plot.window(c(0, n), c(0, nrow + 1))
+        text(0, rev(subset) + 0.1, palette[subset], adj = c(0, 0))
+        y <- rep(subset, each = n)
+        rect(rep(0:(n-1), n), rev(y), rep(1:n, n), rev(y) - 0.5,
+             col = cols[, subset], border = border)
+        palette <- palette[-subset]
+        cols <- cols[, -subset]
+    }
+
+    par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1), cex = 1)
+}
+hcl.swatch()
+@
+
+\end{frame}
+
 \subsection{Recommendations}
@@
 
+\textbf{Colors and palettes:}
 \begin{itemize}
@@
   \item Triangular chroma trajectories useful for distinguishing central colors.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{R packages:}
+\begin{itemize}
+  \item \pkg{colorspace} facilitates exploration, manipulation, and assessment.
+  \item HCL approximations of palettes from \pkg{RColorBrewer}, \pkg{rcartocolor}, \pkg{scico}, \dots
+  \item Prespecified palettes are also easily available in base R.
 \end{itemize}

### 2019-07-10 pkg/colorspace/index.Rmd
added pointers to grDevices::hcl.colors and re-compiled pkgdown site
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@497 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r ggplot2, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
+```{r ggplot2, echo = FALSE, message = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
 library("ggplot2")

### 2019-07-11 pkg/colorspace/paper/Toolbox/slides.Rnw
small tweak
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@496 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
shorter title without 'for Statistical Graphics'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@498 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
+\title{A Toolbox for Manipulating and Assessing Color Palettes}
 \subtitle{\phantom{.}\\Achim Zeileis, Jason C. Fisher, Kurt Hornik, Ross Ihaka, Claire D. McWhite, Paul Murrell, Reto Stauffer, Claus O. Wilke}
-\author{\vspace*{-6.83cm}\hspace*{12.5cm}\includegraphics[width=1.6cm]{palmfooter.png}\\[5.5cm]}
+\author{\vspace*{-6.3cm}\hspace*{12.5cm}\includegraphics[width=1.6cm]{palmfooter.png}\\[5.5cm]} %% -6.83 with "for Statistical Graphics"
 \URL{http://colorspace.R-Forge.R-project.org/}
@@
 
-\section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
+\section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes}
 
@@
 \begin{frame}
-\frametitle{Base R}
+\frametitle{Base R\only<2>{: Why you might not need our package after all}}
 

### 2019-08-23 pkg/colorspace/paper/Toolbox/article.Rmd
updated vignettes with feedback from JSS reviewers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@500 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-  Using the HCL color model general strategies for three types of palettes are implemented:
+  Using the HCL color model, general strategies for three types of palettes are implemented:
   (1)\ Qualitative for coding categorical information, i.e., where no particular
@@
 
-  To aid selection and application of these palettes the package also contains scales
-  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
+  To aid selection and application of these palettes, the package also contains scales
+  for use with \pkg{ggplot2}, \pkg{shiny} and \pkg{tcltk} apps for interactive
   exploration, visualizations of palette properties,
@@
 [@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
-Many systems for systems for statistical and scientific computing provide
-infrastructure for such color palettes, e.g., for \proglang{R} [@color:R] the
-list of useful packages encompasses:
+Many systems for statistical and scientific computing provide
+infrastructure for such color palettes. For example, for \proglang{R} [@color:R] the
+list of useful packages encompasses
 \pkg{RColorBrewer} [@color:RColorBrewer],
@@
 \pkg{wesanderson} [@color:wesanderson],
-or \pkg{scico} [@color:scico] among many others.
+and \pkg{scico} [@color:scico] among many others.
 Furthermore, packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
@@
 HCL (Hue--Chroma--Luminance) color space [@color:Wiki+HCL] whose axes match those of the human visual
-system very well: Hue (= type of color, dominant wavelength),
-chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
+system very well: Hue (type of color, dominant wavelength),
+chroma (colorfulness), luminance (brightness), see Figure\ \ref{fig:hcl-properties}.
 Thus, utilizing this color model the \pkg{colorspace} package can derive general and adaptable
@@
 
-A broad collection of prespecified palettes are shipped in the package. In addition, existing
+A broad collection of prespecified palettes is shipped in the package. In addition, existing
 palettes can be easily tweaked and new or adapted palettes registered. The prespecified
-palettes include suitable HCL color choices that closely approximate of most palettes
+palettes include suitable HCL color choices that closely approximate most palettes
 from packages \pkg{RColorBrewer}, \pkg{rcartocolor}, and \pkg{viridis} by using only a
@@
 (c) visualizations of palette properties, and
-(d) accompanying manipulation utilities (like desaturation, lighten/darken, and emulation of color vision deficiencies).
-
-The remainder of the manuscript is organized as follows:
+(d) accompanying manipulation utilities (like converting to grayscale by desaturation, lighten/darken, and emulation of color vision deficiencies).
+
+The remainder of the paper is organized as follows:
 Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
@@
 Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
-before Section\ \ref{sec:summary} concludes the manuscript.
+before Section\ \ref{sec:summary} concludes the paper.
 
@@
 [@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
-employed, yielding a dark purple and light gray.
+employed, yielding a dark purple and a light gray.
 
@@
 Left: Time series plot of log-prices from \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette.
-Right: Spine plot with survival proportions across passenger classes in the \\code{titanice} data with \\code{sequential\\_hcl(2, \"Purples 3\")} palette."}
+Right: Spine plot with survival proportions across passenger classes in the \\code{titanic} data with \\code{sequential\\_hcl(2, \"Purples 3\")} palette."}
 par(mar = c(5, 4, 1, 2), mfrow = c(1, 2))
@@
 To provide access to the HCL color palettes from within \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
-suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
+suitable discrete and/or continuous \pkg{ggplot2} color scales are provided.
 The scales are named via the scheme
@@
 
-```{r allplots-qualitative, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for \\code{qualitative\\_hcl(4, \"Dark 3\")} palette.
+```{r allplots-qualitative, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for the \\code{qualitative\\_hcl(4, \"Dark 3\")} palette.
 Left: Demo bar plot.
@@
 is (almost) constant in the palette while the hue changes linearly along the color
-"wheel". Ideally, chroma would have also been constant to completely balance the colors.
+"wheel" (from degree 0 to 270). Ideally, chroma would have also been constant to completely balance the colors.
 However, at this luminance the maximum chroma differs across hues so that the palette
@@
 
-```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for \\code{sequential\\_hcl(4, \"Purples 3\")} palette.
+```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for the \\code{sequential\\_hcl(4, \"Purples 3\")} palette.
 Left: Demo heatmap.
@@
 Luminance increases monotonically from dark to light (as required for a proper sequential palette).
-Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
+Chroma is triangular-shaped which allows the viewer to better distinguish the middle colors in the palette
 when compared to a monotonic chroma trajectory.
@@
 
-```{r color-spaces, fig.width = 9.5, fig.height = 4.5, echo = FALSE, out.width = "0.9\\textwidth", fig.cap = "Relationship between three-dimensional color spaces implemented in \\pkg{colorspace}.
+```{r color-spaces, fig.width = 9.5, fig.height = 4.5, echo = FALSE, out.width = "0.9\\textwidth", fig.cap = "Relationships among three-dimensional color spaces implemented in \\pkg{colorspace}.
 Color models that are (or try to be) perceptually-based are displayed with circles, other color models with rectangles."}
@@
   y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
-  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
+  lab = c("polarLAB", "polarLUV\n(HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
   stringsAsFactors = FALSE
@@
 text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
-text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
+text(4.56, 2, "gamma\n= 2.4", cex = 0.7, pos = 3)
 ```
@@
 At the core of the \pkg{colorspace} package are various utilities for computing with
-color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
+color spaces [@color:Wiki+Colorspace], as the name of the package conveys. Thus, the package helps to map various three-dimensional
 representations of color to each other [@color:Ihaka:2003]. A particularly important
@@
 For completeness further standard color models are included as well in the package.
-The connections are illustrated in Figure\ \ref{fig:color-spaces}.
+Their connections are illustrated in Figure\ \ref{fig:color-spaces}.
 Color models that are (or try to be) perceptually-based are displayed with circles
@@
 
-* `RGB()` for the classic Red-Green-Blue color model that mixes three primary colors
+* `RGB()` for the classic Red-Green-Blue color model, which mixes three primary colors
   with different intensities to obtain a spectrum of colors. The advantage of this
@@
 internally always represented by matrices with three columns (corresponding to
-the different three dimensions).
+the three dimensions).
 
@@
 
-For working with the implemented \proglang{S}4 color spaces various utilities are available:
+Various utilities are available for working with the implemented \proglang{S}4 color spaces:
 
@@
 * `writehex()`: Writes hex color strings to a text file.
-* `whitepoint()`: Query and change the white point employed in conversions
-  from CIE XYZ to RGB. Defaults to D65.
+* `whitepoint()`: Query and change the so-called white point employed in conversions
+  from CIE XYZ to RGB. Defaults to D65 that has been specified by the CIE to
+  approximate daylight [@color:Poynton].
 
@@
 
-With `coords(x)` or `coords(y)` the displayed coordinates can also be extracted
-as numeric matrices. And from sRGB we can also coerce to HSV for example:
+The displayed coordinates can also be extracted as numeric matrices by
+`coords(x)` or `coords(y)`. We can also, for example, coerce from sRGB to HSV:
 
@@
 palettes in the second row of Figure\ \ref{fig:hcl-palettes-principles}
-bring out clearly that luminance differences (= light-dark contrasts)
+bring out clearly that luminance differences (light-dark contrasts)
 are crucial for sequential and diverging palettes while qualitative palettes
@@
 To compute the actual color hex codes (representing sRGB coordinates) based on these HCL parameters, the functions
-`qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()` can be used.
+`qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()` can be used which are described in more detail
+in the following sections.
 Either all parameters can be specified "by hand" through the HCL parameters, an entire
@@
 Chroma `c` (or equivalently `c1`) and luminance `l` (or equivalently
-`l1`) are constants.
+`l1`) are constants. Finally, `fixup` indicates
+whether colors with out-of-range coordinates should be corrected
+(as illustrated in Figure\ \ref{fig:allplots-qualitative}).
 
@@
 hue range. The remaining four palettes are taken from @color:Ihaka:2003. They are
-based on the same chroma (= 50) and luminance (= 70) but the hue is restricted to
+based on the same chroma (50) and luminance (70) but the hue is restricted to
 different intervals.
@@
 and then decrease chroma from some value (`c` or `c1`) to zero (i.e., gray) along with
-increasing luminance. For bringing out the extremes (a dark high-chroma color vs.\ a
-light gray) this is already very effective, see `"Blues 2"`, `"Purples 2"`, `"Reds 2"`,
+increasing luminance. This is already very effective for bringing out the extremes
+(a dark high-chroma color vs.\ a light gray), see `"Blues 2"`, `"Purples 2"`, `"Reds 2"`,
 and `"Greens 2"`.
@@
 Qualitative   & Linear  		          & Constant	               & Constant	  \\
-Sequential    & Constant (= single-hue) \emph{or} & Linear (+ power) \emph{or} & Linear (+ power) \\
-	      & Linear (= multi-hue)	          & Triangular (+ power)       & 		  \\
+Sequential    & Constant (single-hue) \emph{or}   & Linear (+ power) \emph{or} & Linear (+ power) \\
+	      & Linear (multi-hue)	          & Triangular (+ power)       & 		  \\
 Diverging     & Constant ($2\times$)	          & Linear (+ power) \emph{or} & Linear (+ power) \\
@@
 where $j$ is the intensity at which $c_{\max}$ is assumed. It is constructed such that
-the slope to the left is minus the slope to the right of $j$:
+the slope to the left is the negative of the slope to the right of $j$:
 
@@
 a custom palette so that it can subsequently be reused with a new dedicated name.
-This is supported by adding a `register = "..."` argument once to a call to
+This is supported by adding a `register` argument once to a call to
 `qualitative_hcl()`, `sequential_hcl()`, or `diverging_hcl()`:
@@
   `alpha` and `rev`.
-* When registering a new palette with an old name that was already available
-  previously, the old palette gets overwritten. We recommend not to overwrite
+* When registering a new palette with a previously-used name, the old palette
+  gets overwritten. We recommend not to overwrite
   the palettes that are predefined in the package (albeit it is technically possible).
@@
 and luminance parameters. Thus, it does not pose any restrictions that
-the two "arms" of the palette need to be balanced and also allows to go
+the two "arms" of the palette need to be balanced and also may go
 through a non-gray neutral color (typically light yellow).
@@
 The flexible specification of HCL-based color palettes in \pkg{colorspace} allows
-to closely approximate color palettes from various other packages:
+one to closely approximate color palettes from various other packages:
 
@@
 (see also Section\ \ref{sec:specplot}) for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
-and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
+and `viridis::viridis(7)`) and two purple/red/yellow palettes (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
 Each panel compares the hue, chroma, and luminance trajectories
 of the original palettes (top swatches, solid lines) and their HCL-based approximations (bottom swatches, dashed lines).
-The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
+The palettes are not identical but very close for most colors. Note also that the chroma trajectories from the HCL
 palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
@@
 Figure\ \ref{fig:cividis} shows the corresponding `specplot()` along with an HCL-based approximation.
-What is unusual about this palette: The hue and chroma trajectories would suggest a diverging palette,
+This palette is unusual: The hue and chroma trajectories would suggest a diverging palette,
 as there are two "arms" wth different hues and a zero-chroma point in the center. However,
@@
 of sequential color palettes of blues, purples, reds, and greens
-(see Figure\ \ref{fig:swatch-brpg}). For all palettes,
-luminance increases monotonically to yield a proper sequential palette. However,
+(see Figure\ \ref{fig:swatch-brpg}).
+
+```{r swatch-brpg, fig.width = 11.5, fig.height = 3, fig.cap = "Variations of blue, purple, red, and green palettes with single hue and monotonic chroma (left), single hue and triangular chroma (center), and multiple hues and triangular chroma (right)."}
+bprg <- c("Blues", "Purples", "Reds", "Greens")
+swatchplot(
+"Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
+"Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
+"Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
+nrow = 5, line = 5)
+```
+
+For all palettes, luminance increases monotonically to yield a proper sequential palette. However,
 the hue and chroma handling is somewhat different to emphasize different parts of the
@@
 
-```{r swatch-brpg, fig.width = 11.5, fig.height = 3, fig.cap = "Variations of blue, purple, red, and green palettes with single hue and monotonic chroma (left), single hue and triangular chroma (center), and multiple hues and triangular chroma (right)."}
-bprg <- c("Blues", "Purples", "Reds", "Greens")
-swatchplot(
-"Single-hue"		= t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
-"Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
-"Multi-hue (advanced)"  = t(sapply(bprg,	   sequential_hcl, n = 7)),
-nrow = 5, line = 5)
-```
-
 ## HCL (and RGB) spectrum {#sec:specplot}
@@
 coordinates. As the hues for low-chroma colors are not (or only poorly) identified,
-by default a smoothing is applied to the hues. Also, to avoid jumps from 0 to 360 or vice
+they are smoothed by default. Also, to avoid jumps from 0 to 360 or vice
 versa, the hue coordinates are shifted suitably.
@@
 if these palettes are constructed very differently (e.g., like the
-highly-saturated base \proglang{R} palettes). To infer the default `type` of projection
+highly saturated base \proglang{R} palettes). To infer the default `type` of projection
 `hclplot()` assesses the luminance trajectory and sets the default correspondingly:
@@
 
-Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
+Note that for $H = 260$ it is possible to go to dark colors (low luminance) with
 high chroma while this is not possible to the same extent for $H = 245$ because of
@@
 Maunga Whau `volcano` data from base \proglang{R}. This heatmap is easily available as
-`demoplot(..., "heatmap")` where `...` is the color vector to be used, e.g.,
+`demoplot(x, "heatmap")` where `x` is the color vector to be used, e.g.,
 
@@
 several graphical user interfaces (GUIs) are provided within the package as
-\pkg{shiny} apps [@color:shiny]. All of these GUIs/apps can either be run locally
-from within \proglang{R} but are also provided online at <http://hclwizard.org/>.
+\pkg{shiny} apps [@color:shiny]. All of these GUIs/apps can be run locally
+from within \proglang{R} and are also provided online at <http://hclwizard.org/>.
 
@@
 palette-generating functions, i.e., starting/ending hue, minimal/maximal chroma,
-minimal maximal luminance, and power transformations that control how
+minimal/maximal luminance, and power transformations that control how
 quickly/slowly chroma and/or luminance are changed through the palette.
@@
 This GUI can be started with `cvd_emulator()`.
-It allows to upload a raster image in
+It supports uploading a raster image in
 JPG or PNG format which is then checked for various kinds of color vision deficiencies
@@
 
-Already this simple example shows that the three RGB rainbow colors have very
-different grayscale levels. This can be brought even more clearly when using 
+Even this simple example suffices to show that the three RGB rainbow colors have very
+different grayscale levels. This deficiency is even clearer when using 
 a full color wheel (of colors with hues in [0, 360] degrees). While the RGB
@@
 colorful. A compromise that sometimes works well is to adjust the luminance coordinate in HCL space but
-to take the chroma coordinate corresponding the HLS transformation.
+to take the chroma coordinate corresponding to the HLS transformation.
 
@@
 Regardless of the chosen color space, the adjustment of the `L` component by a certain `amount` can occur by two methods, relative
-(the default) or absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
-`L * (1 - amount)` (relative). See `?lighten` and `?darken` for more details.
+(the default) or absolute. For example `L - 100 * amount` is used for absolute darkening, or
+`L * (1 - amount)` for relative darkening. See `?lighten` and `?darken` for more details.
 
@@
 ```{r max-chroma-L50}
-max_chroma(seq(0, 360, by = 60), 50)
+max_chroma(h = seq(0, 360, by = 60), l = 50)
 ```
@@
 ```{r max-chroma-H120}
-max_chroma(120, seq(0, 100, by = 20))
+max_chroma(h = 120, l = seq(0, 100, by = 20))
 ```
@@
 
-This manuscript provides an overview of the broad capabilities of the
+This paper provides an overview of the broad capabilities of the
 \pkg{colorspace} package for selecting individual colors or color palettes,
@@
 \pkg{viridis}, or \pkg{scico}) special emphasis is given to flexibility of
-the palettes that can be adjusted to the particular needs of a given
-data visulization. The manuscript also provides various tips and tricks
+the palettes, which can be adjusted to the particular needs of a given
+data visulization. The paper also provides various tips and tricks
 for choosing an effective palette in a given situation. Further useful
@@
 
-Further \proglang{R} packages that can complement the palettes provided
-by \pkg{colorspace} include:
+There are other \proglang{R} packages that can complement the palettes provided by \pkg{colorspace}.
 \pkg{Polychrome} [@color:Polychrome; @color:Polychrome2] implements strategies
@@
 may be helpful in assessing their properties.
-\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions, but
-not between numeric colour spaces; instead, from numeric color spaces to
-English color names.
+\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions,
+not between numeric color spaces, but rather from numeric color spaces to English color names.
 
@@
 and available at <https://github.com/retostauffer/python-colorspace>.
-In the manuscript we focus on the more mature \proglang{R} implementation but
+In the paper we focus on the more mature \proglang{R} implementation but
 replication materials for most examples are not only provided for \proglang{R} but

### 2019-08-26 pkg/colorspace/paper/Toolbox/article.Rmd
added discussion of human perceptual axes and corresponding implementation in HCL
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@501 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+## Human color vision and the HCL color model
+
+It has been hypothesized that human color vision has evolved in three distinct stages:
+
+1. Perception of light/dark contrasts (monochrome only).
+2. Yellow/blue contrasts (usually associated with our notion of warm/cold colors).
+3. Green/red contrasts (helpful for assessing the ripeness of fruit).
+
+See @color:Kaiser+Boynton:1996, @color:Knoblauch:2002, @color:Ihaka:2003, @color:Lumley:2006, @color:Zeileis+Hornik+Murrell:2009
+for more details and references. Thus, colors can be described using a 3-dimensional space as shown in the left panel
+of Figure\ \ref{fig:hcl-axes}.
+However, for describing colors in such a space, it is more natural for humans to employ
+polar coordinates in the color plane (yellow/blue vs.\ green/red, visualized by the dashed circle in Figure\ \ref{fig:hcl-axes})
+plus a third light/dark axis.
+Hence, color models that attempt to capture these perceptual axes are also called perceptually-based color spaces.
+As already argued above, the HCL model captures these dimensions very well, calling them: _hue_, _chroma_, and _luminance_.
+The corresponding sRGB gamut, i.e., the HCL colors that can also be represented in sRGB, is visualized
+in the right panel of Figure\ \ref{fig:hcl-axes} [by @color:Horvath+Lipka:2016]. An animated version of the same plot
+is provided online by @color:Horvath+Lipka:2017.
+
+```{r human-axes, echo = FALSE, fig.width = 4.5, fig.height = 4.5, fig.show = "hide"}
+par(mar = rep(0, 4))
+midgray <- gray(134/255)
+res <- persp(c(-1, 1), c(-1, 1), matrix(c(-1, 1, 1, -1), 2, 2),
+  xlim = c(-0.77, 0.77), ylim = c(-0.77, 0.77), zlim = c(-0.77, 0.77),
+  border = "transparent", box = FALSE, axes = FALSE, phi = 20, theta = 30, r = 50)
+ax <- trans3d(c(1, -1, 0, 0, 0, 0), c(0, 0, 1, -1, 0, 0), c(0, 0, 0, 0, 1, -1), res)
+circ <- (-45:45)/50
+circ <- trans3d(rep(circ, 2), c(sqrt(circ[1]^2 - circ^2), -sqrt(circ[1]^2 - circ^2)), 0, res)
+arrows(ax$x[c(1, 3, 5)], ax$y[c(1, 3, 5)], ax$x[c(2, 4, 6)], ax$y[c(2, 4, 6)], code = 3, length = 0.1, col = midgray, lwd = 3)
+lines(circ$x, circ$y, col = midgray, lty = 2)
+text(ax$x, ax$y, c("red", "green", "yellow", "blue", "light", "dark"), col = midgray, pos = c(4, 2, 4, 2, 3, 1))
+```
+
+\begin{figure}[t!]
+\centering 
+\includegraphics[width=0.42\textwidth]{Figures/human-axes-1.pdf} 
+\includegraphics[width=0.49\textwidth]{Figures/hcl-wikimedia.png}
+\caption{Visualization of axes capturing human color vision (left) and the corresponding HCL color model (right).}\label{fig:hcl-axes}
+\end{figure}
+
+The shape of the HCL space is a distorted double cone which is seen best by looking at
+vertical slices, i.e., chroma-luminance planes for given hues. For example, the left panel
+in Figure\ \ref{fig:hcl-projections} depicts
+the chroma-luminance plane for a certain blue (hue = 255). Along with luminance the colors change from
+dark to light. With increasing chroma the colors become more colorful, where the highest chroma is
+possible for intermediate luminance.
+
+As some colors are relatively dark (e.g., blue and red assume their maximum chroma for relatively low luminances)
+while others are relatively light (e.g., yellow and green), horizontal slices of hue-chroma planes for given hue
+have somewhat irregular shapes. The right panel in Figure\ \ref{fig:hcl-projections} shows such a
+hue-chroma plane for moderately light colors (luminance = 70).
+At that luminance, green and orange can become much more colorful compared to blue or red.
+
+```{r hcl-projections, echo = FALSE, fig.width = 9.5, fig.height = 4.5, dev = "png", fig.cap = "Vertical (left) and horizontal (right) slices of the HCL space yielding a chroma-luminance plane for given hue and a hue-chroma plane for given luminance, respectively."}
+par(mfrow = c(1, 2))
+hclplot(sequential_hcl(5), h = 255, c = 125, size = 0, lwd = 0)
+hclplot(qualitative_hcl(5), l = 70, c = 125, size = 0, lwd = 0)
+```
+
 ## Utilities
 
-Various utilities are available for working with the implemented \proglang{S}4 color spaces:
+Several utilities are available for working with the \proglang{S}4 classes implementing the
+color spaces listed above.
 
@@
   HCL palettes with the more flashy and unbalanced HSV palettes.
+
+Meanwhile, base \proglang{R} has also adopted the HCL-based palettes from \pkg{colorspace} into the `hcl.colors()`
+function in \pkg{grDevices} [@color:grDevices]. This provides all the named palettes introduced in
+\pkg{colorspace} (with the same names, and defaulting to `"Viridis"`) but without the flexibility
+to modify or adapt existing palettes.
 

### 2019-08-26 pkg/colorspace/paper/Toolbox/article.Rmd
incorporate improvements suggested by reviewer 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@502 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 specified through suitable trajectories in the HCL (Hue-Chroma-Luminance) color space.
-A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes}):
+A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes},
+some of these are illustrated in more detail later):
 
@@
 [@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
-employed, yielding a dark purple and a light gray.
+employed: a dark purple that is highlighted against a light gray.
 
@@
 versa, the hue coordinates are shifted suitably.
-By default, the resulting trajectories in the HCL spectrum are visualized by a simple line plot:
+By default, the resulting trajectories
+in the HCL spectrum are visualized by a simple line plot where the
+x-axis gives the ordering of the colors in the palette. The
+y-axis depicts the following information:
 
@@
 
+The app has been influenced considerably by \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].
+Similarities include the selection of a qualitative, sequential, or diverging palette
+from a list of predefined colors along with an example visualization. However, unlike
+\pkg{ColorBrewer.org} our \pkg{shiny} app allows to tweak the HCL parameters underlying
+each palette. This makes the app much more flexible but also more complex, potentially requiring more
+thought and experience. Due to the flexibility, our app cannot automatically judge safety regarding
+color vision deficiencies and printers/photocopiers (as \pkg{ColorBrewer.org} does) but instead it allows to emulate color vision
+deficiencies and desaturation. Finally, \pkg{ColorBrewer.org} is geared towards cartography
+(albeit its palettes are useful much more generally) while our \pkg{shiny} app includes a
+broader range of illustrative displays.
+
 ## Choose individual colors with the HCL color model

### 2019-08-28 pkg/colorspace/paper/Toolbox/article.Rmd
add further Wikepedia link
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@503 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
- changed default palette in specplot from rainbow_hcl(3) to   qualitative_hcl(3) which is darker and has more chroma
- incorporated feedback from Claire's thorough proofreading


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@504 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Brief overview of available predefined palettes in \\pkg{colorspace}."}
+```{r hcl-palettes, echo = TRUE, eval = FALSE}
 library("colorspace")
 hcl_palettes(plot = TRUE)
+```
+```{r hcl-palettes, echo = FALSE, fig.height = 6.75, fig.width = 12, fig.cap = "Brief overview of available predefined palettes in \\pkg{colorspace}."}
+library("colorspace")
+hcl_palettes(plot = TRUE, mar = c(0.1, 7, 0, 0), line = 6)
 ```
@@
 The color vectors returned by the HCL palette functions can usually be passed directly
-to most base graphics function, typically through the `col` argument. Here, the `q4`
+to most base graphics, typically through the `col` argument. Here, the `q4`
 vector created above is used in a time series display (see the left panel of
@@
 
-```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for the \\code{sequential\\_hcl(4, \"Purples 3\")} palette.
+```{r allplots-sequential, echo = FALSE, fig.height = 4.5, fig.width = 14, dev = "png", fig.cap = "Palette visualization and assessment for the \\code{sequential\\_hcl(9, \"Purples 3\")} palette.
 Left: Demo heatmap.
@@
   colors, hence it was widely adopted and still is the basis for color specifications
-  in many systems. For example, the hex color codes are employed in HTML but also in \proglang{R}.
+  in many systems. For example, hex color codes are employed in HTML but also in \proglang{R}.
   However, the RGB model also has some important drawbacks: It does not take into
@@
   and AB plane, respectively. Specifically, the polar coordinates of the LUV
-  model are known as the HCL (Hue-Chroma-Luminance) model. These capture the
+  model are known as the HCL (Hue-Chroma-Luminance) model [@color:Wiki+HCL]. These capture the
   human perceptual axes very well without confounding effects as in the HSV or
@@
 
-Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
-\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
-Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
-[computed in high-level \proglang{R}, @color:R] and
-`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
+Note that since the inception of the color space conversion tools in \pkg{colorspace}
+[in \proglang{C}, @color:Ihaka:2003] other \proglang{R} tools for this purpose became available,
+notably `grDevices::convertColor()` [in high-level \proglang{R}, @color:R] and
+`farver::convert_colour()` [in \proglang{C++}, @color:farver]. For many basic color
 conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable
@@
 
-* `as()` method: Conversions of a `color` object to the various color spaces,
+* `as()` method: Convert a `color` object to the various color spaces,
   e.g., `as(x, "sRGB")`.
@@
   from RGB coordinates or hex color strings.
-* `writehex()`: Writes hex color strings to a text file.
+* `writehex()`: Write hex color strings to a text file.
 * `whitepoint()`: Query and change the so-called white point employed in conversions
@@
 
-When qualtitative palettes are employed for shading areas in statistical displays
+When palettes are employed for shading areas in statistical displays
 (e.g., in bar plots, pie charts, or regions in maps), lighter colors (with moderate
@@
 
-All except the last one are inspired by the \pkg{ColorBrewer.org} palettes with
+All except the last are inspired by the \pkg{ColorBrewer.org} palettes with
 the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
@@
 Note that the palettes differ substantially in the amount of chroma and luminance
-contrasts, respectively. For example, many palettes go from a dark high-chroma color
+contrasts. For example, many palettes go from a dark high-chroma color
 to a neutral low-chroma color (e.g., `"Reds"`, `"Purples"`, `"Greens"`, `"Blues"`) or even
@@
 `diverging_hcl()` codes the underlying numeric values by a triangular luminance
-sequence with different hues in the left and in the right "arm" of the palette. Thus,
+sequence with different hues in the left and in the right "arms" of the palette. Thus,
 it can be seen as a combination of two sequential palettes with some restrictions:
@@
 (or `p1` and `p2`) are used. For more flexible diverging palettes without
-the restrictrictions above (and consequently more parameters) see the `divergingx_hcl()`
+the restrictions above (and consequently more parameters) see the `divergingx_hcl()`
 palettes introduced below.
@@
 
-As emphasized in Figure\ \ref{fig:hcl-palettes-principles}, the luminance is probably the most important
+As emphasized in Figure\ \ref{fig:hcl-palettes-principles}, luminance is probably the most important
 property for defining the type of palette. It is constant for qualitative palettes, monotonic
@@
 
-The trajectories for the hue are also rather intuitive and straightforward for the three
-different types of palettes (constant vs.\ linear). However, the chroma trajectories are probably most complicated
+Hue trajectories are also rather intuitive and straightforward for the three
+different types of palettes (constant vs.\ linear). However, chroma trajectories are probably the most complicated
 and least obvious from the examples above. Hence, the exact mathematical equations underlying
@@
 The three types of trajectories are also depicted in Figure\ \ref{fig:hcl-trajectories}. Note that full intensity $i = 1$ is on
-the left and zero intensity $i = 0$ on the right of each panel. The concrete parameters are:
+the left and zero intensity $i = 0$ is on the right of each panel. The concrete parameters are:
 
@@
 
-And can be used subsequently in `qualitative_hcl()` as well as the
+The palette can be used subsequently in `qualitative_hcl()` as well as the
 qualitative \pkg{ggplot2} color scales (see Section\ \ref{sec:ggplot2}), e.g.,
@@
 * When registering a new palette with a previously-used name, the old palette
-  gets overwritten. We recommend not to overwrite
-  the palettes that are predefined in the package (albeit it is technically possible).
+  gets overwritten. We recommend to not overwrite
+  the palettes that are predefined in the package (albeit technically possible).
 * The registration of a palette is only stored for the current session. When
@@
 
-* \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
+* \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by the \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
   See `demo("brewer", package = "colorspace")`.
-* \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
+* \pkg{CARTO} colors [@color:CARTO] as provided by the \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
   See `demo("carto", package = "colorspace")`.
 * The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
-  as provided by \proglang{R} package \pkg{viridis} [@color:viridis].
+  as provided by the \proglang{R} package \pkg{viridis} [@color:viridis].
   See `demo("viridis", package = "colorspace")`.
-* The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].
+* The scientific color maps of @color:Crameri:2018 as provided by the \proglang{R} package \pkg{scico} [@color:scico].
   See `demo("scico", package = "colorspace")`.
@@
 * _Single-hue (advanced):_ The hue is fixed (as above) but the chroma trajectory is
-  triangular. Compared to the basic single-hue palette above this allows to better
-  distinguish the colors in the middle and not only the extremes.
+  triangular. Compared to the basic single-hue palette above this better
+  distinguishes the colors in the middle and not only the extremes.
 * _Multi-hue (advanced):_ As in the advanced single-hue palette the chroma trajectory
@@
 the HCL space. For example, it is not so clear whether high chroma values are
-close to the maximum possible for a given hue. Thus, it cannot be judged so easily
+close to the maximum possible for a given hue. Thus, it cannot be easily judged
 how the parameters of the hue, chroma, and luminance trajectories can be
@@
 with a palette. It does so by collapsing over one of the coordinates (either the hue $H$
-or the luminance $L$) and displays a heatmap of colors combining the remaining
+or the luminance $L$) and displaying a heatmap of colors combining the remaining
 two dimensions. The coordinates for the given color palette are highlighted
@@
 palettes), collapsing is straightforward. However, when the coordinate that is
-collapsed over is actually not constant in the palette, a simple bivariate linear
+collapsed over is not actually constant in the palette, a simple bivariate linear
 model is used to capture how the collapsed coordinate varies along with the
@@
 
-```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes (similar to \\texttt{\"Blues 2\"} and \\texttt{\"Blues 3\"}).
+```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma planes with variations of blue sequential single-hue palettes (similar to \\texttt{\"Blues 2\"} and \\texttt{\"Blues 3\"}).
 Left: Linear chroma for $H = 260$.
@@
 
-```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with blue multi-hue palette and triangular chroma (left), blue-yellow multi-hue palette and linear chroma (center), and diverging blue-red palette with balanced linear chroma."}
+```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma planes with blue multi-hue palette and triangular chroma (left), blue-yellow multi-hue palette and linear chroma (center), and diverging blue-red palette with balanced linear chroma."}
 par(mfrow = c(1, 3))
@@
 demos with the same sequential heat colors palette: `sequential_hcl(5, "Heat")`.
-All types of demos can, in principle, deal with arbitrarily many colors from any palette
-but clearly the graphics differ in various respects such as:
-
-- Work best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs.\ many colors (e.g., heatmap, perspective, ...).
+All types of demos can, in principle, deal with arbitrarily many colors from any palette,
+but the graphics differ in various respects such as:
+
+- Working best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs.\ many colors (e.g., heatmap, perspective, ...).
 - Intended for categorical data (e.g., bar, pie, ...) vs.\ continuous numeric data (e.g., heatmap, perspective, ...).
@@
 
-Hence, in the following some further illustrations are organized by type of
+Hence, in the following Figures\ \ref{fig:demoplot-qualitative}--\ref{fig:demoplot-diverging}
+some further illustrations are organized by type of
 palette, using suitable demos for the particular palettes.
@@
 techniques and show them in practice for a heatmap with sequential palette.
-Another example with diverging palette is available at
+Another example with a diverging palette is available at
 <http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
@@
 This clearly shows how poorly the RGB rainbow performs, often giving quite
-misleading impressions of what the terrain around Maunga Whau looks like.
+misleading impressions of the terrain around Maunga Whau.
 In contrast, the HCL-based blue-yellow palette works reasonably well in all settings. The most important
@@
 
-The \pkg{shiny} version of this GUI is shown in Figure\ \ref{fig:choose_palette}.
+The \pkg{shiny} version of the palette constructor GUI is shown in Figure\ \ref{fig:choose_palette}.
 It interfaces the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
@@
 from a list of predefined colors along with an example visualization. However, unlike
-\pkg{ColorBrewer.org} our \pkg{shiny} app allows to tweak the HCL parameters underlying
+\pkg{ColorBrewer.org} our \pkg{shiny} app allows tweaking the HCL parameters underlying
 each palette. This makes the app much more flexible but also more complex, potentially requiring more
 thought and experience. Due to the flexibility, our app cannot automatically judge safety regarding
-color vision deficiencies and printers/photocopiers (as \pkg{ColorBrewer.org} does) but instead it allows to emulate color vision
+color vision deficiencies and printers/photocopiers (as \pkg{ColorBrewer.org} does) but instead it allows emulation of color vision
 deficiencies and desaturation. Finally, \pkg{ColorBrewer.org} is geared towards cartography
@@
 In the paper we focus on the more mature \proglang{R} implementation but
-replication materials for most examples are not only provided for \proglang{R} but
-also for \proglang{Python}.
+replication materials for most examples are not only provided for both \proglang{R}
+and \proglang{Python}.
 

### 2019-08-28 pkg/colorspace/R/specplot.R

@@ -180,7 +180,7 @@ specplot <- function(x, y = NULL, rgb = FALSE, hcl = TRUE, fix = TRUE, cex = 1,
 
   # plot spectra and palette?
   if(isTRUE(rgb)) rgb <- hex(sRGB(c(0.8, 0, 0), c(0, 0.8, 0), c(0, 0, 0.8)))
-  if(isTRUE(hcl)) hcl <- rainbow_hcl(3L)
+  if(isTRUE(hcl)) hcl <- qualitative_hcl(3L)
   show_rgb <- !identical(rgb, FALSE)
   show_hcl <- !identical(hcl, FALSE)
 

### 2019-08-28 pkg/colorspace/paper/Toolbox/article.Rmd
rebuilt pkgdown site with modified specplot and version 1.4-2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@505 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r hcl-palettes, echo = TRUE, eval = FALSE}
+```{r hcl-palettes-0, echo = TRUE, eval = FALSE}
 library("colorspace")
@@
 
-```{r hcl-palettes-sequentialm, fig.width = 8.5, fig.height = 9.0, fig.cap = "Prespecified sequential multi-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+```{r hcl-palettes-sequentialm-0, echo = TRUE, eval = FALSE}
 hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
+```
+```{r hcl-palettes-sequentialm, echo = FALSE, fig.width = 8.5, fig.height = 9.0, fig.cap = "Prespecified sequential multi-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE, mar = c(0.1, 7, 0, 0), line = 6)
 ```

### 2019-08-29 pkg/colorspace/paper/Toolbox/article.Rmd
dashed -> gray
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@509 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2019-08-28 pkg/colorspace/paper/Toolbox/article.Rmd
@@
 library("colorspace")
-hcl_palettes(plot = TRUE, mar = c(0.1, 7, 0, 0), line = 6)
+hcl_palettes(plot = TRUE, line = 6)
 ```
@@
 
-```{r hcl-palettes-sequentialm, fig.width = 8.5, fig.height = 9.0, fig.cap = "Prespecified sequential multi-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+```{r hcl-palettes-sequentialm-0, echo = TRUE, eval = FALSE}
 hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE)
+```
+```{r hcl-palettes-sequentialm, echo = FALSE, fig.width = 8.5, fig.height = 9.0, fig.cap = "Prespecified sequential multi-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
+hcl_palettes("sequential (multi-hue)", n = 7, plot = TRUE, line = 6)
 ```
@@
 mylab(0.94, c2 - 5, "c2")
-mylab(0.68, 40, "p1", col = gray(0.5))
-text(0.78, 40, "= 1", col = gray(0.5))
+mylab(0.68, 40, "p1")
+text(0.78, 40, "= 1")
 mylab(0.18, 40, "p1", col = gray(0.5))
@@
 mylab(0.94, c2 - 5, "c2")
-mylab(0.77, 40, "p1", col = gray(0.5))
-text(0.87, 40, "= 1", col = gray(0.5))
+mylab(0.77, 40, "p1")
+text(0.87, 40, "= 1")
 mylab(0.27, 40, "p1", col = gray(0.5))
@@
 * Constant: `c1 = 80`.
-* Linear: `c1 = 80`, `c2 = 10`, `p1 = 1` (solid) vs.\ `p1 = 1.6` (dashed).
-* Triangular: `c1 = 60`, `cmax = 80`, `c2 = 10`, `p1 = 1` (solid) vs.\ `p1 = 1.6` (dashed).
+* Linear: `c1 = 80`, `c2 = 10`, `p1 = 1` (black) vs.\ `p1 = 1.6` (gray).
+* Triangular: `c1 = 60`, `cmax = 80`, `c2 = 10`, `p1 = 1` (black) vs.\ `p1 = 1.6` (gray).
 

### 2019-10-04 pkg/colorspace/index.Rmd
feedback from 2nd round of reviews in JSS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@510 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-```{r hcl, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
+```{r, hcl, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
 library("colorspace")

### 2019-10-04 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from 2nd round of reviews in JSS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@510 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Color is an integral and omnipresent element of many statistical graphics and data
+Color is an integral element of many statistical graphics and data
 visualizations. Therefore, colors should be carefully chosen to support all viewers in
@@
 Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
-Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
+Some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
 before Section\ \ref{sec:summary} concludes the paper.
@@
 To inspect the HCL parameter combinations for a specific palette simply include the `palette` name
-where upper- vs.\ lower-case, spaces, etc.\ are ignored for matching the label, i.e.,
+where upper- vs.\ lower-case, spaces, etc.\ are ignored for matching the label, e.g.,
 `"set2"` matches `"Set 2"`:
@@
 with a palette. This can bring out clearly various aspects, e.g.,
-whether hue is constant, chroma is monotonic or triangular, and whether
+whether hue is constant, whether chroma is monotonic or triangular, and whether
 luminance is approximately constant (as in many qualitative palettes),
@@
 - Intended for categorical data (e.g., bar, pie, ...) vs.\ continuous numeric data (e.g., heatmap, perspective, ...).
-- Shading areas (e.g., map, bar, pie, ...) vs.\ shading points or lines (scatter, lines).
+- Shading areas (e.g., map, bar, pie, ...) vs.\ coloring points or lines (scatter, lines).
 
@@
 may be helpful in assessing their properties.
-\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions,
+\pkg{roloc} [@color:roloc2; @color:roloc] also provides color conversions,
 not between numeric color spaces, but rather from numeric color spaces to English color names.
@@
 In the paper we focus on the more mature \proglang{R} implementation but
-replication materials for most examples are not only provided for both \proglang{R}
-and \proglang{Python}.
+replication materials for most examples are also available for
+\proglang{Python}.
 
@@
 <https://CRAN.R-project.org/>.
+
+# Acknowledgments {#acknowledgments .unnumbered}
+
+The authors would like to thank the journal editors, the associate editor,
+and two anonymous reviewers for their constructive and helpful feedback
+that substantially improved the paper.

### 2019-11-14 pkg/colorspace/R/colorspace.R
point out that [-100,100] are not actually hard limits for A/B and U/V in LAB and LUV
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@511 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -326,9 +326,10 @@ XYZ <-
 #' The \code{L}, \code{A} and \code{B} values give the coordinates of the
 #' colors in the CIE \eqn{L^*a^*b^*}{L*a*b*} space.  This is a transformation
 #' of the 1931 CIE XYZ space which attempts to produce perceptually based axes.
-#' Luminance takes values between 0 and 100, and the other coordinates take
-#' values between -100 and 100.  The \eqn{a} and \eqn{b} coordinates measure
-#' positions on green/red and blue/yellow axes.
+#' Luminance takes values between 0 and 100, and the other coordinates 
+#' typically take values between -100 and 100, although these values can also
+#' be exceeded by highly saturated colors. The \eqn{a} and \eqn{b} coordinates
+#' measure positions on green/red and blue/yellow axes.
 #' 
 #' @param L,A,B these arguments give the L, A and B coordinates of the colors.
 #' The values can be provided in separate \code{L}, \code{A} and \code{B}
@@ -508,8 +509,9 @@ HLS <-
 #' The \code{L}, \code{U} and \code{V} values give the coordinates of the
 #' colors in the CIE (1976) \eqn{L^*u^*v^*}{L*u*v*} space.  This is a
 #' transformation of the 1931 CIE XYZ space which attempts to produce
-#' perceptually based axes.  Luminance takes values between 0 and 100, and the
-#' other coordinates take values between -100 and 100.  The \eqn{a} and \eqn{b}
+#' perceptually based axes.  Luminance takes values between 0 and 100, and the other coordinates 
+#' typically take values between -100 and 100, although these values can also
+#' be exceeded by highly saturated colors. The \eqn{u} and \eqn{v}
 #' coordinates measure positions on green/red and blue/yellow axes.
 #' 
 #' @param L,U,V these arguments give the L, U and V coordinates of the colors.

### 2019-11-14 pkg/colorspace/paper/Toolbox/article.Rmd
improve description of LUV/LAB and polarLUV/polarLAB point out that sometimes polarLAB is also called HCL recompile pkgdown web page
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@513 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   (such as screens and monitors) whereas the latter is usually preferred when
-  working with dyes and pigments. However, the three axes of these two spaces
-  still do not correspond to human perceptual axes.
+  working with dyes and pigments. The L coordinate in both spaces has the same
+  meaning and captures luminace (light-dark contrasts). Both the U and V coordinates
+  as well as the A and B coordinates measure positions on red/green and yellow/blue
+  axes, respectively, albeit in somewhat different ways. While this corresponds
+  to how human color vision likely evolved (see the next section), these two color
+  models still not correspond to perceptual axes that humans use to describe colors.
   See @color:Wiki+CIELUV; @color:Wiki+CIELAB for more details.
-* `polarLUV()` and `polarLAB()` take polar coordinates in the UV plane
+* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
   and AB plane, respectively. Specifically, the polar coordinates of the LUV
-  model are known as the HCL (Hue-Chroma-Luminance) model [@color:Wiki+HCL]. These capture the
-  human perceptual axes very well without confounding effects as in the HSV or
-  HLS approaches. (More details follow below.)
+  model are known as the HCL (Hue-Chroma-Luminance) model [see @color:Wiki+HCL, 
+  which points out that the LAB-based polar coordinates are also sometimes referred to as HCL].
+  The HCL model captures the human perceptual axes very well without confounding
+  effects as in the HSV or HLS approaches. (More details follow below.)
 

### 2019-11-14 pkg/colorspace/R/colorspace.R
corrected typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@514 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -555,8 +555,8 @@ LUV <-
 #' This function creates colors of class ``polarLUV''; a subclass of the
 #' virtual \code{\link{color-class}} class.
 #' 
-#' The polarLUV space is a transformation of the CIE \eqn{L^*u^*u^*}{L*u*u*}
-#' space so that the \eqn{u} and \eqn{u} values are converted to polar
+#' The polarLUV space is a transformation of the CIE \eqn{L^*u^*v^*}{L*u*v*}
+#' space so that the \eqn{u} and \eqn{v} values are converted to polar
 #' coordinates.  The radial component \eqn{C} measures chroma and the angular
 #' coordinate \eqn{H} is measures hue. It is also known as the HCL
 #' (hue-chroma-luminance) space.

### 2019-12-02 pkg/colorspace/R/adjust_transparency.R
rename transparency() to adjust_transparency(), update example tests, and pkgdown web page
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@520 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,4 +1,4 @@
-#' Manage Transparency of Colors
+#' Adjust Transparency of Colors
 #' 
 #' Add, remove, or modify alpha transparency of a vector of colors.
 #'
@@ -8,7 +8,7 @@
 #' corresponds to fully opaque. If a color hex string in R does not provide an explicit
 #' alpha transparency, the color is assumed to be fully opaque.
 #'
-#' The \code{transparency} function can be used to manage the alpha transparency
+#' The \code{transparency} function can be used to adjust the alpha transparency
 #' of a set of colors. It always returns a hex color specification. This hex color
 #' can have the alpha transparency added/removed/modified depending on the
 #' specification of \code{alpha}: \itemize{
@@ -41,23 +41,23 @@
 #' @keywords color
 #' @examples
 #' ## modify transparency of a color (in different formats)
-#' transparency("black",   alpha = c(0, 0.5, 1)) ## name
-#' transparency("#000000", alpha = c(0, 0.5, 1)) ## hex string
-#' transparency(1,         alpha = c(0, 0.5, 1)) ## palette() integer
+#' adjust_transparency("black",   alpha = c(0, 0.5, 1)) ## name
+#' adjust_transparency("#000000", alpha = c(0, 0.5, 1)) ## hex string
+#' adjust_transparency(1,         alpha = c(0, 0.5, 1)) ## palette() integer
 #' 
 #' ## three shades of gray (in different formats:
 #' ## name/opaque, hex/opaque, hex/semi-transparent)
 #' x <- c("gray", "#BEBEBE", "#BEBEBE80")
 #' 
-#' ## manage transparency
-#' transparency(x, alpha = NULL)  ## only if necessary
-#' transparency(x, alpha = TRUE)  ## add
-#' transparency(x, alpha = FALSE) ## remove
-#' transparency(x, alpha = 0.8)   ## modify
-#' @export transparency
+#' ## adjust transparency
+#' adjust_transparency(x, alpha = NULL)  ## only if necessary
+#' adjust_transparency(x, alpha = TRUE)  ## add
+#' adjust_transparency(x, alpha = FALSE) ## remove
+#' adjust_transparency(x, alpha = 0.8)   ## modify
+#' @export adjust_transparency
 #' @importFrom grDevices rgb col2rgb
 
-transparency <- function(col, alpha = TRUE) {
+adjust_transparency <- function(col, alpha = TRUE) {
[…]

### 2019-12-02 pkg/colorspace/paper/Toolbox/article.Rmd
small fixup
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@521 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
-specifications in many systems based on the corresponding hex codes
+specifications in many systems based on the corresponding hexadecimal (or simply hex) codes
 [@color:Wiki+Webcolors], e.g., in HTML but also in \proglang{R}.
@@
 * `darken()` and `lighten()`: Algorithmically lighten or darken colors in HCL and/or HLS space.
+* `adjust_transparency()`: Add, remove, or modify alpha transparency of colors.
 * `max_chroma()`: Compute maximum chroma for given hue and luminance in HCL space.
@@
 )
+```
+
+## Adjust transparency of colors
+
+Alpha transparency is useful for making colors semi-transparent, e.g., for
+overlaying different elements in graphics. An alpha value of 0 (or 00 in hex strings)
+corresponds to fully transparent and an alpha value of 1 (or FF in hex strings)
+corresponds to fully opaque. If a color hex string in \proglang{R} does not provide an explicit
+alpha transparency, the color is assumed to be fully opaque.
+
+color:Wiki+Webcolors
+
+The \code{transparency} function can be used to adjust the alpha transparency
+of a set of colors. It always returns a hex color specification. This hex color
+can have the alpha transparency added/removed/modified depending on the
+specification of \code{alpha}: \itemize{
+\item \code{alpha = NULL}: Returns a hex vector with alpha transparency only if needed.
+      Thus, it keeps the alpha transparency for the colors (if any) but only if
+      different from opaque.
+\item \code{alpha = TRUE}: Returns a hex vector with alpha transparency
+      for all colors, using opaque (FF) as the default if missing.
+\item \code{alpha = FALSE}: Returns a hex vector without alpha transparency for
+      all colors (even if the original colors had non-opaque alpha).
+\item \code{alpha} numeric: Returns a hex vector with alpha transparency for
+      all colors set to the \code{alpha} argument (recycled if necessary).
+}
+
+A character vector with hexadecimal color strings with alpha transparency
+corresponding to \code{alpha} argument.
+
+## modify transparency of a color specified by name (`"black"`), hex string
+(`"#000000"`), or integer position in the palette (`1`).
+
+```{r transparency-black}
+adjust_transparency("black",   alpha = c(0, 0.5, 1))
+adjust_transparency("#000000", alpha = c(0, 0.5, 1))
+adjust_transparency(1,         alpha = c(0, 0.5, 1))
+```
+
+three shades of gray (in different formats:
+name/opaque, hex/opaque, hex/semi-transparent)
+
+adjust transparency: only if necessary (`alpha = NULL`), add (`alpha = TRUE`),
+remove (`alpha = FALSE`), or modify (`alpha = 0.8`)
+
+```{r transparency-black}
+x <- c("gray", "#BEBEBE", "#BEBEBE80")
+adjust_transparency(x, alpha = NULL)
+adjust_transparency(x, alpha = TRUE)
+adjust_transparency(x, alpha = FALSE)
+adjust_transparency(x, alpha = 0.8)
 ```

### 2019-12-02 pkg/colorspace/R/adjust_transparency.R

@@ -8,7 +8,7 @@
 #' corresponds to fully opaque. If a color hex string in R does not provide an explicit
 #' alpha transparency, the color is assumed to be fully opaque.
 #'
-#' The \code{transparency} function can be used to adjust the alpha transparency
+#' The \code{adjust_transparency} function can be used to adjust the alpha transparency
 #' of a set of colors. It always returns a hex color specification. This hex color
 #' can have the alpha transparency added/removed/modified depending on the
 #' specification of \code{alpha}: \itemize{

### 2019-12-02 pkg/colorspace/paper/Toolbox/article.Rmd
mention new adjust_transparency() in toolbox article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@522 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2019-11-14 pkg/colorspace/paper/Toolbox/article.Rmd
@@
 )
+```
+
+## Adjust transparency of colors
+
+Alpha transparency is useful for making colors semi-transparent, e.g., for
+overlaying different elements in graphics [@color:Wiki+Webcolors]. An alpha value
+(or alpha channel) of 0 (or 00 in hex strings) corresponds to fully transparent and
+an alpha value of 1 (or FF in hex strings) corresponds to fully opaque. If a color
+hex string in \proglang{R} does not provide an explicit alpha transparency, the
+color is assumed to be fully opaque.
+
+The `adjust_transparency()` function can be used to adjust the alpha transparency
+of a set of colors. It always returns a hex color specification. This hex color
+can have the alpha transparency added/removed/modified depending on the
+specification of the argument `alpha`:
+
+* `alpha = NULL`: Returns a hex vector with alpha transparency only if needed.
+  Thus, it keeps the alpha transparency for the colors (if any) but only if
+  different from opaque.
+* `alpha = TRUE`: Returns a hex vector with alpha transparency
+  for all colors, using opaque (FF) as the default if missing.
+* `alpha = FALSE`: Returns a hex vector without alpha transparency for
+  all colors (even if the original colors had non-opaque alpha).
+* `alpha` numeric: Returns a hex vector with alpha transparency for
+  all colors set to the `alpha` argument (recycled if necessary).
+
+For illustration, the transparency of a single black color is modified to three
+alpha levels: fully transparent, semi-transparent, and fully opaque, respectively.
+Black can be equivalently specified by name (`"black"`), hex string (`"#000000"`),
+or integer position in the palette (`1`).
+
+```{r transparency-black}
+adjust_transparency("black",   alpha = c(0, 0.5, 1))
+adjust_transparency("#000000", alpha = c(0, 0.5, 1))
+adjust_transparency(1,         alpha = c(0, 0.5, 1))
+```
+
+Subsequently, different settings of `alpha` are illustrated for adjusting
+a vector with three shades of gray, specified by name (`gray`, opaque),
+opaque hex string (`"#BEBEBE"`), and semi-transparent hex string (`"#BEBEBE80"`).
+Four types of adjustment are shown: only if necessary (`alpha = NULL`),
+add (`alpha = TRUE`), remove (`alpha = FALSE`), or modify (`alpha = 0.8`).
+
+```{r transparency-gray}
+x <- c("gray", "#BEBEBE", "#BEBEBE80")
+adjust_transparency(x, alpha = NULL)
+adjust_transparency(x, alpha = TRUE)
+adjust_transparency(x, alpha = FALSE)
+adjust_transparency(x, alpha = 0.8)
 ```

### 2020-09-06 pkg/colorspace/R/colorspace.R
suppress output from writehex in checks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@537 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -806,7 +806,9 @@ readhex <-
 #' @examples
 #' set.seed(1)
 #' x <- RGB(runif(10), runif(10), runif(10))
+#' ## IGNORE_RDIFF_BEGIN
 #' writehex(x, file.path(tempdir(), "random.txt"))
+#' ## IGNORE_RDIFF_END
 #' @export writehex
 writehex <-
   function(x, file="")

### 2020-09-06 pkg/colorspace/R/divergingx_hcl.R
Added three more palettes from Fabio Crameri's scientific color maps (_scico_), namely `"Batlow"` in `sequential_hcl()` and `"Hawaii"` and `"Roma"` in `divergingx_hcl()`.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@540 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -16,8 +16,8 @@
 #'
 #' With this additional flexibility various diverging palettes suggested by
 #' \url{http://ColorBrewer.org/} and CARTO (\url{https://carto.com/carto-colors/}),
-#' can be emulated along with the Zissou 1 palette from \pkg{wesanderson} and
-#' Cividis from \pkg{viridis}.
+#' can be emulated along with the Zissou 1 palette from \pkg{wesanderson},
+#' Cividis from \pkg{viridis}, and Hawaii and Roma from \pkg{scico}.
 #'
 #' Available CARTO palettes: ArmyRose, Earth, Fall, Geyser, TealRose, Temps, and
 #' Tropic (with Tropic also available in \code{diverging_hcl}).
@@ -195,3 +195,7 @@ divex_pals[["Zissou 1"]] <- c(218,  71,  12,  46,  88, 165,  59,  82,  52, 0.2, 
 
 ## Cividis
 divex_pals[["Cividis"]]  <- c(255,  NA,  75,  30,   0,  95,  13,  52,  92, 1.1, 1.0, 1.0,  NA,  47, NA)
+
+## scico
+divex_pals[["Hawaii"]]   <- c(-30,  80, 200,  70,  65,  35,  30,  60,  92, 1.0, 0.7, 1.3, 0.9,  55, 75)
+divex_pals[["Roma"]]     <- c( 10, 120, 265,  80,  25,  80,  25,  92,  25, 0.4, 1.5, 1.0, 1.2,  NA, NA)

### 2020-09-06 pkg/colorspace/R/hcl_palettes.R

@@ -693,6 +693,7 @@ seqm.pals[["BuPu"]]        <- c(320,  200,  40,  5, 15, 98, 1.2, 1.3,  65, 1) # 
 seqm.pals[["Blues"]]       <- c(260,  220,  45,  5, 25, 98, 1.2, 1.3,  70, 1) # ColorBrewer.org: Blues
 seqm.pals[["Lajolla"]]     <- c( 90,  -20,  40,  5, 99,  5, 0.7, 0.8, 100, 1) # scico: lajolla
 seqm.pals[["Turku"]]       <- c( 10,  120,  20,  0, 95,  1, 1.7, 0.8,  55, 1) # scico: turku
+seqm.pals[["Batlow"]]      <- c(270,  -40,  35, 35, 12, 88, 0.6, 1.1,  75, 1) # scico: batlow
 
 dive.pals <- list()
 dive.pals[["Blue-Red"]]    <- c(260,    0,  80, NA, 30, 90, 1.5,  NA,  NA, 1) # Z+KH+PM-09, Fig.6: Blue-Red (high luminance contrast)
@@ -825,7 +826,7 @@ GetPaletteConfig <- function(gui = NULL) {
            # Sequential multiple-hues advanced
            "Purple-Yellow", "YlGnBu", "Greens", "BuGn", "Teal", "Peach", "Blues", "BuPu", "Purples",
            "Purp", "Burg", "Reds", "YlOrRd", "Sunset", "RdPu", "Inferno",
-           "Lajolla", "Turku",
+           "Lajolla", "Turku", "Batlow",
            # Base color maps (for shiny)
            "rainbow", "heat.colors", "topo.colors", "terrain.colors", "cm.colors", "bpy"
        ) # end of variable definition for "take"

### 2020-09-06 pkg/colorspace/demo/scico.R

@@ -2,10 +2,13 @@ if(requireNamespace("scico")) {
 
 library("colorspace")
 
-scico_seq <- c("Oslo", "Lajolla", "Turku")
-for(i in scico_seq) specplot(rev(scico::scico(9, palette = tolower(i))), sequential_hcl(9, i, rev = i == "Lajolla"), main = i)
+scico_seq <- c("Oslo", "Lajolla", "Turku", "Batlow")
+for(i in scico_seq) specplot(rev(scico::scico(9, palette = tolower(i))), sequential_hcl(9, i, rev = i %in% c("Lajolla", "Batlow")), main = i)
 
 scico_div <- c("Broc", "Cork", "Vik", "Berlin", "Lisbon", "Tofino")
 for(i in scico_div) specplot(scico::scico(11, palette = tolower(i)), diverging_hcl(11, i), main = i)
+
+scico_divx <- c("Hawaii", "Roma")
+for(i in scico_divx) specplot(scico::scico(11, palette = tolower(i)), divergingx_hcl(11, i), main = i)
 
 }

### 2020-09-17 pkg/colorspace/paper/Toolbox/article.Rmd
wording slightly improved
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@551 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   it is not a good place to start. See @color:Wiki+sRGB for more details.
-* `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
+* `HSV()` is a simple transformation of either the sRGB or the RGB space that tries to capture the
   perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
@@
   See @color:Wiki+HSV for more details.
-* `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
+* `HLS()` (Hue-Lightness-Saturation) is another transformation of either sRGB or RGB that tries
   to capture the perceptual axes. It does a somewhat better job but the dimensions

### 2020-09-17 pkg/colorspace/paper/Toolbox/article.Rmd
adapted spine plot example to R >= 4.0.0
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@553 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ```{r titanic, eval = FALSE}
-ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
+ttnc <- margin.table(Titanic, c(1, 4))
 spineplot(ttnc, col = sequential_hcl(2, palette = "Purples 3"))
@@
 legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
-ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
+ttnc <- margin.table(Titanic, c(1, 4))
 spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))

### 2020-09-22 pkg/colorspace/paper/ZAMG2020/slides.Rnw
Adding slides for ZAMG internal presentation, September 2020.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@555 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2019-06-25 pkg/colorspace/paper/Toolbox/slides.Rnw
@@
 %\usetheme[nototalframenumber, noslidenumber, license, url]{uibk}
-\usetheme[license, url]{uibk}
+\usetheme[nototalframenumber, license, url]{uibk}
 \renewcommand{\headerimage}[1]{%
@@
 
-\title{A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
-\subtitle{\phantom{.}\\Achim Zeileis, Jason C. Fisher, Kurt Hornik, Ross Ihaka, Claire D. McWhite, Paul Murrell, Reto Stauffer, Claus O. Wilke}
-\author{\vspace*{-6.83cm}\hspace*{12.5cm}\includegraphics[width=1.6cm]{palmfooter.png}\\[5.5cm]}
-\URL{http://colorspace.R-Forge.R-project.org/}
+\title{Robust Color Maps That Work for Most Audiences (Including the U.S. President)}
+\subtitle{Reto Stauffer, Achim Zeileis}
+\author{\vspace*{-6.1cm}\hspace*{12.5cm}\includegraphics[width=1.6cm]{palmfooter.png}\\[5.5cm]\vspace{-1.5em}} %% -6.83 with "for Statistical Graphics"
+\URL{http://hclwizard.org/}%
 
@@
 library("colorspace")
-library("ggplot2")
-theme_set(theme_minimal())
 @
@@
 
-\section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics}
+\section{colorspace: A Toolbox for Manipulating and Assessing Color Palettes} %
 
-\subsection{Motivation}
+\subsection{Motivation}%
 
+% ---------------------------------------------------------------------
 \begin{frame}
@@
 
-\vspace*{-0.3cm}
+\vspace*{2em}
 
-\begin{minipage}{0.54\textwidth}
-\only<1,4>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow.png}}%
-\only<2>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow-deutan.png}}%
-\only<3>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-rainbow-gray.png}}%
-\only<5>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow.png}}%
-\only<6>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow-deutan.png}}%
-\only<7>{\includegraphics[width=\textwidth]{../../vignettes/screenshots/influenza-purpleyellow-gray.png}}
+\begin{minipage}{0.46\textwidth}
+    \includegraphics[width=\textwidth]{images/sharpie-small.png}
 \end{minipage}
-\hspace*{0.6cm}
-\begin{minipage}{0.4\textwidth}
+\pause
+\hfill
+\begin{minipage}{0.51\textwidth}
+    \includegraphics[width=\textwidth]{images/tweet.png}
+\end{minipage}
 
-\textbf{Map:} Influenza severity in Germany (week 8, 2019).\\
-
-\textbf{Source:} Arbeitsgemeinschaft Influenza, Robert-Koch-Institut.\\
-
-\textbf{Reported in:} SPIEGEL Online, Tagesschau, \dots\\
-
-\visible<2->{\textbf{Problems:} Color vision deficiency.}
-\visible<3->{Grayscale.}
-\visible<4->{Flashy.\\}
-
-\visible<5->{\textbf{Alternative:} HCL-based sequential palette.}
-
-\end{minipage}
+    \vspace*{0.5em}
+    \begin{footnotesize}
+    Projected track and wind speed forecast of hurricane Dorian.
+    Screenshot of a video released by the White House (Sep. 4, 2019),
+    tweet by the U.S. president (Sep. 5, 2019).
+    \end{footnotesize}
 
@@
 
+%% ---------------------------------------------------------------------
+%\begin{frame}
+%\frametitle{Motivation}
+%
+%\vspace*{-0.3cm}
+%
+%\begin{minipage}{0.54\textwidth}
+%    \only<1>{\includegraphics[width=\textwidth]{images/dorian-rainbow.png}}%
+%    \only<2-3>{\includegraphics[width=\textwidth]{images/dorian-rainbow-arrow.png}}%
+%    \only<4>{\includegraphics[width=\textwidth]{images/dorian-rainbow-deutan.png}}%
+%    \only<5>{\includegraphics[width=\textwidth]{images/dorian-rainbow-gray.png}}%
+%    \only<6>{\includegraphics[width=\textwidth]{images/dorian-hclrainbow.png}}%
+%    \only<7>{\includegraphics[width=\textwidth]{images/dorian-hclrainbow-gray.png}}
+%    \only<8>{\includegraphics[width=\textwidth]{images/dorian-hclrainbow-deutan.png}}
+%\end{minipage}
+%\hspace*{0.6cm}
+%\begin{minipage}{0.4\textwidth}
+%
+%    \textbf{Map:} Probability of exceeding wind speeds of 63\,km\,h$^{-1}$,
+%    Aug~30 -- Sep~04, 2019.
+%
+%    \medskip
+%    \textbf{Source:} NOAA (\href{https://noaa.gov}{noaa.gov})
+%
+%    \medskip
+%    \visible<3->{\textbf{Problems:} Flashy.}
+%    \visible<4->{Color vision deficiency.}
+%    \visible<5->{Grayscale.}
+%    
+%    \medskip
+%    \visible<6->{\textbf{Alternative:} HCL-based sequential palette.}
+%
+%\end{minipage}
+%
+%\end{frame}
+
+% ---------------------------------------------------------------------
 \begin{frame}
-\frametitle{Motivation}
+    \frametitle{\only<1-4>{Problem}\only<5->{Solution}}
 
-\textbf{Need tools for:}
-\begin{itemize}
-  \item Construction of palettes with better perceptual properties.
-  \item Assessment of color palettes.
-  \item Manipulation of colors.
-\end{itemize}
+\vspace*{-0.3cm}
 
-\pause
-\bigskip
+    \begin{center}
+        \only<1>{\includegraphics[width=.55\textwidth]{images/dorian-rainbow.png}}%
+        \only<2>{\includegraphics[width=.55\textwidth]{images/dorian-rainbow-arrow.png}}%
+        \only<3>{\includegraphics[width=.55\textwidth]{images/dorian-rainbow-deutan.png}}%
+        \only<4>{\includegraphics[width=.55\textwidth]{images/dorian-rainbow-gray.png}}%
+        \only<5>{\includegraphics[width=.55\textwidth]{images/dorian-hclrainbow.png}}%
+        \only<6>{\includegraphics[width=.55\textwidth]{images/dorian-hclrainbow-gray.png}}
+        \only<7>{\includegraphics[width=.55\textwidth]{images/dorian-hclrainbow-deutan.png}}
+    \end{center}
 
-\begin{minipage}{0.5\textwidth}
-\includegraphics[width=\textwidth]{../../vignettes/screenshots/bobross.jpg}
-\end{minipage}
-\hspace*{0.6cm}
-\begin{minipage}{0.36\textwidth}
-\vspace*{-4cm}
-Because Bob Ross would not approve of this!
-\end{minipage}
+    \vspace*{-0.5em}
+    \begin{footnotesize} 
+    Probability of exceeding wind speeds of 39\,mph (63\,km\,h$^{-1}$),
+    Aug~30 -- Sep~04, 2019 (\href{https://noaa.gov}{noaa.gov}).
+    \end{footnotesize}
 
@@
 
+%% ---------------------------------------------------------------------
+%\begin{frame}
+%\frametitle{Motivation}
+%
+%\textbf{Need tools for:}
+%\begin{itemize}
+%  \item Construction of palettes with better perceptual properties.
+%  \item Assessment of color palettes.
+%  \item Manipulation of colors.
+%\end{itemize}
+%
+%\end{frame}
+
 \subsection{Color spaces}
 
-\begin{frame}
-\frametitle{Color spaces}
+%% ---------------------------------------------------------------------
+%\begin{frame}
+%\frametitle{Color spaces}
+%
+%\textbf{Origin of the package:} Convert colors between various three-dimensional representations of color.
+%
+%\medskip
+%
+%\textbf{In particular:} From the perceptually-based HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB, and corresponding hex codes) space.
+%
+%\setkeys{Gin}{width=0.8\textwidth}
+%\begin{center}
+%<<color-spaces, fig = TRUE, width = 9.5, height = 4.5>>=
+%par(mar = rep(0, 4))
+%
+%d <- data.frame(
+%  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
+%  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
+%  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
+%  stringsAsFactors = FALSE
+%)
+%
+%plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+%  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
+%lines(y ~ x, data = d[-c(1, 3, 7, 8),])
+%lines(y ~ x, data = d[c(1, 3, 5),])
+%lines(y ~ x, data = d[c(6, 7, 9),])
+%lines(y ~ x, data = d[c(6, 8, 9),])
+%points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, cex = 12,
+%       bg = c("lightgray", "#F5CF8B", "lightgray", "lightgray"))
+%points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, cex = 12,
+%       bg = ifelse(1:5 == 5, "#F5CF8B", "lightgray"))
+%points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)
+%
+%text(d$x, d$y, d$lab)
+%text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
+%text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
+%@
+%\end{center}
+%
+%\end{frame}
 
-\textbf{Origin of the package:} Convert colors between various three-dimensional representations of color.
-
-\medskip
-
-\textbf{In particular:} From the perceptually-based HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB, and corresponding hex codes) space.
-
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{center}
-<<color-spaces, fig = TRUE, width = 9.5, height = 4.5>>=
-par(mar = rep(0, 4))
-
-d <- data.frame(
-  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
-  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
-  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
-  stringsAsFactors = FALSE
-)
-
-plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
-  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
-lines(y ~ x, data = d[-c(1, 3, 7, 8),])
-lines(y ~ x, data = d[c(1, 3, 5),])
-lines(y ~ x, data = d[c(6, 7, 9),])
-lines(y ~ x, data = d[c(6, 8, 9),])
-points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
-points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
-points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)
-
-text(d$x, d$y, d$lab)
-text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
-text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
-@
-\end{center}
-
-\end{frame}
+%% ---------------------------------------------------------------------
+%\begin{frame}
+%\frametitle{Color spaces}
+%
+%\textbf{Origin of the package:} Convert colors between various three-dimensional representations of color.
+%
+%\medskip
+%
+%\textbf{In particular:} From the perceptually-based HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB, and corresponding hex codes) space.
+%
+%\setkeys{Gin}{width=0.8\textwidth}
+%\begin{center}
+%<<color-spaces, fig = TRUE, width = 9.5, height = 4.5>>=
+%par(mar = rep(0, 4))
+%
+%d <- data.frame(
+%  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
+%  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
+%  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
+%  stringsAsFactors = FALSE
+%)
+%
+%plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+%  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
+%lines(y ~ x, data = d[-c(1, 3, 7, 8),])
+%lines(y ~ x, data = d[c(1, 3, 5),])
+%lines(y ~ x, data = d[c(6, 7, 9),])
+%lines(y ~ x, data = d[c(6, 8, 9),])
+%points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
+%points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
+%points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)
+%
+%text(d$x, d$y, d$lab)
+%text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
+%text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
+%@
+%\end{center}
+%
+%\end{frame}
 
@@
 
+% ---------------------------------------------------------------------
 \begin{frame}
@@
 
+% ---------------------------------------------------------------------
+\begin{frame}
+\frametitle{HCL vs.~RGB}
+
+\setkeys{Gin}{width=1.0\textwidth}
+
+<<specplot-init, include = FALSE>>=
+orig <- c("#008a02", "#00cb00", "#82fd00", "#f5f900", "#f7d600",
+          "#d28400", "#fb7502", "#a80003", "#740001", "#8c0084")
+hcl  <- c("#b0ceb9", "#9fca91", "#a5bc62", "#b3a924", "#c09000",
+          "#c87400", "#c85435", "#b9375c", "#a4176e", "#890075")
+@
+
+    \begin{minipage}{0.48\textwidth}
+<<specplot-orig, echo = FALSE, fig = TRUE, fig.width = 5, fig.height = 4>>=
+specplot(orig, rgb = FALSE, main = "Problem")
+@
+    \end{minipage}
+    \hfill
+    \begin{minipage}{0.48\textwidth}
+<<specplot-hcl, echo = FALSE, fig = TRUE, fig.width = 5, fig.height = 4>>=
+specplot(hcl, rgb = FALSE, main = "Solution")
+@
+    \end{minipage}
+
+\end{frame}
+
 \begin{frame}
@@
 \setkeys{Gin}{width=0.44\textwidth}
-<<hcl-rainbow, fig = TRUE, width = 5.2, height = 6>>=
-specplot(rainbow_hcl(99, c = 60, l = 75), rgb = TRUE)
+<<rgb-rainbow, fig = TRUE, width = 5.2, height = 6>>=
+specplot(rainbow(95), rgb = TRUE)
 @
@@
 \pause
-<<rgb-rainbow, fig = TRUE, width = 5.2, height = 6>>=
-specplot(rainbow(95), rgb = TRUE)
+<<hcl-rainbow, fig = TRUE, width = 5.2, height = 6>>=
+specplot(rainbow_hcl(99, c = 60, l = 75), rgb = TRUE)
 @
@@
 
+\pause
 \bigskip
-\pause
 
@@
 
+\pause
 \bigskip
-\pause
 
 \textbf{Diverging:} For ordered/numeric information around a central neutral value, i.e.,
-where colors diverge from neutral to two extremes. Function: \code{diverging\_hcl()}.
+where colors diverge from neutral to two extremes.
+Function: \code{diverging\_hcl()}.
 
@@
 
-\only<1>{
-\textbf{Sequential:} Luminance contrast is crucial (dark to light or vice versa).
-}
-\only<2>{
+\begin{minipage}[t]{0.49\textwidth}
 \textbf{Blues 2:} Single hue. Decreasing chroma with increasing luminance.
-}
-\only<3>{
-\textbf{Blues 3:} Single hue. Triangular chroma to achieve higher luminance contrast.
-}
-\only<4>{
-\textbf{Blues:} Multi hue. Triangular chroma. High luminance contrast.
-}
 
-\vspace*{-0.4cm}
-
-\begin{minipage}[t]{0.49\textwidth}
-\vspace*{0.46cm}
-
-\setkeys{Gin}{width=\textwidth}
-<<blues, fig = TRUE, width = 4, height = 2.2>>=
-swatchplot(
-  "Blues 2" = sequential_hcl(7, "Blues 2"),
-  "Blues 3" = sequential_hcl(7, "Blues 3"),
-  "Blues"   = sequential_hcl(7, "Blues")
-)
-@
-\end{minipage}
-\only<2>{%
-\begin{minipage}[t]{0.49\textwidth}
 \vspace*{0cm}
@@
 \setkeys{Gin}{width=\textwidth}
-<<blues2-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
+<<blues2-spectrum, fig = TRUE, width = 5.2, height = 4.8>>=
 specplot(sequential_hcl(7, "Blues 2"))
@@
 \end{minipage}%
-}%
-\only<3>{%
+\hfill
 \begin{minipage}[t]{0.49\textwidth}
+\textbf{Blues 3:} Single hue. Triangular chroma to achieve higher luminance contrast.
+
 \vspace*{0cm}
@@
 \setkeys{Gin}{width=\textwidth}
-<<blues3-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
+<<blues3-spectrum, fig = TRUE, width = 5.2, height = 4.8>>=
 specplot(sequential_hcl(7, "Blues 3"))
@@
 \end{minipage}
-}%
-\only<4>{%
-\begin{minipage}[t]{0.49\textwidth}
-\vspace*{0cm}
-
-\setkeys{Gin}{width=\textwidth}
-<<blues-spectrum, fig = TRUE, width = 5.2, height = 5.2>>=
-specplot(sequential_hcl(7, "Blues"))
-@
-\end{minipage}
-}
 
@@
 hcl_palettes(plot = TRUE)
-@
-
-\end{frame}
-
-\subsection{Statistical graphics}
-
-\begin{frame}[fragile]
-\frametitle{Statistical graphics}
-
-\textbf{Base:}
-
-\begin{itemize}
-  \item HCL palette functions return hex color vector.
-  \item Typically passed to \code{col = } argument of base plotting functions.
-\end{itemize}
-
-\bigskip
-
-\textbf{ggplot2:}
-
-\begin{itemize}
-  \item Scales of type \code{scale_<aesthetic>_<datatype>_<colorscale>()}.
-  \item \code{<aesthetic>} is \code{fill} or \code{color}/\code{colour}.
-  \item \code{<datatype>} is \code{discrete} or \code{continuous}.
-  \item \code{<colorscale>} is \code{qualitative}, \code{sequential}, \code{diverging}, or \code{divergingx}.
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Statistical graphics: Base}
-
-<<base-graphics, eval = FALSE, echo = TRUE>>=
-q4 <- qualitative_hcl(4, palette = "Dark 3")
-plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
-legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
-@
-
-\vspace*{-0.6cm}
-
-\setkeys{Gin}{width=0.8\textwidth}
-<<base-graphics-plot, fig = TRUE, width = 10, height = 6>>=
-<<base-graphics>>
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Statistical graphics: Base}
-
-<<base-graphics2, eval = FALSE, echo = TRUE>>=
-ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
-spineplot(ttnc, col = sequential_hcl(2, palette = "Purples 3"))
-@
-
-\vspace*{-0.6cm}
-
-\setkeys{Gin}{width=0.8\textwidth}
-<<base-graphics2-plot, fig = TRUE, width = 10, height = 6>>=
-<<base-graphics2>>
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Statistical graphics: ggplot2}
-
-\setkeys{Gin}{width=0.8\textwidth}
-<<ggplot2-graphics, fig = TRUE, width = 10, height = 5.5, echo = TRUE>>=
-library("ggplot2")
-ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
-  scale_fill_discrete_qualitative(palette = "Dark 3")
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Statistical graphics: ggplot2}
-
-\setkeys{Gin}{width=0.8\textwidth}
-<<ggplot2-graphics2, fig = TRUE, width = 10, height = 5.5, echo = TRUE>>=
-dsamp <- diamonds[1 + 1:1000 * 50, ]
-ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
-  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
 @
@@
 
-\subsection{Color apps}
-
-\begin{frame}[fragile]
-\frametitle{Color apps}
-
-\textbf{Facilitate exploration:} Graphical user interfaces as shiny apps.
-
-\begin{itemize}
-  \item \emph{Palette constructor:} \code{choose\_palette()} or \code{hclwizard()} (also in tcltk).
-  \item \emph{Color picker:} \code{choose\_color()} or \code{hcl\_color\_picker()}.
-  \item \emph{Color vision deficiency emulator:} \code{cvd\_emulator()}.
-\end{itemize}
-
-\bigskip
-
-\textbf{Online versions:} \url{http://hclwizard.org/}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Color apps: \texttt{choose\_palette()} / \texttt{hclwizard()}}
-
-\includegraphics[width=0.9\textwidth]{../../vignettes/screenshots/choose_palette_1.png}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Color apps: \texttt{choose\_color()} / \texttt{hcl\_color\_picker()}}
-
-\includegraphics[width=0.65\textwidth]{../../vignettes/screenshots/choose_color.png}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Color apps: \texttt{cvd\_emulator()}}
-
-\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_1.png}
-\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_2.png}
-
-\end{frame}
-
 \subsection{Base R}
@@
 
+\subsection{Color apps}
+
+\begin{frame}[fragile]
+\frametitle{Color apps}
+
+\textbf{Facilitate exploration:} Graphical user interfaces as shiny apps.
+
+\begin{itemize}
+  \item \emph{Palette constructor:} \code{choose\_palette()} or \code{hclwizard()} (also in tcltk).
+  \item \emph{Color picker:} \code{choose\_color()} or \code{hcl\_color\_picker()}.
+  \item \emph{Color vision deficiency emulator:} \code{cvd\_emulator()}.
+\end{itemize}
+
+\bigskip
+
+\textbf{Online versions:} \url{http://hclwizard.org/}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Color apps: \texttt{choose\_palette()} / \texttt{hclwizard()}}
+
+\includegraphics[width=0.9\textwidth]{../../vignettes/screenshots/choose_palette_1.png}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Color apps: \texttt{choose\_color()} / \texttt{hcl\_color\_picker()}}
+
+\includegraphics[width=0.65\textwidth]{../../vignettes/screenshots/choose_color.png}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Color apps: \texttt{cvd\_emulator()}}
+
+\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_1.png}
+\includegraphics[width=0.34\textwidth]{../../vignettes/screenshots/cvd_emulator_2.png}
+
+\end{frame}
+
 \subsection{Recommendations}
@@
 
+\begin{frame}
+\frametitle{Do it yourself}
+
+    \textbf{Software:} \emph{colorspace}.
+    \begin{itemize}
+      \item R (mature): \url{http://colorspace.R-Forge.R-project.org/}.
+      \item Python (beta): \url{https://github.com/retostauffer/python-colorspace}.
+      \item Web (interactive): \url{http://www.hclwizard.org/}.
+    \end{itemize}
+
+    \bigskip
+
+    \textbf{Take-home messages:}
+    \begin{itemize}
+      \item Choose colors carefully.
+      \item Make areas of interest stand out from background.
+      \item Check robustness.
+      \item Software helps you.
+    \end{itemize}
+
+\end{frame}
+
 \subsection{References}
 
-\begin{frame}
+\begin{frame}\label{lastslide}
 \frametitle{References}
@@
 
+\begin{frame}
+	\frametitle{Thank you for your interest!}
+	\begin{center}
+		\includegraphics[width = .6\textwidth]{images/bobross.jpg}
+	\end{center}
+\end{frame}
+
 \end{document}

### 2020-11-05 pkg/colorspace/paper/Toolbox/article.Rmd
small touch-ups carried over from JSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@556 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+Moreover, the _grDevices_ package in base R gained a new function
+`palette.colors()` [@color:grDevices2] that provides various
+well-established qualitative color palettes that can not be approximated well by
+`qualitative_hcl()` due to pronounced variations in luminance and chroma.
+While a qualitative palette with fixed luminance and chroma is more balanced,
+a certain amount of variations in these properties might be necessary to make
+more colors distinguishable, especially for viewers with color vision deficiencies.
+
 # Palette visualization and assessment {#sec:palette_visualization}

### 2020-11-05 pkg/colorspace/R/adjust_transparency.R
include references to forthcoming JSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@557 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -35,9 +35,9 @@
 #' @return A character vector with hexadecimal color strings with alpha transparency
 #' corresponding to \code{alpha} argument.
 #' @seealso \code{\link[grDevices]{rgb}}, \code{\link[colorspace]{desaturate}}, \code{\link[colorspace]{lighten}}
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords color
 #' @examples
 #' ## modify transparency of a color (in different formats)

### 2020-11-05 pkg/colorspace/R/cvd_image.R

@@ -15,9 +15,9 @@
 #'    output when the GUI is started (development flag). 
 #' @author Reto Stauffer, Claus O. Wilke, Achim Zeileis
 #' @rdname cvd_emulator
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @export
 cvd_emulator <- function(file, overwrite = FALSE, shiny.trace = FALSE) {
 

### 2020-11-05 pkg/colorspace/R/demoplot.R

@@ -15,9 +15,9 @@
 #' @return \code{demoplot} returns invisibly what the respective base graphics
 #' functions return that are called internally.
 #' @seealso \code{\link{specplot}}, \code{\link{hclplot}}
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords hplot
 #' @examples
 #' ## all built-in demos with the same sequential heat color palette

### 2020-11-05 pkg/colorspace/R/divergingx_hcl.R

@@ -15,7 +15,7 @@
 #' between the two arms.
 #'
 #' With this additional flexibility various diverging palettes suggested by
-#' \url{http://ColorBrewer.org/} and CARTO (\url{https://carto.com/carto-colors/}),
+#' \url{https://ColorBrewer2.org/} and CARTO (\url{https://carto.com/carto-colors/}),
 #' can be emulated along with the Zissou 1 palette from \pkg{wesanderson},
 #' Cividis from \pkg{viridis}, and Roma from \pkg{scico}.
 #'
@@ -51,9 +51,9 @@
 #'
 #' @return A character vector with (s)RGB codings of the colors in the palette.
 #' @seealso \code{\link[colorspace]{sequential_hcl}}, \code{\link[colorspace]{diverging_hcl}}
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords color
 #' @examples
 #' ## show emulated CARTO/ColorBrewer.org palettes

### 2020-11-05 pkg/colorspace/R/hcl_color_picker.R

@@ -12,9 +12,9 @@
 #' @param shiny.trace logical: used for debugging the shiny interface.
 #' @author Claus O. Wilke, Reto Stauffer, Achim Zeileis
 #' @seealso \code{\link{choose_palette}}
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords misc
 #'
 #' @return \code{hclcolorpicker} invisibly returns a vector of colors choosen.

### 2020-11-05 pkg/colorspace/R/hcl_palettes.R

@@ -141,9 +141,9 @@
 #' \bold{96}(2), 203--216.
 #' \doi{10.1175/BAMS-D-13-00155.1}
 #'
-#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords color
 #' @examples
 #' ## overview of all _named_ HCL palettes

### 2020-11-05 pkg/colorspace/R/specplot.R

@@ -68,9 +68,9 @@
 #' \bold{96}(2), 203--216.
 #' \doi{10.1175/BAMS-D-13-00155.1}
 #'
-#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords hplot
 #' @examples
 #' ## spectrum of the (in)famous RGB rainbow palette (in both RGB and HCL)

### 2020-11-05 pkg/colorspace/R/swatchplot.R

@@ -32,9 +32,9 @@
 #' @param cex,font numeric vectors of length 1 or 2. Specifications for the
 #' annotation text for the individual palettes and lists of palettes, respectively.
 #' @return \code{swatchplot} invisibly returns a matrix with colors and annotations.
-#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2019).
+#' @references Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
 #' \dQuote{ccolorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
-#' arXiv:1903.06490, arXiv.org E-Print Archive. \url{http://arxiv.org/abs/1903.06490}
+#' \emph{Journal of Statistical Software}, \bold{96}(1), 1--49. \doi{10.18637/jss.v096.i01}
 #' @keywords hplot
 #' @examples
 #' ## swatches of several palette vectors

### 2020-11-07 pkg/colorspace/paper/Toolbox/article.Rmd
incorporate documentation of recently-added scico palettes (Hawaii, Batlow, Roma)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@558 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 * `"YlOrRd"` to `"Blues"` closely match \pkg{ColorBrewer.org} palettes [@color:Harrower+Brewer:2003].
-* `"Lajolla"` and `"Turku"` closely match the scientific color maps of the same name by
-  @color:Crameri:2018 and are intended for a black/dark background.
+* `"Lajolla"` to `"Batlow"` closely match the scientific color maps of the same name by
+  @color:Crameri:2018 and the first two of these are intended for a black/dark background.
 
@@
   (See Section\ \ref{sec:approximations} for more details.)
+* `"Roma"` closely matches the palette of the same name by @color:Crameri:2018.
 

### 2020-11-09 pkg/colorspace/paper/Toolbox/article.Rmd
document the recent additions: binned ggplot2 color scales, and cvd option in swatchplot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@570 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 To provide access to the HCL color palettes from within \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
-suitable discrete and/or continuous \pkg{ggplot2} color scales are provided.
+suitable discrete, continuous, and binned \pkg{ggplot2} color scales are provided.
 The scales are named via the scheme
@@
 - `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`).
-- `<datatype>` is the type of the variable plotted (`discrete` or `continuous`).
+- `<datatype>` is the type of the variable plotted (`discrete`, `continuous`, `binned`).
 - `<colorscale>` sets the type of the color scale used (`qualitative`, `sequential`, `diverging`, `divergingx`).
@@
 consider: `swatchplot("Palette" = sequential_hcl(5))`.
+Optionally, swatches emulating color vision deficiencies (see
+Section\ \ref{sec:color_vision_deficiency}) can be added by setting `cvd = TRUE`.
 

### 2020-11-09 pkg/colorspace/R/swatchplot.R

@@ -58,10 +58,10 @@
 #' )
 #'
 #' ## swatches with color vision deficiency emulation
-#' swatchplot(sequential_hcl(5, "Viridis"), cvd = TRUE)
+#' swatchplot(sequential_hcl(7, "Viridis"), cvd = TRUE)
 #' swatchplot(
-#'   "YlGnBu"  = sequential_hcl(5, "YlGnBu"),
-#'   "Viridis" = sequential_hcl(5, "Viridis"),
+#'   "YlGnBu"  = sequential_hcl(7, "YlGnBu"),
+#'   "Viridis" = sequential_hcl(7, "Viridis"),
 #'   cvd = c("deutan", "desaturate")
 #' )
 #' @export swatchplot

### 2021-04-08 pkg/colorspace/paper/Toolbox/article.Rmd
all URLs for hclwizard.org and colorspace.R-Forge.R-project.org now with https instead of http
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@588 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
     email: \email{Achim.Zeileis@R-project.org}
-    url: https://eeecon.uibk.ac.at/~zeileis/
+    url: https://www.zeileis.org/
   - name: Jason C. Fisher
@@
 The graphics resulting from the demos can also be viewed online at
-<http://colorspace.R-Forge.R-project.org/articles/approximations.html>.
+<https://colorspace.R-Forge.R-project.org/articles/approximations.html>.
 
@@
 Another example with a diverging palette is available at
-<http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
+<https://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
 Finally, CVD emulation is particularly useful for bringing out why the RGB
 rainbow palette is almost always a bad choice in scientific displays. See
-<http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
+<https://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
 illustrations.
@@
 \pkg{shiny} apps [@color:shiny]. All of these GUIs/apps can be run locally
-from within \proglang{R} and are also provided online at <http://hclwizard.org/>.
+from within \proglang{R} and are also provided online at <https://hclwizard.org/>.
 

### 2021-06-24 pkg/colorspace/R/contrast_ratio.R
change default of col2 in contrast_ratio() to 'white', add description to vignette, recompile pkgdown
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@598 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -15,13 +15,11 @@
 #' is defined as \code{ifelse(RGB <= 0.03928, RGB/12.92, ((RGB + 0.055)/1.055)^2.4)} based on
 #' the \code{RGB} coordinates between 0 and 1.
 #' 
-#' @param col vector of any of the three kind of R colors, i.e., either a color
+#' @param col,col2 vectors of any of the three kind of R colors, i.e., either a color
 #' name (an element of \code{\link[grDevices]{colors}}), a hexadecimal string
 #' of the form \code{"#rrggbb"} or \code{"#rrggbbaa"} (see
 #' \code{\link[grDevices]{rgb}}), or an integer \code{i} meaning
-#' \code{palette()[i]}.
-#' @param col2 vector of background colors (as for \code{col} above). If \code{NULL},
-#' all subsequent pairs from \code{col} are used.
+#' \code{palette()[i]}. Both can be vectors and are recycled as necessary.
 #' @param plot logical indicating whether the contrast ratios should also be
 #' visualized by simple color swatches. Can also be a vector of length 2, indicating
 #' whether the foreground color should be visualized on the background color and/or
@@ -56,19 +54,14 @@
 #' @importFrom graphics rect
 #' @importFrom graphics text
 
-contrast_ratio <- function(col, col2 = NULL,
+contrast_ratio <- function(col, col2 = "white",
   plot = FALSE, border = FALSE, cex = 2, off = 0.05, mar = rep(0.5, 4), digits = 2L, ...)
 {
   ## https://www.w3.org/TR/WCAG21/#dfn-contrast-ratio
-  ## determine number of color pairs
-  n <- if(is.null(col2)) length(col) - 1L else max(c(length(col), length(col2)))
-  if(n < 1L) stop("need to specify at least two colors")
+  if(length(col) < 1L || length(col2) < 1L) stop("both 'col' and 'col2' need to specify at least one color")
 
   ## suitably recycle colors if necessary
-  if(is.null(col2)) {
-    col2 <- col[-1L]
-    col <- col[-length(col)]
-  }
+  n <- max(c(length(col), length(col2)))
   if(!(length(col) == n && length(col2) == n)) {
     col <- rep_len(col, n)
     col2 <- rep_len(col2, n)

### 2021-07-02 pkg/colorspace/paper/useR-2021/slides.Rnw
slides for useR-2021 keynote
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@601 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
\usetheme[license, url, foot]{uibk}
\renewcommand{\hrulefill}{}

\title{Choosing Color Palettes for Data Visualization}
\subtitle{Tools and Technologies for Supporting Algorithm Fairness and Inclusion}
\author{Achim Zeileis}
\URL{https://www.zeileis.org/}

%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<preliminaries>>=
library("colorspace")
eu <- window(EuStockMarkets, end = 1998)
@

\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

\begin{document}

\section{Robust Color Palettes for Data Visualization}

\subsection{Motivation}

\begin{frame}[fragile]
\frametitle{Motivation}

\textbf{Colors in data visualization:}
\begin{itemize}
  \item Ubiquitous.
  \item Not always easy to choose.
  \item But also perceived as fun.
\end{itemize}

\bigskip
\pause

\textbf{Potential problems:}
\begin{itemize}
  \item Power of color often overestimated.
  \item Color vision deficiencies affect $\sim 8$\% of male and $\sim 0.5$\% of female viewers.
  \item Other physical or technical limitations.
\end{itemize}

\bigskip
\pause

\textbf{Illustration:} Time series line plot using base graphics.

<<echo=TRUE,eval=FALSE>>=
p <- c(1:3, 5)
plot(EuStockMarkets, log = "y", plot.type = "single", col = p, ...)
legend("topleft", colnames(EuStockMarkets), col = p, ...)
@

\end{frame}

\begin{frame}
\frametitle{Motivation}

\begin{minipage}[t]{.73\textwidth}
\setkeys{Gin}{width=\textwidth}
\vspace*{-0.6cm}

<<EuStockMarkets0, eval=FALSE, echo=FALSE, results=hide>>=
par(mar = c(5, 4, 1, 1))
plot(eu, log = "y", ylab = "EU Stock Prices", plot.type = "single", col = p, xlim = c(1991.6, 1998.4), lwd = 1.5)
@
<<EuStockMarketsLegend1, eval=FALSE, echo=FALSE, results=hide>>=
legend("topleft", colnames(eu), col = p, lty = 1, lwd = 3, bty = "n")
@
<<EuStockMarketsLegend2, eval=FALSE, echo=FALSE, results=hide>>=
text(1998, tail(eu, 1), colnames(eu), pos = 4, col = p, xpd = TRUE)
@

\only<1-2>{%
<<EuStockMarkets1, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- palette.colors(8, "R3")[c(1:3, 5)]
<<EuStockMarkets0>>
<<EuStockMarketsLegend1>>
@
}%
\only<3>{%
<<EuStockMarkets2, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- protan(palette.colors(8, "R3")[c(1:3, 5)])
<<EuStockMarkets0>>
<<EuStockMarketsLegend1>>
@
}%
\only<4>{%
<<EuStockMarkets3, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- protan(palette.colors(8, "R3")[c(1:3, 5)])
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%
\only<5>{%
<<EuStockMarkets4, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- palette.colors(8, "R3")[c(1:3, 5)]
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%
\only<6>{%
<<EuStockMarkets5, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- palette.colors(8, "R4")[c(1:3, 5)]
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%
\only<7>{%
<<EuStockMarkets6, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- protan(palette.colors(8, "R4")[c(1:3, 5)])
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%
\only<8>{%
<<EuStockMarkets7, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- palette.colors(4, "Okabe-Ito")[c(1, 4, 2, 3)]
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%
\only<9>{%
<<EuStockMarkets8, fig=TRUE, height=6, width=8.5, echo=FALSE, results=hide>>=
p <- protan(palette.colors(4, "Okabe-Ito")[c(1, 4, 2, 3)])
<<EuStockMarkets0>>
<<EuStockMarketsLegend2>>
@
}%

\end{minipage}
\hfill
\begin{minipage}[t]{0.26\textwidth}

\small

\textbf{Palette:}
\only<1-5>{R $\le$ 3 default}%
\only<6-7>{R $\ge$ 4 default}%
\only<8-9>{Okabe-Ito}

\medskip

\textbf{Emulation:}
\only<1-2,5-6,8>{None}%
\only<3-4,7,9>{Protanope}

\medskip

\textbf{Labeling:}
\only<1-3>{Legend}%
\only<4-9>{Direct}

\bigskip
\bigskip

\only<2->{\textbf{Comments:}}

\only<2-5>{
\medskip

Too flashy

\medskip

Cyan too light

}

\only<3-5>{

\medskip

Hard to distinguish for protanope viewers

}

\only<6-7>{
\medskip

Similar hues

\medskip

More balanced

brightness

\medskip

Avoid garish colors

}

\only<8-9>{
\medskip

Designed to be robust

against color vision

deficiencies

}

\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Motivation}

%% png("drake.png", height = 570, width = 530, res = 100)
%% contrast_ratio(rev(palette.colors(8, "R3")), rev(palette.colors(8, "R3")), plot = TRUE, main = "old base palette", mar = c(1, 1, 3, 5), off = 0.4)
%% text(0, 7:0 + 0.73, palette.colors(8, "R3"), pos = 4)
%% 
%% png("drake.png", height = 570, width = 530, res = 100)
%% contrast_ratio(rev(palette.colors(8, "R4")), rev(palette.colors(8, "R4")), plot = TRUE, main = "new base palette", mar = c(1, 1, 3, 5), off = 0.4)
%% text(0, 7:0 + 0.73, palette.colors(8, "R4"), pos = 4)
%% dev.off()

\begin{minipage}[t]{0.51\textwidth}

\vspace*{-0.6cm}

\includegraphics[width=\textwidth]{images/palette-bling.png}

{\footnotesize
Source: Mara Averick via Twitter
}

\end{minipage}
\hspace*{0.3cm}
\pause
\begin{minipage}[t]{0.44\textwidth}

\textbf{Base R:} Neglected better color palettes for a long time.

\bigskip

\textbf{Earlier packages:} RColorBrewer, colorspace, ggplot2, viridis, rcartocolor, Polychrome, pals, paletteer, \dots

\bigskip

\textbf{Thus:} Many good palettes easily available.

\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Motivation}

\includegraphics[width=1.0\textwidth]{images/grdevices-qualitative.png}

\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}

\setkeys{Gin}{width=0.6\textwidth}

<<swatchplot-r4, fig=TRUE, height=4, width=6.5, echo=TRUE, results=hide>>=
palette.colors(palette = "R4") |>
  colorspace::swatchplot(cvd = TRUE)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}

\setkeys{Gin}{width=0.6\textwidth}

<<swatchplot-okabeito, fig=TRUE, height=4, width=6.5, echo=TRUE, results=hide>>=
palette.colors(palette = "Okabe-Ito") |>
  colorspace::swatchplot(cvd = TRUE)
@

\end{frame}

\subsection{Color palette construction}

\begin{frame}
\frametitle{Color palette construction}

\begin{minipage}{0.47\textwidth}
\textbf{HCL:} Polar coordinates in CIELUV. Captures perceptual dimensions of the human visual system very well.

\bigskip

\setkeys{Gin}{width=\textwidth}
<<hcl-properties, fig = TRUE, width = 4, height = 2.2>>=
swatchplot(
  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  nrow = 3, line = 5.2, off = 0
)
@
\end{minipage}
\pause
\hspace*{0.6cm}
\begin{minipage}{0.47\textwidth}
\textbf{RGB:} Motivated by how computers/TVs used to generate and still represent color.

\bigskip

\setkeys{Gin}{width=0.9\textwidth}
<<rgb-properties, fig = TRUE, width = 3.6, height = 2.2>>=
swatchplot(
  "Red"       = rgb(0:4/4, 0, 0),
  "Green"     = rgb(0, 0:4/4, 0),
  "Blue"      = rgb(0, 0, 0:4/4),
  nrow = 3, off = 0
)
@
\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Color palette construction}

\setkeys{Gin}{width=\textwidth}
<<hcl-palettes-principles, fig = TRUE, width = 12, height = 1.6>>=
swatchplot(
  "Qualitative (Set 2)"     = rbind("Color" = qualitative_hcl(5, "Set 2"),     "Desaturated" = desaturate(qualitative_hcl(5, "Set 2"))),
  "Sequential (Blues 3)"    = rbind("Color" = sequential_hcl(7, "Blues 3"),    "Desaturated" = desaturate(sequential_hcl(7, "Blues 3"))),
  "Diverging (Green-Brown)" = rbind("Color" = diverging_hcl(7, "Green-Brown"), "Desaturated" = desaturate(diverging_hcl(7, "Green-Brown"))),
  nrow = 3, line = 7)
@

\bigskip
\pause

\textbf{Qualitative:} For categorical information with no particular ordering.

Luminance differences should be limited

\bigskip
\pause

\textbf{Sequential:} For ordered/numeric information from high to low (or vice versa).

\bigskip
\pause

\textbf{Diverging:} For ordered/numeric information diverging from a central neutral value to two extremes.

\end{frame}

\begin{frame}
\frametitle{Color palette construction}

\begin{minipage}[t]{0.47\textwidth}
\vspace*{-0.9cm}
\includegraphics[width=\textwidth]{images/grdevices-sequential.png}
\end{minipage}
\begin{minipage}[t]{0.39\textwidth}
\vspace*{-0.8cm}
\includegraphics[width=\textwidth]{images/grdevices-diverging.png}
\end{minipage}

\end{frame}

\subsection{Risk maps and communication to the public}

\begin{frame}
\frametitle{Risk maps and communication to the public}

\vspace*{-0.3cm}

\begin{center}
\only<1>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orig.png}}%
\only<2>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orig-deutan.png}}%
\only<3>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orig-gray.png}}%
\only<4>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orrd.png}}%
\only<5>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orrd-gray.png}}
\only<6>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orrd-deutan.png}}
\end{center}

\vspace*{-0.5em}
\begin{footnotesize} 
Probability of wind speeds $>$ 39\,mph (63\,km\,h$^{-1}$), 2019-08-30--2019-09-04.
Source: (\href{https://noaa.gov}{noaa.gov})
\end{footnotesize}

\end{frame}

\begin{frame}
\frametitle{Risk maps and communication to the public}

\vspace*{2em}

\begin{minipage}{0.46\textwidth}
\includegraphics[width=\textwidth]{../EGU2020/images/sharpie-small.png}

{\footnotesize
Source: White House (2019-09-04)}
\end{minipage}
\hfill
\begin{minipage}{0.51\textwidth}
\includegraphics[width=\textwidth]{../EGU2020/images/tweet.png}

{\footnotesize
Source: U.S. president via Twitter (2019-09-05)}
\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Colors by designers, painters, and directors?}

\begin{minipage}[t]{0.65\textwidth}
\vspace*{-0.3cm}

\only<1>{\includegraphics[width=\textwidth]{images/almodovar-madre-original.png}}
\only<2>{\includegraphics[width=\textwidth]{images/almodovar-madre-palette.png}}
\only<3>{\includegraphics[width=\textwidth]{images/almodovar-madre-desaturated.png}}
\end{minipage}
\hfill
\begin{minipage}[t]{0.3\textwidth}

{\footnotesize

Movie: \emph{Todo sobre mi madre} (\emph{All About My Mother}, 1999)

\medskip

Picture source: Sony Pictures Classics via MoMA

\medskip

\only<2->{Palette: Hadley Mendelsohn}
}

\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Colors by designers, painters, and directors?}

\vspace*{-0.3cm}

\begin{center}
\only<1>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-madre.png}}%
\only<2>{\includegraphics[width=.55\textwidth]{images/dorian-useR-2021-orrd.png}}%
\end{center}

\vspace*{-0.5em}
\begin{center}
{\footnotesize
Palette: \only<1>{Todo sobre mi madre}\only<2>{OrRd (ColorBrewer.org, HCL version)}\phantom{Org}}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Colors by designers, painters, and directors?}

\begin{minipage}[t]{0.45\textwidth}
\vspace*{-0.3cm}

\only<1>{\includegraphics[width=\textwidth]{images/almodovar-tacones-original.png}}
\only<2>{\includegraphics[width=\textwidth]{images/almodovar-tacones-palette.png}}
\only<3>{\includegraphics[width=\textwidth]{images/almodovar-tacones-deutan.png}}
\end{minipage}
\hfill
\begin{minipage}[t]{0.5\textwidth}

{\footnotesize

Movie: \emph{Tacones lejanos} (\emph{High Heels}, 1991)

\medskip

Picture source: El Deseo S.A. via Twitter

\medskip

\only<2->{Palette: Bibiana Fernandez}
}

\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Wrap-up}

\textbf{Tools:}
\begin{itemize}
  \item \emph{grDevices:} \code{palette.colors()}, \code{hcl.colors()}.
  \item \emph{colorspace:} \code{swatchplot(..., cvd = TRUE)}.
  \item Interactive shiny apps on \url{https://www.hclwizard.org/}.
\end{itemize}

\pause
\bigskip

\textbf{Strategy:}
\begin{itemize}
  \item Check whether color is appropriate for coding your information.
  \item Use appropriate type of palette.
  \item Don't reinvent the wheel, start out from well-established palettes.
  \item Check robustness of palette.
  \item Be careful with palettes with too much chroma.
\end{itemize}

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\small

Zeileis A, Fisher JC, Hornik K, Ihaka R, McWhite CD, Murrell P, Stauffer R, Wilke CO (2020).
  \dquote{colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes.}
  \emph{Journal of Statistical Software}, \textbf{96}(1), 1--49.
  \doi{10.18637/jss.v096.i01}

\bigskip

Zeileis A, Murrell P, Maechler M, Sarkar D (2019).
  \dquote{A New \texttt{palette()} for R.}
  \emph{R~Foundation Blog}, 2019-11-21.
  \url{https://developer.R-project.org/Blog/public/2019/11/21/a-new-palette-for-r/}

\bigskip

Zeileis A, Murrell P (2019).
  \dquote{HCL-Based Color Palettes in grDevices.}
  \emph{R~Foundation Blog}, 2019-04-01.
  \url{https://developer.R-project.org/Blog/public/2019/04/01/hcl-based-color-palettes-in-grdevices/}

\bigskip
\bigskip

Web: \url{https://colorspace.R-Forge.R-project.org/}

\smallskip

Twitter: \href{https://twitter.com/AchimZeileis}{\tt @AchimZeileis}

\end{frame}

\end{document}

### 2021-07-02 pkg/colorspace/paper/useR-2021/almodovar.R

@@ -1,0 +1,47 @@
+library("colorspace")
+
+## Tacones Lejanos (High Heels, 1991)
+## Idea: Bibiana Fernandez (https://www.pinterest.at/pin/396387204673777781/)
+## Credit: https://www.alamy.com/stock-photo-pedro-almodovar-victoria-abril-miriam-diaz-aroca-marisa-paredes-high-31039349.html
+## Images:
+## https://i.pinimg.com/originals/6a/28/4b/6a284b533ac109f9933d2e1dcf63a5d2.jpg
+## https://c7.alamy.com/comp/BPDY05/pedro-almodovar-victoria-abril-miriam-diaz-aroca-marisa-paredes-high-BPDY05.jpg
+## https://twitter.com/juanferrervila/status/1044483192361807872
+tacones0 <- paste0("#", c("334ea9", "5e9aa4", "fceb5f", "423783", "931c14", "b16594"))
+tacones  <- paste0("#", c("1e51bc", "61abb6", "ba91bc", "fff353", "42389b", "ab200c"))
+specplot(tacones0[c(1:2, 6, 3:5)], tacones)
+
+## Todo Sobre Mi Madre (All About My Mother, 1999)
+## Idea: Hadley Mendelsohn (https://www.housebeautiful.com/design-inspiration/a33012615/most-beautiful-movie-color-palettes/)
+## Credit: Sony Pictures Classics via MoMA (https://www.moma.org/calendar/events/2544)
+## Images:
+## https://www.moma.org/d/assets/W1siZiIsIjIwMTYvMTAvMjYvNnNncnR5YmYwX0FsbF9BYm91dF9NeV9Nb3RoZXJfMV9jcm9wcGVkLmpwZyJdLFsicCIsImNvbnZlcnQiLCItcXVhbGl0eSA5MCAtcmVzaXplIDIwMDB4MjAwMFx1MDAzZSJdXQ/All%20About%20My%20Mother_1_cropped.jpg?sha=f46b56461ca7050e
+## https://hips.hearstapps.com/hmg-prod.s3.amazonaws.com/images/todo-sobre-mi-madre-2-1594842994.png
+madre0 <- paste0("#", c("f6e691", "d5c6b3", "cf9788", "d31509", "9d060d", "660003"))
[…]

### 2021-07-02 pkg/colorspace/paper/useR-2021/swatch.R

@@ -1,0 +1,61 @@
+palette.swatch <- function(palette = palette.pals(), n = 8, nrow = 8,
+                           border = "black", cex = 1, ...)
+{
+     cols <- sapply(palette, palette.colors, n = n, recycle = TRUE)
+     ncol <- ncol(cols)
+     nswatch <- min(ncol, nrow)
+     op <- par(mar = rep(0.1, 4),
+               mfrow = c(1, min(5, ceiling(ncol/nrow))),
+     	       cex = cex, ...)
+     on.exit(par(op))
+     while (length(palette)) {
+ 	subset <- seq_len(min(nrow, ncol(cols)))
+ 	plot.new()
+ 	plot.window(c(0, n), c(0.25, nrow + 0.25))
+ 	y <- rev(subset)
+ 	text(0, y + 0.1, palette[subset], adj = c(0, 0))
+ 	y <- rep(y, each = n)
+ 	rect(rep(0:(n-1), n), y, rep(1:n, n), y - 0.5,
+ 	     col = cols[, subset], border = border)
+ 	palette <- palette[-subset]
+ 	cols    <- cols [, -subset, drop = FALSE]
+     }
+}
+
+hcl.swatch <- function(palette = NULL, type = NULL, n = 5, nrow = 11,
+  border = if (n < 15) "black" else NA, ...) {
+    if(is.null(palette)) palette <- hcl.pals(type)
+    cols <- sapply(palette, hcl.colors, n = n)
+    ncol <- ncol(cols)
+    nswatch <- min(ncol, nrow)
+
+    par(mar = rep(0.1, 4),
+        mfrow = c(1, min(5, ceiling(ncol/nrow))),
+        cex = 0.7, ...)
+
+    while (length(palette)) {
+        subset <- 1:min(nrow, ncol(cols))
+        plot.new()
+        plot.window(c(0, n), c(0, nrow + 1))
+        text(0, rev(subset) + 0.1, palette[subset], adj = c(0, 0))
+        y <- rep(subset, each = n)
+        rect(rep(0:(n-1), n), rev(y), rep(1:n, n), rev(y) - 0.5,
+             col = cols[, subset], border = border)
+        palette <- palette[-subset]
+        cols <- cols[, -subset, drop = FALSE]
+    }
+
+    par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1), cex = 1)
+}
+
[…]

### 2021-07-02 pkg/colorspace/paper/useR-2021/slides.Rnw
small improvements, first draft ready
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@602 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   \item Power of color often overestimated.
-  \item Color vision deficiencies affect $\sim 8$\% of male and $\sim 0.5$\% of female viewers.
+  \item Color vision deficiencies ($\sim 8$\% of male and $\sim 0.5$\% of female viewers).
   \item Other physical or technical limitations.
@@
 {\footnotesize
-Source: Mara Averick via Twitter
+\textbf{Source:} Mara Averick via Twitter
 }
@@
 \begin{footnotesize} 
-Probability of wind speeds $>$ 39\,mph (63\,km\,h$^{-1}$), 2019-08-30--2019-09-04.
-Source: (\href{https://noaa.gov}{noaa.gov})
+\textbf{Risk map:} Probability of wind speeds $>$ 39\,mph (63\,km\,h$^{-1}$), 2019-08-30--2019-09-04.
+
+\textbf{Source:} (\href{https://noaa.gov}{noaa.gov})
 \end{footnotesize}
@@
 {\footnotesize
-Source: White House (2019-09-04)}
+\textbf{Source:} White House (2019-09-04)}
 \end{minipage}
@@
 {\footnotesize
-Source: U.S. president via Twitter (2019-09-05)}
+\textbf{Source:} U.S. president via Twitter (2019-09-05)}
 \end{minipage}
@@
 \end{frame}
+
+\subsection{Colors by designers, painters, and directors?}
 
@@
 \hfill
-\begin{minipage}[t]{0.3\textwidth}
+\begin{minipage}[t]{0.32\textwidth}
 
@@
 
-Movie: \emph{Todo sobre mi madre} (\emph{All About My Mother}, 1999)
+\textbf{Movie:} \emph{Todo sobre mi madre} (\emph{All About My Mother}, 1999)
 
@@
 
-Picture source: Sony Pictures Classics via MoMA
+\textbf{Source:} Sony Pictures Classics via MoMA
 
@@
 
-\only<2->{Palette: Hadley Mendelsohn}
+\only<2->{\textbf{Palette:} Hadley Mendelsohn}
 }
@@
 {\footnotesize
-Palette: \only<1>{Todo sobre mi madre}\only<2>{OrRd (ColorBrewer.org, HCL version)}\phantom{Org}}
+\textbf{Palette:} \only<1>{Todo sobre mi madre}\only<2>{OrRd (ColorBrewer.org, HCL version)}\phantom{Org}}
 \end{center}
@@
 
-Movie: \emph{Tacones lejanos} (\emph{High Heels}, 1991)
+\textbf{Movie:} \emph{Tacones lejanos} (\emph{High Heels}, 1991)
 
@@
 
-Picture source: El Deseo S.A. via Twitter
+\textbf{Source:} El Deseo S.A. via Twitter
 
@@
 
-\only<2->{Palette: Bibiana Fernandez}
+\only<2->{\textbf{Palette:} Bibiana Fernandez}
 }
@@
 \end{frame}
+
+\subsection{Wrap-up}
 

### 2021-07-03 pkg/colorspace/paper/useR-2021/slides.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@603 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\section{Robust Color Palettes for Data Visualization}
+\section{Choosing Color Palettes for Data Visualization}
 
@@
 
-\textbf{Base R:} Neglected better color palettes for a long time.
+\textbf{Base R:} Neglected better color
+
+palettes for a long time.
 
@@
 
-\textbf{Earlier packages:} RColorBrewer, colorspace, ggplot2, viridis, rcartocolor, Polychrome, pals, paletteer, \dots
+\textbf{Earlier packages:} RColorBrewer,
+
+colorspace, ggplot2, viridis,
+
+rcartocolor, Polychrome, pals,
+
+paletteer, \dots
 
@@
 
-\textbf{Thus:} Many good palettes easily available.
+\textbf{Thus:} Many good palettes easily
+
+available.
 
@@
 \begin{minipage}{0.47\textwidth}
-\textbf{HCL:} Polar coordinates in CIELUV. Captures perceptual dimensions of the human visual system very well.
+\textbf{HCL:} Polar coordinates in CIELUV.
+
+Captures perceptual dimensions of
+
+the human visual system very well.
 
@@
 \begin{minipage}{0.47\textwidth}
-\textbf{RGB:} Motivated by how computers/TVs used to generate and still represent color.
+\textbf{RGB:} Motivated by how computers
+
+and TVs used to generate and still
+
+represent color.
 
@@
 
-Luminance differences should be limited
+Luminance differences should be limited.
 
@@
 \begin{footnotesize} 
-\textbf{Risk map:} Probability of wind speeds $>$ 39\,mph (63\,km\,h$^{-1}$), 2019-08-30--2019-09-04.
+\textbf{Risk map:} Probability of wind speeds $>$ 39\,mph (63\,km\,h$^{-1}$), 2019-08-30--2019-09-04
 
-\textbf{Source:} (\href{https://noaa.gov}{noaa.gov})
+\textbf{Source:} National Oceanic and Atmospheric Administration (\href{https://noaa.gov}{noaa.gov})
 \end{footnotesize}
@@
 
-\begin{minipage}[t]{0.65\textwidth}
+\begin{minipage}[t]{0.64\textwidth}
 \vspace*{-0.3cm}
@@
 
-Web: \url{https://colorspace.R-Forge.R-project.org/}
+\textbf{Web:} \url{https://colorspace.R-Forge.R-project.org/}
 
@@
 
-Twitter: \href{https://twitter.com/AchimZeileis}{\tt @AchimZeileis}
+\textbf{Twitter:} \href{https://twitter.com/AchimZeileis}{\tt @AchimZeileis}
 
